### 2021-05-27 manuscript_plos.tex
First commit

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

% Adding packages 
\usepackage{bm}
\usepackage[parfill]{parskip}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{cILR: Taxonomic enrichment analysis with isometric log-ratios} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Quang Nguyen\textsuperscript{1,2},
Anne G. Hoen\textsuperscript{1,2\ddag},
H. Robert Frost\textsuperscript{1*\ddag},
\\
\bigskip
\textbf{1} Department of Biomedical Data Science, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA
\\
\textbf{2} Department of Epidemiology, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
%\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\ddag These authors jointly supervised this work.

% Current address notes
%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
% \dag Deceased

% Group/Consortium Author Note
%\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* hildreth.r.frost@dartmouth.edu

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their human host. Previous research has shown that changes in the composition of the microbiome have been associated with important health outcomes such as inflammatory bowel disease, type II diabetes, and obesity. To understand the central role of the microbiome in human health, researchers often relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e. amplicon sequencing) or by profiling all the genomic content of the sample (i.e. whole-genome shotgun sequencing) \cite{cho2012}. Processing raw sequencing data using a variety of bioinformatic pipelines \cite{callahan2016,truong2015} would yield taxonomic abundance tables that would be part of downstream statistical analyses where associations between an outcome/exposure and identified taxa would be found.

However, there exists unique challenges in the analysis of these data tables \cite{li2019,li2015}. First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present. For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016}, allowing for simultaneous model fitting and variable selection. For differential abundance tasks, researchers often utilized multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.

Second, the number of reads obtained is constrained by the sequencing instrument at an arbitrary limit, and applied inconsistently across samples, resulting in a different number of total read counts per sample. Many normalization methods \cite{weiss2017} have been proposed to address these issues, including cross-applying methods from the gene expression literature \cite{mcmurdie2014}. However, these methods rely on assumptions specific to the original bulk RNA-seq data sets such as the presence of housekeeping genes with consistent expression levels \cite{love2014}, which might not be not true in the context of microbiome relative abundance data \cite{quinn2019,quinn2018b}. As such, microbiome data is strictly compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another. Consequently, log-ratio transformations from the compositional data literature are often utilized \cite{aitchison1999}.

Third, the data is highly zero-inflated, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment). Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015}, or applying zero-inflated models \cite{chen2018, kaul2017}. Newer methods developed recently have focused on understanding the different types of zeros in the data, providing more sophisticated heuristics around when pseudocounts can be utilized \cite{kaul2017a}.

Even though the aforementioned problems are challenging, a very approachable method to address some of them is variable aggregation. Aggregated variables can be less sparse than their constituent elements, and hypothesis testing on a reduced number of variables can alleviate the multiple testing burden. This helps increase power in downstream analyses, as well as interpretability. In practice, microbiome researchers perform variable aggregation to pre-defined taxonomic levels (e.g. phylum, family, order) through element-wise summation of counts. However, this approach contains various downsides: first, it does not allow for comparison of enrichment across sets of different sizes, where larger sets naturally contain more counts; second, aggregating compositional variables using component-wise summations can distort the inter-sample distances before and after aggregation due to the non-linearity of this amalgamation in Aitchison space \cite{egozcue2005}.

Set-based analyses are ubiquitous in the gene expression literature \cite{goeman2007} and various methods have been developed for enrichment testing and scoring. More traditional gene-set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest. Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values. The most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size. Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples. More recent methods, such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate sample level enrichment scores more akin to a transformation. Generated scores can then be used in data visualization, as well as flexible scores that can be utilized in downstream analyses.

Here, we developed a new sample-level set-based analysis method specific to microbiome relative abundance data. We leverage the conception of the $Q_1$ competitive hypothesis presented in Tian et al. \cite{tian2005}, which compares the expression of genes within the gene set against the rest of the genes. The competitive hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables. We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a log-ratio-based aggregation method that addresses the downsides of the naive summation-based method presented above. The resulting method, titled competitive isometric log-ratio (cILR), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing. These scores can then act as inputs to differential abundance and predictive modeling tasks downstream.

In this manuscript, we present cILR, provide its formulation and discuss some statistical properties. Then, we illustrate the benefits of cILR as both a sample-level significance test for enrichment of taxa sets and as a feature engineering approach for downstream disease prediction and differential abundance using both real data and simulation studies. We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as existing sample-level enrichment methods in the gene expression literature such as GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}. An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).

\section{Material and Methods} \label{Methods}
\subsection{Competitive Isometric Log-ratio (cILR)}
The cILR method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}. Details on the computational implementation of cILR can be found in the supplemental section. The cILR method takes two inputs:  
\begin{itemize}
    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as 16S rRNA) or whole genome shotgun sequencing. Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing. 
    \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxa $p$ to $m$ sets of interest. These sets can be Linnean taxonomic classifications annotated using databases such as SILVA \cite{quast2013}, or those based on more functionally driven categories such as tropism or ecosystem roles ($A_{i,j} = 1$ indicates that microbe $i$ belongs to set $j$). 
\end{itemize}
The cILR method generates one output:  
\begin{itemize} 
    \item $\mathbf{E}$: $n$ by $m$ matrix indicating the enrichment score of $m$ pre-defined sets identified in $\mathbf{A}$ across $n$ samples. 
\end{itemize}
The procedure is as follows:  
\begin{enumerate}
    \item \textbf{Compute the cILR statistic}: Let $\mathbf{M}$ be a $n$ by $m$ matrix of cILR scores. Let $\mathbf{M}_{i,k}$ be cILR scores for set $k$ of sample $i$:   
    \begin{equation}\label{main_eq}
        \mathbf{M}_{i,k} = \sqrt{\frac{\sum_k A_{ik}(p - \sum_k A_{ik})}{p}} \ln \left( \frac{g(\mathbf{X}_{i,j}|\mathbf{A}_{j,k} = 1)}{g(\mathbf{X}_{i,j}|\mathbf{A}_{j,k} \neq 1))} \right)
    \end{equation}
    where $g(.)$ is the geometric mean. This represents the ratio of the geometric mean of the relative abundance of taxa assigned to set $k$ and remainder taxa. 
    \item \textbf{Estimate the empirical null distribution} Enrichment scores represent the test statistic for the $Q_1$ null hypothesis $H_o$ that relative abundances in $\mathbf{X}$ of members of set $k$ are not enriched compared to those not in set $k$. Since the distribution of cILR under the null vary depending on data characteristics (Figure \ref{fig:kurt_gof}), an empirical null distribution will be estimated from data.
    \begin{itemize}
        \item \textbf{Compute the cILR statistic on permuted and un-permuted $\mathbf{X}$}.  Let $\mathbf{X}_{perm}$ be the column permuted relative abundance matrix, and $\mathbf{M}_{perm}$ be the corresponding cILR scores generated from $\mathbf{X}_{perm}$. Similarly, we have $\mathbf{M}_{unperm}$ be cILR scores generated from $\mathbf{X}$.
        \item \textbf{Estimate correlation-adjusted empirical distribution for each set}. For each set, a fit a parametric distribution to both $\mathbf{M}_{perm}$ and $\mathbf{M}_{unperm}$. The location measure estimated from $\mathbf{M}_{perm}$ and the spread measure estimated from $\mathbf{M}_{unperm}$ will be combined as the correlation-adjusted empirical null distribution $\mathbf{P}_{emp}$ for each set. Two available options are the normal distribution and the mixture normal distribution. For the normal distribution, parameters were estimated using the method of maximum likelihood implemented in the \emph{fitdistr} package \cite{delignette-muller2015}. For the mixture normal distribution, parameters were estimated using an expectation-maximization algorithm implemented in the \emph{mixtools} package \cite{benaglia2009}. 
    \end{itemize}
    \item \textbf{Calculate finalized cILR scores with respect to the empirical null}. Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution. Valid p-values can be calculated by subtracting $\mathbf{E}$ from 1. 
\end{enumerate}

\subsection*{Properties of cILR}
\subsubsection*{cILR and the Isometric Log Ratio Transformation}
The cILR statistic is a special instance of the isometric log-ratio transformation (ILR) \cite{egozcue2003}. The standard ILR is a transformation method to address the negative correlation bias inherent in compositional data by providing an isometry between the $D$-dimensional simplex $\mathbb{S}^D$ and coordinates in the $D-1$ real space $\mathbb{R}^{D-1}$ \cite{egozcue2003,washburne2017}. This is accomplished by projecting the composition onto a chosen orthonormal basis in $\mathbb{R}$, which can be defined by a sequential binary partition (SBP) of the variables (e.g. a rooted phylogenetic tree). The ILR transformed variables are the coordinates of nodes within an SBP tree of the variables. Without loss of generalizability, in a given SBP with node $i$ splitting variables between sets $R$ and $S$, we have the ILR coordinate $x^{*}_{i}$ as: 
\begin{equation}\label{ilr_standard}
    x^{*}_i = \sqrt{\frac{rs}{r+s}} \log\left(\frac{g(X_{j|j \in R})}{g(X_{j|j \in S})}\right)
\end{equation}
where $r$ and $s$ are the cardinalities of sets $R$ and $S$ respectively, $g(z)$ is the geometric mean, and $X_{j}$ are values of the original predictors with indexes defined by membership in $R$ and $S$. The ILR confer many important benefits. First, ILR coordinates exist in real space, whereby common statistical methods can be used. Second, ILR aggregated variables preserve inter-sample distances before and after aggregation \cite{egozcue2005}. Third, ILR variables are not constrained to sum to 0 as that of the centered log-ratio transformation, resulting in a covariance matrix that is not singular \cite{egozcue2003}.

The usage of the ILR statistic is not uncommon in the microbiome literature. They are usually termed ``compositional balances", and were leveraged in many recent approaches in variable transformation \cite{washburne2017,silverman2017,morton2017}. The cILR formulation \eqref{main_eq} is a special case of \eqref{ilr_standard} defined on a node that splits the taxa into two disjoint sets, one representing the set of interest, the other representing the remaining taxa. As such, the cILR transformation inherits the properties of the ILR as a log-ratio method applicable to compositional data sets. However, unlike the ILR and its variants \cite{silverman2017, morton2017, washburne2017}, the axes defined by each cILR set are not orthogonal (since the balances are mutually exclusive between sets and do not belong in the same SBP). Hence, a correlation can exist between cILR aggregated variables.

\subsubsection*{Statistical Properties of cILR}
We can perform significance testing on the cILR statistic which corresponds to the null hypothesis that the center of the subcomposition defined by the set is equal to the center of the subcomposition defined by the complement of the set. This is equivalent to the $Q_1$ competitive null hypothesis in the gene set testing literature \cite{tian2005} where the enrichment of a gene set is defined with respect to genes outside the set.

We can apply prior usage of the ILR statistic in hypothesis testing to cILR by assuming that the null distribution of cILR follows a standard normal distribution \cite{egozcue2005}. However, when applying cILR for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses. Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies. As such, to perform significance testing using cILR, we also estimated the null distribution from observed raw cILR variables.

This assumption is also supported by preliminary simulation studies (detailed below). In panel B of figure \ref{fig:kurt_gof}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness. It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation. Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis). The effect is most dramatic under both high inter-taxa correlation and sparsity. This heterogeneity further supports the decision to estimate an empirical null distribution, similar to Efron \cite{efron2004}.

Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null. To address this issue, we also evaluated a mixture distribution of two normal components. Panel A of figure \ref{fig:kurt_gof} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null. We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.

We performed our empirical null estimation by fitting our distribution of choice and computing relevant parameters on raw cILR scores on taxa-permuted data (equivalent to gene permutation in the gene expression literature). As such, the null distribution is characterized by scores computed on sets of equal size with randomly drawn taxa. However, null distribution based on taxa-permutation is sensitive to inter-taxa correlations within the set \cite{wu2012}. Since the permutation procedure does not preserve correlation structures, estimating parameters from empirical scores on permuted data will underestimate the variance inflation due to correlation. We account for this by combining the mean estimate from permuted data with the variance estimate from unpermuted data, where the inter-taxa correlation structure remains undisturbed. However, this procedure assumes that the variance of cILR is equal under both the null and alternate hypotheses.

\begin{figure}
    \centering
    \caption{Properties of the null distribution of cILR in different simulation scenarios under the global null. Panel \textbf{(B)} presents kurtosis and skewness of cILR scores while panel \textbf{(A)} presents the goodness of fit (as Kolmogorov-Smirnov D statistic) for mixture normal and normal distributions. Panel \textbf{(C)} is a density plot of the shape of the null distribution. Results indicated the necessity of estimating an empirical null and demonstrating that the mixture distribution was the better fit compared to the basic normal.}
    \label{fig:kurt_gof}
\end{figure}

\subsection*{Evaluation}
\subsubsection*{Parametric Simulations}  
To address the performance of cILR under different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}. Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
\begin{equation}
    \mathbf{X}_{ij} =
      \begin{cases}
        0 & \text{with probability $p_j$}\\
        \mathbf{NB}(\mu_j, \phi_j) & \text{with probability $1 - p_j$}\\
      \end{cases}       
\end{equation}

where $\mu_j$ and $\phi_j$ are mean and dispersion parameters, respectively. To incorporate a flexible correlation structure into our simulation model, we utilized the NorTA (Normal to Anything) method \cite{cario1997}. Given an $n$ by $p$ matrix of values $\mathbf{U}$ sampled from multivariate normal distribution with correlation matrix $\mathbf{\rho}$, we can generate target microbiome count vector $\mathbf{X_{.j}}$ for taxa $j$ following the marginal distribution $\mathbf{NB}$ characterized by the negative binomial cumulative distribution function $\mathbb{F_{\mathbf{NB}}}$:
\begin{equation}
    \mathbf{X}_{.j} = \mathbb{F_{\mathbf{NB}}}^{-1}(\Phi_{U_i})
\end{equation}
In this instance, for each taxon $j$, we set elements in $\mathbf{U}_{.j}$ to be zero with probability $p_j$ and applied $\mathbf{NB}^{-1}(\mu_j, \phi_j)$ on non-zero elements to generate our final count matrix $\mathbf{X}$. To ensure that our simulations match closely to real data, we fitted negative binomial distribution using a maximum likelihood approach (with the \emph{fitdistrplus} package in R \cite{delignette-muller2015}) to non-zero counts for each taxon from 16S rRNA profiling of stool samples from the Human Microbiome Project (HMP). We take the median values of the estimated mean and dispersion parameters as the baseline of our simulations. For simplicity, we assumed that inter-taxa correlation follows an exchangeable structure

\noindent \textbf{Single Sample Enrichment}: To assess type I error rate and power for enrichment significance testing at the sample level, we simulated data based on the schema above, and assessed enrichment for one focal set. Type I error was obtained under the global null as the number of samples where the null hypothesis was rejected at $\alpha = 0.05$ over the total number of samples (which represents the total number of hypotheses tested). Power was obtained using the same formulation as type I error rate but under the global alternate. We treated type I error and power as estimates of binomial proportions and utilized the Agresti-Couli \cite{agresti1998} formulation to calculate 95\% confidence intervals. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$) and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). For type I error analysis, we also varied the size of the set (50, 100, 150). For power analyses, set size was kept constant at 100 but different effect sizes (fold change of 1.5, 2, and 3). All sample sizes were set at 10,000.

For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUROC/AUC). This is a strategy used in Frost et al. \cite{frost2020a} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e. whether samples with inflated counts are highly ranked using estimated scores).  DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate. Simulation settings for classification performance were identical to power analyses as detailed in the previous paragraph.

\noindent \textbf{Differential Abundance Analysis}: To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition. Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null. Power is defined similarly, but instead under the global alternate hypothesis. Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error. A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3). Due to the relative nature of microbiome data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets. As such, we applied a compensation procedure as described in Hawinkel et al. \cite{hawinkel2019} to ensure the validity of simulation results. All sample sizes were set at 2,000.

\noindent \textbf{Prediction}: To assess the predictability, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}. For binary outcomes, we use AUC similar to the classification analyses above. For continuous outcomes, we used root mean squared error (RMSE). All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020} suite of packages. Across both learning tasks, we varied sparsity ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation ($\rho = 0, 0.2, 0.5$). Continuous outcomes $Y_{cont}$ were generated as linear combinations of taxa counts.  
\begin{equation}
    Y_{cont} = f(\mathbf{X}) + \mathbf{\epsilon}
\end{equation}
where $\mathbf{\epsilon} \sim N(0, \sigma_{\epsilon}^2)$ and $f(\mathbf{X}) = \beta_0 + \mathbf{X}\mathbf{\beta}$. For each simulation, we set $\beta_0$ to be $\frac{6}{\sqrt{10}}$ similar to \cite{xiao2018}. The degree of model saturation (the number of non zero $\mathbf{\beta}$ values) were varied between 0.1 and 0.5, and signal to noise ratio (SNR = $\frac{\sigma(f(\mathbf{X}))}{\sigma_{\epsilon}}$) was varied between 1.5, 2, and 3.

For binary outcomes, we generate $Y_{binary}$ as Bernoulli draws with probability $p_{binary}$, where 
$$p_{binary} = \frac{1}{1 + \exp(f(\mathbf{X}) + \mathbf{\epsilon})}$$
To ensure a balance of classes, we applied the strategy described in Dong et al. \cite{dong2020} where the associated $\beta$ values are evenly split between positive and negative associations. All data sets generated from prediction tasks have 2,000 samples with 5,000 taxa over 50 sets with a size of 100 taxa per set.

\subsubsection*{Real Datasets}

In addition to simulation analyses, we also evaluated our method using real data sets across both 16S rRNA gene sequencing and whole-genome sequencing. All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.

\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of cILR in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}. We utilized this data set following the approach outlined in Calagaro et al. \cite{calgaro2020}. This data set is special because it is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}. Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false discovery rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment. Microbial tropism annotation at the Genus level was from Beghini et al. \cite{beghini2019} and was downloaded directly from the Github repository associated with Calagaro et al. \cite{matteocalgaro2020}.

\noindent \textbf{Differential Abundance Analsysis}: To assess type I error using cILR scores in differential abundance analysis, we utilized the 16S rRNA gene sequencing of stool samples from the Human Microbiome Project \cite{consortium2012, proctor2019}. Here, we randomly assign samples a label of case or control, and repeated this process 500 times, assessing all candidate methods at each iteration. Type I error is then the number of taxa identified as differentially abundant across all tested taxa. For the true positive rate, we used the same gingival data set as described above. However, instead of testing for aerobic microbes as a group, the true positive rate is the number of aerobic/anaerobic genera identified as differentially abundant across all aerobic or anaerobic genera.

\noindent \textbf{Disease Prediction}: To assess predictive power, we utilized the whole genome sequencing of stool samples of inflammatory bowel disease (IBD) patients from the MetaHIT consortium \cite{nielsen2014}. This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease). Additionally, we also utilized a similar data set from Gevers et al. \cite{gevers2014} which also profiles the gut microbiome of IBD patients and controls but using 16S rRNA gene sequencing. This data set contains 671 16S rRNA gene sequencing samples from a cohort of ABC, where 500 was classified as having IBD.

\subsubsection*{Comparison Methods}

\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank-sum test. We added a pseudocount of 1 to all zero entry values. This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample. For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank-sum test.

\noindent \textbf{Differential Abundance}: Since cILR are sample-level enrichment scores, we performed differential abundance by using a Wilcoxon Rank Sum test and Welch's t-test across case/control status on cILR generated scores. We added a pseudocount of 1 to all zero entry values. For comparison, we chose representative state-of-the-art methods in differential abundance analysis, namely DESeq2 \cite{love2014,mcmurdie2014} and corncob \cite{martin2020}.

\noindent \textbf{Disease Prediction}: We fit random forest on cILR scores, as well as ssGSEA \cite{hanzelmann2013} and GSVA \cite{barbie2009} similar to single sample enrichment section. We added a pseudocount of 1 to all zero entry values. Additionally, we also compared performance using enrichment scores against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.

\section*{Results}
In this section, we present the performance of our proposed method for three applicable microbiome analysis tasks: sample level enrichment, differential abundance, and disease prediction. We obtained these results from both parametric simulations and examples from real data.

\subsection*{Enrichment testing at the sample level}
cILR provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{Methods}. Here, we present empirical results for this application of cILR assessing type I error, power, and classification capacity.

Panel A and B in figure \ref{fig:ss_hypo} demonstrate type I error and power respectively across different simulation conditions. We benchmarked the results of the cILR method against a naive Wilcoxon rank-sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement. All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions. However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted cILR variants show high levels of inflated type I error, where Wilcoxon test performed the worst. On the other hand, adjusted cILR methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.

However, the trade-off for good type I error control is demonstrably lower power, as shown in figure \ref{fig:ss_hypo}B. We included only adjusted cILR methods since power is only meaningful when type I error is properly controlled. Results indicated that both sparsity and inter-taxa correlation impacts power, correlation more so than sparsity. The adjusted cILR approaches were able to maintain power above 0.8 even in low effect size settings as long as there was negligible inter-taxa correlation within the set. However, in higher correlation scenarios ($\rho = 0.2$ and $0.5$), power was dramatically reduced to 0.25 and below and would require higher effect sizes to maintain power at 0.8, especially at higher sparsity levels.

These results were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}. Here, we tested the enrichment of aerobic microbes at each sample using approaches similar to our parametric simulations. As expected in Figure \ref{fig:ss_hypo}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods. Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05. Power analysis (Figure \ref{fig:ss_hypo}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.

To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set. This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated. Figure \ref{fig:ss_auc} presents this result. We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $U$ test statistic from the Wilcoxon rank-sum test. Across both simulations (Figure \ref{fig:ss_auc}A) and real-data applications (Figure \ref{fig:ss_auc}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios. In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2). Notably, the W-statistic provided the least information for classifying samples with inflated taxa.

\begin{figure}
    \centering
    \caption{Type I error rate \textbf{(A)}, and power \textbf{(B)} across different parametric simulation scenarios. Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach. False-positive rate \textbf{(C)} and true positive rate \textbf{(D)} evaluation of similar methods on real 16S rRNA data from the oral microbiome of the gingival site. For \textbf{(A)} and \textbf{(B)}, enrichment of a specified set was tested at the sample level using cILR and the Wilcoxon rank-sum test at $\alpha$ of 0.05. For \textbf{(C)} and \textbf{(D)}, the set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples. Adjusted cILR demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate. However, this resulted in lower power for adjusted cILR methods.}
    \label{fig:ss_hypo}
\end{figure}

\begin{figure}
    \centering
    \caption{Classification performance via AUC of cILR, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{Methods}. Performance scores measure whether scores can highly rank samples that are known to have inflated abundance. In the gingival data set presented in panel \textbf{(B)}, samples from the supragingival site are assumed to have an inflated abundance of aerobic microbes. Error bars are the 95\% DeLong confidence intervals for AUC \cite{delong1988}} 
    \label{fig:ss_auc}
\end{figure}

\subsection*{Differential abundance analysis}
Here we assessed the ability to use cILR with a standard difference of means statistical test (such as Welch's t-test or a Wilcoxon rank-sum test) to perform differential abundance analysis. We compare the performance of cILR related methods to two commonly used approaches in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.

\subsection*{Disease Prediction} 
We can conceive of cILR as a feature engineering step before model fit and prediction. Here we assessed the predictive performance of a standard machine learning model (random forest \cite{breiman2001}) using cILR scores compared to scores generated by other single sample enrichment methods in the gene set testing literature (ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}). Additionally, we also compared our results with just using the centered log-ratio (CLR) transformation on taxa sets aggregated via count summations.

Figure \ref{fig:sim_pred_auc} and \ref{fig:sim_pred_rsq} showed results from simulation studies as detailed in the \nameref{Methods} section under both classification and regression tasks. As expected, performance across all assessed methods increased with a higher signal-to-noise ratio, however, they remained consistent across different levels of model saturation and inter-taxa correlation. Importantly, cILR methods outperformed both GSVA and ssGSEA, where the difference in AUC increases as a function of sparsity. Conversely, cILR scores did not perform as well as the traditional CLR approach, however, the performance gap decreased at higher sparsity levels. Similar patterns were observed for regression results (Figure \ref{fig:sim_pred_rsq}). Most interestingly, regression performance, in general, decreased at higher saturation levels (proportion of sets associated with the outcome). cILR remained more performant than both ssGSEA and GSVA, but was still less than the standard CLR, especially at higher signal-to-noise ratios. Similarly, cILR still performed much better in higher sparsity situations, where the results were almost identical to the CLR approach, especially in high saturation scenarios.

In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:pred_data}. The learning task across both data sets is to classify patients who are diagnosed with inflammatory bowel disease (includes both Crohn's disease and ulcerative colitis) and healthy controls. The Gevers et al. \cite{gevers2014} data set is a 16S rRNA sequencing data set while the Nielsen et al. \cite{nielsen2014} is a whole-genome shotgun sequencing data set. Similar to simulation experiments, across both data sets cILR methods provide much better performance than both GSVA and ssGSEA. Interestingly, the standard CLR approach only outperformed cILR in the 16S rRNA data set but was marginally worse in the WGS data set. Additionally, we observed that using the normal distribution for cILR generates better predictive performance compared to the other variants.

These results demonstrated that cILR generated scores are informative features in disease prediction tasks. Simulation results indicated that cILR methods perform much better than both ssGSEA and GSVA, but not as good as the standard CLR approach. Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.

\begin{figure}
    \caption{AUC scores of a random forest model for a binary outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio. Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome). cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
    \label{fig:sim_pred_auc}
\end{figure}

\begin{figure}
    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio. Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome). cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
    \label{fig:sim_pred_rsq}
\end{figure}

\begin{figure}
    \caption{Classification performance of a standard random forest model using cILR scores compared against existing methods in gene set testing literature and the standard centered-log ratio transformation approach. The learning task involves predicting patients with inflammatory bowel disease (including Crohn's disease and ulcerative colitis) versus controls. Data sets used span both 16S rRNA sequencing (Gevers et al. \cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al. \cite{nielsen2014}). cILR marginally outperforms GSVA and ssGSEA in both data sets, but only performs better than CLR in the WGS data set.}
    \label{fig:pred_data}
\end{figure}

\section*{Discussion}
\subsection*{Set-based analysis can enhance microbiome data analysis}
Gene-set testing (or pathway analysis) is an important tool in the analysis of high-dimensional genomic data sets \cite{khatri2012}. By focusing on a group of variables rather than individuals, this approach reduces the hypothesis testing burden, thereby increasing power as well as replicability. Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also increase interpretability.

In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate variables to taxonomic classification levels such as genus, family, or phylum. The process of aggregation involves the element-wise summation of counts of taxa that belong to the same taxonomic group. Even though this allows for a reduction in the number of overall variables (from thousands to only hundreds), there still exist two disadvantages: first, it doesn't preserve inter-sample distances before and after aggregation \cite{egozcue2005}, and second, it doesn't allow for enrichment testing and comparison across sets of different sizes.

As such, there is a need for microbiome researchers to adopt set enrichment methods to perform a more robust set-based analysis that allows for enrichment testing of functionally relevant sets of microbes. Some software suites, such as \emph{MicrobiomeAnalyst}, have begun to provide an option to perform such analyses with curated taxa sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is based on a hypergeometric test of significant hypotheses and is sensitive to the method of differential abundance used. More recent methods in set-based testing generate enrichment scores at the sample level \cite{hanzelmann2013, frost2020a}. This approach allows for the flexible incorporation of different statistical techniques downstream, but unfortunately is not currently tailored to the compositional nature of microbiome relative abundance data \cite{gloor2017}.

Here, we developed cILR, a novel microbiome-specific method to perform set-based enrichment analysis at the sample level. Our method is based on the isometric log-ratio transformation \cite{egozcue2003}, and measures set enrichment relative to those outside the set, which corresponds to a $Q_1$ null hypothesis in the gene set analysis literature \cite{tian2005}. cILR generates sample-specific enrichment scores, enabling cILR to act as both a significance test for enrichment at the sample level, and as inputs for downstream analyses such as differential abundance testing, predictive modeling, and data visualization.

\subsection*{Significance testing with cILR}
We can use cILR to test for enrichment of taxa-sets at the sample level. Inter-taxa correlation can result in an inflated variance of resulting test statistics \cite{wu2012}, which results in an inflated type I error if not properly controlled for. Our method addresses this issue by combining the mean estimate of scores computed on permuted data and the variance estimate of scores computed on unpermuted data, where correlation is undisrupted. As a consequence, adjusted cILR methods control very well for type I error at the appropriate $\alpha$ level (Figure \ref{fig:ss_hypo}). The conservativeness of the test persisted across different sparsity levels, as well as inter-taxa correlation. Conversely, the unadjusted cILR methods and the naive Wilcoxon rank-sum test has inflated type I error even at low correlation levels ($\rho = 0.2$). However, the trade-off for proper type I error control is lower power. Simulation results showed that power is sensitive to sparsity across all correlation levels.

However, if there is no correlation between members of a set, cILR was still able to maintain 80\% power even with low effect size (fold change of 1.5), long as the set of interest is more densely populated, such as a set of core taxa that are prevalent across all samples. Unfortunately, this relationship doesn't hold for higher correlation levels. This means that in sets where associated taxa are highly correlated, a higher fold change is required for cILR to detect enrichment.

We also assessed cILR using different distribution forms, the normal distribution, and the mixture normal distribution. The difference in performance across the two choices was minimal, where both were able to control for type I error, but the normal distribution was slightly better when it comes to power only in certain simulation settings. This runs contrary to our distribution comparison analysis in Figure \ref{fig:kurt_gof}, where the mixture distribution was the superior fit. We hypothesized that this might be due to the difficulty in fitting mixture distributions using the expectation-maximization algorithm (EM), as the convergence rate is slow when there is a lot of overlap between the mixtures and the mixing coefficients of one of the components are low \cite{naim}. Furthermore, there is a lot of degeneracy of parameter values in our variance adjustment procedure for the mixture distribution(Supplemental Material). These distribution fitting problems might introduce issues where the mixture distribution might not be the best fit. As such, improvements to mixture distribution fitting might allow for better performance to be extracted.

These results are replicated in our real data analysis, where we attempted to test for enrichment for a group of aerobic taxa among samples in the gingival site from HMP \cite{consortium2012}. Even though our labeling is not perfect, it is encouraging to see that the false discovery rate is low (approximately 0.05) when tested among a set of hypotheses where there is a mixture of true null and alternate situations. We believe that a practitioner can identify many interesting samples to follow up on by applying multiple testing adjustment procedures (such as Benjamini-Hochberg \cite{benjamini1995}) with cILR p-values and using less stringent criteria (e.g. 0.1 false discovery rate).

In short, we demonstrated that cILR can be used to test for significance testing at the sample level. Even though there is not a lot of difference between the distribution choices, ultimately cILR controls for type I error well despite tradeoffs in power. Even then, cILR will still be able to detect enriched sets in high correlation and sparsity situations insofar as the effect size is large. Researchers can utilize cILR to test for enrichment of a certain set of taxa of interest to perform follow-up experiments or to identify the prevalence of certain sub-population of taxa (for example, oxygen preference as in our real data example). The conservativeness of the test ensures confidence in detected samples.

\subsection*{Using cILR as inputs for downstream analyses}
Since cILR generates enrichment scores for each set at the sample level, they can be used in downstream analyses alongside standard statistical approaches. In this section we assessed the utility of cILR generated scores in two common analysis tasks in microbiome research: differential abundance testing and disease prediction.

For disease prediction, we benchmarked a basic random forest model \cite{breiman2001} to predict whether or not a patient has IBD given only their microbiome profile using cILR and relevant methods as inputs. Across both simulations and real data analyses, cILR scores perform better than alternative single sample scoring methods in the gene set testing literature (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}). This is consistent with previous analysis (Figure \ref{fig:ss_auc}), where we assessed the informativeness of cILR generated scores by using AUC to measure whether samples with high scores correspond with the true label of an inflated set. Results indicated that cILR scores are more adaptive to microbiome data sets than GSVA and ssGSEA. However, cILR scores underperform the standard approach of applying the CLR transformation to count aggregated data, except for the WGS real data analysis. Despite this, there are still benefits to using cILR scores in the context of downstream modeling beyond random forest predictive models. First, CLR transformed variables have a sum to zero constraints, making the covariance matrix is singular \cite{gloor2017}, impacting methods that rely on matrix decomposition such as principal component regression or canonical correlation analysis. Second, cILR provides an additional value of significance testing for enrichment for sample screening. Third, CLR transformed variables still involve the count summation procedure for variable aggregation, therefore comparison between sets of different sizes remains unviable.

Despite such drawbacks, cILR methods still produce competitive performance values across both regression and classification tasks even in low signal-to-noise settings. Most interestingly, the performance gap between cILR and CLR was smaller in high sparsity settings ($p = 0.6$), which is a feature of microbiome relative abundance data \cite{li2019}. Furthermore, these predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), and good performance also indicated that cILR scores can also capture the joint distribution of considers sets, especially with regards to their relationship with a defined outcome such as IBD status.

Unsurprisingly, different variants of cILR did not differ significantly in performance similar to significance testing experiments. Even though we expect Z-scores to perform better due to the added information of directionality of enrichment, using CDF values did not alter performance in both the single sample AUC evaluations (Figure \ref{fig:ss_auc}) as well as the random forest evaluations (Figure \ref{fig:sim_pred_auc}).

In short, users can apply cILR generated enrichment scores as inputs to downstream analyses, especially predictive models and differential abundance analysis. cILR scores are informative in discriminating samples between classes even in low effect size situations. Even though cILR did not outperform a standard analysis of count aggregation followed by the CLR transformation, the additional benefits listed above would prove to be attractive in certain use cases, especially in high sparsity situations.

\subsection*{Limitations and future directions}
There are various limitations to using cILR to generate sample-level taxonomic aggregation for microbiome data. First, significance testing at the sample level is not powered to assess inflated sets with low effect size. Practitioners can be confident in the validity of a significant result (following proper multiple testing correction) due to cILR's type I error control, but should not expect cILR to be able to identify all samples with inflated counts. Second, even though cILR is flexible and addresses the compositional nature of microbiome data, the performance of cILR might not be consistently better than competing methods. Careful evaluation of methods should be done before application to ensure that the value of cILR scores is appropriately extracted in suitable models. Third, cILR did not directly address the zero-inflatedness of microbiome data and utilizes a pseudocount to ensure log operations are valid. As such, practitioners are encouraged to choose a different pseudocount or an appropriate method of choice.

As such, we hope to address the aforementioned issues with cILR in later projects. First, we hope to refine the distribution fitting procedure for the mixture distribution due to its superior fit and address the identifiability problem that might underlie the lack of power in our experiments. Second, we hope to incorporate some of the sophisticated model-based zero correction methods \cite{martin-fernandez2012,kaul2017a}. Third, we can account for the difference in importance of certain taxa by incorporating taxa-specific weights. Finally, we also hope to curate more interest apriori sets that are functionally informative, which would generate more interesting insight and improve interpretability compared to using taxonomic categories such as Phylum or Genus.

\section*{Conclusion}
Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomic data sets. However, there has not been a lot of set-based analysis methods developed specifically for microbiome relative abundance data. In this manuscript, we introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level. We demonstrated that our method can control for type I error for significance testing at the sample level, while generated scores are also valid inputs in downstream analyses, including disease prediction and differential abundance.

\section*{Supporting information}

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Fig.}
\label{S1_Fig}
{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).

\paragraph*{S2 Fig.}
\label{S2_Fig}
{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 File.}
\label{S1_File}
{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Video.}
\label{S1_Video}
{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Appendix.}
\label{S1_Appendix}
{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Table.}
\label{S1_Table}
{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\section*{Acknowledgments}
Cras egestas velit mauris, eu mollis turpis pellentesque sit amet. Interdum et malesuada fames ac ante ipsum primis in faucibus. Nam id pretium nisi. Sed ac quam id nisi malesuada congue. Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
%

%\bibliography{tax_agg}{}

\begin{thebibliography}{10}

    \bibitem{cho2012}
    Cho I, Blaser MJ.
    \newblock The Human Microbiome: At the Interface of Health and Disease.
    \newblock Nature Reviews Genetics. 2012;13(4):260--270.
    \newblock doi:{10.1038/nrg3182}.

    \bibitem{callahan2016}
    Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
    \newblock {{DADA2}}: {{High}}-Resolution Sample Inference from {{Illumina}}
      Amplicon Data.
    \newblock Nature Methods. 2016;13(7):581--583.
    \newblock doi:{10.1038/nmeth.3869}.

    \bibitem{truong2015}
    Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
    \newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
    \newblock Nature Methods. 2015;12(10):902--903.
    \newblock doi:{10.1038/nmeth.3589}.

    \bibitem{li2019}
    Li H.
    \newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
      {{Metagenomics}}.
    \newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
      2019. p. 977--550.

    \bibitem{li2015}
    Li H.
    \newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
      Compositional Data Analysis}}.
    \newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
    \newblock doi:{10.1146/annurev-statistics-010814-020351}.

    \bibitem{shi2016}
    Shi P, Zhang A, Li H.
    \newblock Regression Analysis for Microbiome Compositional Data.
    \newblock The Annals of Applied Statistics. 2016;10(2):1019--1040.
    \newblock doi:{10.1214/16-AOAS928}.

    \bibitem{sankaran2014}
    Sankaran K, Holmes S.
    \newblock {{structSSI}}: {{Simultaneous}} and {{Selective Inference}} for
      {{Grouped}} or {{Hierarchically Structured Data}}.
    \newblock Journal of statistical software. 2014;59(13):1--21.
    \newblock doi:{10.18637/jss.v059.i13}.

    \bibitem{benjamini1995}
    Benjamini Y, Hochberg Y.
    \newblock Controlling the {{False Discovery Rate}}: {{A Practical}} and
      {{Powerful Approach}} to {{Multiple Testing}}.
    \newblock Journal of the Royal Statistical Society Series B (Methodological).
      1995;57(1):289--300.

    \bibitem{chen2018}
    Chen J, King E, Deek R, Wei Z, Yu Y, Grill D, et~al.
    \newblock An Omnibus Test for Differential Distribution Analysis of Microbiome
      Sequencing Data.
    \newblock Bioinformatics. 2018;34(4):643--651.
    \newblock doi:{10.1093/bioinformatics/btx650}.

    \bibitem{weiss2017}
    Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
    \newblock Normalization and Microbial Differential Abundance Strategies Depend
      upon Data Characteristics.
    \newblock Microbiome. 2017;5(1).
    \newblock doi:{10.1186/s40168-017-0237-y}.

    \bibitem{mcmurdie2014}
    McMurdie PJ, Holmes S.
    \newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
      Inadmissible}}.
    \newblock PLOS Computational Biology. 2014;10(4):e1003531.
    \newblock doi:{10.1371/journal.pcbi.1003531}.

    \bibitem{love2014}
    Love MI, Huber W, Anders S.
    \newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
      Data with {{DESeq2}}.
    \newblock Genome Biology. 2014;15(12):550.
    \newblock doi:{10.1186/s13059-014-0550-8}.

    \bibitem{quinn2019}
    Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
    \newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
    \newblock GigaScience. 2019;8(giz107).
    \newblock doi:{10.1093/gigascience/giz107}.

    \bibitem{quinn2018b}
    Quinn TP, Erb I, Richardson MF, Crowley TM.
    \newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
    \newblock Bioinformatics. 2018;34(16):2870--2878.
    \newblock doi:{10.1093/bioinformatics/bty175}.

    \bibitem{gloor2017}
    Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
    \newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
      Optional}}.
    \newblock Frontiers in Microbiology. 2017;8.
    \newblock doi:{10.3389/fmicb.2017.02224}.

    \bibitem{aitchison1999}
    Aitchison J.
    \newblock A {{Concise Guide}} to {{Compositional Data Analysis}}. 1999; p. 134.

    \bibitem{kurtz2015}
    Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
    \newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
      Ecological Networks}}.
    \newblock PLOS Computational Biology. 2015;11(5):e1004226.
    \newblock doi:{10.1371/journal.pcbi.1004226}.

    \bibitem{kaul2017}
    Kaul A, Davidov O, Peddada SD.
    \newblock Structural Zeros in High-Dimensional Data with Applications to
      Microbiome Studies.
    \newblock Biostatistics. 2017;18(3):422--433.
    \newblock doi:{10.1093/biostatistics/kxw053}.

    \bibitem{kaul2017a}
    Kaul A, Mandal S, Davidov O, Peddada SD.
    \newblock Analysis of {{Microbiome Data}} in the {{Presence}} of {{Excess
      Zeros}}.
    \newblock Frontiers in Microbiology. 2017;8.
    \newblock doi:{10.3389/fmicb.2017.02114}.

    \bibitem{egozcue2005}
    Egozcue JJ, {Pawlowsky-Glahn} V.
    \newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
      Analysis}}.
    \newblock Mathematical Geology. 2005;37(7):795--828.
    \newblock doi:{10.1007/s11004-005-7381-9}.

    \bibitem{goeman2007}
    Goeman JJ, B{\"u}hlmann P.
    \newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
      Issues.
    \newblock Bioinformatics. 2007;23(8):980--987.
    \newblock doi:{10.1093/bioinformatics/btm051}.

    \bibitem{subramanian2005}
    Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
    \newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
      Interpreting Genome-Wide Expression Profiles.
    \newblock Proceedings of the National Academy of Sciences.
      2005;102(43):15545--15550.
    \newblock doi:{10.1073/pnas.0506580102}.

    \bibitem{hanzelmann2013}
    H{\"a}nzelmann S, Castelo R, Guinney J.
    \newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
      {{RNA}}-{{Seq}} Data.
    \newblock BMC Bioinformatics. 2013;14(1):7.
    \newblock doi:{10.1186/1471-2105-14-7}.

    \bibitem{frost2020a}
    Frost HR.
    \newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
      Method for Cell-Specific Gene Set Scoring.
    \newblock Nucleic Acids Research. 2020;48(16):e94--e94.
    \newblock doi:{10.1093/nar/gkaa582}.

    \bibitem{tian2005}
    Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
    \newblock Discovering Statistically Significant Pathways in Expression
      Profiling Studies.
    \newblock Proceedings of the National Academy of Sciences.
      2005;102(38):13544--13549.
    \newblock doi:{10.1073/pnas.0506577102}.

    \bibitem{egozcue2003}
    Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcelo-Vidal} C.
    \newblock Isometric {{Logratio Transformations}} for {{Compositional Data
      Analysis}}.
    \newblock Mathematical Geology. 2003; p.~22.

    \bibitem{barbie2009}
    Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
    \newblock Systematic {{RNA}} Interference Reveals That Oncogenic
      {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
    \newblock Nature. 2009;462(7269):108--112.
    \newblock doi:{10.1038/nature08460}.

    \bibitem{quast2013}
    Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
    \newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
      Processing and Web-Based Tools.
    \newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
    \newblock doi:{10.1093/nar/gks1219}.

    \bibitem{delignette-muller2015}
    {Delignette-Muller} ML, Dutang C.
    \newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
    \newblock Journal of Statistical Software. 2015;64(4):1--34.

    \bibitem{benaglia2009}
    Benaglia T, Chauveau D, Hunter DR, Young D.
    \newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
    \newblock Journal of Statistical Software. 2009;32(6):1--29.

    \bibitem{washburne2017}
    Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL, Mukherjee S, et~al.
    \newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
      Associations in Microbiome Datasets.
    \newblock PeerJ. 2017;5:e2969.
    \newblock doi:{10.7717/peerj.2969}.

    \bibitem{silverman2017}
    Silverman JD, Washburne AD, Mukherjee S, David LA.
    \newblock A Phylogenetic Transform Enhances Analysis of Compositional
      Microbiota Data.
    \newblock eLife. 2017;6:e21887.
    \newblock doi:{10.7554/eLife.21887}.

    \bibitem{morton2017}
    Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
      et~al.
    \newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
    \newblock mSystems. 2017;2(1).
    \newblock doi:{10.1128/mSystems.00162-16}.

    \bibitem{efron2004}
    Efron B.
    \newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
    \newblock Journal of the American Statistical Association.
      2004;99(465):96--104.
    \newblock doi:{10.1198/016214504000000089}.

    \bibitem{wu2012}
    Wu D, Smyth GK.
    \newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
      Correlation.
    \newblock Nucleic Acids Research. 2012;40(17):e133.
    \newblock doi:{10.1093/nar/gks461}.

    \bibitem{calgaro2020}
    Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
    \newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
      and Metagenomics Applied to Microbiome Data.
    \newblock Genome Biology. 2020;21(1):191.
    \newblock doi:{10.1186/s13059-020-02104-1}.

    \bibitem{cario1997}
    Cario MC.
    \newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
      Distributions}} and {{Correlation Matrix}}. 1997; p.~19.

    \bibitem{agresti1998}
    Agresti A, Coull BA.
    \newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
      Estimation}} of {{Binomial Proportions}}.
    \newblock The American Statistician. 1998;52(2):119--126.
    \newblock doi:{10.2307/2685469}.

    \bibitem{delong1988}
    DeLong ER, DeLong DM, {Clarke-Pearson} DL.
    \newblock Comparing the Areas under Two or More Correlated Receiver Operating
      Characteristic Curves: A Nonparametric Approach.
    \newblock Biometrics. 1988;44(3):837--845.

    \bibitem{hawinkel2019}
    Hawinkel S, Mattiello F, Bijnens L, Thas O.
    \newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
      Control the False Discovery Rate.
    \newblock Briefings in Bioinformatics. 2019;20(1):210--221.
    \newblock doi:{10.1093/bib/bbx104}.

    \bibitem{breiman2001}
    Breiman L.
    \newblock Random {{Forests}}.
    \newblock Machine Learning. 2001;45(1):5--32.
    \newblock doi:{10.1023/A:1010933404324}.

    \bibitem{kuhn2020}
    Kuhn M, Wickham H.
    \newblock Tidymodels: {{Easily}} Install and Load the 'tidymodels' Packages;
      2020.

    \bibitem{xiao2018}
    Xiao J, Chen L, Yu Y, Zhang X, Chen J.
    \newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
      {{Predictive Modeling}} of {{Microbial Community Data}}.
    \newblock Frontiers in Microbiology. 2018;9.
    \newblock doi:{10.3389/fmicb.2018.03112}.

    \bibitem{dong2020}
    Dong M, Li L, Chen M, Kusalik A, Xu W.
    \newblock Predictive Analysis Methods for Human Microbiome Data with
      Application to {{Parkinson}}'s Disease.
    \newblock PLOS ONE. 2020;15(8):e0237779.
    \newblock doi:{10.1371/journal.pone.0237779}.

    \bibitem{pasolli2017}
    Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
    \newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
    \newblock Nature Methods. 2017;14(11):1023--1024.
    \newblock doi:{10.1038/nmeth.4468}.

    \bibitem{schiffer2019}
    Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
    \newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
      through Bioconductor.
    \newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.

    \bibitem{gonzalez2018}
    Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
      Ackermann G, et~al.
    \newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
    \newblock Nature Methods. 2018;15(10):796--798.
    \newblock doi:{10.1038/s41592-018-0141-9}.

    \bibitem{consortium2012}
    Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
      et~al.
    \newblock Structure, Function and Diversity of the Healthy Human Microbiome.
    \newblock Nature. 2012;486(7402):207--214.
    \newblock doi:{10.1038/nature11234}.

    \bibitem{proctor2019}
    Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
    \newblock The {{Integrative Human Microbiome Project}}.
    \newblock Nature. 2019;569(7758):641--648.
    \newblock doi:{10.1038/s41586-019-1238-8}.

    \bibitem{thurnheer2016}
    Thurnheer T, Bostanci N, Belibasakis GN.
    \newblock Microbial Dynamics during Conversion from Supragingival to
      Subgingival Biofilms in an in Vitro Model.
    \newblock Molecular Oral Microbiology. 2016;31(2):125--135.
    \newblock doi:{10.1111/omi.12108}.

    \bibitem{beghini2019}
    Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
    \newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
      in the {{New York City Health}} and {{Nutrition Examination Study}}.
    \newblock Annals of Epidemiology. 2019;34:18--25.e3.
    \newblock doi:{10.1016/j.annepidem.2019.03.005}.

    \bibitem{matteocalgaro2020}
    Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
    \newblock Zenodo.

    \bibitem{nielsen2014}
    Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
    \newblock Identification and Assembly of Genomes and Genetic Elements in
      Complex Metagenomic Samples without Using Reference Genomes.
    \newblock Nature Biotechnology. 2014;32(8):822--828.
    \newblock doi:{10.1038/nbt.2939}.

    \bibitem{gevers2014}
    Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
      et~al.
    \newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
      {{Disease}}.
    \newblock Cell Host \& Microbe. 2014;15(3):382--392.
    \newblock doi:{10.1016/j.chom.2014.02.005}.

    \bibitem{martin2020}
    Martin BD, Witten D, Willis AD.
    \newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
      Regression.
    \newblock The Annals of Applied Statistics. 2020;14(1):94--115.
    \newblock doi:{10.1214/19-AOAS1283}.

    \bibitem{khatri2012}
    Khatri P, Sirota M, Butte AJ.
    \newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
      {{Outstanding Challenges}}.
    \newblock PLOS Computational Biology. 2012;8(2):e1002375.
    \newblock doi:{10.1371/journal.pcbi.1002375}.

    \bibitem{ashburner2000}
    Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
    \newblock Gene {{Ontology}}: Tool for the Unification of Biology.
    \newblock Nature genetics. 2000;25(1):25--29.
    \newblock doi:{10.1038/75556}.

    \bibitem{chong2020}
    Chong J, Liu P, Zhou G, Xia J.
    \newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
      Functional, and Meta-Analysis of Microbiome Data.
    \newblock Nature Protocols. 2020;15(3):799--821.
    \newblock doi:{10.1038/s41596-019-0264-1}.

    \bibitem{naim}
    Naim I, Gildea D.
    \newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
      {{Unbalanced Mixing Coefficients}}; p.~8.

    \bibitem{martin-fernandez2012}
    {Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
      {Palarea-Albaladejo} J.
    \newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
      {{Classical}} and Robust Approaches.
    \newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
    \newblock doi:{10.1016/j.csda.2012.02.012}.

    \end{thebibliography}

\end{document}

### 2021-05-27 supplemental.tex
First commit

\documentclass{article}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{float}
\usepackage[parfill]{parskip}

\author{Quang P. Nguyen}
\title{Supplementary Materials}
\begin{document}
\maketitle
\captionsetup[figure]{labelfont={bf},name={Figure},labelsep=period, margin=1cm}

\section{Performance Evaluation} 
\section{Distribution fitting}
\section{p-value behaviour}

\newpage
\bibliography{tax_agg}{}
\bibliographystyle{plain}

\end{document}

### 2021-05-27 tax_agg.tex
First commit

--- 2021-05-27 manuscript_plos.tex
@@
-% Template for PLoS
-% Version 3.5 March 2018
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
+\documentclass{article}
+\usepackage{amsfonts}
+\usepackage{graphicx}
+\usepackage[margin=1in]{geometry}
+\usepackage{bm}
+\usepackage{amsmath}
+\usepackage{authblk}
+\usepackage{caption}
+\usepackage{hyperref}
+\usepackage{float}
+\usepackage[parfill]{parskip}
+%\renewcommand{\familydefault}{\sfdefault}
 
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-%\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-
-%% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
-
-%% END MACROS SECTION
-
-% Adding packages 
-\usepackage{bm}
-\usepackage[parfill]{parskip}
+\title{cILR: Taxonomic Enrichment Analysis with Isometric Log-Ratios}
+\author[1,2]{Quang P. Nguyen}
+\author[1,2]{Anne G. Hoen}
+\author[1]{H. Robert Frost}
+\affil[1]{Department of Biomedical Data Science, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA}
+\affil[2]{Department of Epidemiology, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA}
+\date{}                     %% if you don't need a date to appear
+\setcounter{Maxaffil}{0}
+\renewcommand\Affilfont{\itshape\small}
 
 \begin{document}
-\vspace*{0.2in}
+\maketitle
+\captionsetup[figure]{labelfont={bf},name={Figure},labelsep=period, margin=1cm}
 
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{cILR: Taxonomic enrichment analysis with isometric log-ratios} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Quang Nguyen\textsuperscript{1,2},
-Anne G. Hoen\textsuperscript{1,2\ddag},
-H. Robert Frost\textsuperscript{1*\ddag},
-\\
-\bigskip
-\textbf{1} Department of Biomedical Data Science, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA
-\\
-\textbf{2} Department of Epidemiology, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA
-\\
-\bigskip
+\begin{abstract}
+    \noindent Research in human associated microbiomes primarily involve high-throughput sequencing of samples, resulting in count tables of taxa abundances.
+Unfortunately, the analysis microbiome data is complex as it is high-dimensional, sparse, and strictly compositional.
+Various statistical tools have been developed to address these issues, but a very approachable method is to aggregate taxa into pre-defined sets, which is called gene set testing or enrichment analysis in the genomics literature.
+However, there has been a considerable lack of interest in developing set-based analysis methods tailored specifically for microbiome relative abundance data.
+Here, we present a new sample-level taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis in the gene set testing literature.
+Our approach, titled competitive isometric log ratio (cILR), generates enrichment scores per sample as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
+We also provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
+Herein we demonstrate under both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios, while additionally providing informative scores that can be inputs to downstream differential abundance and predictive tasks with good performance.
+These results demonstrate how our approach can be used to generate compositionally valid taxonomic aggregation and enrichment.
 
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-%\Yinyang These authors contributed equally to this work.
+\end{abstract}
 
-% Additional Equal Contribution Note
-% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-\ddag These authors jointly supervised this work.
-
-% Current address notes
-%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
-% \textcurrency b Insert second current address 
-% \textcurrency c Insert third current address
-
-% Deceased author note
-% \dag Deceased
-
-% Group/Consortium Author Note
-%\textpilcrow Membership list can be found in the Acknowledgments section.
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* hildreth.r.frost@dartmouth.edu
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-\section*{Author summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
-
-\linenumbers
-
-% Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
+\section*{Background}  
 The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their human host.
-Previous research has shown that changes in the composition of the microbiome have been associated with important health outcomes such as inflammatory bowel disease, type II diabetes, and obesity.
+Previous research has shown that changes in the composition of the microbiome have been associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
 To understand the central role of the microbiome in human health, researchers often relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
@@
 
-\section{Material and Methods} \label{Methods}
-\subsection{Competitive Isometric Log-ratio (cILR)}
+\section*{Materials and Methods} \label{Methods}
+\subsection*{Competitive Isometric Log-ratio (cILR)}
 The cILR method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}.
@@
 
+\begin{figure} [H]
+    \centering
+    \includegraphics[width=\linewidth]{figures/kurtosis_skewness_gof.png}
+    \caption{Properties of the null distribution of cILR in different simulation scenarios under the global null.
+Panel \textbf{(B)} presents kurtosis and skewness of cILR scores while panel \textbf{(A)} presents the goodness of fit (as Kolmogorov-Smirnov D statistic) for mixture normal and normal distributions.
+Panel \textbf{(C)} is a density plot of the shape of the null distribution.
+Results indicated the necessity of estimating an empirical null and demonstrating that the mixture distribution was the better fit compared to the basic normal.}
+    \label{fig:kurt_gof}
+\end{figure}
+
 Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null.
@@
 However, this procedure assumes that the variance of cILR is equal under both the null and alternate hypotheses.
-
-\begin{figure}
-    \centering
-    \caption{Properties of the null distribution of cILR in different simulation scenarios under the global null.
-Panel \textbf{(B)} presents kurtosis and skewness of cILR scores while panel \textbf{(A)} presents the goodness of fit (as Kolmogorov-Smirnov D statistic) for mixture normal and normal distributions.
-Panel \textbf{(C)} is a density plot of the shape of the null distribution.
-Results indicated the necessity of estimating an empirical null and demonstrating that the mixture distribution was the better fit compared to the basic normal.}
-    \label{fig:kurt_gof}
-\end{figure}
 
@@
 For binary outcomes, we generate $Y_{binary}$ as Bernoulli draws with probability $p_{binary}$, where 
-$$p_{binary} = \frac{1}{1 + \exp(f(\mathbf{X}) + \mathbf{\epsilon})}$$
+\begin{equation}
+    p_{binary} = \frac{1}{1 + \exp(f(\mathbf{X}) + \mathbf{\epsilon})}
+\end{equation}
 To ensure a balance of classes, we applied the strategy described in Dong et al.
@@
 \cite{gevers2014} which also profiles the gut microbiome of IBD patients and controls but using 16S rRNA gene sequencing.
-This data set contains 671 16S rRNA gene sequencing samples from a cohort of ABC, where 500 was classified as having IBD.
+This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages < 17) from the RISK cohort enrolled in North America.
+Of the 671 samples obtained, 500 samples belong to patients with IBD.
 
@@
 
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
+    \caption{Type I error rate \textbf{(A)}, and power \textbf{(B)} across different parametric simulation scenarios.
+Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach.
+False-positive rate \textbf{(C)} and true positive rate \textbf{(D)} evaluation of similar methods on real 16S rRNA data from the oral microbiome of the gingival site.
+For \textbf{(A)} and \textbf{(B)}, enrichment of a specified set was tested at the sample level using cILR and the Wilcoxon rank-sum test at $\alpha$ of 0.05.
+For \textbf{(C)} and \textbf{(D)}, the set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples.
+Adjusted cILR demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate.
+However, this resulted in lower power for adjusted cILR methods.}
+    \label{fig:ss_hypo}
+\end{figure}
+
 However, the trade-off for good type I error control is demonstrably lower power, as shown in figure \ref{fig:ss_hypo}B.
-We included only adjusted cILR methods since power is only meaningful when type I error is properly controlled.
-Results indicated that both sparsity and inter-taxa correlation impacts power, correlation more so than sparsity.
-The adjusted cILR approaches were able to maintain power above 0.8 even in low effect size settings as long as there was negligible inter-taxa correlation within the set.
-However, in higher correlation scenarios ($\rho = 0.2$ and $0.5$), power was dramatically reduced to 0.25 and below and would require higher effect sizes to maintain power at 0.8, especially at higher sparsity levels.
+In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un adjusted ones.
+However, in higher correlation scenarios, the difference in power mus much more dramatic.
+At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted cILR was only perfomring at 50\% power, while unadjusted cILR and wilcoxon rank sum test were able to reach 80\%.
+These results indicated that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of cILR.
 
-These results were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}.
+These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}.
 Here, we tested the enrichment of aerobic microbes at each sample using approaches similar to our parametric simulations.
@@
 
-\begin{figure}
+\begin{figure}[H]
     \centering
-    \caption{Type I error rate \textbf{(A)}, and power \textbf{(B)} across different parametric simulation scenarios.
-Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach.
-False-positive rate \textbf{(C)} and true positive rate \textbf{(D)} evaluation of similar methods on real 16S rRNA data from the oral microbiome of the gingival site.
-For \textbf{(A)} and \textbf{(B)}, enrichment of a specified set was tested at the sample level using cILR and the Wilcoxon rank-sum test at $\alpha$ of 0.05.
-For \textbf{(C)} and \textbf{(D)}, the set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples.
-Adjusted cILR demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate.
-However, this resulted in lower power for adjusted cILR methods.}
-    \label{fig:ss_hypo}
-\end{figure}
-
-\begin{figure}
-    \centering
+    \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
     \caption{Classification performance via AUC of cILR, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{Methods}.
@@
 
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{figures/sim_pred_auc.png}
+    \caption{AUC scores of a random forest model for a binary outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio.
+Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome).
+cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
+    \label{fig:sim_pred_auc}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{figures/sim_pred_rsq.png}
+    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio.
+Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome).
+cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
+    \label{fig:sim_pred_rsq}
+\end{figure}
+
 In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:pred_data}.
@@
 
-These results demonstrated that cILR generated scores are informative features in disease prediction tasks.
-Simulation results indicated that cILR methods perform much better than both ssGSEA and GSVA, but not as good as the standard CLR approach.
-Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
-
-\begin{figure}
-    \caption{AUC scores of a random forest model for a binary outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio.
-Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome).
-cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
-    \label{fig:sim_pred_auc}
-\end{figure}
-
-\begin{figure}
-    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio.
-Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome).
-cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
-    \label{fig:sim_pred_rsq}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[H]
+    \centering
+    \includegraphics[scale = 0.6]{figures/data_prediction_plot.png}
     \caption{Classification performance of a standard random forest model using cILR scores compared against existing methods in gene set testing literature and the standard centered-log ratio transformation approach.
@@
 \end{figure}
+
+These results demonstrated that cILR generated scores are informative features in disease prediction tasks.
+Simulation results indicated that cILR methods perform much better than both ssGSEA and GSVA, but not as good as the standard CLR approach.
+Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
 
@@
 This runs contrary to our distribution comparison analysis in Figure \ref{fig:kurt_gof}, where the mixture distribution was the superior fit.
-We hypothesized that this might be due to the difficulty in fitting mixture distributions using the expectation-maximization algorithm (EM), as the convergence rate is slow when there is a lot of overlap between the mixtures and the mixing coefficients of one of the components are low \cite{naim}.
+We hypothesized that this might be due to the difficulty in fitting mixture distributions using the expectation-maximization algorithm (EM), as the convergence rate is slow when there is a lot of overlap between the mixtures and the mixing coefficients of one of the components are low \cite{naim2012}.
 Furthermore, there is a lot of degeneracy of parameter values in our variance adjustment procedure for the mixture distribution(Supplemental Material).
@@
 
-\section*{Supporting information}
-
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Fig.}
-\label{S1_Fig}
-{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
-
-\paragraph*{S2 Fig.}
-\label{S2_Fig}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 File.}
-\label{S1_File}
-{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 Video.}
-\label{S1_Video}
-{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 Appendix.}
-\label{S1_Appendix}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 Table.}
-\label{S1_Table}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\section*{Acknowledgments}
-Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
-Interdum et malesuada fames ac ante ipsum primis in faucibus.
-Nam id pretium nisi.
-Sed ac quam id nisi malesuada congue.
-Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.
-
-\nolinenumbers
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-%\bibliography{tax_agg}{}
-
-\begin{thebibliography}{10}
-
-    \bibitem{cho2012}
-    Cho I, Blaser MJ.
-    \newblock The Human Microbiome: At the Interface of Health and Disease.
-    \newblock Nature Reviews Genetics. 2012;13(4):260--270.
-    \newblock doi:{10.1038/nrg3182}.
-
-    \bibitem{callahan2016}
-    Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
-    \newblock {{DADA2}}: {{High}}-Resolution Sample Inference from {{Illumina}}
-      Amplicon Data.
-    \newblock Nature Methods. 2016;13(7):581--583.
-    \newblock doi:{10.1038/nmeth.3869}.
-
-    \bibitem{truong2015}
-    Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
-    \newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
-    \newblock Nature Methods. 2015;12(10):902--903.
-    \newblock doi:{10.1038/nmeth.3589}.
-
-    \bibitem{li2019}
-    Li H.
-    \newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
-      {{Metagenomics}}.
-    \newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
-      2019. p. 977--550.
-
-    \bibitem{li2015}
-    Li H.
-    \newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
-      Compositional Data Analysis}}.
-    \newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
-    \newblock doi:{10.1146/annurev-statistics-010814-020351}.
-
-    \bibitem{shi2016}
-    Shi P, Zhang A, Li H.
-    \newblock Regression Analysis for Microbiome Compositional Data.
-    \newblock The Annals of Applied Statistics. 2016;10(2):1019--1040.
-    \newblock doi:{10.1214/16-AOAS928}.
-
-    \bibitem{sankaran2014}
-    Sankaran K, Holmes S.
-    \newblock {{structSSI}}: {{Simultaneous}} and {{Selective Inference}} for
-      {{Grouped}} or {{Hierarchically Structured Data}}.
-    \newblock Journal of statistical software. 2014;59(13):1--21.
-    \newblock doi:{10.18637/jss.v059.i13}.
-
-    \bibitem{benjamini1995}
-    Benjamini Y, Hochberg Y.
-    \newblock Controlling the {{False Discovery Rate}}: {{A Practical}} and
-      {{Powerful Approach}} to {{Multiple Testing}}.
-    \newblock Journal of the Royal Statistical Society Series B (Methodological).
-      1995;57(1):289--300.
-
-    \bibitem{chen2018}
-    Chen J, King E, Deek R, Wei Z, Yu Y, Grill D, et~al.
-    \newblock An Omnibus Test for Differential Distribution Analysis of Microbiome
-      Sequencing Data.
-    \newblock Bioinformatics. 2018;34(4):643--651.
-    \newblock doi:{10.1093/bioinformatics/btx650}.
-
-    \bibitem{weiss2017}
-    Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
-    \newblock Normalization and Microbial Differential Abundance Strategies Depend
-      upon Data Characteristics.
-    \newblock Microbiome. 2017;5(1).
-    \newblock doi:{10.1186/s40168-017-0237-y}.
-
-    \bibitem{mcmurdie2014}
-    McMurdie PJ, Holmes S.
-    \newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
-      Inadmissible}}.
-    \newblock PLOS Computational Biology. 2014;10(4):e1003531.
-    \newblock doi:{10.1371/journal.pcbi.1003531}.
-
-    \bibitem{love2014}
-    Love MI, Huber W, Anders S.
-    \newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
-      Data with {{DESeq2}}.
-    \newblock Genome Biology. 2014;15(12):550.
-    \newblock doi:{10.1186/s13059-014-0550-8}.
-
-    \bibitem{quinn2019}
-    Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
-    \newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
-    \newblock GigaScience. 2019;8(giz107).
-    \newblock doi:{10.1093/gigascience/giz107}.
-
-    \bibitem{quinn2018b}
-    Quinn TP, Erb I, Richardson MF, Crowley TM.
-    \newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
-    \newblock Bioinformatics. 2018;34(16):2870--2878.
-    \newblock doi:{10.1093/bioinformatics/bty175}.
-
-    \bibitem{gloor2017}
-    Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
-    \newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
-      Optional}}.
-    \newblock Frontiers in Microbiology. 2017;8.
-    \newblock doi:{10.3389/fmicb.2017.02224}.
-
-    \bibitem{aitchison1999}
-    Aitchison J.
-    \newblock A {{Concise Guide}} to {{Compositional Data Analysis}}. 1999; p. 134.
-
-    \bibitem{kurtz2015}
-    Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
-    \newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
-      Ecological Networks}}.
-    \newblock PLOS Computational Biology. 2015;11(5):e1004226.
-    \newblock doi:{10.1371/journal.pcbi.1004226}.
-
-    \bibitem{kaul2017}
-    Kaul A, Davidov O, Peddada SD.
-    \newblock Structural Zeros in High-Dimensional Data with Applications to
-      Microbiome Studies.
-    \newblock Biostatistics. 2017;18(3):422--433.
-    \newblock doi:{10.1093/biostatistics/kxw053}.
-
-    \bibitem{kaul2017a}
-    Kaul A, Mandal S, Davidov O, Peddada SD.
-    \newblock Analysis of {{Microbiome Data}} in the {{Presence}} of {{Excess
-      Zeros}}.
-    \newblock Frontiers in Microbiology. 2017;8.
-    \newblock doi:{10.3389/fmicb.2017.02114}.
-
-    \bibitem{egozcue2005}
-    Egozcue JJ, {Pawlowsky-Glahn} V.
-    \newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
-      Analysis}}.
-    \newblock Mathematical Geology. 2005;37(7):795--828.
-    \newblock doi:{10.1007/s11004-005-7381-9}.
-
-    \bibitem{goeman2007}
-    Goeman JJ, B{\"u}hlmann P.
-    \newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
-      Issues.
-    \newblock Bioinformatics. 2007;23(8):980--987.
-    \newblock doi:{10.1093/bioinformatics/btm051}.
-
-    \bibitem{subramanian2005}
-    Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
-    \newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
-      Interpreting Genome-Wide Expression Profiles.
-    \newblock Proceedings of the National Academy of Sciences.
-      2005;102(43):15545--15550.
-    \newblock doi:{10.1073/pnas.0506580102}.
-
-    \bibitem{hanzelmann2013}
-    H{\"a}nzelmann S, Castelo R, Guinney J.
-    \newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
-      {{RNA}}-{{Seq}} Data.
-    \newblock BMC Bioinformatics. 2013;14(1):7.
-    \newblock doi:{10.1186/1471-2105-14-7}.
-
-    \bibitem{frost2020a}
-    Frost HR.
-    \newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
-      Method for Cell-Specific Gene Set Scoring.
-    \newblock Nucleic Acids Research. 2020;48(16):e94--e94.
-    \newblock doi:{10.1093/nar/gkaa582}.
-
-    \bibitem{tian2005}
-    Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
-    \newblock Discovering Statistically Significant Pathways in Expression
-      Profiling Studies.
-    \newblock Proceedings of the National Academy of Sciences.
-      2005;102(38):13544--13549.
-    \newblock doi:{10.1073/pnas.0506577102}.
-
-    \bibitem{egozcue2003}
-    Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcelo-Vidal} C.
-    \newblock Isometric {{Logratio Transformations}} for {{Compositional Data
-      Analysis}}.
-    \newblock Mathematical Geology. 2003; p.~22.
-
-    \bibitem{barbie2009}
-    Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
-    \newblock Systematic {{RNA}} Interference Reveals That Oncogenic
-      {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
-    \newblock Nature. 2009;462(7269):108--112.
-    \newblock doi:{10.1038/nature08460}.
-
-    \bibitem{quast2013}
-    Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
-    \newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
-      Processing and Web-Based Tools.
-    \newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
-    \newblock doi:{10.1093/nar/gks1219}.
-
-    \bibitem{delignette-muller2015}
-    {Delignette-Muller} ML, Dutang C.
-    \newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
-    \newblock Journal of Statistical Software. 2015;64(4):1--34.
-
-    \bibitem{benaglia2009}
-    Benaglia T, Chauveau D, Hunter DR, Young D.
-    \newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
-    \newblock Journal of Statistical Software. 2009;32(6):1--29.
-
-    \bibitem{washburne2017}
-    Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL, Mukherjee S, et~al.
-    \newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
-      Associations in Microbiome Datasets.
-    \newblock PeerJ. 2017;5:e2969.
-    \newblock doi:{10.7717/peerj.2969}.
-
-    \bibitem{silverman2017}
-    Silverman JD, Washburne AD, Mukherjee S, David LA.
-    \newblock A Phylogenetic Transform Enhances Analysis of Compositional
-      Microbiota Data.
-    \newblock eLife. 2017;6:e21887.
-    \newblock doi:{10.7554/eLife.21887}.
-
-    \bibitem{morton2017}
-    Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
-      et~al.
-    \newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
-    \newblock mSystems. 2017;2(1).
-    \newblock doi:{10.1128/mSystems.00162-16}.
-
-    \bibitem{efron2004}
-    Efron B.
-    \newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
-    \newblock Journal of the American Statistical Association.
-      2004;99(465):96--104.
-    \newblock doi:{10.1198/016214504000000089}.
-
-    \bibitem{wu2012}
-    Wu D, Smyth GK.
-    \newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
-      Correlation.
-    \newblock Nucleic Acids Research. 2012;40(17):e133.
-    \newblock doi:{10.1093/nar/gks461}.
-
-    \bibitem{calgaro2020}
-    Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
-    \newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
-      and Metagenomics Applied to Microbiome Data.
-    \newblock Genome Biology. 2020;21(1):191.
-    \newblock doi:{10.1186/s13059-020-02104-1}.
-
-    \bibitem{cario1997}
-    Cario MC.
-    \newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
-      Distributions}} and {{Correlation Matrix}}. 1997; p.~19.
-
-    \bibitem{agresti1998}
-    Agresti A, Coull BA.
-    \newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
-      Estimation}} of {{Binomial Proportions}}.
-    \newblock The American Statistician. 1998;52(2):119--126.
-    \newblock doi:{10.2307/2685469}.
-
-    \bibitem{delong1988}
-    DeLong ER, DeLong DM, {Clarke-Pearson} DL.
-    \newblock Comparing the Areas under Two or More Correlated Receiver Operating
-      Characteristic Curves: A Nonparametric Approach.
-    \newblock Biometrics. 1988;44(3):837--845.
-
-    \bibitem{hawinkel2019}
-    Hawinkel S, Mattiello F, Bijnens L, Thas O.
-    \newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
-      Control the False Discovery Rate.
-    \newblock Briefings in Bioinformatics. 2019;20(1):210--221.
-    \newblock doi:{10.1093/bib/bbx104}.
-
-    \bibitem{breiman2001}
-    Breiman L.
-    \newblock Random {{Forests}}.
-    \newblock Machine Learning. 2001;45(1):5--32.
-    \newblock doi:{10.1023/A:1010933404324}.
-
-    \bibitem{kuhn2020}
-    Kuhn M, Wickham H.
-    \newblock Tidymodels: {{Easily}} Install and Load the 'tidymodels' Packages;
-      2020.
-
-    \bibitem{xiao2018}
-    Xiao J, Chen L, Yu Y, Zhang X, Chen J.
-    \newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
-      {{Predictive Modeling}} of {{Microbial Community Data}}.
-    \newblock Frontiers in Microbiology. 2018;9.
-    \newblock doi:{10.3389/fmicb.2018.03112}.
-
-    \bibitem{dong2020}
-    Dong M, Li L, Chen M, Kusalik A, Xu W.
-    \newblock Predictive Analysis Methods for Human Microbiome Data with
-      Application to {{Parkinson}}'s Disease.
-    \newblock PLOS ONE. 2020;15(8):e0237779.
-    \newblock doi:{10.1371/journal.pone.0237779}.
-
-    \bibitem{pasolli2017}
-    Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
-    \newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
-    \newblock Nature Methods. 2017;14(11):1023--1024.
-    \newblock doi:{10.1038/nmeth.4468}.
-
-    \bibitem{schiffer2019}
-    Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
-    \newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
-      through Bioconductor.
-    \newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
-
-    \bibitem{gonzalez2018}
-    Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
-      Ackermann G, et~al.
-    \newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
-    \newblock Nature Methods. 2018;15(10):796--798.
-    \newblock doi:{10.1038/s41592-018-0141-9}.
-
-    \bibitem{consortium2012}
-    Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
-      et~al.
-    \newblock Structure, Function and Diversity of the Healthy Human Microbiome.
-    \newblock Nature. 2012;486(7402):207--214.
-    \newblock doi:{10.1038/nature11234}.
-
-    \bibitem{proctor2019}
-    Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
-    \newblock The {{Integrative Human Microbiome Project}}.
-    \newblock Nature. 2019;569(7758):641--648.
-    \newblock doi:{10.1038/s41586-019-1238-8}.
-
-    \bibitem{thurnheer2016}
-    Thurnheer T, Bostanci N, Belibasakis GN.
-    \newblock Microbial Dynamics during Conversion from Supragingival to
-      Subgingival Biofilms in an in Vitro Model.
-    \newblock Molecular Oral Microbiology. 2016;31(2):125--135.
-    \newblock doi:{10.1111/omi.12108}.
-
-    \bibitem{beghini2019}
-    Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
-    \newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
-      in the {{New York City Health}} and {{Nutrition Examination Study}}.
-    \newblock Annals of Epidemiology. 2019;34:18--25.e3.
-    \newblock doi:{10.1016/j.annepidem.2019.03.005}.
-
-    \bibitem{matteocalgaro2020}
-    Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
-    \newblock Zenodo.
-
-    \bibitem{nielsen2014}
-    Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
-    \newblock Identification and Assembly of Genomes and Genetic Elements in
-      Complex Metagenomic Samples without Using Reference Genomes.
-    \newblock Nature Biotechnology. 2014;32(8):822--828.
-    \newblock doi:{10.1038/nbt.2939}.
-
-    \bibitem{gevers2014}
-    Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
-      et~al.
-    \newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
-      {{Disease}}.
-    \newblock Cell Host \& Microbe. 2014;15(3):382--392.
-    \newblock doi:{10.1016/j.chom.2014.02.005}.
-
-    \bibitem{martin2020}
-    Martin BD, Witten D, Willis AD.
-    \newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
-      Regression.
-    \newblock The Annals of Applied Statistics. 2020;14(1):94--115.
-    \newblock doi:{10.1214/19-AOAS1283}.
-
-    \bibitem{khatri2012}
-    Khatri P, Sirota M, Butte AJ.
-    \newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
-      {{Outstanding Challenges}}.
-    \newblock PLOS Computational Biology. 2012;8(2):e1002375.
-    \newblock doi:{10.1371/journal.pcbi.1002375}.
-
-    \bibitem{ashburner2000}
-    Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
-    \newblock Gene {{Ontology}}: Tool for the Unification of Biology.
-    \newblock Nature genetics. 2000;25(1):25--29.
-    \newblock doi:{10.1038/75556}.
-
-    \bibitem{chong2020}
-    Chong J, Liu P, Zhou G, Xia J.
-    \newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
-      Functional, and Meta-Analysis of Microbiome Data.
-    \newblock Nature Protocols. 2020;15(3):799--821.
-    \newblock doi:{10.1038/s41596-019-0264-1}.
-
-    \bibitem{naim}
-    Naim I, Gildea D.
-    \newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
-      {{Unbalanced Mixing Coefficients}}; p.~8.
-
-    \bibitem{martin-fernandez2012}
-    {Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
-      {Palarea-Albaladejo} J.
-    \newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
-      {{Classical}} and Robust Approaches.
-    \newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
-    \newblock doi:{10.1016/j.csda.2012.02.012}.
-
-    \end{thebibliography}
-
+\newpage
+\bibliography{tax_agg}{}
+\bibliographystyle{plos2015}
 \end{document}

### 2021-05-28 tax_agg.tex
Fixed intro section

@@
 \begin{abstract}
-    \noindent Research in human associated microbiomes primarily involve high-throughput sequencing of samples, resulting in count tables of taxa abundances.
-Unfortunately, the analysis microbiome data is complex as it is high-dimensional, sparse, and strictly compositional.
-Various statistical tools have been developed to address these issues, but a very approachable method is to aggregate taxa into pre-defined sets, which is called gene set testing or enrichment analysis in the genomics literature.
-However, there has been a considerable lack of interest in developing set-based analysis methods tailored specifically for microbiome relative abundance data.
-Here, we present a new sample-level taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis in the gene set testing literature.
-Our approach, titled competitive isometric log ratio (cILR), generates enrichment scores per sample as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
-We also provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
+    \noindent Research in human associated microbiomes primarily involve the analysis of taxonomic count tables generated via high-throughput sequencing.
+It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional.
+An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets.
+Set-based analysis is ubiquitous in the genomics literature, and have demonstrable impact in improving interpretability and power of downstream analysis.
+Current practice in the microbiome field, however, performs aggregation by summing the abundance of respective taxa.
+This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
+Unfortunately, there is a lack of interest in developing more sophisticated set-based analysis methods specific to microbiome data.
+Here, we attempt to fill this gap with a new sample-level taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature.
+Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
+We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
 Herein we demonstrate under both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios, while additionally providing informative scores that can be inputs to downstream differential abundance and predictive tasks with good performance.
-These results demonstrate how our approach can be used to generate compositionally valid taxonomic aggregation and enrichment.
-
+These results demonstrate how our approach can enable researchers to generate meaningful aggregation scores for sets of taxa.
 \end{abstract}
@@
 
-Even though the aforementioned problems are challenging, a very approachable method to address some of them is variable aggregation.
-Aggregated variables can be less sparse than their constituent elements, and hypothesis testing on a reduced number of variables can alleviate the multiple testing burden.
-This helps increase power in downstream analyses, as well as interpretability.
-In practice, microbiome researchers perform variable aggregation to pre-defined taxonomic levels (e.g.
-phylum, family, order) through element-wise summation of counts.
-However, this approach contains various downsides: first, it does not allow for comparison of enrichment across sets of different sizes, where larger sets naturally contain more counts; second, aggregating compositional variables using component-wise summations can distort the inter-sample distances before and after aggregation due to the non-linearity of this amalgamation in Aitchison space \cite{egozcue2005}.
-
-Set-based analyses are ubiquitous in the gene expression literature \cite{goeman2007} and various methods have been developed for enrichment testing and scoring.
+Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}.
+Aggregated sets can be less sparse than their constituent elements, and by focusing on groups rather tahn individuals, this approach can reduce the burden of testing multiple hypotheses, thereby increasing power and reproducibility.
+Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations.
+There exists a diverse set of available methods developed in the gene set testing literature.
 More traditional gene-set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest.
 Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values.
-The most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size.
+As such, the most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size.
 Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples.
-More recent methods, such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate sample level enrichment scores more akin to a transformation.
-Generated scores can then be used in data visualization, as well as flexible scores that can be utilized in downstream analyses.
+Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample and is more akin to a transformation.
+This strategy allows for the flexible incorporation of different statistical techniques downstream, such as predictive models, as well as for visualization purposes.
 
-Here, we developed a new sample-level set-based analysis method specific to microbiome relative abundance data.
+In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate variables to taxonomic classification levels such as genus, family, or phylum.
+Standard practice involves summing the abundance of relevant taxa of the same group \cite{mclaren2019}.
+Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, increase bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}.
+
+As such, there is a need for microbiome researchers to adopt set enrichment methods to perform a more robust set-based analysis that allows for enrichment testing of functionally relevant sets of microbes.
+Unfortunately, there is a lack of interest in extending existing methods to be more specific to microbiome relative abundance data.
+Some software suites, such as \emph{MicrobiomeAnalyst}, do provide researchers tools to perform enrichment testing with curated taxa sets \cite{chong2020}.
+However, the approach used in \emph{MicrobiomeAnalyst} is an overrepresentation analysis and therefore similarly sensitive to the method of differential abundance used.
+
+Here, we developed a novel method that allows for generating enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}.
 We leverage the conception of the $Q_1$ competitive hypothesis presented in Tian et al.
@@
 
-In this manuscript, we present cILR, provide its formulation and discuss some statistical properties.
-Then, we illustrate the benefits of cILR as both a sample-level significance test for enrichment of taxa sets and as a feature engineering approach for downstream disease prediction and differential abundance using both real data and simulation studies.
-We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as existing sample-level enrichment methods in the gene expression literature such as GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}.
+In the following sections, we provide the formulation of cILR and discuss some statistical properties.
+Then, we illustrate significance testing at the sample level using cILR and demonstrating type I error and power under different simulation scenarios and real data applications.
+Then we assessed the informativeness of cILR generated scores, and evaluated how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis.
+We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are equivalent single sample enrichment analysis methods in the genomics literature.
 An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).
@@
 \section*{Discussion}
-\subsection*{Set-based analysis can enhance microbiome data analysis}
-Gene-set testing (or pathway analysis) is an important tool in the analysis of high-dimensional genomic data sets \cite{khatri2012}.
-By focusing on a group of variables rather than individuals, this approach reduces the hypothesis testing burden, thereby increasing power as well as replicability.
-Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also increase interpretability.
-
-In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate variables to taxonomic classification levels such as genus, family, or phylum.
-The process of aggregation involves the element-wise summation of counts of taxa that belong to the same taxonomic group.
-Even though this allows for a reduction in the number of overall variables (from thousands to only hundreds), there still exist two disadvantages: first, it doesn't preserve inter-sample distances before and after aggregation \cite{egozcue2005}, and second, it doesn't allow for enrichment testing and comparison across sets of different sizes.
-
-As such, there is a need for microbiome researchers to adopt set enrichment methods to perform a more robust set-based analysis that allows for enrichment testing of functionally relevant sets of microbes.
-Some software suites, such as \emph{MicrobiomeAnalyst}, have begun to provide an option to perform such analyses with curated taxa sets \cite{chong2020}.
-However, the approach used in \emph{MicrobiomeAnalyst} is based on a hypergeometric test of significant hypotheses and is sensitive to the method of differential abundance used.
-More recent methods in set-based testing generate enrichment scores at the sample level \cite{hanzelmann2013, frost2020a}.
-This approach allows for the flexible incorporation of different statistical techniques downstream, but unfortunately is not currently tailored to the compositional nature of microbiome relative abundance data \cite{gloor2017}.
-
-Here, we developed cILR, a novel microbiome-specific method to perform set-based enrichment analysis at the sample level.
-Our method is based on the isometric log-ratio transformation \cite{egozcue2003}, and measures set enrichment relative to those outside the set, which corresponds to a $Q_1$ null hypothesis in the gene set analysis literature \cite{tian2005}.
-cILR generates sample-specific enrichment scores, enabling cILR to act as both a significance test for enrichment at the sample level, and as inputs for downstream analyses such as differential abundance testing, predictive modeling, and data visualization.
 

### 2021-05-28 tax_agg.tex
Updates from Overleaf

@@
 Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}.
-Aggregated sets can be less sparse than their constituent elements, and by focusing on groups rather tahn individuals, this approach can reduce the burden of testing multiple hypotheses, thereby increasing power and reproducibility.
+Aggregated sets can be less sparse than their constituent elements, and testing on a smaller number of variables can reduce the multiple testing burden, thereby increasing power and reproducibility.
 Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations.
-There exists a diverse set of available methods developed in the gene set testing literature.
-More traditional gene-set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest.
+There exists a diverse set of available methods developed to perform such analyses.
+More traditional set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest \cite{goeman2007}.
 Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values.
-As such, the most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size.
+The most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size.
 Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples.
 Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample and is more akin to a transformation.
-This strategy allows for the flexible incorporation of different statistical techniques downstream, such as predictive models, as well as for visualization purposes.
+This strategy allows for the flexible incorporation of different statistical techniques downstream, such as predictive models, as well as for visualization purposes in ordination plots.
 
-In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate variables to taxonomic classification levels such as genus, family, or phylum.
-Standard practice involves summing the abundance of relevant taxa of the same group \cite{mclaren2019}.
+In microbiome research, even though no explicit enrichment analysis is performed, standard practice often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum by simple summation of abundances \cite{mclaren2019}.
 Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, increase bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}.
-
-As such, there is a need for microbiome researchers to adopt set enrichment methods to perform a more robust set-based analysis that allows for enrichment testing of functionally relevant sets of microbes.
+As such, there is a need for microbiome researchers to adopt set enrichment methods to perform a more robust analysis that allows for enrichment testing of functionally relevant sets of microbes.
 Unfortunately, there is a lack of interest in extending existing methods to be more specific to microbiome relative abundance data.
@@
 The competitive hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables.
-We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a log-ratio-based aggregation method that addresses the downsides of the naive summation-based method presented above.
+We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a multiplicative aggregation method that addresses the downsides of the naive summation-based method presented above \cite{mclaren2019, silverman2017}.
 The resulting method, titled competitive isometric log-ratio (cILR), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing.
@@
 The cILR method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}.
-Details on the computational implementation of cILR can be found in the supplemental section.
+Details on the computational implementation of cILR can be found in the supplemental materials.
 The cILR method takes two inputs:
 \begin{itemize}
-    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as 16S rRNA) or whole genome shotgun sequencing.
+    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing.
 Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing.
-    \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxa $p$ to $m$ sets of interest.
+    \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxon $p$ to $m$ sets of interest.
 These sets can be Linnean taxonomic classifications annotated using databases such as SILVA \cite{quast2013}, or those based on more functionally driven categories such as tropism or ecosystem roles ($A_{i,j} = 1$ indicates that microbe $i$ belongs to set $j$).
@@
 The usage of the ILR statistic is not uncommon in the microbiome literature.
-They are usually termed ``compositional balances", and were leveraged in many recent approaches in variable transformation \cite{washburne2017,silverman2017,morton2017}.
+They are usually termed ``compositional balances", and have been leveraged in many recent approaches in variable transformation \cite{washburne2017,silverman2017,morton2017}.
 The cILR formulation \eqref{main_eq} is a special case of \eqref{ilr_standard} defined on a node that splits the taxa into two disjoint sets, one representing the set of interest, the other representing the remaining taxa.

### 2021-05-28 tax_agg.tex

@@
 The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their human host.
-Previous research has shown that changes in the composition of the microbiome have been associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
-To understand the central role of the microbiome in human health, researchers often relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
+Previous research has shown that changes in the composition of the microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
+To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
 amplicon sequencing) or by profiling all the genomic content of the sample (i.e.
 whole-genome shotgun sequencing) \cite{cho2012}.
-Processing raw sequencing data using a variety of bioinformatic pipelines \cite{callahan2016,truong2015} would yield taxonomic abundance tables that would be part of downstream statistical analyses where associations between an outcome/exposure and identified taxa would be found.
+Processing raw sequencing data using a variety of bioinformatic pipelines \cite{callahan2016,truong2015} yields taxonomic abundance tables that can be used in downstream statistical analyses to identify associations between an outcome/exposure and identified taxa.
 
@@
 For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016}, allowing for simultaneous model fitting and variable selection.
-For differential abundance tasks, researchers often utilized multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.
+For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.
 
@@
 This assumption is also supported by preliminary simulation studies (detailed below).
-In panel B of figure \ref{fig:kurt_gof}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
+In panel A of figure \ref{fig:kurt_gof}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
 It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation.
@@
 To address this issue, we also evaluated a mixture distribution of two normal components.
-Panel A of figure \ref{fig:kurt_gof} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
+Panel B of figure \ref{fig:kurt_gof} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
 We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.
@@
 \subsubsection*{Parametric Simulations}  
-To address the performance of cILR under different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
+To address the performance of cILR for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
 Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
@@
 Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3).
-Due to the relative nature of microbiome data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets.
+Due to the compositional nature of microbiome data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets.
 As such, we applied a compensation procedure as described in Hawinkel et al.
@@
 
-In addition to simulation analyses, we also evaluated our method using real data sets across both 16S rRNA gene sequencing and whole-genome sequencing.
+In addition to simulation analyses, we also evaluated our method using real data sets based on both 16S rRNA gene sequencing and whole-genome sequencing.
 All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.
@@
 Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false discovery rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment.
-Microbial tropism annotation at the Genus level was from Beghini et al.
-\cite{beghini2019} and was downloaded directly from the Github repository associated with Calagaro et al.
+Microbial tropism annotation at the genus level was from Beghini et al.
+\cite{beghini2019} and was downloaded directly from the GitHub repository associated with Calagaro et al.
 \cite{matteocalgaro2020}.
@@
 \cite{gevers2014} which also profiles the gut microbiome of IBD patients and controls but using 16S rRNA gene sequencing.
-This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages < 17) from the RISK cohort enrolled in North America.
+This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada.
 Of the 671 samples obtained, 500 samples belong to patients with IBD.
@@
 
-\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank-sum test.
+\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank sum test.
 We added a pseudocount of 1 to all zero entry values.
 This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample.
-For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank-sum test.
+For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test.
 
@@
 Panel A and B in figure \ref{fig:ss_hypo} demonstrate type I error and power respectively across different simulation conditions.
-We benchmarked the results of the cILR method against a naive Wilcoxon rank-sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
+We benchmarked the results of the cILR method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
 All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions.
@@
 False-positive rate \textbf{(C)} and true positive rate \textbf{(D)} evaluation of similar methods on real 16S rRNA data from the oral microbiome of the gingival site.
-For \textbf{(A)} and \textbf{(B)}, enrichment of a specified set was tested at the sample level using cILR and the Wilcoxon rank-sum test at $\alpha$ of 0.05.
+For \textbf{(A)} and \textbf{(B)}, enrichment of a specified set was tested at the sample level using cILR and the Wilcoxon rank sum test at $\alpha$ of 0.05.
 For \textbf{(C)} and \textbf{(D)}, the set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples.
@@
 However, the trade-off for good type I error control is demonstrably lower power, as shown in figure \ref{fig:ss_hypo}B.
-In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un adjusted ones.
-However, in higher correlation scenarios, the difference in power mus much more dramatic.
-At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted cILR was only perfomring at 50\% power, while unadjusted cILR and wilcoxon rank sum test were able to reach 80\%.
-These results indicated that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of cILR.
+In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un-adjusted ones.
+However, in higher correlation scenarios, the difference in power is much more dramatic.
+At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted cILR was only performing at 50\% power, while unadjusted cILR and wilcoxon rank sum test were able to reach 80\%.
+These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of cILR.
+Most importantly, cILR demonstrate higher power in all scenarios where type I error is properly controlled.
 
 These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}.
-Here, we tested the enrichment of aerobic microbes at each sample using approaches similar to our parametric simulations.
+Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations.
 As expected in Figure \ref{fig:ss_hypo}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods.
@@
 Figure \ref{fig:ss_auc} presents this result.
-We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $U$ test statistic from the Wilcoxon rank-sum test.
+We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $U$ test statistic from the Wilcoxon rank sum test.
 Across both simulations (Figure \ref{fig:ss_auc}A) and real-data applications (Figure \ref{fig:ss_auc}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
@@
 \subsection*{Differential abundance analysis}
-Here we assessed the ability to use cILR with a standard difference of means statistical test (such as Welch's t-test or a Wilcoxon rank-sum test) to perform differential abundance analysis.
-We compare the performance of cILR related methods to two commonly used approaches in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
+Here we assessed the ability to use cILR with a standard difference of means statistical test (such as Welch's t-test or a Wilcoxon rank sum test) to perform differential abundance analysis.
+We compared the performance of cILR related methods to two commonly used approaches in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
 
 \subsection*{Disease Prediction} 
-We can conceive of cILR as a feature engineering step before model fit and prediction.
+We can conceive of cILR as a feature engineering step before model fitting and prediction.
 Here we assessed the predictive performance of a standard machine learning model (random forest \cite{breiman2001}) using cILR scores compared to scores generated by other single sample enrichment methods in the gene set testing literature (ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}).
-Additionally, we also compared our results with just using the centered log-ratio (CLR) transformation on taxa sets aggregated via count summations.
+Additionally, we also compared our results with the use of the centered log-ratio (CLR) transformation on taxa sets aggregated via count summations.
 
@@
 \cite{nielsen2014} is a whole-genome shotgun sequencing data set.
-Similar to simulation experiments, across both data sets cILR methods provide much better performance than both GSVA and ssGSEA.
+Similar to simulation experiments, across both data sets cILR methods provide much better performance than either GSVA or ssGSEA.
 Interestingly, the standard CLR approach only outperformed cILR in the 16S rRNA data set but was marginally worse in the WGS data set.
@@
 
-These results demonstrated that cILR generated scores are informative features in disease prediction tasks.
-Simulation results indicated that cILR methods perform much better than both ssGSEA and GSVA, but not as good as the standard CLR approach.
+These results demonstrate that cILR generated scores are informative features in disease prediction tasks.
+Simulation results indicate that cILR methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach.
 Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
@@
 
-\subsection*{Significance testing with cILR}
+\subsection*{Inference with cILR}
 We can use cILR to test for enrichment of taxa-sets at the sample level.
@@
 The conservativeness of the test persisted across different sparsity levels, as well as inter-taxa correlation.
-Conversely, the unadjusted cILR methods and the naive Wilcoxon rank-sum test has inflated type I error even at low correlation levels ($\rho = 0.2$).
+Conversely, the unadjusted cILR methods and the naive Wilcoxon rank sum test has inflated type I error even at low correlation levels ($\rho = 0.2$).
 However, the trade-off for proper type I error control is lower power.

### 2021-05-28 tax_agg.tex

@@
     \centering
-    \includegraphics[width=\textwidth]{figures/sim_pred_auc.png}
+    \includegraphics[width = \textwidth]{figures/sim_pred_combined.png}
     \caption{AUC scores of a random forest model for a binary outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio.
@@
 
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth]{figures/sim_pred_rsq.png}
-    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio.
-Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome).
+%\begin{figure}[H]
+%    \centering
+%    \includegraphics[width=\textwidth]{figures/sim_pred_rsq.png}
+%    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on cILR,     %ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data %across sparsity levels, correlation, and signal-to-noise ratio.
+Panel \textbf{(A)} and \textbf{(B)} %represent results across different levels of model saturation (proportion of sets associated with the %outcome).
 cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
-    \label{fig:sim_pred_rsq}
-\end{figure}
+%    \label{fig:sim_pred_rsq}
+%\end{figure}
 

### 2021-05-28 tax_agg.tex
Updates from Overleaf

@@
     \item \textbf{Estimate the empirical null distribution} Enrichment scores represent the test statistic for the $Q_1$ null hypothesis $H_o$ that relative abundances in $\mathbf{X}$ of members of set $k$ are not enriched compared to those not in set $k$.
-Since the distribution of cILR under the null vary depending on data characteristics (Figure \ref{fig:kurt_gof}), an empirical null distribution will be estimated from data.
+Since the distribution of cILR under the null vary depending on data characteristics (Figure \ref{fig:1}), an empirical null distribution will be estimated from data.
     \begin{itemize}
@@
 This assumption is also supported by preliminary simulation studies (detailed below).
-In panel A of figure \ref{fig:kurt_gof}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
+In panel A of figure \ref{fig:1}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
 It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation.
@@
 Results indicated the necessity of estimating an empirical null and demonstrating that the mixture distribution was the better fit compared to the basic normal.}
-    \label{fig:kurt_gof}
+    \label{fig:1}
 \end{figure}
@@
 To address this issue, we also evaluated a mixture distribution of two normal components.
-Panel B of figure \ref{fig:kurt_gof} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
+Panel B of figure \ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
 We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.
@@
 
-Panel A and B in figure \ref{fig:ss_hypo} demonstrate type I error and power respectively across different simulation conditions.
+Panel A and B in figure \ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
 We benchmarked the results of the cILR method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
@@
 However, this resulted in lower power for adjusted cILR methods.}
-    \label{fig:ss_hypo}
+    \label{fig:2}
 \end{figure}
 
-However, the trade-off for good type I error control is demonstrably lower power, as shown in figure \ref{fig:ss_hypo}B.
+However, the trade-off for good type I error control is demonstrably lower power, as shown in figure \ref{fig:2}B.
 In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un-adjusted ones.
@@
 Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations.
-As expected in Figure \ref{fig:ss_hypo}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods.
+As expected in Figure \ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods.
 Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05.
-Power analysis (Figure \ref{fig:ss_hypo}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
+Power analysis (Figure \ref{fig:2}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
 
@@
 This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated.
-Figure \ref{fig:ss_auc} presents this result.
+Figure \ref{fig:3} presents this result.
 We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $U$ test statistic from the Wilcoxon rank sum test.
-Across both simulations (Figure \ref{fig:ss_auc}A) and real-data applications (Figure \ref{fig:ss_auc}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
+Across both simulations (Figure \ref{fig:3}A) and real-data applications (Figure \ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
 In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
@@
 Error bars are the 95\% DeLong confidence intervals for AUC \cite{delong1988}}
-    \label{fig:ss_auc}
+    \label{fig:3}
 \end{figure}
@@
 
-Figure \ref{fig:sim_pred_auc} and \ref{fig:sim_pred_rsq} showed results from simulation studies as detailed in the \nameref{Methods} section under both classification and regression tasks.
+Figure \ref{fig:4} and \ref{fig:4} showed results from simulation studies as detailed in the \nameref{Methods} section under both classification and regression tasks.
 As expected, performance across all assessed methods increased with a higher signal-to-noise ratio, however, they remained consistent across different levels of model saturation and inter-taxa correlation.
@@
 Conversely, cILR scores did not perform as well as the traditional CLR approach, however, the performance gap decreased at higher sparsity levels.
-Similar patterns were observed for regression results (Figure \ref{fig:sim_pred_rsq}).
+Similar patterns were observed for regression results (Figure \ref{fig:4}).
 Most interestingly, regression performance, in general, decreased at higher saturation levels (proportion of sets associated with the outcome).
@@
     \centering
-    \includegraphics[width = \textwidth]{figures/sim_pred_combined.png}
+    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
     \caption{AUC scores of a random forest model for a binary outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio.
@@
 cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
-    \label{fig:sim_pred_auc}
+    \label{fig:4}
 \end{figure}
@@
 
-In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:pred_data}.
+In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:5}.
 The learning task across both data sets is to classify patients who are diagnosed with inflammatory bowel disease (includes both Crohn's disease and ulcerative colitis) and healthy controls.
@@
 cILR marginally outperforms GSVA and ssGSEA in both data sets, but only performs better than CLR in the WGS data set.}
-    \label{fig:pred_data}
+    \label{fig:5}
 \end{figure}
@@
 Our method addresses this issue by combining the mean estimate of scores computed on permuted data and the variance estimate of scores computed on unpermuted data, where correlation is undisrupted.
-As a consequence, adjusted cILR methods control very well for type I error at the appropriate $\alpha$ level (Figure \ref{fig:ss_hypo}).
+As a consequence, adjusted cILR methods control very well for type I error at the appropriate $\alpha$ level (Figure \ref{fig:2}).
 The conservativeness of the test persisted across different sparsity levels, as well as inter-taxa correlation.
@@
 The difference in performance across the two choices was minimal, where both were able to control for type I error, but the normal distribution was slightly better when it comes to power only in certain simulation settings.
-This runs contrary to our distribution comparison analysis in Figure \ref{fig:kurt_gof}, where the mixture distribution was the superior fit.
+This runs contrary to our distribution comparison analysis in Figure \ref{fig:1}, where the mixture distribution was the superior fit.
 We hypothesized that this might be due to the difficulty in fitting mixture distributions using the expectation-maximization algorithm (EM), as the convergence rate is slow when there is a lot of overlap between the mixtures and the mixing coefficients of one of the components are low \cite{naim2012}.
@@
 Across both simulations and real data analyses, cILR scores perform better than alternative single sample scoring methods in the gene set testing literature (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}).
-This is consistent with previous analysis (Figure \ref{fig:ss_auc}), where we assessed the informativeness of cILR generated scores by using AUC to measure whether samples with high scores correspond with the true label of an inflated set.
+This is consistent with previous analysis (Figure \ref{fig:3}), where we assessed the informativeness of cILR generated scores by using AUC to measure whether samples with high scores correspond with the true label of an inflated set.
 Results indicated that cILR scores are more adaptive to microbiome data sets than GSVA and ssGSEA.
@@
 Unsurprisingly, different variants of cILR did not differ significantly in performance similar to significance testing experiments.
-Even though we expect Z-scores to perform better due to the added information of directionality of enrichment, using CDF values did not alter performance in both the single sample AUC evaluations (Figure \ref{fig:ss_auc}) as well as the random forest evaluations (Figure \ref{fig:sim_pred_auc}).
+Even though we expect Z-scores to perform better due to the added information of directionality of enrichment, using CDF values did not alter performance in both the single sample AUC evaluations (Figure \ref{fig:3}) as well as the random forest evaluations (Figure \ref{fig:4}).
 

### 2021-05-28 manuscript_plos.tex

@@
 {\Large
-\textbf\newline{cILR: Taxonomic enrichment analysis with isometric log-ratios} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{cILR: Competitive isometric log-ratio for taxonomic enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 \\
-Quang Nguyen\textsuperscript{1,2},
+Quang P. Nguyen\textsuperscript{1,2},
 Anne G. Hoen\textsuperscript{1,2\ddag},
-H. Robert Frost\textsuperscript{1*\ddag},
+H. Robert Frost\textsuperscript{1*\ddag\dag},
 \\
@@
 % Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-\ddag These authors jointly supervised this work.
-
+\ddag These authors jointly supervised this work.\\
+\dag Corresponding author
 % Current address notes

### 2021-05-28 tax_agg.tex
Updates from Overleaf

@@
 
+\begin{figure}[H]
+    \centering
+    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
+    \caption{Placeholder} 
+    \label{fig:4}
+\end{figure}
+
 \subsection*{Disease Prediction} 
@@
 
-Figure \ref{fig:4} and \ref{fig:4} showed results from simulation studies as detailed in the \nameref{Methods} section under both classification and regression tasks.
+Figure \ref{fig:6} and \ref{fig:6} showed results from simulation studies as detailed in the \nameref{Methods} section under both classification and regression tasks.
 As expected, performance across all assessed methods increased with a higher signal-to-noise ratio, however, they remained consistent across different levels of model saturation and inter-taxa correlation.
@@
 Conversely, cILR scores did not perform as well as the traditional CLR approach, however, the performance gap decreased at higher sparsity levels.
-Similar patterns were observed for regression results (Figure \ref{fig:4}).
+Similar patterns were observed for regression results (Figure \ref{fig:6}).
 Most interestingly, regression performance, in general, decreased at higher saturation levels (proportion of sets associated with the outcome).
@@
 cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
-    \label{fig:4}
+    \label{fig:6}
 \end{figure}
@@
 
-In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:5}.
+In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:7}.
 The learning task across both data sets is to classify patients who are diagnosed with inflammatory bowel disease (includes both Crohn's disease and ulcerative colitis) and healthy controls.
@@
 cILR marginally outperforms GSVA and ssGSEA in both data sets, but only performs better than CLR in the WGS data set.}
-    \label{fig:5}
+    \label{fig:7}
 \end{figure}
@@
 Unsurprisingly, different variants of cILR did not differ significantly in performance similar to significance testing experiments.
-Even though we expect Z-scores to perform better due to the added information of directionality of enrichment, using CDF values did not alter performance in both the single sample AUC evaluations (Figure \ref{fig:3}) as well as the random forest evaluations (Figure \ref{fig:4}).
+Even though we expect Z-scores to perform better due to the added information of directionality of enrichment, using CDF values did not alter performance in both the single sample AUC evaluations (Figure \ref{fig:6}) as well as the random forest evaluations (Figure \ref{fig:6}).
 

### 2021-06-02 manuscript_plos.tex

@@
 \section*{Introduction}
+\section*{Background}  
 The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their human host.
-Previous research has shown that changes in the composition of the microbiome have been associated with important health outcomes such as inflammatory bowel disease, type II diabetes, and obesity.
-To understand the central role of the microbiome in human health, researchers often relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
+Previous research has shown that changes in the composition of the microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
+To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
 amplicon sequencing) or by profiling all the genomic content of the sample (i.e.
 whole-genome shotgun sequencing) \cite{cho2012}.
-Processing raw sequencing data using a variety of bioinformatic pipelines \cite{callahan2016,truong2015} would yield taxonomic abundance tables that would be part of downstream statistical analyses where associations between an outcome/exposure and identified taxa would be found.
+Processing raw sequencing data using a variety of bioinformatic pipelines \cite{callahan2016,truong2015} yields taxonomic abundance tables that can be used in downstream statistical analyses to identify associations between an outcome/exposure and identified taxa.
 
@@
 
-Even though the aforementioned problems are challenging, a very approachable method to address some of them is variable aggregation.
-Aggregated variables can be less sparse than their constituent elements, and hypothesis testing on a reduced number of variables can alleviate the multiple testing burden.
-This helps increase power in downstream analyses, as well as interpretability.
-In practice, microbiome researchers perform variable aggregation to pre-defined taxonomic levels (e.g.
-phylum, family, order) through element-wise summation of counts.
-However, this approach contains various downsides: first, it does not allow for comparison of enrichment across sets of different sizes, where larger sets naturally contain more counts; second, aggregating compositional variables using component-wise summations can distort the inter-sample distances before and after aggregation due to the non-linearity of this amalgamation in Aitchison space \cite{egozcue2005}.
-
-Set-based analyses are ubiquitous in the gene expression literature \cite{goeman2007} and various methods have been developed for enrichment testing and scoring.
-More traditional gene-set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest.
+Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}.
+Aggregated sets can be less sparse than their constituent elements, and testing on a smaller number of variables can reduce the multiple testing burden, thereby increasing power and reproducibility.
+Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations.
+There exists a diverse set of available methods developed to perform such analyses.
+More traditional set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest \cite{goeman2007}.
 Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values.
@@
 Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples.
-More recent methods, such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate sample level enrichment scores more akin to a transformation.
-Generated scores can then be used in data visualization, as well as flexible scores that can be utilized in downstream analyses.
-
-Here, we developed a new sample-level set-based analysis method specific to microbiome relative abundance data.
+Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample and is more akin to a transformation.
+This strategy allows for the flexible incorporation of different statistical techniques downstream, such as predictive models, as well as for visualization purposes in ordination plots.
+
+In microbiome research, even though no explicit enrichment analysis is performed, standard practice often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum by simple summation of abundances \cite{mclaren2019}.
+Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, increase bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}.
+As such, there is a need for microbiome researchers to adopt set enrichment methods to perform a more robust analysis that allows for enrichment testing of functionally relevant sets of microbes.
+Unfortunately, there is a lack of interest in extending existing methods to be more specific to microbiome relative abundance data.
+Some software suites, such as \emph{MicrobiomeAnalyst}, do provide researchers tools to perform enrichment testing with curated taxa sets \cite{chong2020}.
+However, the approach used in \emph{MicrobiomeAnalyst} is an overrepresentation analysis and therefore similarly sensitive to the method of differential abundance used.
+
+Here, we developed a novel method that allows for generating enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}.
 We leverage the conception of the $Q_1$ competitive hypothesis presented in Tian et al.
@@
 
-In this manuscript, we present cILR, provide its formulation and discuss some statistical properties.
-Then, we illustrate the benefits of cILR as both a sample-level significance test for enrichment of taxa sets and as a feature engineering approach for downstream disease prediction and differential abundance using both real data and simulation studies.
-We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as existing sample-level enrichment methods in the gene expression literature such as GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}.
+In the following sections, we provide the formulation of cILR and discuss some statistical properties.
+Then, we illustrate significance testing at the sample level using cILR and demonstrating type I error and power under different simulation scenarios and real data applications.
+Then we assessed the informativeness of cILR generated scores, and evaluated how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis.
+We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are equivalent single sample enrichment analysis methods in the genomics literature.
 An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).
 
-\section{Material and Methods} \label{Methods}
-\subsection{Competitive Isometric Log-ratio (cILR)}
+\section*{Materials and Methods} \label{Methods}
+\subsection*{Competitive Isometric Log-ratio (cILR)}
 The cILR method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}.
-Details on the computational implementation of cILR can be found in the supplemental section.
+Details on the computational implementation of cILR can be found in the supplemental materials.
 The cILR method takes two inputs:
 \begin{itemize}
-    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as 16S rRNA) or whole genome shotgun sequencing.
+    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing.
 Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing.
-    \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxa $p$ to $m$ sets of interest.
+    \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxon $p$ to $m$ sets of interest.
 These sets can be Linnean taxonomic classifications annotated using databases such as SILVA \cite{quast2013}, or those based on more functionally driven categories such as tropism or ecosystem roles ($A_{i,j} = 1$ indicates that microbe $i$ belongs to set $j$).
@@
 This assumption is also supported by preliminary simulation studies (detailed below).
-In panel B of figure \ref{fig:kurt_gof}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
+In panel A of figure \ref{fig:1}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
 It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation.
@@
 
+\begin{figure} [H]
+    \centering
+    \includegraphics[width=\linewidth]{figures/kurtosis_skewness_gof.png}
+    \caption{Properties of the null distribution of cILR in different simulation scenarios under the global null.
+Panel \textbf{(B)} presents kurtosis and skewness of cILR scores while panel \textbf{(A)} presents the goodness of fit (as Kolmogorov-Smirnov D statistic) for mixture normal and normal distributions.
+Panel \textbf{(C)} is a density plot of the shape of the null distribution.
+Results indicated the necessity of estimating an empirical null and demonstrating that the mixture distribution was the better fit compared to the basic normal.}
+    \label{fig:1}
+\end{figure}
+
 Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null.
 To address this issue, we also evaluated a mixture distribution of two normal components.
-Panel A of figure \ref{fig:kurt_gof} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
+Panel B of figure \ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
 We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.
@@
 
-\begin{figure}
-    \centering
-    \caption{Properties of the null distribution of cILR in different simulation scenarios under the global null.
-Panel \textbf{(B)} presents kurtosis and skewness of cILR scores while panel \textbf{(A)} presents the goodness of fit (as Kolmogorov-Smirnov D statistic) for mixture normal and normal distributions.
-Panel \textbf{(C)} is a density plot of the shape of the null distribution.
-Results indicated the necessity of estimating an empirical null and demonstrating that the mixture distribution was the better fit compared to the basic normal.}
-    \label{fig:kurt_gof}
-\end{figure}
-
 \subsection*{Evaluation}
 \subsubsection*{Parametric Simulations}  
-To address the performance of cILR under different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
+To address the performance of cILR for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
 Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
@@
 \cite{gevers2014} which also profiles the gut microbiome of IBD patients and controls but using 16S rRNA gene sequencing.
-This data set contains 671 16S rRNA gene sequencing samples from a cohort of ABC, where 500 was classified as having IBD.
+This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada.
+Of the 671 samples obtained, 500 samples belong to patients with IBD.
 
@@
 
-\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank-sum test.
+\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank sum test.
 We added a pseudocount of 1 to all zero entry values.
 This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample.
-For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank-sum test.
+For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test.
+All three approaches were applied directly on count data (after pseudocount).
+For GSVA, the Poisson kernel was used.
 
@@
 For comparison, we chose representative state-of-the-art methods in differential abundance analysis, namely DESeq2 \cite{love2014,mcmurdie2014} and corncob \cite{martin2020}.
+For DESeq2, we performed a likelihood ratio test against an intercept only reduced model with dispersion estimated with local fit.
+For corncob, we also performed a likelihood ratio test against an intercept only reduced model without bootstrapping.
 
@@
 We added a pseudocount of 1 to all zero entry values.
-Additionally, we also compared performance using enrichment scores against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.
+Additionally, we also compared performance using cILR against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.
 
@@
 
-Panel A and B in figure \ref{fig:ss_hypo} demonstrate type I error and power respectively across different simulation conditions.
-We benchmarked the results of the cILR method against a naive Wilcoxon rank-sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
+\subsubsection*{Simulation studies}
+Panel A and B in figure \ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
+We benchmarked the results of the cILR method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
 All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions.
@@
 
-However, the trade-off for good type I error control is demonstrably lower power, as shown in figure \ref{fig:ss_hypo}B.
-We included only adjusted cILR methods since power is only meaningful when type I error is properly controlled.
-Results indicated that both sparsity and inter-taxa correlation impacts power, correlation more so than sparsity.
-The adjusted cILR approaches were able to maintain power above 0.8 even in low effect size settings as long as there was negligible inter-taxa correlation within the set.
-However, in higher correlation scenarios ($\rho = 0.2$ and $0.5$), power was dramatically reduced to 0.25 and below and would require higher effect sizes to maintain power at 0.8, especially at higher sparsity levels.
-
-These results were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}.
-Here, we tested the enrichment of aerobic microbes at each sample using approaches similar to our parametric simulations.
-As expected in Figure \ref{fig:ss_hypo}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods.
-Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05.
-Power analysis (Figure \ref{fig:ss_hypo}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
-
-To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set.
-This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated.
-Figure \ref{fig:ss_auc} presents this result.
-We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $U$ test statistic from the Wilcoxon rank-sum test.
-Across both simulations (Figure \ref{fig:ss_auc}A) and real-data applications (Figure \ref{fig:ss_auc}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
-In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
-Notably, the W-statistic provided the least information for classifying samples with inflated taxa.
-
-\begin{figure}
+\begin{figure}[H]
     \centering
+    \includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
     \caption{Type I error rate \textbf{(A)}, and power \textbf{(B)} across different parametric simulation scenarios.
@@
 However, this resulted in lower power for adjusted cILR methods.}
-    \label{fig:ss_hypo}
+    \label{fig:2}
 \end{figure}
 
-\begin{figure}
+However, the trade-off for good type I error control is demonstrably lower power, as shown in figure \ref{fig:2}B.
+In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un-adjusted ones.
+However, in higher correlation scenarios, the difference in power is much more dramatic.
+At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted cILR was only performing at 50\% power, while unadjusted cILR and wilcoxon rank sum test were able to reach 80\%.
+These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of cILR.
+Most importantly, cILR demonstrate higher power in all scenarios where type I error is properly controlled.
+
+To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set.
+This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated.
+Figure \ref{fig:3} presents this result.
+We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
+Across both simulations (Figure \ref{fig:3}A) and real-data applications (Figure \ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
+In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
+Notably, the W-statistic provided the least information for classifying samples with inflated taxa.
+
+\subsubsection*{Real data evaluations}  
+These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}.
+Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations.
+As expected in Figure \ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods.
+Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05.
+Power analysis (Figure \ref{fig:2}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
+
+\begin{figure}[H]
     \centering
+    \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
     \caption{Classification performance via AUC of cILR, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{Methods}.
@@
 \subsection*{Differential abundance analysis}
-Here we assessed the ability to use cILR with a standard difference of means statistical test (such as Welch's t-test or a Wilcoxon rank-sum test) to perform differential abundance analysis.
-We compare the performance of cILR related methods to two commonly used approaches in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
+cILR generates sample-specific scores representing the degree of enrichment of a pre-defined set.
+As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test).
+We compared the performance of this approach with cILR and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
+
+\subsubsection*{Simulation studies}
+\subsubsection*{Real data evaluations}
+We also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP.
+For type I error evaluations, we use stool samples and randomly assign them with case/control status and calculated type I error as the proportion of genera identified as significantly different.
+For true positive rate evaluations, we use the gingival data set as detailed in the previous section, and calculated the true positive rate as the proportion of genera labeled as either anaerobic or aerobic that were found to be significant.
+We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of cILR were controlling for type I error at the defined $\alpha$ threshold of 0.05.
+Conversely, true positive rate for all methods were roughly more similar.
+As expected, using the Wilcoxon rank sum test demonstrated have power true positive rate, but the difference was not large.
+Surprisingly, using Welch's t-test demonstrated similar true positive rate as both corncob and DESeq2, while also controlling for type I error.
+Choice of underlying distribution for cILR did not seem to alter performance.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
+    \caption{Differential abundance analysis using corncob, DESeq2 and cILR with either Wilcoxon rank sum test or Welch's t-test.
+Panel \textbf{(A)} shows type I error results as the proportion of significant genera after 500 iterations where case/control status was assigned randomly to each sample.
+Panel \textbf{(B)} shows true positive rate results as the proportion of significant genera who are either obligate anaerobes or aerobes.
+Both evaluations use 16S rRNA gene sequencing data from HMP.
+Type I error evaluation used stool samples while the true positive rate evaluation used samples from the gingival site.
+Results showed that cILR associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2}
+    \label{fig:4}
+\end{figure}
 
 \subsection*{Disease Prediction}   
-We can conceive of cILR as a feature engineering step before model fit and prediction.
-Here we assessed the predictive performance of a standard machine learning model (random forest \cite{breiman2001}) using cILR scores compared to scores generated by other single sample enrichment methods in the gene set testing literature (ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}).
-Additionally, we also compared our results with just using the centered log-ratio (CLR) transformation on taxa sets aggregated via count summations.
-
-Figure \ref{fig:sim_pred_auc} and \ref{fig:sim_pred_rsq} showed results from simulation studies as detailed in the \nameref{Methods} section under both classification and regression tasks.
-As expected, performance across all assessed methods increased with a higher signal-to-noise ratio, however, they remained consistent across different levels of model saturation and inter-taxa correlation.
-Importantly, cILR methods outperformed both GSVA and ssGSEA, where the difference in AUC increases as a function of sparsity.
-Conversely, cILR scores did not perform as well as the traditional CLR approach, however, the performance gap decreased at higher sparsity levels.
-Similar patterns were observed for regression results (Figure \ref{fig:sim_pred_rsq}).
-Most interestingly, regression performance, in general, decreased at higher saturation levels (proportion of sets associated with the outcome).
-cILR remained more performant than both ssGSEA and GSVA, but was still less than the standard CLR, especially at higher signal-to-noise ratios.
-Similarly, cILR still performed much better in higher sparsity situations, where the results were almost identical to the CLR approach, especially in high saturation scenarios.
-
-In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:pred_data}.
-The learning task across both data sets is to classify patients who are diagnosed with inflammatory bowel disease (includes both Crohn's disease and ulcerative colitis) and healthy controls.
-The Gevers et al.
-\cite{gevers2014} data set is a 16S rRNA sequencing data set while the Nielsen et al.
-\cite{nielsen2014} is a whole-genome shotgun sequencing data set.
-Similar to simulation experiments, across both data sets cILR methods provide much better performance than both GSVA and ssGSEA.
-Interestingly, the standard CLR approach only outperformed cILR in the 16S rRNA data set but was marginally worse in the WGS data set.
-Additionally, we observed that using the normal distribution for cILR generates better predictive performance compared to the other variants.
-
-These results demonstrated that cILR generated scores are informative features in disease prediction tasks.
-Simulation results indicated that cILR methods perform much better than both ssGSEA and GSVA, but not as good as the standard CLR approach.
-Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
-
-\begin{figure}
-    \caption{AUC scores of a random forest model for a binary outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio.
-Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome).
+Since cILR can generate informative scores that can discriminate between samples with inflated counts for a set (Fig.
+\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models.
+In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating cILR, ssGSEA, and GSVA).
+Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxa sets aggregated via abundance summations.
+
+\subsubsection*{Simulation studies}
+Fig.
+\ref{fig:6} showed results for simulation studies as detailed in the \nameref{Methods} section.
+Panel A.
+presents results for a regression learning task with a continuous outcome while panel B.
+presents results for a classification task with a binary outcome.
+As expected, performance across all assessed methods increased with a higher signal-to-noise ratio.
+Both CLR and cILR approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks.
+This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while cILR and CLR did not experience a similar level of impact.
+As such, performance gap widens with increasing correlation and sparsity.
+Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.
+
+Unfortunately, cILR did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
+This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation.
+However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and cILR/CLR.
+Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity.
+Additionally, we did not observe any performance difference between the different variations of cILR.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
+    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
+Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure.
+Panel \textbf{(B)} presents performance on a classification task with AUC as the evaluation measure.
+cILR approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
+    \label{fig:6}
+\end{figure}
+
+%\begin{figure}[H]
+%    \centering
+%    \includegraphics[width=\textwidth]{figures/sim_pred_rsq.png}
+%    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on cILR,     %ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data %across sparsity levels, correlation, and signal-to-noise ratio.
+Panel \textbf{(A)} and \textbf{(B)} %represent results across different levels of model saturation (proportion of sets associated with the %outcome).
 cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
-    \label{fig:sim_pred_auc}
-\end{figure}
-
-\begin{figure}
-    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio.
-Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome).
-cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
-    \label{fig:sim_pred_rsq}
-\end{figure}
-
-\begin{figure}
-    \caption{Classification performance of a standard random forest model using cILR scores compared against existing methods in gene set testing literature and the standard centered-log ratio transformation approach.
-The learning task involves predicting patients with inflammatory bowel disease (including Crohn's disease and ulcerative colitis) versus controls.
-Data sets used span both 16S rRNA sequencing (Gevers et al.
+%    \label{fig:sim_pred_rsq}
+%\end{figure}
+
+\subsubsection*{Real data evaluations}
+In addition to parametric simulations, we also assessed the performance of using cILR scores in predictive models with real data sets.
+Fig.
+\ref{fig:7} presents results for two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
+The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al.
+\cite{gevers2014} data set uses 16S rRNA gene sequencing, while the Nielsen et al \cite{nielsen2014} data set uses whole genome shotgun sequencing.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[scale = 0.6]{figures/data_prediction_plot.png}
+    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
+Data sets used span both 16S rRNA gene sequencing (Gevers et al.
 \cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al.
 \cite{nielsen2014}).
-cILR marginally outperforms GSVA and ssGSEA in both data sets, but only performs better than CLR in the WGS data set.}
-    \label{fig:pred_data}
+cILR performs better than GSVA and ssGSEA but not as well as CLR, with the exception of the whole genome sequencing data set.}
+    \label{fig:7}
 \end{figure}
 
+Similar to simulation experiments, we also fitted a naive random forest model using CILR, ssGSEA, GSVA, or CLR transformed variables as inputs, and use AUC as the performance criteria.
+Results also replicated that of the simulations, where across both data sets cILR and CLR methods provide much better performance than both GSVA or ssGSEA.
+Interestingly, the cILR approach performed better than CLR in the whole genome data set but did not perform as well in the 16S rRNA gene sequencing data set.
+However, these results indicate that cILR generated scores are informative, providing competitive performance when acting as inputs to disease predictive models.
+Most importantly, performance values are consistent across both simulated and real data sets.
+
+These results demonstrate that cILR generated scores are informative features in disease prediction tasks.
+Simulation results indicate that cILR methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach.
+Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
+
 \section*{Discussion}
-\subsection*{Set-based analysis can enhance microbiome data analysis}
-Gene-set testing (or pathway analysis) is an important tool in the analysis of high-dimensional genomic data sets \cite{khatri2012}.
-By focusing on a group of variables rather than individuals, this approach reduces the hypothesis testing burden, thereby increasing power as well as replicability.
-Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also increase interpretability.
-
-In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate variables to taxonomic classification levels such as genus, family, or phylum.
-The process of aggregation involves the element-wise summation of counts of taxa that belong to the same taxonomic group.
-Even though this allows for a reduction in the number of overall variables (from thousands to only hundreds), there still exist two disadvantages: first, it doesn't preserve inter-sample distances before and after aggregation \cite{egozcue2005}, and second, it doesn't allow for enrichment testing and comparison across sets of different sizes.
-
-As such, there is a need for microbiome researchers to adopt set enrichment methods to perform a more robust set-based analysis that allows for enrichment testing of functionally relevant sets of microbes.
-Some software suites, such as \emph{MicrobiomeAnalyst}, have begun to provide an option to perform such analyses with curated taxa sets \cite{chong2020}.
-However, the approach used in \emph{MicrobiomeAnalyst} is based on a hypergeometric test of significant hypotheses and is sensitive to the method of differential abundance used.
-More recent methods in set-based testing generate enrichment scores at the sample level \cite{hanzelmann2013, frost2020a}.
-This approach allows for the flexible incorporation of different statistical techniques downstream, but unfortunately is not currently tailored to the compositional nature of microbiome relative abundance data \cite{gloor2017}.
-
-Here, we developed cILR, a novel microbiome-specific method to perform set-based enrichment analysis at the sample level.
-Our method is based on the isometric log-ratio transformation \cite{egozcue2003}, and measures set enrichment relative to those outside the set, which corresponds to a $Q_1$ null hypothesis in the gene set analysis literature \cite{tian2005}.
-cILR generates sample-specific enrichment scores, enabling cILR to act as both a significance test for enrichment at the sample level, and as inputs for downstream analyses such as differential abundance testing, predictive modeling, and data visualization.
-
-\subsection*{Significance testing with cILR}
-We can use cILR to test for enrichment of taxa-sets at the sample level.
-Inter-taxa correlation can result in an inflated variance of resulting test statistics \cite{wu2012}, which results in an inflated type I error if not properly controlled for.
-Our method addresses this issue by combining the mean estimate of scores computed on permuted data and the variance estimate of scores computed on unpermuted data, where correlation is undisrupted.
-As a consequence, adjusted cILR methods control very well for type I error at the appropriate $\alpha$ level (Figure \ref{fig:ss_hypo}).
-The conservativeness of the test persisted across different sparsity levels, as well as inter-taxa correlation.
-Conversely, the unadjusted cILR methods and the naive Wilcoxon rank-sum test has inflated type I error even at low correlation levels ($\rho = 0.2$).
-However, the trade-off for proper type I error control is lower power.
-Simulation results showed that power is sensitive to sparsity across all correlation levels.
-
-However, if there is no correlation between members of a set, cILR was still able to maintain 80\% power even with low effect size (fold change of 1.5), long as the set of interest is more densely populated, such as a set of core taxa that are prevalent across all samples.
-Unfortunately, this relationship doesn't hold for higher correlation levels.
-This means that in sets where associated taxa are highly correlated, a higher fold change is required for cILR to detect enrichment.
-
-We also assessed cILR using different distribution forms, the normal distribution, and the mixture normal distribution.
-The difference in performance across the two choices was minimal, where both were able to control for type I error, but the normal distribution was slightly better when it comes to power only in certain simulation settings.
-This runs contrary to our distribution comparison analysis in Figure \ref{fig:kurt_gof}, where the mixture distribution was the superior fit.
-We hypothesized that this might be due to the difficulty in fitting mixture distributions using the expectation-maximization algorithm (EM), as the convergence rate is slow when there is a lot of overlap between the mixtures and the mixing coefficients of one of the components are low \cite{naim}.
-Furthermore, there is a lot of degeneracy of parameter values in our variance adjustment procedure for the mixture distribution(Supplemental Material).
-These distribution fitting problems might introduce issues where the mixture distribution might not be the best fit.
-As such, improvements to mixture distribution fitting might allow for better performance to be extracted.
-
-These results are replicated in our real data analysis, where we attempted to test for enrichment for a group of aerobic taxa among samples in the gingival site from HMP \cite{consortium2012}.
-Even though our labeling is not perfect, it is encouraging to see that the false discovery rate is low (approximately 0.05) when tested among a set of hypotheses where there is a mixture of true null and alternate situations.
-We believe that a practitioner can identify many interesting samples to follow up on by applying multiple testing adjustment procedures (such as Benjamini-Hochberg \cite{benjamini1995}) with cILR p-values and using less stringent criteria (e.g.
-0.1 false discovery rate).
-
-In short, we demonstrated that cILR can be used to test for significance testing at the sample level.
-Even though there is not a lot of difference between the distribution choices, ultimately cILR controls for type I error well despite tradeoffs in power.
-Even then, cILR will still be able to detect enriched sets in high correlation and sparsity situations insofar as the effect size is large.
-Researchers can utilize cILR to test for enrichment of a certain set of taxa of interest to perform follow-up experiments or to identify the prevalence of certain sub-population of taxa (for example, oxygen preference as in our real data example).
-The conservativeness of the test ensures confidence in detected samples.
-
-\subsection*{Using cILR as inputs for downstream analyses}
-Since cILR generates enrichment scores for each set at the sample level, they can be used in downstream analyses alongside standard statistical approaches.
-In this section we assessed the utility of cILR generated scores in two common analysis tasks in microbiome research: differential abundance testing and disease prediction.
-
-For disease prediction, we benchmarked a basic random forest model \cite{breiman2001} to predict whether or not a patient has IBD given only their microbiome profile using cILR and relevant methods as inputs.
-Across both simulations and real data analyses, cILR scores perform better than alternative single sample scoring methods in the gene set testing literature (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}).
-This is consistent with previous analysis (Figure \ref{fig:ss_auc}), where we assessed the informativeness of cILR generated scores by using AUC to measure whether samples with high scores correspond with the true label of an inflated set.
-Results indicated that cILR scores are more adaptive to microbiome data sets than GSVA and ssGSEA.
-However, cILR scores underperform the standard approach of applying the CLR transformation to count aggregated data, except for the WGS real data analysis.
-Despite this, there are still benefits to using cILR scores in the context of downstream modeling beyond random forest predictive models.
-First, CLR transformed variables have a sum to zero constraints, making the covariance matrix is singular \cite{gloor2017}, impacting methods that rely on matrix decomposition such as principal component regression or canonical correlation analysis.
-Second, cILR provides an additional value of significance testing for enrichment for sample screening.
-Third, CLR transformed variables still involve the count summation procedure for variable aggregation, therefore comparison between sets of different sizes remains unviable.
-
-Despite such drawbacks, cILR methods still produce competitive performance values across both regression and classification tasks even in low signal-to-noise settings.
-Most interestingly, the performance gap between cILR and CLR was smaller in high sparsity settings ($p = 0.6$), which is a feature of microbiome relative abundance data \cite{li2019}.
-Furthermore, these predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), and good performance also indicated that cILR scores can also capture the joint distribution of considers sets, especially with regards to their relationship with a defined outcome such as IBD status.
-
-Unsurprisingly, different variants of cILR did not differ significantly in performance similar to significance testing experiments.
-Even though we expect Z-scores to perform better due to the added information of directionality of enrichment, using CDF values did not alter performance in both the single sample AUC evaluations (Figure \ref{fig:ss_auc}) as well as the random forest evaluations (Figure \ref{fig:sim_pred_auc}).
-
-In short, users can apply cILR generated enrichment scores as inputs to downstream analyses, especially predictive models and differential abundance analysis.
-cILR scores are informative in discriminating samples between classes even in low effect size situations.
-Even though cILR did not outperform a standard analysis of count aggregation followed by the CLR transformation, the additional benefits listed above would prove to be attractive in certain use cases, especially in high sparsity situations.
+
+\subsection*{Inference with cILR}
+The formulation of cILR as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}.
+This allows for conducting inference with cILR even at the sample-level.
+We assessed the usage of cILR in this type of analysis by evaluating type I error and power across both simulation studies and real data applications.
+Most importantly, we demonstrated that our adjusted cILR approach was able to address the issue of variance inflation due to correlation \cite{wu2012} by controlling for type I error at the appropriate $\alpha$ level across different levels of simulated inter-taxa correlation (Fig.
+\ref{fig:2}) while conversely unadjusted cILR and the naive Wilcoxon rank sum test showed much higher rates of error.
+This is further encouraged in real data analysis where the false discovery rate was around $0.05$ when a collection of true null and true alternate hypotheses were tested.
+Unfortunately the trade-off between good type I error control is lower power.
+The conservativeness of the test attenuates with higher sparsity and correlation, where power was not approaching even 50\% even at the highest effect sizes.
+However, when the degree of such data features are reasonable, cILR will still be able to detect a reasonable proportion of samples with inflated counts.
+
+We also observed that choosing different distributional forms did not alter performance values for cILR.
+This runs contrary to our comparison analysis in Fig.
+\ref{fig:1}, where we demonstrated that the mixture normal distribution had superior fit compared to the simple normal for raw cILR scores computed under the global null.
+We hypothesized that this might be due to the difficulty in fitting mixture distribution to data using the expectation maximization algorithm, as the convergence rate is slow when there is high overlap between the mixtures, resulting in a small mixing coefficient for one of the components \cite{naim2012}.
+As such, in our implementation of cILR, in order to ensure convergence for the estimating procedure we increased the number of iterations while relaxing the tolerance parameter.
+Furthermore, there are also possible problems with our adjustment procedure for the mixture distribution that might impact overall fit.
+In order to combine the scale and location estimate of two mixture distributions, we fixed the overall mean, standard deviation, mixing coefficient and component-wise means and used an optimization procedure to find the component-wise variances.
+However, this means that we have one equation for the overall variance but two possible parameters to estimate.
+As such, there is no guaranteed unique solution to component-wise variances.
+We hypothesized that the instability and degeneracy in component-wise variance estimates might impact the fidelity of estimates at the tails of the distribution, thereby affecting inference.
+
+Despite these concerns, empirical results still indicate that cILR can confidently identify samples with inflated counts.
+The conservativeness of the correlation adjustment procedure ensures that significant results can be trusted by practitioners, even if cILR might not be able to exhaustively identify all samples with inflated counts.
+In situations where either the data is less sparse (e.g.
+containing a lot of core taxa that are prevalent across all samples), there is less inter-taxa correlation within the set (e.g.
+taxa that do not participate in common pathways but have shared characteristics like pathogenicity), or if the effect size is large, then cILR will still be able to produce reasonable power.
+A practitioner can use cILR to screen for samples for subsequent analysis that might involve significant costs, or perform hypothesis generation using a less stringent criteria alongside a multiple testing adjustment procedure (such as Benjamini-Hochberg \cite{benjamini1995}).
+
+\subsection*{Downstream analysis}
+The sample-level enrichment scores generated by the cILR method can be used in downstream analyses commonly performed in microbiome research: differential abundance testing and disease prediction.
+
+\subsubsection*{Differential abundance analysis}
+For differential abundance testing, we evaluated whether using cILR alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes.
+We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014}.
+We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome data \cite{mcmurdie2014}.
+Conversely, corncob is a newer method developed specifically for microbiome data sets, where taxonomic counts are modeled directly newly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.
+In our analyses, we found that both corncob and DESeq2 has inflated levels of type I error.
+
+\subsubsection*{Predictive models}
+For disease prediction, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using cILR generated scores as inputs.
+Similar to our inference analysis, we compared cILR against both ssGSEA and GSVA.
+Additionally, we also evaluated cILR with the approach where counts of a set were aggregated using sums and then centered log-ratio transformed.
+This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}.
+Results indicated that cILR produce good performance values across both real data analysis and simulation scenarios.
+Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicate that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
+Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome data sets.
+This is consistent with our sample ranking analysis (Fig.
+\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
+However, cILR did not outperform the CLR approach across all our simulation studies, and only marginally performed better in the real data analysis with WGS data.
+
+However, in simulation studies, this performance gap between CLR and cILR decreases with higher sparsity and correlation, especially in low effect saturation scenarios.
+Additionally, there are also downsides to applying CLR.
+First, the singular covariance matrix of CLR transformed variables is singular due to a sum to zero constraint \cite{gloor2017}, preventing the proper usage of approaches that rely on matrix decomposition.
+Second, the procedure still relies on using summation of counts prior to transformation, which means that we still can't compare across sets of different sizes, and any bias might still be propagated \cite{mclaren2019}.
+As such, despite benefits in performance for a naive random forest model, there is still space for using cILR as primary inputs into predictive models.
+
+Similar to other experiments in downstream usage of cILR, performance did not change with different underlying distributions, output types, or correlation status.
+This is surprising since we expect z-scores to perform better as they are able to capture the direction of an association.
+The fact that this effect persisted even onto our real data analysis suggests that this is not due to a deficiency of our simulation design.
+As such, practitioners who wish to use cILR in predictive models might be suited to use the settings that is the fastest to compute.
+
+Ultimately, results indicate that cILR can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity.
+Even though there exists situations where it might not provide maximum predictive values, the flexibility of cILR in various types of analyses enable ven though in some scenarios it might not provide maximum predictive values.
 
 \subsection*{Limitations and future directions} 
-There are various limitations to using cILR to generate sample-level taxonomic aggregation for microbiome data.
-First, significance testing at the sample level is not powered to assess inflated sets with low effect size.
-Practitioners can be confident in the validity of a significant result (following proper multiple testing correction) due to cILR's type I error control, but should not expect cILR to be able to identify all samples with inflated counts.
-Second, even though cILR is flexible and addresses the compositional nature of microbiome data, the performance of cILR might not be consistently better than competing methods.
-Careful evaluation of methods should be done before application to ensure that the value of cILR scores is appropriately extracted in suitable models.
-Third, cILR did not directly address the zero-inflatedness of microbiome data and utilizes a pseudocount to ensure log operations are valid.
-As such, practitioners are encouraged to choose a different pseudocount or an appropriate method of choice.
-
-As such, we hope to address the aforementioned issues with cILR in later projects.
-First, we hope to refine the distribution fitting procedure for the mixture distribution due to its superior fit and address the identifiability problem that might underlie the lack of power in our experiments.
-Second, we hope to incorporate some of the sophisticated model-based zero correction methods \cite{martin-fernandez2012,kaul2017a}.
-Third, we can account for the difference in importance of certain taxa by incorporating taxa-specific weights.
-Finally, we also hope to curate more interest apriori sets that are functionally informative, which would generate more interesting insight and improve interpretability compared to using taxonomic categories such as Phylum or Genus.
+There are various limitations to our evaluation of cILR.
+First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome data.
+There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation section of prior studies.
+Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is flawed.
+This is because the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut.
+As such, results from power analyses using this data set is only relative between the comparison methods, and absolute values should should not carry meaning.
+Finally, we assumed that taxa within a set all be equally associated with the outcome.
+This limits our ability to evaluate the performance of cILR when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.
+
+However, our evaluation still showed various drawbacks of the cILR method.
+First, inference with cILR is limited in being able to exhaustively detect all samples with significant inflated counts for a set in situations where there is a high degree of sparsity and inter-taxa correlation.
+Second, for downstream analyses, cILR might not always better than competing methods, especially when being used as inputs to predictive models.
+We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated with adjusting the mixture normal distribution.
+As such, we hope to refine the null distribution estimating procedure by either choosing a better distributional form, or to further constrain the optimization procedure of the mixture normal distribution by fixing the third and fourth moments.
+
+In addition, there are possible extensions cILR can we can consider to provide more flexibility across different types of scenarios in data analysis.
+First, cILR did not address the sparsity of microbiome data and relies on pseudocount to ensure log operations are valid.
+We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
+Second, cILR also treated all taxa within the set as equally contributing to the set.
+Incorporate taxa-specific weights can reduce the influence of outliers, such as rare or highly invariant taxa.
+Finally, curating sets based on apriori characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.
 
 \section*{Conclusion}
-Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomic data sets.
+Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomics data sets.
 However, there has not been a lot of set-based analysis methods developed specifically for microbiome relative abundance data.
@@
 
-%\bibliography{tax_agg}{}
-
-\begin{thebibliography}{10}
-
-    \bibitem{cho2012}
-    Cho I, Blaser MJ.
-    \newblock The Human Microbiome: At the Interface of Health and Disease.
-    \newblock Nature Reviews Genetics. 2012;13(4):260--270.
-    \newblock doi:{10.1038/nrg3182}.
-
-    \bibitem{callahan2016}
-    Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
-    \newblock {{DADA2}}: {{High}}-Resolution Sample Inference from {{Illumina}}
-      Amplicon Data.
-    \newblock Nature Methods. 2016;13(7):581--583.
-    \newblock doi:{10.1038/nmeth.3869}.
-
-    \bibitem{truong2015}
-    Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
-    \newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
-    \newblock Nature Methods. 2015;12(10):902--903.
-    \newblock doi:{10.1038/nmeth.3589}.
-
-    \bibitem{li2019}
-    Li H.
-    \newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
-      {{Metagenomics}}.
-    \newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
-      2019. p. 977--550.
-
-    \bibitem{li2015}
-    Li H.
-    \newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
-      Compositional Data Analysis}}.
-    \newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
-    \newblock doi:{10.1146/annurev-statistics-010814-020351}.
-
-    \bibitem{shi2016}
-    Shi P, Zhang A, Li H.
-    \newblock Regression Analysis for Microbiome Compositional Data.
-    \newblock The Annals of Applied Statistics. 2016;10(2):1019--1040.
-    \newblock doi:{10.1214/16-AOAS928}.
-
-    \bibitem{sankaran2014}
-    Sankaran K, Holmes S.
-    \newblock {{structSSI}}: {{Simultaneous}} and {{Selective Inference}} for
-      {{Grouped}} or {{Hierarchically Structured Data}}.
-    \newblock Journal of statistical software. 2014;59(13):1--21.
-    \newblock doi:{10.18637/jss.v059.i13}.
-
-    \bibitem{benjamini1995}
-    Benjamini Y, Hochberg Y.
-    \newblock Controlling the {{False Discovery Rate}}: {{A Practical}} and
-      {{Powerful Approach}} to {{Multiple Testing}}.
-    \newblock Journal of the Royal Statistical Society Series B (Methodological).
-      1995;57(1):289--300.
-
-    \bibitem{chen2018}
-    Chen J, King E, Deek R, Wei Z, Yu Y, Grill D, et~al.
-    \newblock An Omnibus Test for Differential Distribution Analysis of Microbiome
-      Sequencing Data.
-    \newblock Bioinformatics. 2018;34(4):643--651.
-    \newblock doi:{10.1093/bioinformatics/btx650}.
-
-    \bibitem{weiss2017}
-    Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
-    \newblock Normalization and Microbial Differential Abundance Strategies Depend
-      upon Data Characteristics.
-    \newblock Microbiome. 2017;5(1).
-    \newblock doi:{10.1186/s40168-017-0237-y}.
-
-    \bibitem{mcmurdie2014}
-    McMurdie PJ, Holmes S.
-    \newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
-      Inadmissible}}.
-    \newblock PLOS Computational Biology. 2014;10(4):e1003531.
-    \newblock doi:{10.1371/journal.pcbi.1003531}.
-
-    \bibitem{love2014}
-    Love MI, Huber W, Anders S.
-    \newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
-      Data with {{DESeq2}}.
-    \newblock Genome Biology. 2014;15(12):550.
-    \newblock doi:{10.1186/s13059-014-0550-8}.
-
-    \bibitem{quinn2019}
-    Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
-    \newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
-    \newblock GigaScience. 2019;8(giz107).
-    \newblock doi:{10.1093/gigascience/giz107}.
-
-    \bibitem{quinn2018b}
-    Quinn TP, Erb I, Richardson MF, Crowley TM.
-    \newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
-    \newblock Bioinformatics. 2018;34(16):2870--2878.
-    \newblock doi:{10.1093/bioinformatics/bty175}.
-
-    \bibitem{gloor2017}
-    Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
-    \newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
-      Optional}}.
-    \newblock Frontiers in Microbiology. 2017;8.
-    \newblock doi:{10.3389/fmicb.2017.02224}.
-
-    \bibitem{aitchison1999}
-    Aitchison J.
-    \newblock A {{Concise Guide}} to {{Compositional Data Analysis}}. 1999; p. 134.
-
-    \bibitem{kurtz2015}
-    Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
-    \newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
-      Ecological Networks}}.
-    \newblock PLOS Computational Biology. 2015;11(5):e1004226.
-    \newblock doi:{10.1371/journal.pcbi.1004226}.
-
-    \bibitem{kaul2017}
-    Kaul A, Davidov O, Peddada SD.
-    \newblock Structural Zeros in High-Dimensional Data with Applications to
-      Microbiome Studies.
-    \newblock Biostatistics. 2017;18(3):422--433.
-    \newblock doi:{10.1093/biostatistics/kxw053}.
-
-    \bibitem{kaul2017a}
-    Kaul A, Mandal S, Davidov O, Peddada SD.
-    \newblock Analysis of {{Microbiome Data}} in the {{Presence}} of {{Excess
-      Zeros}}.
-    \newblock Frontiers in Microbiology. 2017;8.
-    \newblock doi:{10.3389/fmicb.2017.02114}.
-
-    \bibitem{egozcue2005}
-    Egozcue JJ, {Pawlowsky-Glahn} V.
-    \newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
-      Analysis}}.
-    \newblock Mathematical Geology. 2005;37(7):795--828.
-    \newblock doi:{10.1007/s11004-005-7381-9}.
-
-    \bibitem{goeman2007}
-    Goeman JJ, B{\"u}hlmann P.
-    \newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
-      Issues.
-    \newblock Bioinformatics. 2007;23(8):980--987.
-    \newblock doi:{10.1093/bioinformatics/btm051}.
-
-    \bibitem{subramanian2005}
-    Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
-    \newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
-      Interpreting Genome-Wide Expression Profiles.
-    \newblock Proceedings of the National Academy of Sciences.
-      2005;102(43):15545--15550.
-    \newblock doi:{10.1073/pnas.0506580102}.
-
-    \bibitem{hanzelmann2013}
-    H{\"a}nzelmann S, Castelo R, Guinney J.
-    \newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
-      {{RNA}}-{{Seq}} Data.
-    \newblock BMC Bioinformatics. 2013;14(1):7.
-    \newblock doi:{10.1186/1471-2105-14-7}.
-
-    \bibitem{frost2020a}
-    Frost HR.
-    \newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
-      Method for Cell-Specific Gene Set Scoring.
-    \newblock Nucleic Acids Research. 2020;48(16):e94--e94.
-    \newblock doi:{10.1093/nar/gkaa582}.
-
-    \bibitem{tian2005}
-    Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
-    \newblock Discovering Statistically Significant Pathways in Expression
-      Profiling Studies.
-    \newblock Proceedings of the National Academy of Sciences.
-      2005;102(38):13544--13549.
-    \newblock doi:{10.1073/pnas.0506577102}.
-
-    \bibitem{egozcue2003}
-    Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcelo-Vidal} C.
-    \newblock Isometric {{Logratio Transformations}} for {{Compositional Data
-      Analysis}}.
-    \newblock Mathematical Geology. 2003; p.~22.
-
-    \bibitem{barbie2009}
-    Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
-    \newblock Systematic {{RNA}} Interference Reveals That Oncogenic
-      {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
-    \newblock Nature. 2009;462(7269):108--112.
-    \newblock doi:{10.1038/nature08460}.
-
-    \bibitem{quast2013}
-    Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
-    \newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
-      Processing and Web-Based Tools.
-    \newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
-    \newblock doi:{10.1093/nar/gks1219}.
-
-    \bibitem{delignette-muller2015}
-    {Delignette-Muller} ML, Dutang C.
-    \newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
-    \newblock Journal of Statistical Software. 2015;64(4):1--34.
-
-    \bibitem{benaglia2009}
-    Benaglia T, Chauveau D, Hunter DR, Young D.
-    \newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
-    \newblock Journal of Statistical Software. 2009;32(6):1--29.
-
-    \bibitem{washburne2017}
-    Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL, Mukherjee S, et~al.
-    \newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
-      Associations in Microbiome Datasets.
-    \newblock PeerJ. 2017;5:e2969.
-    \newblock doi:{10.7717/peerj.2969}.
-
-    \bibitem{silverman2017}
-    Silverman JD, Washburne AD, Mukherjee S, David LA.
-    \newblock A Phylogenetic Transform Enhances Analysis of Compositional
-      Microbiota Data.
-    \newblock eLife. 2017;6:e21887.
-    \newblock doi:{10.7554/eLife.21887}.
-
-    \bibitem{morton2017}
-    Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
-      et~al.
-    \newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
-    \newblock mSystems. 2017;2(1).
-    \newblock doi:{10.1128/mSystems.00162-16}.
-
-    \bibitem{efron2004}
-    Efron B.
-    \newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
-    \newblock Journal of the American Statistical Association.
-      2004;99(465):96--104.
-    \newblock doi:{10.1198/016214504000000089}.
-
-    \bibitem{wu2012}
-    Wu D, Smyth GK.
-    \newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
-      Correlation.
-    \newblock Nucleic Acids Research. 2012;40(17):e133.
-    \newblock doi:{10.1093/nar/gks461}.
-
-    \bibitem{calgaro2020}
-    Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
-    \newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
-      and Metagenomics Applied to Microbiome Data.
-    \newblock Genome Biology. 2020;21(1):191.
-    \newblock doi:{10.1186/s13059-020-02104-1}.
-
-    \bibitem{cario1997}
-    Cario MC.
-    \newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
-      Distributions}} and {{Correlation Matrix}}. 1997; p.~19.
-
-    \bibitem{agresti1998}
-    Agresti A, Coull BA.
-    \newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
-      Estimation}} of {{Binomial Proportions}}.
-    \newblock The American Statistician. 1998;52(2):119--126.
-    \newblock doi:{10.2307/2685469}.
-
-    \bibitem{delong1988}
-    DeLong ER, DeLong DM, {Clarke-Pearson} DL.
-    \newblock Comparing the Areas under Two or More Correlated Receiver Operating
-      Characteristic Curves: A Nonparametric Approach.
-    \newblock Biometrics. 1988;44(3):837--845.
-
-    \bibitem{hawinkel2019}
-    Hawinkel S, Mattiello F, Bijnens L, Thas O.
-    \newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
-      Control the False Discovery Rate.
-    \newblock Briefings in Bioinformatics. 2019;20(1):210--221.
-    \newblock doi:{10.1093/bib/bbx104}.
-
-    \bibitem{breiman2001}
-    Breiman L.
-    \newblock Random {{Forests}}.
-    \newblock Machine Learning. 2001;45(1):5--32.
-    \newblock doi:{10.1023/A:1010933404324}.
-
-    \bibitem{kuhn2020}
-    Kuhn M, Wickham H.
-    \newblock Tidymodels: {{Easily}} Install and Load the 'tidymodels' Packages;
-      2020.
-
-    \bibitem{xiao2018}
-    Xiao J, Chen L, Yu Y, Zhang X, Chen J.
-    \newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
-      {{Predictive Modeling}} of {{Microbial Community Data}}.
-    \newblock Frontiers in Microbiology. 2018;9.
-    \newblock doi:{10.3389/fmicb.2018.03112}.
-
-    \bibitem{dong2020}
-    Dong M, Li L, Chen M, Kusalik A, Xu W.
-    \newblock Predictive Analysis Methods for Human Microbiome Data with
-      Application to {{Parkinson}}'s Disease.
-    \newblock PLOS ONE. 2020;15(8):e0237779.
-    \newblock doi:{10.1371/journal.pone.0237779}.
-
-    \bibitem{pasolli2017}
-    Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
-    \newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
-    \newblock Nature Methods. 2017;14(11):1023--1024.
-    \newblock doi:{10.1038/nmeth.4468}.
-
-    \bibitem{schiffer2019}
-    Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
-    \newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
-      through Bioconductor.
-    \newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
-
-    \bibitem{gonzalez2018}
-    Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
-      Ackermann G, et~al.
-    \newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
-    \newblock Nature Methods. 2018;15(10):796--798.
-    \newblock doi:{10.1038/s41592-018-0141-9}.
-
-    \bibitem{consortium2012}
-    Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
-      et~al.
-    \newblock Structure, Function and Diversity of the Healthy Human Microbiome.
-    \newblock Nature. 2012;486(7402):207--214.
-    \newblock doi:{10.1038/nature11234}.
-
-    \bibitem{proctor2019}
-    Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
-    \newblock The {{Integrative Human Microbiome Project}}.
-    \newblock Nature. 2019;569(7758):641--648.
-    \newblock doi:{10.1038/s41586-019-1238-8}.
-
-    \bibitem{thurnheer2016}
-    Thurnheer T, Bostanci N, Belibasakis GN.
-    \newblock Microbial Dynamics during Conversion from Supragingival to
-      Subgingival Biofilms in an in Vitro Model.
-    \newblock Molecular Oral Microbiology. 2016;31(2):125--135.
-    \newblock doi:{10.1111/omi.12108}.
-
-    \bibitem{beghini2019}
-    Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
-    \newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
-      in the {{New York City Health}} and {{Nutrition Examination Study}}.
-    \newblock Annals of Epidemiology. 2019;34:18--25.e3.
-    \newblock doi:{10.1016/j.annepidem.2019.03.005}.
-
-    \bibitem{matteocalgaro2020}
-    Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
-    \newblock Zenodo.
-
-    \bibitem{nielsen2014}
-    Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
-    \newblock Identification and Assembly of Genomes and Genetic Elements in
-      Complex Metagenomic Samples without Using Reference Genomes.
-    \newblock Nature Biotechnology. 2014;32(8):822--828.
-    \newblock doi:{10.1038/nbt.2939}.
-
-    \bibitem{gevers2014}
-    Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
-      et~al.
-    \newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
-      {{Disease}}.
-    \newblock Cell Host \& Microbe. 2014;15(3):382--392.
-    \newblock doi:{10.1016/j.chom.2014.02.005}.
-
-    \bibitem{martin2020}
-    Martin BD, Witten D, Willis AD.
-    \newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
-      Regression.
-    \newblock The Annals of Applied Statistics. 2020;14(1):94--115.
-    \newblock doi:{10.1214/19-AOAS1283}.
-
-    \bibitem{khatri2012}
-    Khatri P, Sirota M, Butte AJ.
-    \newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
-      {{Outstanding Challenges}}.
-    \newblock PLOS Computational Biology. 2012;8(2):e1002375.
-    \newblock doi:{10.1371/journal.pcbi.1002375}.
-
-    \bibitem{ashburner2000}
-    Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
-    \newblock Gene {{Ontology}}: Tool for the Unification of Biology.
-    \newblock Nature genetics. 2000;25(1):25--29.
-    \newblock doi:{10.1038/75556}.
-
-    \bibitem{chong2020}
-    Chong J, Liu P, Zhou G, Xia J.
-    \newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
-      Functional, and Meta-Analysis of Microbiome Data.
-    \newblock Nature Protocols. 2020;15(3):799--821.
-    \newblock doi:{10.1038/s41596-019-0264-1}.
-
-    \bibitem{naim}
-    Naim I, Gildea D.
-    \newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
-      {{Unbalanced Mixing Coefficients}}; p.~8.
-
-    \bibitem{martin-fernandez2012}
-    {Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
-      {Palarea-Albaladejo} J.
-    \newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
-      {{Classical}} and Robust Approaches.
-    \newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
-    \newblock doi:{10.1016/j.csda.2012.02.012}.
-
-    \end{thebibliography}
+\bibliography{tax_agg}{}
 

### 2021-06-02 tax_agg.tex

@@
 For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test.
+All three approaches were applied directly on count data (after pseudocount).
+For GSVA, the Poisson kernel was used.
 
@@
 
+\subsubsection*{Simulation studies}
 Panel A and B in figure \ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
@@
 
+To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set.
+This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated.
+Figure \ref{fig:3} presents this result.
+We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
+Across both simulations (Figure \ref{fig:3}A) and real-data applications (Figure \ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
+In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
+Notably, the W-statistic provided the least information for classifying samples with inflated taxa.
+
+\subsubsection*{Real data evaluations}  
 These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}.
@@
 Power analysis (Figure \ref{fig:2}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
-
-To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set.
-This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated.
-Figure \ref{fig:3} presents this result.
-We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $U$ test statistic from the Wilcoxon rank sum test.
-Across both simulations (Figure \ref{fig:3}A) and real-data applications (Figure \ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
-In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
-Notably, the W-statistic provided the least information for classifying samples with inflated taxa.
 
@@
 \subsection*{Differential abundance analysis}
-Here we assessed the ability to use cILR with a standard difference of means statistical test (such as Welch's t-test or a Wilcoxon rank sum test) to perform differential abundance analysis.
-We compared the performance of cILR related methods to two commonly used approaches in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
+cILR generates sample-specific scores representing the degree of enrichment of a pre-defined set.
+As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test).
+We compared the performance of this approach with cILR and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
+
+\subsubsection*{Simulation studies}
+\subsubsection*{Real data evaluations}
+We also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP.
+For type I error evaluations, we use stool samples and randomly assign them with case/control status and calculated type I error as the proportion of genera identified as significantly different.
+For true positive rate evaluations, we use the gingival data set as detailed in the previous section, and calculated the true positive rate as the proportion of genera labeled as either anaerobic or aerobic that were found to be significant.
+We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of cILR were controlling for type I error at the defined $\alpha$ threshold of 0.05.
+Conversely, true positive rate for all methods were roughly more similar.
+As expected, using the Wilcoxon rank sum test demonstrated have power true positive rate, but the difference was not large.
+Surprisingly, using Welch's t-test demonstrated similar true positive rate as both corncob and DESeq2, while also controlling for type I error.
+Choice of underlying distribution for cILR did not seem to alter performance.
 
@@
     \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
-    \caption{Placeholder} 
+    \caption{Differential abundance analysis using corncob, DESeq2 and cILR with either Wilcoxon rank sum test or Welch's t-test.
+Panel \textbf{(A)} shows type I error results as the proportion of significant genera after 500 iterations where case/control status was assigned randomly to each sample.
+Panel \textbf{(B)} shows true positive rate results as the proportion of significant genera who are either obligate anaerobes or aerobes.
+Both evaluations use 16S rRNA gene sequencing data from HMP.
+Type I error evaluation used stool samples while the true positive rate evaluation used samples from the gingival site.
+Results showed that cILR associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2}
     \label{fig:4}
@@
 \subsection*{Disease Prediction}   
-We can conceive of cILR as a feature engineering step before model fitting and prediction.
-Here we assessed the predictive performance of a standard machine learning model (random forest \cite{breiman2001}) using cILR scores compared to scores generated by other single sample enrichment methods in the gene set testing literature (ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}).
-Additionally, we also compared our results with the use of the centered log-ratio (CLR) transformation on taxa sets aggregated via count summations.
+Since cILR can generate informative scores that can discriminate between samples with inflated counts for a set (Fig.
+\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models.
+In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating cILR, ssGSEA, and GSVA).
+Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxa sets aggregated via abundance summations.
 
-Figure \ref{fig:6} and \ref{fig:6} showed results from simulation studies as detailed in the \nameref{Methods} section under both classification and regression tasks.
-As expected, performance across all assessed methods increased with a higher signal-to-noise ratio, however, they remained consistent across different levels of model saturation and inter-taxa correlation.
-Importantly, cILR methods outperformed both GSVA and ssGSEA, where the difference in AUC increases as a function of sparsity.
-Conversely, cILR scores did not perform as well as the traditional CLR approach, however, the performance gap decreased at higher sparsity levels.
-Similar patterns were observed for regression results (Figure \ref{fig:6}).
-Most interestingly, regression performance, in general, decreased at higher saturation levels (proportion of sets associated with the outcome).
-cILR remained more performant than both ssGSEA and GSVA, but was still less than the standard CLR, especially at higher signal-to-noise ratios.
-Similarly, cILR still performed much better in higher sparsity situations, where the results were almost identical to the CLR approach, especially in high saturation scenarios.
+\subsubsection*{Simulation studies}
+Fig.
+\ref{fig:6} showed results for simulation studies as detailed in the \nameref{Methods} section.
+Panel A.
+presents results for a regression learning task with a continuous outcome while panel B.
+presents results for a classification task with a binary outcome.
+As expected, performance across all assessed methods increased with a higher signal-to-noise ratio.
+Both CLR and cILR approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks.
+This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while cILR and CLR did not experience a similar level of impact.
+As such, performance gap widens with increasing correlation and sparsity.
+Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.
+
+Unfortunately, cILR did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
+This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation.
+However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and cILR/CLR.
+Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity.
+Additionally, we did not observe any performance difference between the different variations of cILR.
 
@@
     \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
-    \caption{AUC scores of a random forest model for a binary outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio.
-Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome).
-cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
+    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
+Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure.
+Panel \textbf{(B)} presents performance on a classification task with AUC as the evaluation measure.
+cILR approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
     \label{fig:6}
@@
 
-In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:7}.
-The learning task across both data sets is to classify patients who are diagnosed with inflammatory bowel disease (includes both Crohn's disease and ulcerative colitis) and healthy controls.
-The Gevers et al.
-\cite{gevers2014} data set is a 16S rRNA sequencing data set while the Nielsen et al.
-\cite{nielsen2014} is a whole-genome shotgun sequencing data set.
-Similar to simulation experiments, across both data sets cILR methods provide much better performance than either GSVA or ssGSEA.
-Interestingly, the standard CLR approach only outperformed cILR in the 16S rRNA data set but was marginally worse in the WGS data set.
-Additionally, we observed that using the normal distribution for cILR generates better predictive performance compared to the other variants.
+\subsubsection*{Real data evaluations}
+In addition to parametric simulations, we also assessed the performance of using cILR scores in predictive models with real data sets.
+Fig.
+\ref{fig:7} presents results for two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
+The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al.
+\cite{gevers2014} data set uses 16S rRNA gene sequencing, while the Nielsen et al \cite{nielsen2014} data set uses whole genome shotgun sequencing.
 
@@
     \includegraphics[scale = 0.6]{figures/data_prediction_plot.png}
-    \caption{Classification performance of a standard random forest model using cILR scores compared against existing methods in gene set testing literature and the standard centered-log ratio transformation approach.
-The learning task involves predicting patients with inflammatory bowel disease (including Crohn's disease and ulcerative colitis) versus controls.
-Data sets used span both 16S rRNA sequencing (Gevers et al.
+    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
+Data sets used span both 16S rRNA gene sequencing (Gevers et al.
 \cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al.
 \cite{nielsen2014}).
-cILR marginally outperforms GSVA and ssGSEA in both data sets, but only performs better than CLR in the WGS data set.}
+cILR performs better than GSVA and ssGSEA but not as well as CLR, with the exception of the whole genome sequencing data set.}
     \label{fig:7}
 \end{figure}
+
+Similar to simulation experiments, we also fitted a naive random forest model using CILR, ssGSEA, GSVA, or CLR transformed variables as inputs, and use AUC as the performance criteria.
+Results also replicated that of the simulations, where across both data sets cILR and CLR methods provide much better performance than both GSVA or ssGSEA.
+Interestingly, the cILR approach performed better than CLR in the whole genome data set but did not perform as well in the 16S rRNA gene sequencing data set.
+However, these results indicate that cILR generated scores are informative, providing competitive performance when acting as inputs to disease predictive models.
+Most importantly, performance values are consistent across both simulated and real data sets.
 
@@
 \subsection*{Inference with cILR}
-We can use cILR to test for enrichment of taxa-sets at the sample level.
-Inter-taxa correlation can result in an inflated variance of resulting test statistics \cite{wu2012}, which results in an inflated type I error if not properly controlled for.
-Our method addresses this issue by combining the mean estimate of scores computed on permuted data and the variance estimate of scores computed on unpermuted data, where correlation is undisrupted.
-As a consequence, adjusted cILR methods control very well for type I error at the appropriate $\alpha$ level (Figure \ref{fig:2}).
-The conservativeness of the test persisted across different sparsity levels, as well as inter-taxa correlation.
-Conversely, the unadjusted cILR methods and the naive Wilcoxon rank sum test has inflated type I error even at low correlation levels ($\rho = 0.2$).
-However, the trade-off for proper type I error control is lower power.
-Simulation results showed that power is sensitive to sparsity across all correlation levels.
+The formulation of cILR as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}.
+This allows for conducting inference with cILR even at the sample-level.
+We assessed the usage of cILR in this type of analysis by evaluating type I error and power across both simulation studies and real data applications.
+Most importantly, we demonstrated that our adjusted cILR approach was able to address the issue of variance inflation due to correlation \cite{wu2012} by controlling for type I error at the appropriate $\alpha$ level across different levels of simulated inter-taxa correlation (Fig.
+\ref{fig:2}) while conversely unadjusted cILR and the naive Wilcoxon rank sum test showed much higher rates of error.
+This is further encouraged in real data analysis where the false discovery rate was around $0.05$ when a collection of true null and true alternate hypotheses were tested.
+Unfortunately the trade-off between good type I error control is lower power.
+The conservativeness of the test attenuates with higher sparsity and correlation, where power was not approaching even 50\% even at the highest effect sizes.
+However, when the degree of such data features are reasonable, cILR will still be able to detect a reasonable proportion of samples with inflated counts.
 
-However, if there is no correlation between members of a set, cILR was still able to maintain 80\% power even with low effect size (fold change of 1.5), long as the set of interest is more densely populated, such as a set of core taxa that are prevalent across all samples.
-Unfortunately, this relationship doesn't hold for higher correlation levels.
-This means that in sets where associated taxa are highly correlated, a higher fold change is required for cILR to detect enrichment.
+We also observed that choosing different distributional forms did not alter performance values for cILR.
+This runs contrary to our comparison analysis in Fig.
+\ref{fig:1}, where we demonstrated that the mixture normal distribution had superior fit compared to the simple normal for raw cILR scores computed under the global null.
+We hypothesized that this might be due to the difficulty in fitting mixture distribution to data using the expectation maximization algorithm, as the convergence rate is slow when there is high overlap between the mixtures, resulting in a small mixing coefficient for one of the components \cite{naim2012}.
+As such, in our implementation of cILR, in order to ensure convergence for the estimating procedure we increased the number of iterations while relaxing the tolerance parameter.
+Furthermore, there are also possible problems with our adjustment procedure for the mixture distribution that might impact overall fit.
+In order to combine the scale and location estimate of two mixture distributions, we fixed the overall mean, standard deviation, mixing coefficient and component-wise means and used an optimization procedure to find the component-wise variances.
+However, this means that we have one equation for the overall variance but two possible parameters to estimate.
+As such, there is no guaranteed unique solution to component-wise variances.
+We hypothesized that the instability and degeneracy in component-wise variance estimates might impact the fidelity of estimates at the tails of the distribution, thereby affecting inference.
 
-We also assessed cILR using different distribution forms, the normal distribution, and the mixture normal distribution.
-The difference in performance across the two choices was minimal, where both were able to control for type I error, but the normal distribution was slightly better when it comes to power only in certain simulation settings.
-This runs contrary to our distribution comparison analysis in Figure \ref{fig:1}, where the mixture distribution was the superior fit.
-We hypothesized that this might be due to the difficulty in fitting mixture distributions using the expectation-maximization algorithm (EM), as the convergence rate is slow when there is a lot of overlap between the mixtures and the mixing coefficients of one of the components are low \cite{naim2012}.
-Furthermore, there is a lot of degeneracy of parameter values in our variance adjustment procedure for the mixture distribution(Supplemental Material).
-These distribution fitting problems might introduce issues where the mixture distribution might not be the best fit.
-As such, improvements to mixture distribution fitting might allow for better performance to be extracted.
+Despite these concerns, empirical results still indicate that cILR can confidently identify samples with inflated counts.
+The conservativeness of the correlation adjustment procedure ensures that significant results can be trusted by practitioners, even if cILR might not be able to exhaustively identify all samples with inflated counts.
+In situations where either the data is less sparse (e.g.
+containing a lot of core taxa that are prevalent across all samples), there is less inter-taxa correlation within the set (e.g.
+taxa that do not participate in common pathways but have shared characteristics like pathogenicity), or if the effect size is large, then cILR will still be able to produce reasonable power.
+A practitioner can use cILR to screen for samples for subsequent analysis that might involve significant costs, or perform hypothesis generation using a less stringent criteria alongside a multiple testing adjustment procedure (such as Benjamini-Hochberg \cite{benjamini1995}).
 
-These results are replicated in our real data analysis, where we attempted to test for enrichment for a group of aerobic taxa among samples in the gingival site from HMP \cite{consortium2012}.
-Even though our labeling is not perfect, it is encouraging to see that the false discovery rate is low (approximately 0.05) when tested among a set of hypotheses where there is a mixture of true null and alternate situations.
-We believe that a practitioner can identify many interesting samples to follow up on by applying multiple testing adjustment procedures (such as Benjamini-Hochberg \cite{benjamini1995}) with cILR p-values and using less stringent criteria (e.g.
-0.1 false discovery rate).
+\subsection*{Downstream analysis}
+The sample-level enrichment scores generated by the cILR method can be used in downstream analyses commonly performed in microbiome research: differential abundance testing and disease prediction.
 
-In short, we demonstrated that cILR can be used to test for significance testing at the sample level.
-Even though there is not a lot of difference between the distribution choices, ultimately cILR controls for type I error well despite tradeoffs in power.
-Even then, cILR will still be able to detect enriched sets in high correlation and sparsity situations insofar as the effect size is large.
-Researchers can utilize cILR to test for enrichment of a certain set of taxa of interest to perform follow-up experiments or to identify the prevalence of certain sub-population of taxa (for example, oxygen preference as in our real data example).
-The conservativeness of the test ensures confidence in detected samples.
+\subsubsection*{Differential abundance analysis}
+For differential abundance testing, we evaluated whether using cILR alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes.
+We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014}.
+We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome data \cite{mcmurdie2014}.
+Conversely, corncob is a newer method developed specifically for microbiome data sets, where taxonomic counts are modeled directly newly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.
+In our analyses, we found that both corncob and DESeq2 has inflated levels of type I error.
 
-\subsection*{Using cILR as inputs for downstream analyses}
-Since cILR generates enrichment scores for each set at the sample level, they can be used in downstream analyses alongside standard statistical approaches.
-In this section we assessed the utility of cILR generated scores in two common analysis tasks in microbiome research: differential abundance testing and disease prediction.
+\subsubsection*{Predictive models}
+For disease prediction, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using cILR generated scores as inputs.
+Similar to our inference analysis, we compared cILR against both ssGSEA and GSVA.
+Additionally, we also evaluated cILR with the approach where counts of a set were aggregated using sums and then centered log-ratio transformed.
+This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}.
+Results indicated that cILR produce good performance values across both real data analysis and simulation scenarios.
+Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicate that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
+Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome data sets.
+This is consistent with our sample ranking analysis (Fig.
+\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
+However, cILR did not outperform the CLR approach across all our simulation studies, and only marginally performed better in the real data analysis with WGS data.
 
-For disease prediction, we benchmarked a basic random forest model \cite{breiman2001} to predict whether or not a patient has IBD given only their microbiome profile using cILR and relevant methods as inputs.
-Across both simulations and real data analyses, cILR scores perform better than alternative single sample scoring methods in the gene set testing literature (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}).
-This is consistent with previous analysis (Figure \ref{fig:3}), where we assessed the informativeness of cILR generated scores by using AUC to measure whether samples with high scores correspond with the true label of an inflated set.
-Results indicated that cILR scores are more adaptive to microbiome data sets than GSVA and ssGSEA.
-However, cILR scores underperform the standard approach of applying the CLR transformation to count aggregated data, except for the WGS real data analysis.
-Despite this, there are still benefits to using cILR scores in the context of downstream modeling beyond random forest predictive models.
-First, CLR transformed variables have a sum to zero constraints, making the covariance matrix is singular \cite{gloor2017}, impacting methods that rely on matrix decomposition such as principal component regression or canonical correlation analysis.
-Second, cILR provides an additional value of significance testing for enrichment for sample screening.
-Third, CLR transformed variables still involve the count summation procedure for variable aggregation, therefore comparison between sets of different sizes remains unviable.
+However, in simulation studies, this performance gap between CLR and cILR decreases with higher sparsity and correlation, especially in low effect saturation scenarios.
+Additionally, there are also downsides to applying CLR.
+First, the singular covariance matrix of CLR transformed variables is singular due to a sum to zero constraint \cite{gloor2017}, preventing the proper usage of approaches that rely on matrix decomposition.
+Second, the procedure still relies on using summation of counts prior to transformation, which means that we still can't compare across sets of different sizes, and any bias might still be propagated \cite{mclaren2019}.
+As such, despite benefits in performance for a naive random forest model, there is still space for using cILR as primary inputs into predictive models.
 
-Despite such drawbacks, cILR methods still produce competitive performance values across both regression and classification tasks even in low signal-to-noise settings.
-Most interestingly, the performance gap between cILR and CLR was smaller in high sparsity settings ($p = 0.6$), which is a feature of microbiome relative abundance data \cite{li2019}.
-Furthermore, these predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), and good performance also indicated that cILR scores can also capture the joint distribution of considers sets, especially with regards to their relationship with a defined outcome such as IBD status.
+Similar to other experiments in downstream usage of cILR, performance did not change with different underlying distributions, output types, or correlation status.
+This is surprising since we expect z-scores to perform better as they are able to capture the direction of an association.
+The fact that this effect persisted even onto our real data analysis suggests that this is not due to a deficiency of our simulation design.
+As such, practitioners who wish to use cILR in predictive models might be suited to use the settings that is the fastest to compute.
 
-Unsurprisingly, different variants of cILR did not differ significantly in performance similar to significance testing experiments.
-Even though we expect Z-scores to perform better due to the added information of directionality of enrichment, using CDF values did not alter performance in both the single sample AUC evaluations (Figure \ref{fig:6}) as well as the random forest evaluations (Figure \ref{fig:6}).
-
-In short, users can apply cILR generated enrichment scores as inputs to downstream analyses, especially predictive models and differential abundance analysis.
-cILR scores are informative in discriminating samples between classes even in low effect size situations.
-Even though cILR did not outperform a standard analysis of count aggregation followed by the CLR transformation, the additional benefits listed above would prove to be attractive in certain use cases, especially in high sparsity situations.
+Ultimately, results indicate that cILR can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity.
+Even though there exists situations where it might not provide maximum predictive values, the flexibility of cILR in various types of analyses enable ven though in some scenarios it might not provide maximum predictive values.
 
 \subsection*{Limitations and future directions} 
-There are various limitations to using cILR to generate sample-level taxonomic aggregation for microbiome data.
-First, significance testing at the sample level is not powered to assess inflated sets with low effect size.
-Practitioners can be confident in the validity of a significant result (following proper multiple testing correction) due to cILR's type I error control, but should not expect cILR to be able to identify all samples with inflated counts.
-Second, even though cILR is flexible and addresses the compositional nature of microbiome data, the performance of cILR might not be consistently better than competing methods.
-Careful evaluation of methods should be done before application to ensure that the value of cILR scores is appropriately extracted in suitable models.
-Third, cILR did not directly address the zero-inflatedness of microbiome data and utilizes a pseudocount to ensure log operations are valid.
-As such, practitioners are encouraged to choose a different pseudocount or an appropriate method of choice.
+There are various limitations to our evaluation of cILR.
+First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome data.
+There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation section of prior studies.
+Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is flawed.
+This is because the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut.
+As such, results from power analyses using this data set is only relative between the comparison methods, and absolute values should should not carry meaning.
+Finally, we assumed that taxa within a set all be equally associated with the outcome.
+This limits our ability to evaluate the performance of cILR when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.
 
-As such, we hope to address the aforementioned issues with cILR in later projects.
-First, we hope to refine the distribution fitting procedure for the mixture distribution due to its superior fit and address the identifiability problem that might underlie the lack of power in our experiments.
-Second, we hope to incorporate some of the sophisticated model-based zero correction methods \cite{martin-fernandez2012,kaul2017a}.
-Third, we can account for the difference in importance of certain taxa by incorporating taxa-specific weights.
-Finally, we also hope to curate more interest apriori sets that are functionally informative, which would generate more interesting insight and improve interpretability compared to using taxonomic categories such as Phylum or Genus.
+However, our evaluation still showed various drawbacks of the cILR method.
+First, inference with cILR is limited in being able to exhaustively detect all samples with significant inflated counts for a set in situations where there is a high degree of sparsity and inter-taxa correlation.
+Second, for downstream analyses, cILR might not always better than competing methods, especially when being used as inputs to predictive models.
+We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated with adjusting the mixture normal distribution.
+As such, we hope to refine the null distribution estimating procedure by either choosing a better distributional form, or to further constrain the optimization procedure of the mixture normal distribution by fixing the third and fourth moments.
+
+In addition, there are possible extensions cILR can we can consider to provide more flexibility across different types of scenarios in data analysis.
+First, cILR did not address the sparsity of microbiome data and relies on pseudocount to ensure log operations are valid.
+We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
+Second, cILR also treated all taxa within the set as equally contributing to the set.
+Incorporate taxa-specific weights can reduce the influence of outliers, such as rare or highly invariant taxa.
+Finally, curating sets based on apriori characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.
 
 \section*{Conclusion}
-Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomic data sets.
+Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomics data sets.
 However, there has not been a lot of set-based analysis methods developed specifically for microbiome relative abundance data.

### 2021-06-07 manuscript_plos.tex
Newest version

@@
 \section*{Abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+Research in human associated microbiomes primarily involve the analysis of taxonomic count tables generated via high-throughput sequencing.
+It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional.
+An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets.
+Set-based analysis is ubiquitous in the genomics literature, and have demonstrable impact in improving interpretability and power of downstream analysis.
+Current practice in the microbiome field, however, performs aggregation by summing the abundance of respective taxa.
+This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
+Unfortunately, there is a lack of interest in developing more sophisticated set-based analysis methods specific to microbiome data.
+Here, we attempt to fill this gap with a new sample-level taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature.
+Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
+We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
+Herein we demonstrate under both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios, while additionally providing informative scores that can be inputs to downstream differential abundance and predictive tasks with good performance.
+These results demonstrate how our approach can enable researchers to generate meaningful aggregation scores for sets of taxa.
 
@@
 \section*{Author summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+The study of human associated microbiomes rely on genomic surveys via high-throughput sequencing.
+However, microbiome data is sparse and high dimensional which prevents the application of standard statistical techniques.
+One approach to address this problem is to perform analyses at the level of taxon sets.
+Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability.
+Unfortunately, there is not a lot of research in developing new set-based tools for microbiome data specifically, given that compared to other omics data types microbiome data is strictly compositional.
+We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis.
+Our scores can be used for inference purposes, as well as inputs to other downstream analyses such as differential abundance and predictive models.
+We demonstrated the performance of our method against competing approaches across both real data analyses and simulation studies.
 
@@
 \section*{Introduction}
-\section*{Background}  
 The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their human host.
@@
 Unfortunately, there is a lack of interest in extending existing methods to be more specific to microbiome relative abundance data.
-Some software suites, such as \emph{MicrobiomeAnalyst}, do provide researchers tools to perform enrichment testing with curated taxa sets \cite{chong2020}.
+Some software suites, such as \emph{MicrobiomeAnalyst}, do provide researchers tools to perform enrichment testing with curated taxon sets \cite{chong2020}.
 However, the approach used in \emph{MicrobiomeAnalyst} is an overrepresentation analysis and therefore similarly sensitive to the method of differential abundance used.
@@
 
-\section*{Materials and Methods} \label{Methods}
+\section*{Materials and Methods} \label{methods}
 \subsection*{Competitive Isometric Log-ratio (cILR)}
@@
     \item \textbf{Estimate the empirical null distribution} Enrichment scores represent the test statistic for the $Q_1$ null hypothesis $H_o$ that relative abundances in $\mathbf{X}$ of members of set $k$ are not enriched compared to those not in set $k$.
-Since the distribution of cILR under the null vary depending on data characteristics (Figure \ref{fig:1}), an empirical null distribution will be estimated from data.
+Since the distribution of cILR under the null vary depending on data characteristics (Fig~\ref{fig:1}), an empirical null distribution will be estimated from data.
     \begin{itemize}
@@
 They are usually termed ``compositional balances", and have been leveraged in many recent approaches in variable transformation \cite{washburne2017,silverman2017,morton2017}.
-The cILR formulation \eqref{main_eq} is a special case of \eqref{ilr_standard} defined on a node that splits the taxa into two disjoint sets, one representing the set of interest, the other representing the remaining taxa.
+The cILR formulation in Eq~\eqref{main_eq} is a special case of Eq~\eqref{ilr_standard} defined on a node that splits the taxa into two disjoint sets, one representing the set of interest, the other representing the remaining taxa.
 As such, the cILR transformation inherits the properties of the ILR as a log-ratio method applicable to compositional data sets.
@@
 This assumption is also supported by preliminary simulation studies (detailed below).
-In panel A of figure \ref{fig:1}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
+In panel A of Fig~\ref{fig:1}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
 It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation.
@@
 
-\begin{figure} [H]
+\begin{figure} [!ht]
     \centering
     \includegraphics[width=\linewidth]{figures/kurtosis_skewness_gof.png}
-    \caption{Properties of the null distribution of cILR in different simulation scenarios under the global null.
+    \caption{{\bf Properties of the null distribution of cILR under the global null simulations}.
 Panel \textbf{(B)} presents kurtosis and skewness of cILR scores while panel \textbf{(A)} presents the goodness of fit (as Kolmogorov-Smirnov D statistic) for mixture normal and normal distributions.
@@
 To address this issue, we also evaluated a mixture distribution of two normal components.
-Panel B of figure \ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
+Panel B of Fig~\ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
 We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.
@@
 \subsection*{Enrichment testing at the sample level}
-cILR provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{Methods}.
+cILR provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{methods}.
 Here, we present empirical results for this application of cILR assessing type I error, power, and classification capacity.
@@
 \subsubsection*{Simulation studies}
-Panel A and B in figure \ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
+Panel A and B in Fig~\ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
 We benchmarked the results of the cILR method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
@@
 
-\begin{figure}[H]
+\begin{figure}[!ht]
     \centering
@@
 
-However, the trade-off for good type I error control is demonstrably lower power, as shown in figure \ref{fig:2}B.
+However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B.
 In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un-adjusted ones.
@@
 This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated.
-Figure \ref{fig:3} presents this result.
+Fig~\ref{fig:3} presents this result.
 We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
-Across both simulations (Figure \ref{fig:3}A) and real-data applications (Figure \ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
+Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
 In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
@@
 \subsubsection*{Real data evaluations}  
-These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}.
+These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{methods}.
 Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations.
-As expected in Figure \ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods.
+As expected in Fig~\ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods.
 Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05.
-Power analysis (Figure \ref{fig:2}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
-
-\begin{figure}[H]
+Power analysis (Fig~\ref{fig:2}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
+
+\begin{figure}[!ht]
     \centering
     \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
-    \caption{Classification performance via AUC of cILR, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{Methods}.
+    \caption{Classification performance via AUC of cILR, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}.
 Performance scores measure whether scores can highly rank samples that are known to have inflated abundance.
@@
 \subsection*{Disease Prediction}   
-Since cILR can generate informative scores that can discriminate between samples with inflated counts for a set (Fig.
-\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models.
+Since cILR can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models.
 In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating cILR, ssGSEA, and GSVA).
-Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxa sets aggregated via abundance summations.
+Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.
 
 \subsubsection*{Simulation studies}
-Fig.
-\ref{fig:6} showed results for simulation studies as detailed in the \nameref{Methods} section.
+Fig~\ref{fig:6} showed results for simulation studies as detailed in the \nameref{methods} section.
 Panel A.
@@
 
-%\begin{figure}[H]
+%\begin{figure}[!ht]
 %    \centering
@@
 In addition to parametric simulations, we also assessed the performance of using cILR scores in predictive models with real data sets.
-Fig.
-\ref{fig:7} presents results for two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
+Fig~\ref{fig:7} presents results for two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
 The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al.
@@
 We assessed the usage of cILR in this type of analysis by evaluating type I error and power across both simulation studies and real data applications.
-Most importantly, we demonstrated that our adjusted cILR approach was able to address the issue of variance inflation due to correlation \cite{wu2012} by controlling for type I error at the appropriate $\alpha$ level across different levels of simulated inter-taxa correlation (Fig.
-\ref{fig:2}) while conversely unadjusted cILR and the naive Wilcoxon rank sum test showed much higher rates of error.
+Most importantly, we demonstrated that our adjusted cILR approach was able to address the issue of variance inflation due to correlation \cite{wu2012} by controlling for type I error at the appropriate $\alpha$ level across different levels of simulated inter-taxa correlation (Fig~\ref{fig:2}) while conversely unadjusted cILR and the naive Wilcoxon rank sum test showed much higher rates of error.
 This is further encouraged in real data analysis where the false discovery rate was around $0.05$ when a collection of true null and true alternate hypotheses were tested.
@@
 We also observed that choosing different distributional forms did not alter performance values for cILR.
-This runs contrary to our comparison analysis in Fig.
-\ref{fig:1}, where we demonstrated that the mixture normal distribution had superior fit compared to the simple normal for raw cILR scores computed under the global null.
+This runs contrary to our comparison analysis in Fig~\ref{fig:1}, where we demonstrated that the mixture normal distribution had superior fit compared to the simple normal for raw cILR scores computed under the global null.
 We hypothesized that this might be due to the difficulty in fitting mixture distribution to data using the expectation maximization algorithm, as the convergence rate is slow when there is high overlap between the mixtures, resulting in a small mixing coefficient for one of the components \cite{naim2012}.
@@
 Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome data sets.
-This is consistent with our sample ranking analysis (Fig.
-\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
+This is consistent with our sample ranking analysis (Fig~\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
 However, cILR did not outperform the CLR approach across all our simulation studies, and only marginally performed better in the real data analysis with WGS data.
@@
 
-\paragraph*{S2 Fig.}
-\label{S2_Fig}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
 \paragraph*{S1 File.}
@@
 
-\paragraph*{S1 Video.}
-\label{S1_Video}
-{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
 \paragraph*{S1 Appendix.}
@@
 
-\paragraph*{S1 Table.}
-\label{S1_Table}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
 \section*{Acknowledgments}
@@
 
-\bibliography{tax_agg}{}
+% \bibliography{tax_agg}{}
+
+\begin{thebibliography}{10}
+
+\bibitem{proctor2019}
+Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
+\newblock The {{Integrative Human Microbiome Project}}.
+\newblock Nature. 2019;569(7758):641--648.
+\newblock doi:{10.1038/s41586-019-1238-8}.
+
+\bibitem{sharma2019}
+Sharma S, Tripathi P.
+\newblock Gut Microbiome and Type 2 Diabetes: Where We Are and Where to Go?
+\newblock The Journal of Nutritional Biochemistry. 2019;63:101--108.
+\newblock doi:{10.1016/j.jnutbio.2018.10.003}.
+
+\bibitem{aoun2020}
+Aoun A, Darwish F, Hamod N.
+\newblock The {{Influence}} of the {{Gut Microbiome}} on {{Obesity}} in
+  {{Adults}} and the {{Role}} of {{Probiotics}}, {{Prebiotics}}, and
+  {{Synbiotics}} for {{Weight Loss}}.
+\newblock Preventive Nutrition and Food Science. 2020;25(2):113--123.
+\newblock doi:{10.3746/pnf.2020.25.2.113}.
+
+\bibitem{cho2012}
+Cho I, Blaser MJ.
+\newblock The Human Microbiome: At the Interface of Health and Disease.
+\newblock Nature Reviews Genetics. 2012;13(4):260--270.
+\newblock doi:{10.1038/nrg3182}.
+
+\bibitem{callahan2016}
+Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
+\newblock {{DADA2}}: {{High}}-Resolution Sample Inference from {{Illumina}}
+  Amplicon Data.
+\newblock Nature Methods. 2016;13(7):581--583.
+\newblock doi:{10.1038/nmeth.3869}.
+
+\bibitem{truong2015}
+Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
+\newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
+\newblock Nature Methods. 2015;12(10):902--903.
+\newblock doi:{10.1038/nmeth.3589}.
+
+\bibitem{li2019}
+Li H.
+\newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
+  {{Metagenomics}}.
+\newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
+  2019. p. 977--550.
+
+\bibitem{li2015}
+Li H.
+\newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
+  Compositional Data Analysis}}.
+\newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
+\newblock doi:{10.1146/annurev-statistics-010814-020351}.
+
+\bibitem{shi2016}
+Shi P, Zhang A, Li H.
+\newblock Regression Analysis for Microbiome Compositional Data.
+\newblock The Annals of Applied Statistics. 2016;10(2):1019--1040.
+\newblock doi:{10.1214/16-AOAS928}.
+
+\bibitem{sankaran2014}
+Sankaran K, Holmes S.
+\newblock {{structSSI}}: {{Simultaneous}} and {{Selective Inference}} for
+  {{Grouped}} or {{Hierarchically Structured Data}}.
+\newblock Journal of statistical software. 2014;59(13):1--21.
+\newblock doi:{10.18637/jss.v059.i13}.
+
+\bibitem{benjamini1995}
+Benjamini Y, Hochberg Y.
+\newblock Controlling the {{False Discovery Rate}}: {{A Practical}} and
+  {{Powerful Approach}} to {{Multiple Testing}}.
+\newblock Journal of the Royal Statistical Society Series B (Methodological).
+  1995;57(1):289--300.
+
+\bibitem{chen2018}
+Chen J, King E, Deek R, Wei Z, Yu Y, Grill D, et~al.
+\newblock An Omnibus Test for Differential Distribution Analysis of Microbiome
+  Sequencing Data.
+\newblock Bioinformatics. 2018;34(4):643--651.
+\newblock doi:{10.1093/bioinformatics/btx650}.
+
+\bibitem{weiss2017}
+Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
+\newblock Normalization and Microbial Differential Abundance Strategies Depend
+  upon Data Characteristics.
+\newblock Microbiome. 2017;5(1).
+\newblock doi:{10.1186/s40168-017-0237-y}.
+
+\bibitem{mcmurdie2014}
+McMurdie PJ, Holmes S.
+\newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
+  Inadmissible}}.
+\newblock PLOS Computational Biology. 2014;10(4):e1003531.
+\newblock doi:{10.1371/journal.pcbi.1003531}.
+
+\bibitem{love2014}
+Love MI, Huber W, Anders S.
+\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
+  Data with {{DESeq2}}.
+\newblock Genome Biology. 2014;15(12):550.
+\newblock doi:{10.1186/s13059-014-0550-8}.
+
+\bibitem{quinn2019}
+Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
+\newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
+\newblock GigaScience. 2019;8(giz107).
+\newblock doi:{10.1093/gigascience/giz107}.
+
+\bibitem{quinn2018b}
+Quinn TP, Erb I, Richardson MF, Crowley TM.
+\newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
+\newblock Bioinformatics. 2018;34(16):2870--2878.
+\newblock doi:{10.1093/bioinformatics/bty175}.
+
+\bibitem{gloor2017}
+Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
+\newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
+  Optional}}.
+\newblock Frontiers in Microbiology. 2017;8.
+\newblock doi:{10.3389/fmicb.2017.02224}.
+
+\bibitem{aitchison1999}
+Aitchison J.
+\newblock A {{Concise Guide}} to {{Compositional Data Analysis}}. 1999; p. 134.
+
+\bibitem{kurtz2015}
+Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
+\newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
+  Ecological Networks}}.
+\newblock PLOS Computational Biology. 2015;11(5):e1004226.
+\newblock doi:{10.1371/journal.pcbi.1004226}.
+
+\bibitem{kaul2017}
+Kaul A, Davidov O, Peddada SD.
+\newblock Structural Zeros in High-Dimensional Data with Applications to
+  Microbiome Studies.
+\newblock Biostatistics. 2017;18(3):422--433.
+\newblock doi:{10.1093/biostatistics/kxw053}.
+
+\bibitem{kaul2017a}
+Kaul A, Mandal S, Davidov O, Peddada SD.
+\newblock Analysis of {{Microbiome Data}} in the {{Presence}} of {{Excess
+  Zeros}}.
+\newblock Frontiers in Microbiology. 2017;8.
+\newblock doi:{10.3389/fmicb.2017.02114}.
+
+\bibitem{khatri2012}
+Khatri P, Sirota M, Butte AJ.
+\newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
+  {{Outstanding Challenges}}.
+\newblock PLOS Computational Biology. 2012;8(2):e1002375.
+\newblock doi:{10.1371/journal.pcbi.1002375}.
+
+\bibitem{goeman2007}
+Goeman JJ, B{\"u}hlmann P.
+\newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
+  Issues.
+\newblock Bioinformatics. 2007;23(8):980--987.
+\newblock doi:{10.1093/bioinformatics/btm051}.
+
+\bibitem{subramanian2005}
+Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
+\newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
+  Interpreting Genome-Wide Expression Profiles.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(43):15545--15550.
+\newblock doi:{10.1073/pnas.0506580102}.
+
+\bibitem{ashburner2000}
+Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
+\newblock Gene {{Ontology}}: Tool for the Unification of Biology.
+\newblock Nature genetics. 2000;25(1):25--29.
+\newblock doi:{10.1038/75556}.
+
+\bibitem{hanzelmann2013}
+H{\"a}nzelmann S, Castelo R, Guinney J.
+\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
+  {{RNA}}-{{Seq}} Data.
+\newblock BMC Bioinformatics. 2013;14(1):7.
+\newblock doi:{10.1186/1471-2105-14-7}.
+
+\bibitem{frost2020a}
+Frost HR.
+\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
+  Method for Cell-Specific Gene Set Scoring.
+\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
+\newblock doi:{10.1093/nar/gkaa582}.
+
+\bibitem{mclaren2019}
+McLaren MR, Willis AD, Callahan BJ.
+\newblock Consistent and Correctable Bias in Metagenomic Sequencing
+  Experiments.
+\newblock eLife. 2019;8:e46923.
+\newblock doi:{10.7554/eLife.46923}.
+
+\bibitem{egozcue2005}
+Egozcue JJ, {Pawlowsky-Glahn} V.
+\newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2005;37(7):795--828.
+\newblock doi:{10.1007/s11004-005-7381-9}.
+
+\bibitem{chong2020}
+Chong J, Liu P, Zhou G, Xia J.
+\newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
+  Functional, and Meta-Analysis of Microbiome Data.
+\newblock Nature Protocols. 2020;15(3):799--821.
+\newblock doi:{10.1038/s41596-019-0264-1}.
+
+\bibitem{tian2005}
+Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
+\newblock Discovering Statistically Significant Pathways in Expression
+  Profiling Studies.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(38):13544--13549.
+\newblock doi:{10.1073/pnas.0506577102}.
+
+\bibitem{egozcue2003}
+Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcelo-Vidal} C.
+\newblock Isometric {{Logratio Transformations}} for {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2003; p.~22.
+
+\bibitem{silverman2017}
+Silverman JD, Washburne AD, Mukherjee S, David LA.
+\newblock A Phylogenetic Transform Enhances Analysis of Compositional
+  Microbiota Data.
+\newblock eLife. 2017;6:e21887.
+\newblock doi:{10.7554/eLife.21887}.
+
+\bibitem{barbie2009}
+Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
+\newblock Systematic {{RNA}} Interference Reveals That Oncogenic
+  {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
+\newblock Nature. 2009;462(7269):108--112.
+\newblock doi:{10.1038/nature08460}.
+
+\bibitem{quast2013}
+Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
+\newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
+  Processing and Web-Based Tools.
+\newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
+\newblock doi:{10.1093/nar/gks1219}.
+
+\bibitem{delignette-muller2015}
+{Delignette-Muller} ML, Dutang C.
+\newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
+\newblock Journal of Statistical Software. 2015;64(4):1--34.
+
+\bibitem{benaglia2009}
+Benaglia T, Chauveau D, Hunter DR, Young D.
+\newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
+\newblock Journal of Statistical Software. 2009;32(6):1--29.
+
+\bibitem{washburne2017}
+Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL, Mukherjee S, et~al.
+\newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
+  Associations in Microbiome Datasets.
+\newblock PeerJ. 2017;5:e2969.
+\newblock doi:{10.7717/peerj.2969}.
+
+\bibitem{morton2017}
+Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
+  et~al.
+\newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
+\newblock mSystems. 2017;2(1).
+\newblock doi:{10.1128/mSystems.00162-16}.
+
+\bibitem{efron2004}
+Efron B.
+\newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
+\newblock Journal of the American Statistical Association.
+  2004;99(465):96--104.
+\newblock doi:{10.1198/016214504000000089}.
+
+\bibitem{wu2012}
+Wu D, Smyth GK.
+\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
+  Correlation.
+\newblock Nucleic Acids Research. 2012;40(17):e133.
+\newblock doi:{10.1093/nar/gks461}.
+
+\bibitem{calgaro2020}
+Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
+\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
+  and Metagenomics Applied to Microbiome Data.
+\newblock Genome Biology. 2020;21(1):191.
+\newblock doi:{10.1186/s13059-020-02104-1}.
+
+\bibitem{cario1997}
+Cario MC.
+\newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
+  Distributions}} and {{Correlation Matrix}}. 1997; p.~19.
+
+\bibitem{agresti1998}
+Agresti A, Coull BA.
+\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
+  Estimation}} of {{Binomial Proportions}}.
+\newblock The American Statistician. 1998;52(2):119--126.
+\newblock doi:{10.2307/2685469}.
+
+\bibitem{delong1988}
+DeLong ER, DeLong DM, {Clarke-Pearson} DL.
+\newblock Comparing the Areas under Two or More Correlated Receiver Operating
+  Characteristic Curves: A Nonparametric Approach.
+\newblock Biometrics. 1988;44(3):837--845.
+
+\bibitem{hawinkel2019}
+Hawinkel S, Mattiello F, Bijnens L, Thas O.
+\newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
+  Control the False Discovery Rate.
+\newblock Briefings in Bioinformatics. 2019;20(1):210--221.
+\newblock doi:{10.1093/bib/bbx104}.
+
+\bibitem{breiman2001}
+Breiman L.
+\newblock Random {{Forests}}.
+\newblock Machine Learning. 2001;45(1):5--32.
+\newblock doi:{10.1023/A:1010933404324}.
+
+\bibitem{kuhn2020}
+Kuhn M, Wickham H.
+\newblock Tidymodels: {{Easily}} Install and Load the 'tidymodels' Packages;
+  2020.
+
+\bibitem{xiao2018}
+Xiao J, Chen L, Yu Y, Zhang X, Chen J.
+\newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
+  {{Predictive Modeling}} of {{Microbial Community Data}}.
+\newblock Frontiers in Microbiology. 2018;9.
+\newblock doi:{10.3389/fmicb.2018.03112}.
+
+\bibitem{dong2020}
+Dong M, Li L, Chen M, Kusalik A, Xu W.
+\newblock Predictive Analysis Methods for Human Microbiome Data with
+  Application to {{Parkinson}}'s Disease.
+\newblock PLOS ONE. 2020;15(8):e0237779.
+\newblock doi:{10.1371/journal.pone.0237779}.
+
+\bibitem{pasolli2017}
+Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
+\newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
+\newblock Nature Methods. 2017;14(11):1023--1024.
+\newblock doi:{10.1038/nmeth.4468}.
+
+\bibitem{schiffer2019}
+Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
+\newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
+  through Bioconductor.
+\newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
+
+\bibitem{gonzalez2018}
+Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
+  Ackermann G, et~al.
+\newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
+\newblock Nature Methods. 2018;15(10):796--798.
+\newblock doi:{10.1038/s41592-018-0141-9}.
+
+\bibitem{consortium2012}
+Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
+  et~al.
+\newblock Structure, Function and Diversity of the Healthy Human Microbiome.
+\newblock Nature. 2012;486(7402):207--214.
+\newblock doi:{10.1038/nature11234}.
+
+\bibitem{thurnheer2016}
+Thurnheer T, Bostanci N, Belibasakis GN.
+\newblock Microbial Dynamics during Conversion from Supragingival to
+  Subgingival Biofilms in an in Vitro Model.
+\newblock Molecular Oral Microbiology. 2016;31(2):125--135.
+\newblock doi:{10.1111/omi.12108}.
+
+\bibitem{beghini2019}
+Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
+\newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
+  in the {{New York City Health}} and {{Nutrition Examination Study}}.
+\newblock Annals of Epidemiology. 2019;34:18--25.e3.
+\newblock doi:{10.1016/j.annepidem.2019.03.005}.
+
+\bibitem{matteocalgaro2020}
+Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
+\newblock Zenodo.
+
+\bibitem{nielsen2014}
+Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
+\newblock Identification and Assembly of Genomes and Genetic Elements in
+  Complex Metagenomic Samples without Using Reference Genomes.
+\newblock Nature Biotechnology. 2014;32(8):822--828.
+\newblock doi:{10.1038/nbt.2939}.
+
+\bibitem{gevers2014}
+Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
+  et~al.
+\newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
+  {{Disease}}.
+\newblock Cell Host \& Microbe. 2014;15(3):382--392.
+\newblock doi:{10.1016/j.chom.2014.02.005}.
+
+\bibitem{martin2020}
+Martin BD, Witten D, Willis AD.
+\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
+  Regression.
+\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
+\newblock doi:{10.1214/19-AOAS1283}.
+
+\bibitem{naim2012}
+Naim I, Gildea D.
+\newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
+  {{Unbalanced Mixing Coefficients}}.
+\newblock Proceedings of the 29th International Coference on International
+  Conference on Machine Learning. 2012; p.~8.
+\newblock doi:{10.5555/3042573.3042756}.
+
+\bibitem{wu2016}
+Wu C, Chen J, Kim J, Pan W.
+\newblock An Adaptive Association Test for Microbiome Data.
+\newblock Genome Medicine. 2016;8(1):56.
+\newblock doi:{10.1186/s13073-016-0302-3}.
+
+\bibitem{martin-fernandez2012}
+{Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
+  {Palarea-Albaladejo} J.
+\newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
+  {{Classical}} and Robust Approaches.
+\newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
+\newblock doi:{10.1016/j.csda.2012.02.012}.
+
+\end{thebibliography}
 

### 2021-06-14 supplemental.tex
Updates from Overleaf

@@
 \usepackage[parfill]{parskip}
+\newcommand{\overbar}[1]{\mkern 2mu\overline{\mkern-2mu#1\mkern-20mu}\mkern 20mu}
 
-\author{Quang P. Nguyen}
-\title{Supplementary Materials}
+\title{Supplementary Materials for cILR: Taxonomic enrichment analysis with isometric log ratios}
 \begin{document}
+\author{Quang P. Nguyen, Anne G. Hoen, H. Robert Frost}
 \maketitle
@@
 
-\section{Performance Evaluation} 
-\section{Distribution fitting}
-\section{p-value behaviour}
+\section{Addressing variance inflation due to correlation}
+In order to perform inference with competitive isometric log ratio (cILR), we estimated the null distribution empirically following Efron et al.
+\cite{efron2004}.
+We chose two distributional forms for the null: the normal distribution and a two-component mixture normal distribution.
+Subsequently, we estimated the parameters of the respective distributions from a column (or variable) permuted raw score matrix.
+This means that our null distribution for a given set is equivalent to scores computed for sets of similar sizes but containing randomly chosen taxa.
+For the normal distribution, we estimated parameters using the maximum likelihood using the \emph{fitdistrplus} package \cite{delignette-muller2015}.
+For the mixture normal distribution, we utilized the expectation-maximization procedure from the package \emph{mixtools} \cite{benaglia2009}.
+
+However, when variables are correlated, the variance of the sample mean of the variables is inflated \cite{wu2012}.
+This is a significant problem since evidence shows that there is strong inter-taxa correlation within the microbiome \cite{kurtz2015}.
+Without loss of generalizability, for a set of variables $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:
+\begin{equation} \label{eq:1}
+    Var(\bar{x}) = \frac{1}{m^2}\left(\sum_{i = 1}(\sigma_i^2) + \sum_{i < j}\rho_{ij}\sigma_i\sigma_j\right)
+\end{equation}
+where $\sigma_i$ is the standard deviation of variable $i$ while $\rho_{ij}$ is the correlation between $i$ and $j$.
+The second term of (\ref{eq:1}) is the correlation dependent variance component, which goes to 0 if there is no correlation.
+Wu et al.
+\cite{wu2012} also showed that a null distribution estimated from a column permuted matrix does not account for this variance inflation, since the permutation procedure disrupts the natural correlation structure of the original variables.
+
+The cILR statistic follows a similar pattern.
+Since the geometric mean of a set of variables is equivalent to the exponential of the arithmetic mean of their logarithms, we can re-write cILR for a set $k$ with size $\kappa$ as follows:
+\begin{equation}\label{eq:2}
+    cILR_k = \sqrt{\frac{\kappa(p - \kappa)}{\kappa + (p - \kappa)}} \left( \overbar{\log{X_{j|j \in K}}} - \overbar{\log{X_{j|j \notin K}}} \right)   
+\end{equation}
+where $p$ is the overall number of taxa, $j$ is the index of a taxa and $K$ is the set of indices of taxa in set $k$.
+As such, the variance of the cILR test statistic is therefore dependent on the variance inflation of both mean components $\overbar{\log{X_{j|j \in K}}}$ and $\overbar{\log{X_{j|j \notin K}}}$.
+
+In order to address this variance inflation due to correlation issue, we chose to combine the location (or mean) estimate from the column permuted raw score matrix with the spread (or variance) estimate from the original un-permuted scores.
+This allows us to take advantage of the null distribution formed via variable permutation while accounting for the variance which was estimated from scores where the correlation structure has not been disrupted.
+As such, this procedure assumes that the
+
+\section{Performance evaluation}
+
+\section{p-value distribution}
 
@@
 \bibliography{tax_agg}{}
-\bibliographystyle{plain}
+\bibliographystyle{plos2015}
 

### 2021-06-15 supplemental.tex
First section of supplemental done

@@
 However, when variables are correlated, the variance of the sample mean of the variables is inflated \cite{wu2012}.
-This is a significant problem since evidence shows that there is strong inter-taxa correlation within the microbiome \cite{kurtz2015}.
 Without loss of generalizability, for a set of variables $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:
@@
 Wu et al.
-\cite{wu2012} also showed that a null distribution estimated from a column permuted matrix does not account for this variance inflation, since the permutation procedure disrupts the natural correlation structure of the original variables.
+\cite{wu2012} showed that a null distribution estimated from a column permuted matrix does not account for this variance inflation, since the permutation procedure disrupts the natural correlation structure of the original variables.
 
@@
 where $p$ is the overall number of taxa, $j$ is the index of a taxa and $K$ is the set of indices of taxa in set $k$.
-As such, the variance of the cILR test statistic is therefore dependent on the variance inflation of both mean components $\overbar{\log{X_{j|j \in K}}}$ and $\overbar{\log{X_{j|j \notin K}}}$.
+The cILR statistic then looks similar to a t-statistic for difference in means of log-transformed proportions.
+As such, the pooled variance of cILR is dependent on the variance inflation of both mean components $\overbar{\log{X_{j|j \in K}}}$ and $\overbar{\log{X_{j|j \notin K}}}$.
 
-In order to address this variance inflation due to correlation issue, we chose to combine the location (or mean) estimate from the column permuted raw score matrix with the spread (or variance) estimate from the original un-permuted scores.
-This allows us to take advantage of the null distribution formed via variable permutation while accounting for the variance which was estimated from scores where the correlation structure has not been disrupted.
-As such, this procedure assumes that the
+It is important to address this problem since there is strong inter-taxa correlation within the microbiome \cite{kurtz2015}.
+Even though Wu et al.
+proposed a correlation adjusted t-statistic for the CAMERA method \cite{wu2012}, it is difficult to adapt such modifications to the cILR statistic.
+This is because CAMERA requires an estimation procedure for pair-wise correlations of the $\log{X_j}$, as well as making the assumption that the marginal variance of each $\log{X_j}$ is equivalent to each other.
+
+Instead, our strategy is to combine the location (or mean) estimate from the column permuted raw score matrix with the spread (or variance) estimate from the original un-permuted scores.
+This still allows us to leverage the null generated via column permutation while using the proper variance estimate taken from scores where the correlation structure has not been disrupted.
+As such, this procedure assumes that the variance of the test statistic under the alternate hypothesis is the same as that of the null.
+
+For the normal distribution, it is straightforward to combine mean and variance estimates from the respective raw score matrices.
+For the mixture normal distribution, however, due to the fact that the distribution is made of component-wise means, variances and mixing coefficients, we decided to take an optimization based approach to identifying the component wise variances $\sigma_1$ and $\sigma_2$ such the overall mean and variance estimates come from the respective raw cILR score matrices as detailed above.
+We can write the optimization problem as follows:
+
+\begin{equation} \label{eq:3}
+\begin{aligned}
+    \min_{\sigma_1, \sigma_2} \quad & \sqrt{\left(\sqrt{(\sigma_1 + \mu'_1 - M')\lambda'_1 + (\sigma_2 + \mu'_2 - M')\lambda'_2} - SD\right)^2}\\
+    \textrm{s.t.}  \quad & \sigma_1 \geq 10^{-5}, \sigma_2 \geq 10^{-5}    \\
+    \textrm{where} \quad & M', SD, \mu'_1, \mu'_2, \lambda'_1, \lambda'_2 \text{ are constants} \\
+\end{aligned}
+\end{equation}
+$\lambda'_1$, $\lambda'_2$, $\mu'_1$, $\mu'_2$, and $M'$ are component-wise mixing coefficients, component-wise means, and overall mean of the mixture distribution estimated from column-permuted scores while $SD$ is the overall standard deviation of the mixture distribution estimated from unpermuted scores.
+We solve for this optimization problem using a quasi Newton method with box constraints (L-BFGS-B) with the default finite-difference approximation of the gradient, as implemented in the \emph{optim} function in R.
 
@@
 
-\section{p-value distribution}
+\section{Distribution of p-values}
 

### 2021-06-16 supplemental.tex
Updates from Overleaf

@@
 \section{Performance evaluation} 
+We implemented cILR in the package \emph{teaR} on GitHub with \emph{Rcpp} implementations for geometric mean computations.
+We evaluated computational time using the \emph{bench} package in R.
+Sample size and number of sets were varied across different evaluation conditions.
+All set sizes were kept constant at 50.
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/performance.png}
+    \caption{Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
+Evaluation was performed on simulated data sets.
+For sample size analysis, only 10 sets were evaluated.
+For taxa set analysis, sample size was fixed at 1,000.
+Across all evaluations, the size of each taxa set was also fixed at 50.}
+    \label{fig:s1}
+\end{figure}
+
+In general, increasing sample size and number of evaluated sets both increase computational time.
+The mixture normal distribution takes longer to compute compared to the normal distribution, and adjusted methods take longer but was most apparent when combined with the mixture distribution.
+Overall, a reasonable data set of 1,000 samples and 100 sets (5,000 taxa) would take around 200 seconds (or approximately 3 minutes) for the longest approach (correlation adjusted mixture normal distribution).
+Benchmark was performed on a single core using a personal Windows laptop (Specifications: Intel Core i7-9750H (2.6-4.5 GHz - 12MB Cache) with 16GB of RAM)
 

### 2021-07-16 manuscript_plos.tex

@@
 \subsubsection*{Simulation studies}
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
+    \caption{Placeholder} 
+    \label{fig:4}
+\end{figure}
 \subsubsection*{Real data evaluations}
-We also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP.
+We also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP (Fig~\ref{fig:5}).
 For type I error evaluations, we use stool samples and randomly assign them with case/control status and calculated type I error as the proportion of genera identified as significantly different.
@@
 Results showed that cILR associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2}
-    \label{fig:4}
+    \label{fig:5}
 \end{figure}

### 2021-07-16 manuscript_plos.tex

@@
 \section*{Abstract}
-Research in human associated microbiomes primarily involve the analysis of taxonomic count tables generated via high-throughput sequencing.
+Research in human associated microbiomes primarily involves the analysis of taxonomic count tables generated via high-throughput sequencing.
 It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional.
 An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets.
-Set-based analysis is ubiquitous in the genomics literature, and have demonstrable impact in improving interpretability and power of downstream analysis.
-Current practice in the microbiome field, however, performs aggregation by summing the abundance of respective taxa.
+Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis.
+Unfortunately, there is a lack of interest in developing more sophisticated set-based analysis methods specific to microbiome data, where current practice often employs abundance summation as a technique for aggregation.
 This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
-Unfortunately, there is a lack of interest in developing more sophisticated set-based analysis methods specific to microbiome data.
-Here, we attempt to fill this gap with a new sample-level taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature.
+Here, we attempt to fill this gap with a new single sample taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature.
 Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
 We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
-Herein we demonstrate under both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios, while additionally providing informative scores that can be inputs to downstream differential abundance and predictive tasks with good performance.
-These results demonstrate how our approach can enable researchers to generate meaningful aggregation scores for sets of taxa.
+Herein we demonstrate under both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios, while additionally providing informative scores that can be inputs to downstream differential abundance and prediction tasks.
 
@@
 We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis.
-Our scores can be used for inference purposes, as well as inputs to other downstream analyses such as differential abundance and predictive models.
-We demonstrated the performance of our method against competing approaches across both real data analyses and simulation studies.
+Our scores can be used for statistical inference, and as inputs to other downstream analyses such as differential abundance and prediction models.
+We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.
 
@@
 Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples.
-Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample and is more akin to a transformation.
-This strategy allows for the flexible incorporation of different statistical techniques downstream, such as predictive models, as well as for visualization purposes in ordination plots.
+Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample level and are more akin to a transformation.
+This strategy allows for the flexible incorporation of different statistical techniques downstream, such as prediction models, as well as for visualization purposes in ordination plots.
 
 In microbiome research, even though no explicit enrichment analysis is performed, standard practice often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum by simple summation of abundances \cite{mclaren2019}.
-Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, increase bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}.
-As such, there is a need for microbiome researchers to adopt set enrichment methods to perform a more robust analysis that allows for enrichment testing of functionally relevant sets of microbes.
-Unfortunately, there is a lack of interest in extending existing methods to be more specific to microbiome relative abundance data.
+Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, it increases bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}.
+As such, there is a need for microbiome researchers to adopt more robust sample-level set enrichment methods.
+Unfortunately, limited work has been done to extend existing methods to be more specific to microbiome relative abundance data.
 Some software suites, such as \emph{MicrobiomeAnalyst}, do provide researchers tools to perform enrichment testing with curated taxon sets \cite{chong2020}.
-However, the approach used in \emph{MicrobiomeAnalyst} is an overrepresentation analysis and therefore similarly sensitive to the method of differential abundance used.
-
-Here, we developed a novel method that allows for generating enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}.
-We leverage the conception of the $Q_1$ competitive hypothesis presented in Tian et al.
-\cite{tian2005}, which compares the expression of genes within the gene set against the rest of the genes.
-The competitive hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables.
+However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used.
+
+Here, we developed a novel method that generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}.
+We leverage the concept of the $Q_1$ competitive hypothesis presented in Tian et al.
+\cite{tian2005}, which tests the null that the value of variables within a specific set is equal to the value of measured variables not in the set.
+The competitive null hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables.
 We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a multiplicative aggregation method that addresses the downsides of the naive summation-based method presented above \cite{mclaren2019, silverman2017}.
@@
 In the following sections, we provide the formulation of cILR and discuss some statistical properties.
-Then, we illustrate significance testing at the sample level using cILR and demonstrating type I error and power under different simulation scenarios and real data applications.
-Then we assessed the informativeness of cILR generated scores, and evaluated how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis.
-We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are equivalent single sample enrichment analysis methods in the genomics literature.
+We illustrate significance testing at the sample level using cILR and evaluate type I error and power under different simulation scenarios and real data applications.
+We assess the informativeness of cILR generated scores, and evaluate how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis.
+We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods.
 An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).
@@
 
-\noindent \textbf{Prediction}: To assess the predictability, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}.
+\noindent \textbf{Prediction}: To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}.
 For binary outcomes, we use AUC similar to the classification analyses above.
@@
 This data set is special because it is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}.
-Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false discovery rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment.
+Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false positive rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment.
 Microbial tropism annotation at the genus level was from Beghini et al.

### 2021-07-26 supplemental.tex

@@
 \section{Distribution of p-values}
+We also demonstrate the validity our inference procedure by assessing the distribution of p-values.
+We used simulated data sets as described in the main manuscript, under different data sparsity levels (0, 0.6) and inter-taxa correlation within the set ($\rho$ = 0, 0.5).
+We evaluated p-values of the correlation adjusted method for $\rho = 0.5$ case and the p-values of the unadjusted method for $\rho = 0$ case.
 

### 2021-07-26 supplemental.tex

@@
 
+In general, sparsity has no effect on the distribution of p-values, where it remains uniform at zero correlation values.
+However, for higher correlation values, p-values deviate more from uniform under the global null, with the mixture normal distribution being further away from uniform compared to the normal distribution.
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/pval_distr.png}
+    \caption{Q-Q plot of 10,000 p-values compared against a uniform distribution bounded between 0 and 1.
+Evaluation was performed on simulated null data sets of 10,000 samples testing for enrichment of a set of size 50.
+For correlation of 0.5, p-values represent correlation adjusted cILR while for correlation of 0, p-values represent unadjusted cILR.}
+    \label{fig:s2}
+\end{figure}
+
 \newpage

### 2021-07-30 manuscript_plos.tex

@@
 \subsection*{Evaluation}
+All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/cILR\_analysis}{qpmnguyen/cILR\_analysis}).
+
 \subsubsection*{Parametric Simulations}  
@@
 For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUROC/AUC).
-This is a strategy used in Frost et al.
-\cite{frost2020a} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
+This is a strategy used in Frost \cite{frost2020a} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
 whether samples with inflated counts are highly ranked using estimated scores).
@@
 \noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank sum test.
-We added a pseudocount of 1 to all zero entry values.
+We added a pseudocount of 1 to all values.
 This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample.
@@
 \noindent \textbf{Differential Abundance}: Since cILR are sample-level enrichment scores, we performed differential abundance by using a Wilcoxon Rank Sum test and Welch's t-test across case/control status on cILR generated scores.
-We added a pseudocount of 1 to all zero entry values.
+We added a pseudocount of 1 to all values.
 For comparison, we chose representative state-of-the-art methods in differential abundance analysis, namely DESeq2 \cite{love2014,mcmurdie2014} and corncob \cite{martin2020}.
@@
 \noindent \textbf{Disease Prediction}: We fit random forest on cILR scores, as well as ssGSEA \cite{hanzelmann2013} and GSVA \cite{barbie2009} similar to single sample enrichment section.
-We added a pseudocount of 1 to all zero entry values.
+We added a pseudocount of 1 to all values.
 Additionally, we also compared performance using cILR against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.
@@
     \includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
-    \caption{Type I error rate \textbf{(A)}, and power \textbf{(B)} across different parametric simulation scenarios.
+    \caption{Sample-level inference with cILR under parametric simulations (\textbf{(A)} and \textbf{B}), and real data analysis (\textbf{(C)} and \textbf{(D)}).
+In simulation analyses, panel \textbf{(A)} shows type I error rate, and panel \textbf{(B)} shows power for single sample enrichment test for a specified set and was compared against a Wilcoxon rank sum test at $\alpha$ of 0.05.
+In real data analysis, panel \textbf{(C)} shows the false-positive rate, and panel \textbf{(D)} shows the true positive rate.
+For this analysis, 16S rRNA data from the oral microbiome of the gingival site was used.
+The set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples.
 Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach.
-False-positive rate \textbf{(C)} and true positive rate \textbf{(D)} evaluation of similar methods on real 16S rRNA data from the oral microbiome of the gingival site.
-For \textbf{(A)} and \textbf{(B)}, enrichment of a specified set was tested at the sample level using cILR and the Wilcoxon rank sum test at $\alpha$ of 0.05.
-For \textbf{(C)} and \textbf{(D)}, the set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples.
 Adjusted cILR demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate.
@@
     \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
-    \caption{Placeholder} 
+    \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
+For cILR methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by cILR (across different output types and distributional assumptions).
+cILR associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
     \label{fig:4}
 \end{figure}
+
+Fig.~\ref{fig:4} present results for simulation studies for both type I error (panel A) and power (panel B) evaluations.
+All methods control for type I error well across both sparsity and correlation levels, where the estimated rate was consistently around the 0.05 pre-defined threshold.
+Results were similar across all evaluated methods, although in some instances, for example in medium correlation setting ($\rho = 0.2$), the unadjusted cILR resulted in higher type I error, regardless of difference in means test and distribution of choice.
+
+The difference between the methods is more noticeable when evaluating power.
+All cILR associated variants showed much higher power even when the effect size is limited (fold change is 1.5), and there is a noticeable gap in performance between cILR and both DESeq2 and corncob.
+Surprisingly, this effect is consistent across correlation levels and sparsity, even though we expectedly see performance in power drop as a function of sparsity especially in low effect size settings.
+
 \subsubsection*{Real data evaluations}
-We also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP (Fig~\ref{fig:5}).
+In addition to simulation studies, we also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP (Fig~\ref{fig:5}).
 For type I error evaluations, we use stool samples and randomly assign them with case/control status and calculated type I error as the proportion of genera identified as significantly different.
 For true positive rate evaluations, we use the gingival data set as detailed in the previous section, and calculated the true positive rate as the proportion of genera labeled as either anaerobic or aerobic that were found to be significant.
+
 We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of cILR were controlling for type I error at the defined $\alpha$ threshold of 0.05.
-Conversely, true positive rate for all methods were roughly more similar.
-As expected, using the Wilcoxon rank sum test demonstrated have power true positive rate, but the difference was not large.
-Surprisingly, using Welch's t-test demonstrated similar true positive rate as both corncob and DESeq2, while also controlling for type I error.
-Choice of underlying distribution for cILR did not seem to alter performance.
+This is surprising given the consistency of preserving type I error for both corncob and DESeq2 in all simulation evaluations.
+
+In true positive experiments with data from the gingival site, estimated rates were more similar across the different methods.
+As expected, using the Wilcoxon rank sum test resulted in lower true positive rate compared to remaining methods, but the difference was not noticable.
+This is also surprising given that in simulation studies, both corncob and DESeq2 showed markedly lower power across all effect sizes.
 
@@
 Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.
-
 \subsubsection*{Simulation studies}
-Fig~\ref{fig:6} showed results for simulation studies as detailed in the \nameref{methods} section.
-Panel A.
-presents results for a regression learning task with a continuous outcome while panel B.
-presents results for a classification task with a binary outcome.
+Fig~\ref{fig:6} shows results for simulation studies as detailed in the \nameref{methods} section.
+Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome.
 As expected, performance across all assessed methods increased with a higher signal-to-noise ratio.
@@
 This is further encouraged in real data analysis where the false discovery rate was around $0.05$ when a collection of true null and true alternate hypotheses were tested.
-Unfortunately the trade-off between good type I error control is lower power.
+Unfortunately the trade-off of good type I error control is lower power.
 The conservativeness of the test attenuates with higher sparsity and correlation, where power was not approaching even 50\% even at the highest effect sizes.
@@
 \subsubsection*{Differential abundance analysis}
-For differential abundance testing, we evaluated whether using cILR alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes.
-We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014}.
+For differential abundance testing, we evaluated whether using cILR scores alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes.
+We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014} applied on data where taxa were aggregated using the naive sum method.
 We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome data \cite{mcmurdie2014}.
-Conversely, corncob is a newer method developed specifically for microbiome data sets, where taxonomic counts are modeled directly newly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.
-In our analyses, we found that both corncob and DESeq2 has inflated levels of type I error.
+Conversely, corncob is a newer method developed specifically for microbiome data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.
+
+Surprisingly, we found some conflicting results when evaluating comparisons across parametric simulations and real data analysis.
+The performance of cILR was consistent across two evaluation criteria, demonstrating good type I error and respectable power.
+However, corncob and DESeq2 showed opposite effects: in simulation experiments, both methods show good type I error control but low power, while in real data analyses, conversely both corncob and DESeq2 showed inflated type I error but comparable power with respect to cILR methods.
+Despite such discrepancy, results still indicate good performance of cILR scores when used as inputs to downstream differential abundance analysis compared to using aggregated raw counts, even in methods designed to handle that type of data such as corncob and DESeq2.
+
+We hypothesized that the above behavior can arise due to issues with performing inference in the presence taxon-specific protocol biases \cite{mclaren2019}.
+According to McLaren et al., the observed relative abundance of taxa is different than the true relative abundance due to protocol bias, and importantly this bias is specific to each taxon \cite{mclaren2019}.
+This is especially true in the context of sum-based aggregations, where the resulting bias of the aggregated taxon are dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}).
+Conceptually, this means that measurement error for a taxon aggregate is different across samples as relative abundance of contributing changes, leading to issues when attempting to perform inference.
+As such, we expect methods like corncob or DESeq2 when performed on such aggregates to have inflated type I error compared to our taxon-ratio based approach.
+
+The bias model also helps explain differences in performance of DESeq2 and corncob in simulation analyses compared to real data.
+Our simulation protocol does not explicitly include bias in our formulation, and all taxa were generated from the same underlying distribution with similar variances across all samples (the only difference is in the mean value where a taxa is expected to have inflated counts).
+As such, we do not expect our simulated taxa to have any taxon-specific biases, which is not the case in real data settings.
+Therefore, we can expect DESeq2 and corncob to retain their expected type I error control in simulation analyses compared to real data.
+It is still surprising to see lower power for both methods in simulation analyses, which might be due to the fact that the evaluation protocol only considers default settings for both methods and did not attempt to optimize performance.
+
+The fact that the performance of cILR remains consistent across both simulation and real data analysis shows that cILR is invariant to taxon-specific biases.
+Furthermore, our evaluation indicates that even a simple difference in means test when used in combination with cILR scores can preserve type I error while maintaining good power.
+As such, a practioner can use cILR as a pre-processing step prior to performing a differential abundance test.
 
@@
 This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}.
-Results indicated that cILR produce good performance values across both real data analysis and simulation scenarios.
-Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicate that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
+Results indicated that cILR produces good performance values across both real data analysis and simulation scenarios.
+Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
 Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome data sets.
 This is consistent with our sample ranking analysis (Fig~\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
-However, cILR did not outperform the CLR approach across all our simulation studies, and only marginally performed better in the real data analysis with WGS data.
+However, cILR did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data.
 
@@
 Ultimately, results indicate that cILR can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity.
-Even though there exists situations where it might not provide maximum predictive values, the flexibility of cILR in various types of analyses enable ven though in some scenarios it might not provide maximum predictive values.
+Even though there exists situations where it might not provide maximum predictive values, the flexibility of cILR in various types of analyses enable even though in some scenarios it might not provide maximum predictive values.
 
@@
 First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome data.
-There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation section of prior studies.
-Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is flawed.
+There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation of prior studies.
+Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is not perfect.
 This is because the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut.
-As such, results from power analyses using this data set is only relative between the comparison methods, and absolute values should should not carry meaning.
-Finally, we assumed that taxa within a set all be equally associated with the outcome.
+As such, results from power analyses using this data set is only relative between the comparison methods.
+Finally, we assumed that taxa within a set are all equally associated with the outcome.
 This limits our ability to evaluate the performance of cILR when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.
 
-However, our evaluation still showed various drawbacks of the cILR method.
+Our evaluation showed various drawbacks of the cILR method.
 First, inference with cILR is limited in being able to exhaustively detect all samples with significant inflated counts for a set in situations where there is a high degree of sparsity and inter-taxa correlation.
-Second, for downstream analyses, cILR might not always better than competing methods, especially when being used as inputs to predictive models.
+Second, for downstream analyses, cILR might not always perform better than competing methods, especially when being used to generate inputs to predictive models.
 We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated with adjusting the mixture normal distribution.
@@
 
-In addition, there are possible extensions cILR can we can consider to provide more flexibility across different types of scenarios in data analysis.
-First, cILR did not address the sparsity of microbiome data and relies on pseudocount to ensure log operations are valid.
+In addition, there are possible extensions cILR can we can consider to provide more flexibility across different data analysis scenarios in data analysis.
+First, cILR did not address the sparsity of microbiome data and relies on a pseudocount to ensure log operations are valid.
 We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
 Second, cILR also treated all taxa within the set as equally contributing to the set.
-Incorporate taxa-specific weights can reduce the influence of outliers, such as rare or highly invariant taxa.
+Incorporation of taxa-specific weights can reduce the influence of outliers, such as rare or highly invariant taxa.
 Finally, curating sets based on apriori characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.
@@
 Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomics data sets.
-However, there has not been a lot of set-based analysis methods developed specifically for microbiome relative abundance data.
+However, limited work has been done developing set based methods specifically for microbiome relative abundance data.
 In this manuscript, we introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level.
@@
 \label{S1_Fig}
-{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+{\bf Computational performance of cILR.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
+Evaluation was performed on simulated data sets.
+For sample size analysis, only 10 sets were evaluated.
+For taxa set analysis, sample size was fixed at 1,000.
+Across all evaluations, the size of each taxa set was also fixed at 50.
+
+\paragraph*{S2 Fig.}
+\label{S2_Fig}
+{\bf Distribution of p-values.} Q-Q plot of 10,000 p-values compared against a uniform distribution bounded between 0 and 1.
+Evaluation was performed on simulated null data sets of 10,000 samples testing for enrichment of a set of size 50.
+For correlation of 0.5, p-values represent correlation adjusted cILR while for correlation of 0, p-values represent unadjusted cILR.
 
@@
 \label{S1_File}
-{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 Appendix.}
-\label{S1_Appendix}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
+{\bf Supplemental derivations.} Includes additional details on addressing variance inflation due to correlation in cILR, as well as computational performance and p-value distribution of the method.
 
 \section*{Acknowledgments}
-Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
-Interdum et malesuada fames ac ante ipsum primis in faucibus.
-Nam id pretium nisi.
-Sed ac quam id nisi malesuada congue.
-Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.
+The authors thank Becky Lebeaux, Modupe Coker, Erika Dade, Jie Zhou, and Weston Viles for insightful comments and suggestions that greatly improved the paper.
+This research is supported by funding from the National Library of Medicine (NLM) (grants NLM R01LM012723 and NLM K01LM012426).
+None of the funding bodies had a role in the design of the study or the collection, analysis, and interpretation of data and in writing of the manuscript.
 
 \nolinenumbers
-
 % Either type in your references using

### 2021-08-20 manuscript_plos.tex

@@
 \section*{Abstract}
-Research in human associated microbiomes primarily involves the analysis of taxonomic count tables generated via high-throughput sequencing.
+Research in human associated microbiomes often involves the analysis of taxonomic count tables generated via high-throughput sequencing.
 It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional.
@@
 Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis.
-Unfortunately, there is a lack of interest in developing more sophisticated set-based analysis methods specific to microbiome data, where current practice often employs abundance summation as a technique for aggregation.
+Unfortunately, there is a lack of sophisticated set-based analysis methods specific to microbiome data, where current practice often employs abundance summation as a technique for aggregation.
 This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
@@
 We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
-Herein we demonstrate under both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios, while additionally providing informative scores that can be inputs to downstream differential abundance and prediction tasks.
+Herein we demonstrate using both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios.
+Additionally, it provides informative scores that can be inputs to downstream differential abundance and prediction tasks.
 
@@
 \section*{Author summary}
-The study of human associated microbiomes rely on genomic surveys via high-throughput sequencing.
+The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing.
 However, microbiome data is sparse and high dimensional which prevents the application of standard statistical techniques.
@@
 Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability.
-Unfortunately, there is not a lot of research in developing new set-based tools for microbiome data specifically, given that compared to other omics data types microbiome data is strictly compositional.
+Unfortunately, there is not a lot of research in developing new set-based tools for microbiome data specifically, given that compared to other 'omics data types microbiome data is strictly compositional.
 We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis.
@@
 \section*{Introduction}
-The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their human host.
-Previous research has shown that changes in the composition of the microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
+The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host.
+Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
 To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
@@
 whole-genome shotgun sequencing) \cite{cho2012}.
-Processing raw sequencing data using a variety of bioinformatic pipelines \cite{callahan2016,truong2015} yields taxonomic abundance tables that can be used in downstream statistical analyses to identify associations between an outcome/exposure and identified taxa.
-
-However, there exists unique challenges in the analysis of these data tables \cite{li2019,li2015}.
+Raw sequencing data is then processed through a variety of bioinformatic pipelines \cite{callahan2016, truong2015}, yielding various data products, one of which are taxonomic tables which can be used to study associations between members of the microbiome and an exposure or outcome of interest.
+
+However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019,li2015}.
 First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present.

### 2021-09-07 manuscript_plos.tex
Fixing issues Annie talked about

@@
 Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis.
-Unfortunately, there is a lack of sophisticated set-based analysis methods specific to microbiome data, where current practice often employs abundance summation as a technique for aggregation.
+Unfortunately, there is a lack of sophisticated set-based analysis methods specific to microbiome taxonomic data, where current practice often employs abundance summation as a technique for aggregation.
 This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
@@
 The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing.
-However, microbiome data is sparse and high dimensional which prevents the application of standard statistical techniques.
+However, microbiome taxonomic data is sparse and high dimensional which prevents the application of standard statistical techniques.
 One approach to address this problem is to perform analyses at the level of taxon sets.
 Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability.
-Unfortunately, there is not a lot of research in developing new set-based tools for microbiome data specifically, given that compared to other 'omics data types microbiome data is strictly compositional.
+Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given that compared to other 'omics data types microbiome taxonomic data is strictly compositional.
 We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis.
@@
 
-Second, the number of reads obtained is constrained by the sequencing instrument at an arbitrary limit, and applied inconsistently across samples, resulting in a different number of total read counts per sample.
+Second, the number of reads obtained is constrained by the sequencing instrument at an arbitrary limit, and is inconsistent across samples, resulting in a variable number of total read counts per sample.
 Many normalization methods \cite{weiss2017} have been proposed to address these issues, including cross-applying methods from the gene expression literature \cite{mcmurdie2014}.
 However, these methods rely on assumptions specific to the original bulk RNA-seq data sets such as the presence of housekeeping genes with consistent expression levels \cite{love2014}, which might not be not true in the context of microbiome relative abundance data \cite{quinn2019,quinn2018b}.
-As such, microbiome data is strictly compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another.
+As such, microbiome taxonomic data is strictly compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another.
 Consequently, log-ratio transformations from the compositional data literature are often utilized \cite{aitchison1999}.
 
-Third, the data is highly zero-inflated, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment).
+Third, the data are highly zero-inflated, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment).
 Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015}, or applying zero-inflated models \cite{chen2018, kaul2017}.
@@
 Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}.
-Aggregated sets can be less sparse than their constituent elements, and testing on a smaller number of variables can reduce the multiple testing burden, thereby increasing power and reproducibility.
+Aggregated sets are less sparse than their constituent elements, and testing on a smaller number of variables reduces the multiple testing burden, thereby increasing power and reproducibility.
 Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations.
@@
 Unfortunately, limited work has been done to extend existing methods to be more specific to microbiome relative abundance data.
-Some software suites, such as \emph{MicrobiomeAnalyst}, do provide researchers tools to perform enrichment testing with curated taxon sets \cite{chong2020}.
+Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}.
 However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used.
 
-Here, we developed a novel method that generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}.
+Here, we present a novel method that generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}.
 We leverage the concept of the $Q_1$ competitive hypothesis presented in Tian et al.
@@
 We assess the informativeness of cILR generated scores, and evaluate how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis.
-We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods.
+We compare the performance of cILR in these respective tasks against standard microbiome taxonomic data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods.
 An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).
@@
 Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution.
-Valid p-values can be calculated by subtracting $\mathbf{E}$ from 1.
+P-values can be calculated by subtracting $\mathbf{E}$ from 1.
 \end{enumerate}
@@
 This assumption is also supported by preliminary simulation studies (detailed below).
-In panel A of Fig~\ref{fig:1}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
+In panel A of Fig~\ref{fig:1}, we simulated microbiome taxonomic count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
 It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation.
@@
 Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3).
-Due to the compositional nature of microbiome data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets.
+Due to the compositional nature of microbiome taxonomic data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets.
 As such, we applied a compensation procedure as described in Hawinkel et al.
@@
 \cite{calgaro2020}.
-This data set is special because it is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}.
+This data set is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}.
 Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false positive rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment.
@@
 
-Unfortunately, cILR did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
+In this analysis, cILR unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
 This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation.
@@
 We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014} applied on data where taxa were aggregated using the naive sum method.
-We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome data \cite{mcmurdie2014}.
-Conversely, corncob is a newer method developed specifically for microbiome data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.
+We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}.
+Conversely, corncob is a newer method developed specifically for microbiome taxonomic data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.
 
@@
 Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
-Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome data sets.
+Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome taxonomic data sets.
 This is consistent with our sample ranking analysis (Fig~\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
@@
 There are various limitations to our evaluation of cILR.
-First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome data.
+First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome taxonomic data.
 There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation of prior studies.
@@
 In addition, there are possible extensions cILR can we can consider to provide more flexibility across different data analysis scenarios in data analysis.
-First, cILR did not address the sparsity of microbiome data and relies on a pseudocount to ensure log operations are valid.
+First, cILR did not address the sparsity of microbiome taxonomic data and relies on a pseudocount to ensure log operations are valid.
 We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
 Second, cILR also treated all taxa within the set as equally contributing to the set.
-Incorporation of taxa-specific weights can reduce the influence of outliers, such as rare or highly invariant taxa.
-Finally, curating sets based on apriori characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.
+Incorporation of taxa-specific weights could reduce the influence of outliers, such as rare or highly invariant taxa.
+Finally, curating sets based on \emph{apriori} characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.
 
@@
 However, limited work has been done developing set based methods specifically for microbiome relative abundance data.
-In this manuscript, we introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level.
+We introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level.
 We demonstrated that our method can control for type I error for significance testing at the sample level, while generated scores are also valid inputs in downstream analyses, including disease prediction and differential abundance.

### 2021-09-07 manuscript_plos.tex
Updates from Overleaf

@@
     \centering
-    \includegraphics[width=\linewidth]{figures/kurtosis_skewness_gof.png}
+    %\includegraphics[width=\linewidth]{figures/kurtosis_skewness_gof.png}
     \caption{{\bf Properties of the null distribution of cILR under the global null simulations}.
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
+    %\includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
     \caption{Sample-level inference with cILR under parametric simulations (\textbf{(A)} and \textbf{B}), and real data analysis (\textbf{(C)} and \textbf{(D)}).
@@
     \centering
-    \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
+    %\includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
     \caption{Classification performance via AUC of cILR, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}.
@@
     \centering
-    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
+    %\includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
     \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
@@
     \centering
-    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
+    %\includegraphics[width = \textwidth]{figures/data_diff_ab.png}
     \caption{Differential abundance analysis using corncob, DESeq2 and cILR with either Wilcoxon rank sum test or Welch's t-test.
@@
     \centering
-    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
+    %\includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
     \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
@@
     \centering
-    \includegraphics[scale = 0.6]{figures/data_prediction_plot.png}
+    %\includegraphics[scale = 0.6]{figures/data_prediction_plot.png}
     \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
@@
 
+\section*{Acknowledgments}
+The authors thank Becky Lebeaux, Modupe Coker, Erika Dade, Jie Zhou, and Weston Viles for insightful comments and suggestions that greatly improved the paper.
+This research is supported by funding from the National Library of Medicine (NLM) (grants NLM R01LM012723 and NLM K01LM012426).
+None of the funding bodies had a role in the design of the study or the collection, analysis, and interpretation of data and in writing of the manuscript.
+
+\nolinenumbers
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here. See http://journals.plos.org/plosone/s/latex for 
+% step-by-step instructions.
+%
+
+% \bibliography{tax_agg}{}
+
+\begin{thebibliography}{10}
+
+\bibitem{proctor2019}
+Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
+\newblock The {{Integrative Human Microbiome Project}}.
+\newblock Nature. 2019;569(7758):641--648.
+\newblock doi:{10.1038/s41586-019-1238-8}.
+
+\bibitem{sharma2019}
+Sharma S, Tripathi P.
+\newblock Gut Microbiome and Type 2 Diabetes: Where We Are and Where to Go?
+\newblock The Journal of Nutritional Biochemistry. 2019;63:101--108.
+\newblock doi:{10.1016/j.jnutbio.2018.10.003}.
+
+\bibitem{aoun2020}
+Aoun A, Darwish F, Hamod N.
+\newblock The {{Influence}} of the {{Gut Microbiome}} on {{Obesity}} in
+  {{Adults}} and the {{Role}} of {{Probiotics}}, {{Prebiotics}}, and
+  {{Synbiotics}} for {{Weight Loss}}.
+\newblock Preventive Nutrition and Food Science. 2020;25(2):113--123.
+\newblock doi:{10.3746/pnf.2020.25.2.113}.
+
+\bibitem{cho2012}
+Cho I, Blaser MJ.
+\newblock The Human Microbiome: At the Interface of Health and Disease.
+\newblock Nature Reviews Genetics. 2012;13(4):260--270.
+\newblock doi:{10.1038/nrg3182}.
+
+\bibitem{callahan2016}
+Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
+\newblock {{DADA2}}: {{High}}-Resolution Sample Inference from {{Illumina}}
+  Amplicon Data.
+\newblock Nature Methods. 2016;13(7):581--583.
+\newblock doi:{10.1038/nmeth.3869}.
+
+\bibitem{truong2015}
+Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
+\newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
+\newblock Nature Methods. 2015;12(10):902--903.
+\newblock doi:{10.1038/nmeth.3589}.
+
+\bibitem{li2019}
+Li H.
+\newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
+  {{Metagenomics}}.
+\newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
+  2019. p. 977--550.
+
+\bibitem{li2015}
+Li H.
+\newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
+  Compositional Data Analysis}}.
+\newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
+\newblock doi:{10.1146/annurev-statistics-010814-020351}.
+
+\bibitem{shi2016}
+Shi P, Zhang A, Li H.
+\newblock Regression Analysis for Microbiome Compositional Data.
+\newblock The Annals of Applied Statistics. 2016;10(2):1019--1040.
+\newblock doi:{10.1214/16-AOAS928}.
+
+\bibitem{sankaran2014}
+Sankaran K, Holmes S.
+\newblock {{structSSI}}: {{Simultaneous}} and {{Selective Inference}} for
+  {{Grouped}} or {{Hierarchically Structured Data}}.
+\newblock Journal of statistical software. 2014;59(13):1--21.
+\newblock doi:{10.18637/jss.v059.i13}.
+
+\bibitem{benjamini1995}
+Benjamini Y, Hochberg Y.
+\newblock Controlling the {{False Discovery Rate}}: {{A Practical}} and
+  {{Powerful Approach}} to {{Multiple Testing}}.
+\newblock Journal of the Royal Statistical Society Series B (Methodological).
+  1995;57(1):289--300.
+
+\bibitem{chen2018}
+Chen J, King E, Deek R, Wei Z, Yu Y, Grill D, et~al.
+\newblock An Omnibus Test for Differential Distribution Analysis of Microbiome
+  Sequencing Data.
+\newblock Bioinformatics. 2018;34(4):643--651.
+\newblock doi:{10.1093/bioinformatics/btx650}.
+
+\bibitem{weiss2017}
+Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
+\newblock Normalization and Microbial Differential Abundance Strategies Depend
+  upon Data Characteristics.
+\newblock Microbiome. 2017;5(1).
+\newblock doi:{10.1186/s40168-017-0237-y}.
+
+\bibitem{mcmurdie2014}
+McMurdie PJ, Holmes S.
+\newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
+  Inadmissible}}.
+\newblock PLOS Computational Biology. 2014;10(4):e1003531.
+\newblock doi:{10.1371/journal.pcbi.1003531}.
+
+\bibitem{love2014}
+Love MI, Huber W, Anders S.
+\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
+  Data with {{DESeq2}}.
+\newblock Genome Biology. 2014;15(12):550.
+\newblock doi:{10.1186/s13059-014-0550-8}.
+
+\bibitem{quinn2019}
+Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
+\newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
+\newblock GigaScience. 2019;8(giz107).
+\newblock doi:{10.1093/gigascience/giz107}.
+
+\bibitem{quinn2018b}
+Quinn TP, Erb I, Richardson MF, Crowley TM.
+\newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
+\newblock Bioinformatics. 2018;34(16):2870--2878.
+\newblock doi:{10.1093/bioinformatics/bty175}.
+
+\bibitem{gloor2017}
+Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
+\newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
+  Optional}}.
+\newblock Frontiers in Microbiology. 2017;8.
+\newblock doi:{10.3389/fmicb.2017.02224}.
+
+\bibitem{aitchison1999}
+Aitchison J.
+\newblock A {{Concise Guide}} to {{Compositional Data Analysis}}. 1999; p. 134.
+
+\bibitem{kurtz2015}
+Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
+\newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
+  Ecological Networks}}.
+\newblock PLOS Computational Biology. 2015;11(5):e1004226.
+\newblock doi:{10.1371/journal.pcbi.1004226}.
+
+\bibitem{kaul2017}
+Kaul A, Davidov O, Peddada SD.
+\newblock Structural Zeros in High-Dimensional Data with Applications to
+  Microbiome Studies.
+\newblock Biostatistics. 2017;18(3):422--433.
+\newblock doi:{10.1093/biostatistics/kxw053}.
+
+\bibitem{kaul2017a}
+Kaul A, Mandal S, Davidov O, Peddada SD.
+\newblock Analysis of {{Microbiome Data}} in the {{Presence}} of {{Excess
+  Zeros}}.
+\newblock Frontiers in Microbiology. 2017;8.
+\newblock doi:{10.3389/fmicb.2017.02114}.
+
+\bibitem{khatri2012}
+Khatri P, Sirota M, Butte AJ.
+\newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
+  {{Outstanding Challenges}}.
+\newblock PLOS Computational Biology. 2012;8(2):e1002375.
+\newblock doi:{10.1371/journal.pcbi.1002375}.
+
+\bibitem{goeman2007}
+Goeman JJ, B{\"u}hlmann P.
+\newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
+  Issues.
+\newblock Bioinformatics. 2007;23(8):980--987.
+\newblock doi:{10.1093/bioinformatics/btm051}.
+
+\bibitem{subramanian2005}
+Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
+\newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
+  Interpreting Genome-Wide Expression Profiles.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(43):15545--15550.
+\newblock doi:{10.1073/pnas.0506580102}.
+
+\bibitem{ashburner2000}
+Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
+\newblock Gene {{Ontology}}: Tool for the Unification of Biology.
+\newblock Nature genetics. 2000;25(1):25--29.
+\newblock doi:{10.1038/75556}.
+
+\bibitem{hanzelmann2013}
+H{\"a}nzelmann S, Castelo R, Guinney J.
+\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
+  {{RNA}}-{{Seq}} Data.
+\newblock BMC Bioinformatics. 2013;14(1):7.
+\newblock doi:{10.1186/1471-2105-14-7}.
+
+\bibitem{frost2020a}
+Frost HR.
+\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
+  Method for Cell-Specific Gene Set Scoring.
+\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
+\newblock doi:{10.1093/nar/gkaa582}.
+
+\bibitem{mclaren2019}
+McLaren MR, Willis AD, Callahan BJ.
+\newblock Consistent and Correctable Bias in Metagenomic Sequencing
+  Experiments.
+\newblock eLife. 2019;8:e46923.
+\newblock doi:{10.7554/eLife.46923}.
+
+\bibitem{egozcue2005}
+Egozcue JJ, {Pawlowsky-Glahn} V.
+\newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2005;37(7):795--828.
+\newblock doi:{10.1007/s11004-005-7381-9}.
+
+\bibitem{chong2020}
+Chong J, Liu P, Zhou G, Xia J.
+\newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
+  Functional, and Meta-Analysis of Microbiome Data.
+\newblock Nature Protocols. 2020;15(3):799--821.
+\newblock doi:{10.1038/s41596-019-0264-1}.
+
+\bibitem{tian2005}
+Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
+\newblock Discovering Statistically Significant Pathways in Expression
+  Profiling Studies.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(38):13544--13549.
+\newblock doi:{10.1073/pnas.0506577102}.
+
+\bibitem{egozcue2003}
+Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcelo-Vidal} C.
+\newblock Isometric {{Logratio Transformations}} for {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2003; p.~22.
+
+\bibitem{silverman2017}
+Silverman JD, Washburne AD, Mukherjee S, David LA.
+\newblock A Phylogenetic Transform Enhances Analysis of Compositional
+  Microbiota Data.
+\newblock eLife. 2017;6:e21887.
+\newblock doi:{10.7554/eLife.21887}.
+
+\bibitem{barbie2009}
+Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
+\newblock Systematic {{RNA}} Interference Reveals That Oncogenic
+  {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
+\newblock Nature. 2009;462(7269):108--112.
+\newblock doi:{10.1038/nature08460}.
+
+\bibitem{quast2013}
+Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
+\newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
+  Processing and Web-Based Tools.
+\newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
+\newblock doi:{10.1093/nar/gks1219}.
+
+\bibitem{delignette-muller2015}
+{Delignette-Muller} ML, Dutang C.
+\newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
+\newblock Journal of Statistical Software. 2015;64(4):1--34.
+
+\bibitem{benaglia2009}
+Benaglia T, Chauveau D, Hunter DR, Young D.
+\newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
+\newblock Journal of Statistical Software. 2009;32(6):1--29.
+
+\bibitem{washburne2017}
+Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL, Mukherjee S, et~al.
+\newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
+  Associations in Microbiome Datasets.
+\newblock PeerJ. 2017;5:e2969.
+\newblock doi:{10.7717/peerj.2969}.
+
+\bibitem{morton2017}
+Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
+  et~al.
+\newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
+\newblock mSystems. 2017;2(1).
+\newblock doi:{10.1128/mSystems.00162-16}.
+
+\bibitem{efron2004}
+Efron B.
+\newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
+\newblock Journal of the American Statistical Association.
+  2004;99(465):96--104.
+\newblock doi:{10.1198/016214504000000089}.
+
+\bibitem{wu2012}
+Wu D, Smyth GK.
+\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
+  Correlation.
+\newblock Nucleic Acids Research. 2012;40(17):e133.
+\newblock doi:{10.1093/nar/gks461}.
+
+\bibitem{calgaro2020}
+Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
+\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
+  and Metagenomics Applied to Microbiome Data.
+\newblock Genome Biology. 2020;21(1):191.
+\newblock doi:{10.1186/s13059-020-02104-1}.
+
+\bibitem{cario1997}
+Cario MC.
+\newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
+  Distributions}} and {{Correlation Matrix}}. 1997; p.~19.
+
+\bibitem{agresti1998}
+Agresti A, Coull BA.
+\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
+  Estimation}} of {{Binomial Proportions}}.
+\newblock The American Statistician. 1998;52(2):119--126.
+\newblock doi:{10.2307/2685469}.
+
+\bibitem{delong1988}
+DeLong ER, DeLong DM, {Clarke-Pearson} DL.
+\newblock Comparing the Areas under Two or More Correlated Receiver Operating
+  Characteristic Curves: A Nonparametric Approach.
+\newblock Biometrics. 1988;44(3):837--845.
+
+\bibitem{hawinkel2019}
+Hawinkel S, Mattiello F, Bijnens L, Thas O.
+\newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
+  Control the False Discovery Rate.
+\newblock Briefings in Bioinformatics. 2019;20(1):210--221.
+\newblock doi:{10.1093/bib/bbx104}.
+
+\bibitem{breiman2001}
+Breiman L.
+\newblock Random {{Forests}}.
+\newblock Machine Learning. 2001;45(1):5--32.
+\newblock doi:{10.1023/A:1010933404324}.
+
+\bibitem{kuhn2020}
+Kuhn M, Wickham H.
+\newblock Tidymodels: {{Easily}} Install and Load the 'tidymodels' Packages;
+  2020.
+
+\bibitem{xiao2018}
+Xiao J, Chen L, Yu Y, Zhang X, Chen J.
+\newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
+  {{Predictive Modeling}} of {{Microbial Community Data}}.
+\newblock Frontiers in Microbiology. 2018;9.
+\newblock doi:{10.3389/fmicb.2018.03112}.
+
+\bibitem{dong2020}
+Dong M, Li L, Chen M, Kusalik A, Xu W.
+\newblock Predictive Analysis Methods for Human Microbiome Data with
+  Application to {{Parkinson}}'s Disease.
+\newblock PLOS ONE. 2020;15(8):e0237779.
+\newblock doi:{10.1371/journal.pone.0237779}.
+
+\bibitem{pasolli2017}
+Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
+\newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
+\newblock Nature Methods. 2017;14(11):1023--1024.
+\newblock doi:{10.1038/nmeth.4468}.
+
+\bibitem{schiffer2019}
+Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
+\newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
+  through Bioconductor.
+\newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
+
+\bibitem{gonzalez2018}
+Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
+  Ackermann G, et~al.
+\newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
+\newblock Nature Methods. 2018;15(10):796--798.
+\newblock doi:{10.1038/s41592-018-0141-9}.
+
+\bibitem{consortium2012}
+Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
+  et~al.
+\newblock Structure, Function and Diversity of the Healthy Human Microbiome.
+\newblock Nature. 2012;486(7402):207--214.
+\newblock doi:{10.1038/nature11234}.
+
+\bibitem{thurnheer2016}
+Thurnheer T, Bostanci N, Belibasakis GN.
+\newblock Microbial Dynamics during Conversion from Supragingival to
+  Subgingival Biofilms in an in Vitro Model.
+\newblock Molecular Oral Microbiology. 2016;31(2):125--135.
+\newblock doi:{10.1111/omi.12108}.
+
+\bibitem{beghini2019}
+Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
+\newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
+  in the {{New York City Health}} and {{Nutrition Examination Study}}.
+\newblock Annals of Epidemiology. 2019;34:18--25.e3.
+\newblock doi:{10.1016/j.annepidem.2019.03.005}.
+
+\bibitem{matteocalgaro2020}
+Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
+\newblock Zenodo.
+
+\bibitem{nielsen2014}
+Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
+\newblock Identification and Assembly of Genomes and Genetic Elements in
+  Complex Metagenomic Samples without Using Reference Genomes.
+\newblock Nature Biotechnology. 2014;32(8):822--828.
+\newblock doi:{10.1038/nbt.2939}.
+
+\bibitem{gevers2014}
+Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
+  et~al.
+\newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
+  {{Disease}}.
+\newblock Cell Host \& Microbe. 2014;15(3):382--392.
+\newblock doi:{10.1016/j.chom.2014.02.005}.
+
+\bibitem{martin2020}
+Martin BD, Witten D, Willis AD.
+\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
+  Regression.
+\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
+\newblock doi:{10.1214/19-AOAS1283}.
+
+\bibitem{naim2012}
+Naim I, Gildea D.
+\newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
+  {{Unbalanced Mixing Coefficients}}.
+\newblock Proceedings of the 29th International Coference on International
+  Conference on Machine Learning. 2012; p.~8.
+\newblock doi:{10.5555/3042573.3042756}.
+
+\bibitem{wu2016}
+Wu C, Chen J, Kim J, Pan W.
+\newblock An Adaptive Association Test for Microbiome Data.
+\newblock Genome Medicine. 2016;8(1):56.
+\newblock doi:{10.1186/s13073-016-0302-3}.
+
+\bibitem{martin-fernandez2012}
+{Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
+  {Palarea-Albaladejo} J.
+\newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
+  {{Classical}} and Robust Approaches.
+\newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
+\newblock doi:{10.1016/j.csda.2012.02.012}.
+
+\end{thebibliography}
+
 \section*{Supporting information}
@@
 
-\section*{Acknowledgments}
-The authors thank Becky Lebeaux, Modupe Coker, Erika Dade, Jie Zhou, and Weston Viles for insightful comments and suggestions that greatly improved the paper.
-This research is supported by funding from the National Library of Medicine (NLM) (grants NLM R01LM012723 and NLM K01LM012426).
-None of the funding bodies had a role in the design of the study or the collection, analysis, and interpretation of data and in writing of the manuscript.
-
-\nolinenumbers
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-% \bibliography{tax_agg}{}
-
-\begin{thebibliography}{10}
-
-\bibitem{proctor2019}
-Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
-\newblock The {{Integrative Human Microbiome Project}}.
-\newblock Nature. 2019;569(7758):641--648.
-\newblock doi:{10.1038/s41586-019-1238-8}.
-
-\bibitem{sharma2019}
-Sharma S, Tripathi P.
-\newblock Gut Microbiome and Type 2 Diabetes: Where We Are and Where to Go?
-\newblock The Journal of Nutritional Biochemistry. 2019;63:101--108.
-\newblock doi:{10.1016/j.jnutbio.2018.10.003}.
-
-\bibitem{aoun2020}
-Aoun A, Darwish F, Hamod N.
-\newblock The {{Influence}} of the {{Gut Microbiome}} on {{Obesity}} in
-  {{Adults}} and the {{Role}} of {{Probiotics}}, {{Prebiotics}}, and
-  {{Synbiotics}} for {{Weight Loss}}.
-\newblock Preventive Nutrition and Food Science. 2020;25(2):113--123.
-\newblock doi:{10.3746/pnf.2020.25.2.113}.
-
-\bibitem{cho2012}
-Cho I, Blaser MJ.
-\newblock The Human Microbiome: At the Interface of Health and Disease.
-\newblock Nature Reviews Genetics. 2012;13(4):260--270.
-\newblock doi:{10.1038/nrg3182}.
-
-\bibitem{callahan2016}
-Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
-\newblock {{DADA2}}: {{High}}-Resolution Sample Inference from {{Illumina}}
-  Amplicon Data.
-\newblock Nature Methods. 2016;13(7):581--583.
-\newblock doi:{10.1038/nmeth.3869}.
-
-\bibitem{truong2015}
-Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
-\newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
-\newblock Nature Methods. 2015;12(10):902--903.
-\newblock doi:{10.1038/nmeth.3589}.
-
-\bibitem{li2019}
-Li H.
-\newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
-  {{Metagenomics}}.
-\newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
-  2019. p. 977--550.
-
-\bibitem{li2015}
-Li H.
-\newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
-  Compositional Data Analysis}}.
-\newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
-\newblock doi:{10.1146/annurev-statistics-010814-020351}.
-
-\bibitem{shi2016}
-Shi P, Zhang A, Li H.
-\newblock Regression Analysis for Microbiome Compositional Data.
-\newblock The Annals of Applied Statistics. 2016;10(2):1019--1040.
-\newblock doi:{10.1214/16-AOAS928}.
-
-\bibitem{sankaran2014}
-Sankaran K, Holmes S.
-\newblock {{structSSI}}: {{Simultaneous}} and {{Selective Inference}} for
-  {{Grouped}} or {{Hierarchically Structured Data}}.
-\newblock Journal of statistical software. 2014;59(13):1--21.
-\newblock doi:{10.18637/jss.v059.i13}.
-
-\bibitem{benjamini1995}
-Benjamini Y, Hochberg Y.
-\newblock Controlling the {{False Discovery Rate}}: {{A Practical}} and
-  {{Powerful Approach}} to {{Multiple Testing}}.
-\newblock Journal of the Royal Statistical Society Series B (Methodological).
-  1995;57(1):289--300.
-
-\bibitem{chen2018}
-Chen J, King E, Deek R, Wei Z, Yu Y, Grill D, et~al.
-\newblock An Omnibus Test for Differential Distribution Analysis of Microbiome
-  Sequencing Data.
-\newblock Bioinformatics. 2018;34(4):643--651.
-\newblock doi:{10.1093/bioinformatics/btx650}.
-
-\bibitem{weiss2017}
-Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
-\newblock Normalization and Microbial Differential Abundance Strategies Depend
-  upon Data Characteristics.
-\newblock Microbiome. 2017;5(1).
-\newblock doi:{10.1186/s40168-017-0237-y}.
-
-\bibitem{mcmurdie2014}
-McMurdie PJ, Holmes S.
-\newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
-  Inadmissible}}.
-\newblock PLOS Computational Biology. 2014;10(4):e1003531.
-\newblock doi:{10.1371/journal.pcbi.1003531}.
-
-\bibitem{love2014}
-Love MI, Huber W, Anders S.
-\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
-  Data with {{DESeq2}}.
-\newblock Genome Biology. 2014;15(12):550.
-\newblock doi:{10.1186/s13059-014-0550-8}.
-
-\bibitem{quinn2019}
-Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
-\newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
-\newblock GigaScience. 2019;8(giz107).
-\newblock doi:{10.1093/gigascience/giz107}.
-
-\bibitem{quinn2018b}
-Quinn TP, Erb I, Richardson MF, Crowley TM.
-\newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
-\newblock Bioinformatics. 2018;34(16):2870--2878.
-\newblock doi:{10.1093/bioinformatics/bty175}.
-
-\bibitem{gloor2017}
-Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
-\newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
-  Optional}}.
-\newblock Frontiers in Microbiology. 2017;8.
-\newblock doi:{10.3389/fmicb.2017.02224}.
-
-\bibitem{aitchison1999}
-Aitchison J.
-\newblock A {{Concise Guide}} to {{Compositional Data Analysis}}. 1999; p. 134.
-
-\bibitem{kurtz2015}
-Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
-\newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
-  Ecological Networks}}.
-\newblock PLOS Computational Biology. 2015;11(5):e1004226.
-\newblock doi:{10.1371/journal.pcbi.1004226}.
-
-\bibitem{kaul2017}
-Kaul A, Davidov O, Peddada SD.
-\newblock Structural Zeros in High-Dimensional Data with Applications to
-  Microbiome Studies.
-\newblock Biostatistics. 2017;18(3):422--433.
-\newblock doi:{10.1093/biostatistics/kxw053}.
-
-\bibitem{kaul2017a}
-Kaul A, Mandal S, Davidov O, Peddada SD.
-\newblock Analysis of {{Microbiome Data}} in the {{Presence}} of {{Excess
-  Zeros}}.
-\newblock Frontiers in Microbiology. 2017;8.
-\newblock doi:{10.3389/fmicb.2017.02114}.
-
-\bibitem{khatri2012}
-Khatri P, Sirota M, Butte AJ.
-\newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
-  {{Outstanding Challenges}}.
-\newblock PLOS Computational Biology. 2012;8(2):e1002375.
-\newblock doi:{10.1371/journal.pcbi.1002375}.
-
-\bibitem{goeman2007}
-Goeman JJ, B{\"u}hlmann P.
-\newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
-  Issues.
-\newblock Bioinformatics. 2007;23(8):980--987.
-\newblock doi:{10.1093/bioinformatics/btm051}.
-
-\bibitem{subramanian2005}
-Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
-\newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
-  Interpreting Genome-Wide Expression Profiles.
-\newblock Proceedings of the National Academy of Sciences.
-  2005;102(43):15545--15550.
-\newblock doi:{10.1073/pnas.0506580102}.
-
-\bibitem{ashburner2000}
-Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
-\newblock Gene {{Ontology}}: Tool for the Unification of Biology.
-\newblock Nature genetics. 2000;25(1):25--29.
-\newblock doi:{10.1038/75556}.
-
-\bibitem{hanzelmann2013}
-H{\"a}nzelmann S, Castelo R, Guinney J.
-\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
-  {{RNA}}-{{Seq}} Data.
-\newblock BMC Bioinformatics. 2013;14(1):7.
-\newblock doi:{10.1186/1471-2105-14-7}.
-
-\bibitem{frost2020a}
-Frost HR.
-\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
-  Method for Cell-Specific Gene Set Scoring.
-\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
-\newblock doi:{10.1093/nar/gkaa582}.
-
-\bibitem{mclaren2019}
-McLaren MR, Willis AD, Callahan BJ.
-\newblock Consistent and Correctable Bias in Metagenomic Sequencing
-  Experiments.
-\newblock eLife. 2019;8:e46923.
-\newblock doi:{10.7554/eLife.46923}.
-
-\bibitem{egozcue2005}
-Egozcue JJ, {Pawlowsky-Glahn} V.
-\newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
-  Analysis}}.
-\newblock Mathematical Geology. 2005;37(7):795--828.
-\newblock doi:{10.1007/s11004-005-7381-9}.
-
-\bibitem{chong2020}
-Chong J, Liu P, Zhou G, Xia J.
-\newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
-  Functional, and Meta-Analysis of Microbiome Data.
-\newblock Nature Protocols. 2020;15(3):799--821.
-\newblock doi:{10.1038/s41596-019-0264-1}.
-
-\bibitem{tian2005}
-Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
-\newblock Discovering Statistically Significant Pathways in Expression
-  Profiling Studies.
-\newblock Proceedings of the National Academy of Sciences.
-  2005;102(38):13544--13549.
-\newblock doi:{10.1073/pnas.0506577102}.
-
-\bibitem{egozcue2003}
-Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcelo-Vidal} C.
-\newblock Isometric {{Logratio Transformations}} for {{Compositional Data
-  Analysis}}.
-\newblock Mathematical Geology. 2003; p.~22.
-
-\bibitem{silverman2017}
-Silverman JD, Washburne AD, Mukherjee S, David LA.
-\newblock A Phylogenetic Transform Enhances Analysis of Compositional
-  Microbiota Data.
-\newblock eLife. 2017;6:e21887.
-\newblock doi:{10.7554/eLife.21887}.
-
-\bibitem{barbie2009}
-Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
-\newblock Systematic {{RNA}} Interference Reveals That Oncogenic
-  {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
-\newblock Nature. 2009;462(7269):108--112.
-\newblock doi:{10.1038/nature08460}.
-
-\bibitem{quast2013}
-Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
-\newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
-  Processing and Web-Based Tools.
-\newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
-\newblock doi:{10.1093/nar/gks1219}.
-
-\bibitem{delignette-muller2015}
-{Delignette-Muller} ML, Dutang C.
-\newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
-\newblock Journal of Statistical Software. 2015;64(4):1--34.
-
-\bibitem{benaglia2009}
-Benaglia T, Chauveau D, Hunter DR, Young D.
-\newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
-\newblock Journal of Statistical Software. 2009;32(6):1--29.
-
-\bibitem{washburne2017}
-Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL, Mukherjee S, et~al.
-\newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
-  Associations in Microbiome Datasets.
-\newblock PeerJ. 2017;5:e2969.
-\newblock doi:{10.7717/peerj.2969}.
-
-\bibitem{morton2017}
-Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
-  et~al.
-\newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
-\newblock mSystems. 2017;2(1).
-\newblock doi:{10.1128/mSystems.00162-16}.
-
-\bibitem{efron2004}
-Efron B.
-\newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
-\newblock Journal of the American Statistical Association.
-  2004;99(465):96--104.
-\newblock doi:{10.1198/016214504000000089}.
-
-\bibitem{wu2012}
-Wu D, Smyth GK.
-\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
-  Correlation.
-\newblock Nucleic Acids Research. 2012;40(17):e133.
-\newblock doi:{10.1093/nar/gks461}.
-
-\bibitem{calgaro2020}
-Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
-\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
-  and Metagenomics Applied to Microbiome Data.
-\newblock Genome Biology. 2020;21(1):191.
-\newblock doi:{10.1186/s13059-020-02104-1}.
-
-\bibitem{cario1997}
-Cario MC.
-\newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
-  Distributions}} and {{Correlation Matrix}}. 1997; p.~19.
-
-\bibitem{agresti1998}
-Agresti A, Coull BA.
-\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
-  Estimation}} of {{Binomial Proportions}}.
-\newblock The American Statistician. 1998;52(2):119--126.
-\newblock doi:{10.2307/2685469}.
-
-\bibitem{delong1988}
-DeLong ER, DeLong DM, {Clarke-Pearson} DL.
-\newblock Comparing the Areas under Two or More Correlated Receiver Operating
-  Characteristic Curves: A Nonparametric Approach.
-\newblock Biometrics. 1988;44(3):837--845.
-
-\bibitem{hawinkel2019}
-Hawinkel S, Mattiello F, Bijnens L, Thas O.
-\newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
-  Control the False Discovery Rate.
-\newblock Briefings in Bioinformatics. 2019;20(1):210--221.
-\newblock doi:{10.1093/bib/bbx104}.
-
-\bibitem{breiman2001}
-Breiman L.
-\newblock Random {{Forests}}.
-\newblock Machine Learning. 2001;45(1):5--32.
-\newblock doi:{10.1023/A:1010933404324}.
-
-\bibitem{kuhn2020}
-Kuhn M, Wickham H.
-\newblock Tidymodels: {{Easily}} Install and Load the 'tidymodels' Packages;
-  2020.
-
-\bibitem{xiao2018}
-Xiao J, Chen L, Yu Y, Zhang X, Chen J.
-\newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
-  {{Predictive Modeling}} of {{Microbial Community Data}}.
-\newblock Frontiers in Microbiology. 2018;9.
-\newblock doi:{10.3389/fmicb.2018.03112}.
-
-\bibitem{dong2020}
-Dong M, Li L, Chen M, Kusalik A, Xu W.
-\newblock Predictive Analysis Methods for Human Microbiome Data with
-  Application to {{Parkinson}}'s Disease.
-\newblock PLOS ONE. 2020;15(8):e0237779.
-\newblock doi:{10.1371/journal.pone.0237779}.
-
-\bibitem{pasolli2017}
-Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
-\newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
-\newblock Nature Methods. 2017;14(11):1023--1024.
-\newblock doi:{10.1038/nmeth.4468}.
-
-\bibitem{schiffer2019}
-Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
-\newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
-  through Bioconductor.
-\newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
-
-\bibitem{gonzalez2018}
-Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
-  Ackermann G, et~al.
-\newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
-\newblock Nature Methods. 2018;15(10):796--798.
-\newblock doi:{10.1038/s41592-018-0141-9}.
-
-\bibitem{consortium2012}
-Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
-  et~al.
-\newblock Structure, Function and Diversity of the Healthy Human Microbiome.
-\newblock Nature. 2012;486(7402):207--214.
-\newblock doi:{10.1038/nature11234}.
-
-\bibitem{thurnheer2016}
-Thurnheer T, Bostanci N, Belibasakis GN.
-\newblock Microbial Dynamics during Conversion from Supragingival to
-  Subgingival Biofilms in an in Vitro Model.
-\newblock Molecular Oral Microbiology. 2016;31(2):125--135.
-\newblock doi:{10.1111/omi.12108}.
-
-\bibitem{beghini2019}
-Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
-\newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
-  in the {{New York City Health}} and {{Nutrition Examination Study}}.
-\newblock Annals of Epidemiology. 2019;34:18--25.e3.
-\newblock doi:{10.1016/j.annepidem.2019.03.005}.
-
-\bibitem{matteocalgaro2020}
-Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
-\newblock Zenodo.
-
-\bibitem{nielsen2014}
-Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
-\newblock Identification and Assembly of Genomes and Genetic Elements in
-  Complex Metagenomic Samples without Using Reference Genomes.
-\newblock Nature Biotechnology. 2014;32(8):822--828.
-\newblock doi:{10.1038/nbt.2939}.
-
-\bibitem{gevers2014}
-Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
-  et~al.
-\newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
-  {{Disease}}.
-\newblock Cell Host \& Microbe. 2014;15(3):382--392.
-\newblock doi:{10.1016/j.chom.2014.02.005}.
-
-\bibitem{martin2020}
-Martin BD, Witten D, Willis AD.
-\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
-  Regression.
-\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
-\newblock doi:{10.1214/19-AOAS1283}.
-
-\bibitem{naim2012}
-Naim I, Gildea D.
-\newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
-  {{Unbalanced Mixing Coefficients}}.
-\newblock Proceedings of the 29th International Coference on International
-  Conference on Machine Learning. 2012; p.~8.
-\newblock doi:{10.5555/3042573.3042756}.
-
-\bibitem{wu2016}
-Wu C, Chen J, Kim J, Pan W.
-\newblock An Adaptive Association Test for Microbiome Data.
-\newblock Genome Medicine. 2016;8(1):56.
-\newblock doi:{10.1186/s13073-016-0302-3}.
-
-\bibitem{martin-fernandez2012}
-{Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
-  {Palarea-Albaladejo} J.
-\newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
-  {{Classical}} and Robust Approaches.
-\newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
-\newblock doi:{10.1016/j.csda.2012.02.012}.
-
-\end{thebibliography}
-
 \end{document}

### 2021-11-30 manuscript_revised.tex
Updates from Overleaf

--- 2021-08-20 manuscript_plos.tex
@@
 {\Large
-\textbf\newline{cILR: Competitive isometric log-ratio for taxonomic enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{CBEA: Competitive balances for taxonomic enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 \section*{Abstract}
-Research in human associated microbiomes often involves the analysis of taxonomic count tables generated via high-throughput sequencing.
+Research in human associated microbiomes primarily involves the analysis of taxonomic count tables generated via high-throughput sequencing.
 It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional.
@@
 Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis.
-Unfortunately, there is a lack of sophisticated set-based analysis methods specific to microbiome data, where current practice often employs abundance summation as a technique for aggregation.
+Unfortunately, there is a lack of interest in developing more sophisticated set-based analysis methods specific to microbiome data, where current practice often employs abundance summation as a technique for aggregation.
 This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
@@
 We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
-Herein we demonstrate using both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios.
-Additionally, it provides informative scores that can be inputs to downstream differential abundance and prediction tasks.
+Herein we demonstrate under both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios, while additionally providing informative scores that can be inputs to downstream differential abundance and prediction tasks.
 
@@
 \section*{Author summary}
-The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing.
+The study of human associated microbiomes rely on genomic surveys via high-throughput sequencing.
 However, microbiome data is sparse and high dimensional which prevents the application of standard statistical techniques.
@@
 Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability.
-Unfortunately, there is not a lot of research in developing new set-based tools for microbiome data specifically, given that compared to other 'omics data types microbiome data is strictly compositional.
+Unfortunately, there is not a lot of research in developing new set-based tools for microbiome data specifically, given that compared to other omics data types microbiome data is strictly compositional.
 We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis.
@@
 \section*{Introduction}
-The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host.
-Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
+The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their human host.
+Previous research has shown that changes in the composition of the microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
 To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
@@
 whole-genome shotgun sequencing) \cite{cho2012}.
-Raw sequencing data is then processed through a variety of bioinformatic pipelines \cite{callahan2016, truong2015}, yielding various data products, one of which are taxonomic tables which can be used to study associations between members of the microbiome and an exposure or outcome of interest.
-
-However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019,li2015}.
+Processing raw sequencing data using a variety of bioinformatic pipelines \cite{callahan2016,truong2015} yields taxonomic abundance tables that can be used in downstream statistical analyses to identify associations between an outcome/exposure and identified taxa.
+
+However, there exists unique challenges in the analysis of these data tables \cite{li2019,li2015}.
 First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present.

### 2021-11-30 cbea.tex
Renaming manuscript

--- 2021-09-07 manuscript_plos.tex
@@
 
-\begin{figure} [!ht]
+\begin{figure} [!h]
     \centering
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!h]
     \centering
@@
 \subsubsection*{Simulation studies}
-\begin{figure}[!ht]
+\begin{figure}[!h]
     \centering
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!h]
     \centering
-    \includegraphics[scale = 0.6]{figures/data_prediction_plot.png}
+    \includegraphics[width = \linewidth]{figures/data_prediction_plot.png}
     \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
@@
 
+\section*{Acknowledgments}
+The authors thank Becky Lebeaux, Modupe Coker, Erika Dade, Jie Zhou, and Weston Viles for insightful comments and suggestions that greatly improved the paper.
+
+\nolinenumbers
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here. See http://journals.plos.org/plosone/s/latex for 
+% step-by-step instructions.
+%
+
+% \bibliography{tax_agg}{}
+
+\begin{thebibliography}{10}
+
+\bibitem{proctor2019}
+Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
+\newblock The {{Integrative Human Microbiome Project}}.
+\newblock Nature. 2019;569(7758):641--648.
+\newblock doi:{10.1038/s41586-019-1238-8}.
+
+\bibitem{sharma2019}
+Sharma S, Tripathi P.
+\newblock Gut Microbiome and Type 2 Diabetes: Where We Are and Where to Go?
+\newblock The Journal of Nutritional Biochemistry. 2019;63:101--108.
+\newblock doi:{10.1016/j.jnutbio.2018.10.003}.
+
+\bibitem{aoun2020}
+Aoun A, Darwish F, Hamod N.
+\newblock The {{Influence}} of the {{Gut Microbiome}} on {{Obesity}} in
+  {{Adults}} and the {{Role}} of {{Probiotics}}, {{Prebiotics}}, and
+  {{Synbiotics}} for {{Weight Loss}}.
+\newblock Preventive Nutrition and Food Science. 2020;25(2):113--123.
+\newblock doi:{10.3746/pnf.2020.25.2.113}.
+
+\bibitem{cho2012}
+Cho I, Blaser MJ.
+\newblock The Human Microbiome: At the Interface of Health and Disease.
+\newblock Nature Reviews Genetics. 2012;13(4):260--270.
+\newblock doi:{10.1038/nrg3182}.
+
+\bibitem{callahan2016}
+Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
+\newblock {{DADA2}}: {{High}}-Resolution Sample Inference from {{Illumina}}
+  Amplicon Data.
+\newblock Nature Methods. 2016;13(7):581--583.
+\newblock doi:{10.1038/nmeth.3869}.
+
+\bibitem{truong2015}
+Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
+\newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
+\newblock Nature Methods. 2015;12(10):902--903.
+\newblock doi:{10.1038/nmeth.3589}.
+
+\bibitem{li2019}
+Li H.
+\newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
+  {{Metagenomics}}.
+\newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
+  2019. p. 977--550.
+
+\bibitem{li2015}
+Li H.
+\newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
+  Compositional Data Analysis}}.
+\newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
+\newblock doi:{10.1146/annurev-statistics-010814-020351}.
+
+\bibitem{shi2016}
+Shi P, Zhang A, Li H.
+\newblock Regression Analysis for Microbiome Compositional Data.
+\newblock The Annals of Applied Statistics. 2016;10(2):1019--1040.
+\newblock doi:{10.1214/16-AOAS928}.
+
+\bibitem{sankaran2014}
+Sankaran K, Holmes S.
+\newblock {{structSSI}}: {{Simultaneous}} and {{Selective Inference}} for
+  {{Grouped}} or {{Hierarchically Structured Data}}.
+\newblock Journal of statistical software. 2014;59(13):1--21.
+\newblock doi:{10.18637/jss.v059.i13}.
+
+\bibitem{benjamini1995}
+Benjamini Y, Hochberg Y.
+\newblock Controlling the {{False Discovery Rate}}: {{A Practical}} and
+  {{Powerful Approach}} to {{Multiple Testing}}.
+\newblock Journal of the Royal Statistical Society Series B (Methodological).
+  1995;57(1):289--300.
+
+\bibitem{chen2018}
+Chen J, King E, Deek R, Wei Z, Yu Y, Grill D, et~al.
+\newblock An Omnibus Test for Differential Distribution Analysis of Microbiome
+  Sequencing Data.
+\newblock Bioinformatics. 2018;34(4):643--651.
+\newblock doi:{10.1093/bioinformatics/btx650}.
+
+\bibitem{weiss2017}
+Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
+\newblock Normalization and Microbial Differential Abundance Strategies Depend
+  upon Data Characteristics.
+\newblock Microbiome. 2017;5(1).
+\newblock doi:{10.1186/s40168-017-0237-y}.
+
+\bibitem{mcmurdie2014}
+McMurdie PJ, Holmes S.
+\newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
+  Inadmissible}}.
+\newblock PLOS Computational Biology. 2014;10(4):e1003531.
+\newblock doi:{10.1371/journal.pcbi.1003531}.
+
+\bibitem{love2014}
+Love MI, Huber W, Anders S.
+\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
+  Data with {{DESeq2}}.
+\newblock Genome Biology. 2014;15(12):550.
+\newblock doi:{10.1186/s13059-014-0550-8}.
+
+\bibitem{quinn2019}
+Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
+\newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
+\newblock GigaScience. 2019;8(giz107).
+\newblock doi:{10.1093/gigascience/giz107}.
+
+\bibitem{quinn2018b}
+Quinn TP, Erb I, Richardson MF, Crowley TM.
+\newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
+\newblock Bioinformatics. 2018;34(16):2870--2878.
+\newblock doi:{10.1093/bioinformatics/bty175}.
+
+\bibitem{gloor2017}
+Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
+\newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
+  Optional}}.
+\newblock Frontiers in Microbiology. 2017;8.
+\newblock doi:{10.3389/fmicb.2017.02224}.
+
+\bibitem{aitchison1999}
+Aitchison J.
+\newblock A {{Concise Guide}} to {{Compositional Data Analysis}}. 1999; p. 134.
+
+\bibitem{kurtz2015}
+Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
+\newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
+  Ecological Networks}}.
+\newblock PLOS Computational Biology. 2015;11(5):e1004226.
+\newblock doi:{10.1371/journal.pcbi.1004226}.
+
+\bibitem{kaul2017}
+Kaul A, Davidov O, Peddada SD.
+\newblock Structural Zeros in High-Dimensional Data with Applications to
+  Microbiome Studies.
+\newblock Biostatistics. 2017;18(3):422--433.
+\newblock doi:{10.1093/biostatistics/kxw053}.
+
+\bibitem{kaul2017a}
+Kaul A, Mandal S, Davidov O, Peddada SD.
+\newblock Analysis of {{Microbiome Data}} in the {{Presence}} of {{Excess
+  Zeros}}.
+\newblock Frontiers in Microbiology. 2017;8.
+\newblock doi:{10.3389/fmicb.2017.02114}.
+
+\bibitem{khatri2012}
+Khatri P, Sirota M, Butte AJ.
+\newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
+  {{Outstanding Challenges}}.
+\newblock PLOS Computational Biology. 2012;8(2):e1002375.
+\newblock doi:{10.1371/journal.pcbi.1002375}.
+
+\bibitem{goeman2007}
+Goeman JJ, B{\"u}hlmann P.
+\newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
+  Issues.
+\newblock Bioinformatics. 2007;23(8):980--987.
+\newblock doi:{10.1093/bioinformatics/btm051}.
+
+\bibitem{subramanian2005}
+Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
+\newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
+  Interpreting Genome-Wide Expression Profiles.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(43):15545--15550.
+\newblock doi:{10.1073/pnas.0506580102}.
+
+\bibitem{ashburner2000}
+Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
+\newblock Gene {{Ontology}}: Tool for the Unification of Biology.
+\newblock Nature genetics. 2000;25(1):25--29.
+\newblock doi:{10.1038/75556}.
+
+\bibitem{hanzelmann2013}
+H{\"a}nzelmann S, Castelo R, Guinney J.
+\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
+  {{RNA}}-{{Seq}} Data.
+\newblock BMC Bioinformatics. 2013;14(1):7.
+\newblock doi:{10.1186/1471-2105-14-7}.
+
+\bibitem{frost2020a}
+Frost HR.
+\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
+  Method for Cell-Specific Gene Set Scoring.
+\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
+\newblock doi:{10.1093/nar/gkaa582}.
+
+\bibitem{mclaren2019}
+McLaren MR, Willis AD, Callahan BJ.
+\newblock Consistent and Correctable Bias in Metagenomic Sequencing
+  Experiments.
+\newblock eLife. 2019;8:e46923.
+\newblock doi:{10.7554/eLife.46923}.
+
+\bibitem{egozcue2005}
+Egozcue JJ, {Pawlowsky-Glahn} V.
+\newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2005;37(7):795--828.
+\newblock doi:{10.1007/s11004-005-7381-9}.
+
+\bibitem{chong2020}
+Chong J, Liu P, Zhou G, Xia J.
+\newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
+  Functional, and Meta-Analysis of Microbiome Data.
+\newblock Nature Protocols. 2020;15(3):799--821.
+\newblock doi:{10.1038/s41596-019-0264-1}.
+
+\bibitem{tian2005}
+Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
+\newblock Discovering Statistically Significant Pathways in Expression
+  Profiling Studies.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(38):13544--13549.
+\newblock doi:{10.1073/pnas.0506577102}.
+
+\bibitem{egozcue2003}
+Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcelo-Vidal} C.
+\newblock Isometric {{Logratio Transformations}} for {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2003; p.~22.
+
+\bibitem{silverman2017}
+Silverman JD, Washburne AD, Mukherjee S, David LA.
+\newblock A Phylogenetic Transform Enhances Analysis of Compositional
+  Microbiota Data.
+\newblock eLife. 2017;6:e21887.
+\newblock doi:{10.7554/eLife.21887}.
+
+\bibitem{barbie2009}
+Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
+\newblock Systematic {{RNA}} Interference Reveals That Oncogenic
+  {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
+\newblock Nature. 2009;462(7269):108--112.
+\newblock doi:{10.1038/nature08460}.
+
+\bibitem{quast2013}
+Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
+\newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
+  Processing and Web-Based Tools.
+\newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
+\newblock doi:{10.1093/nar/gks1219}.
+
+\bibitem{delignette-muller2015}
+{Delignette-Muller} ML, Dutang C.
+\newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
+\newblock Journal of Statistical Software. 2015;64(4):1--34.
+
+\bibitem{benaglia2009}
+Benaglia T, Chauveau D, Hunter DR, Young D.
+\newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
+\newblock Journal of Statistical Software. 2009;32(6):1--29.
+
+\bibitem{washburne2017}
+Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL, Mukherjee S, et~al.
+\newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
+  Associations in Microbiome Datasets.
+\newblock PeerJ. 2017;5:e2969.
+\newblock doi:{10.7717/peerj.2969}.
+
+\bibitem{morton2017}
+Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
+  et~al.
+\newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
+\newblock mSystems. 2017;2(1).
+\newblock doi:{10.1128/mSystems.00162-16}.
+
+\bibitem{efron2004}
+Efron B.
+\newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
+\newblock Journal of the American Statistical Association.
+  2004;99(465):96--104.
+\newblock doi:{10.1198/016214504000000089}.
+
+\bibitem{wu2012}
+Wu D, Smyth GK.
+\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
+  Correlation.
+\newblock Nucleic Acids Research. 2012;40(17):e133.
+\newblock doi:{10.1093/nar/gks461}.
+
+\bibitem{calgaro2020}
+Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
+\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
+  and Metagenomics Applied to Microbiome Data.
+\newblock Genome Biology. 2020;21(1):191.
+\newblock doi:{10.1186/s13059-020-02104-1}.
+
+\bibitem{cario1997}
+Cario MC.
+\newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
+  Distributions}} and {{Correlation Matrix}}. 1997; p.~19.
+
+\bibitem{agresti1998}
+Agresti A, Coull BA.
+\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
+  Estimation}} of {{Binomial Proportions}}.
+\newblock The American Statistician. 1998;52(2):119--126.
+\newblock doi:{10.2307/2685469}.
+
+\bibitem{delong1988}
+DeLong ER, DeLong DM, {Clarke-Pearson} DL.
+\newblock Comparing the Areas under Two or More Correlated Receiver Operating
+  Characteristic Curves: A Nonparametric Approach.
+\newblock Biometrics. 1988;44(3):837--845.
+
+\bibitem{hawinkel2019}
+Hawinkel S, Mattiello F, Bijnens L, Thas O.
+\newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
+  Control the False Discovery Rate.
+\newblock Briefings in Bioinformatics. 2019;20(1):210--221.
+\newblock doi:{10.1093/bib/bbx104}.
+
+\bibitem{breiman2001}
+Breiman L.
+\newblock Random {{Forests}}.
+\newblock Machine Learning. 2001;45(1):5--32.
+\newblock doi:{10.1023/A:1010933404324}.
+
+\bibitem{kuhn2020}
+Kuhn M, Wickham H.
+\newblock Tidymodels: {{Easily}} Install and Load the 'tidymodels' Packages;
+  2020.
+
+\bibitem{xiao2018}
+Xiao J, Chen L, Yu Y, Zhang X, Chen J.
+\newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
+  {{Predictive Modeling}} of {{Microbial Community Data}}.
+\newblock Frontiers in Microbiology. 2018;9.
+\newblock doi:{10.3389/fmicb.2018.03112}.
+
+\bibitem{dong2020}
+Dong M, Li L, Chen M, Kusalik A, Xu W.
+\newblock Predictive Analysis Methods for Human Microbiome Data with
+  Application to {{Parkinson}}'s Disease.
+\newblock PLOS ONE. 2020;15(8):e0237779.
+\newblock doi:{10.1371/journal.pone.0237779}.
+
+\bibitem{pasolli2017}
+Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
+\newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
+\newblock Nature Methods. 2017;14(11):1023--1024.
+\newblock doi:{10.1038/nmeth.4468}.
+
+\bibitem{schiffer2019}
+Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
+\newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
+  through Bioconductor.
+\newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
+
+\bibitem{gonzalez2018}
+Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
+  Ackermann G, et~al.
+\newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
+\newblock Nature Methods. 2018;15(10):796--798.
+\newblock doi:{10.1038/s41592-018-0141-9}.
+
+\bibitem{consortium2012}
+Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
+  et~al.
+\newblock Structure, Function and Diversity of the Healthy Human Microbiome.
+\newblock Nature. 2012;486(7402):207--214.
+\newblock doi:{10.1038/nature11234}.
+
+\bibitem{thurnheer2016}
+Thurnheer T, Bostanci N, Belibasakis GN.
+\newblock Microbial Dynamics during Conversion from Supragingival to
+  Subgingival Biofilms in an in Vitro Model.
+\newblock Molecular Oral Microbiology. 2016;31(2):125--135.
+\newblock doi:{10.1111/omi.12108}.
+
+\bibitem{beghini2019}
+Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
+\newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
+  in the {{New York City Health}} and {{Nutrition Examination Study}}.
+\newblock Annals of Epidemiology. 2019;34:18--25.e3.
+\newblock doi:{10.1016/j.annepidem.2019.03.005}.
+
+\bibitem{matteocalgaro2020}
+Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
+\newblock Zenodo.
+
+\bibitem{nielsen2014}
+Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
+\newblock Identification and Assembly of Genomes and Genetic Elements in
+  Complex Metagenomic Samples without Using Reference Genomes.
+\newblock Nature Biotechnology. 2014;32(8):822--828.
+\newblock doi:{10.1038/nbt.2939}.
+
+\bibitem{gevers2014}
+Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
+  et~al.
+\newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
+  {{Disease}}.
+\newblock Cell Host \& Microbe. 2014;15(3):382--392.
+\newblock doi:{10.1016/j.chom.2014.02.005}.
+
+\bibitem{martin2020}
+Martin BD, Witten D, Willis AD.
+\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
+  Regression.
+\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
+\newblock doi:{10.1214/19-AOAS1283}.
+
+\bibitem{naim2012}
+Naim I, Gildea D.
+\newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
+  {{Unbalanced Mixing Coefficients}}.
+\newblock Proceedings of the 29th International Coference on International
+  Conference on Machine Learning. 2012; p.~8.
+\newblock doi:{10.5555/3042573.3042756}.
+
+\bibitem{wu2016}
+Wu C, Chen J, Kim J, Pan W.
+\newblock An Adaptive Association Test for Microbiome Data.
+\newblock Genome Medicine. 2016;8(1):56.
+\newblock doi:{10.1186/s13073-016-0302-3}.
+
+\bibitem{martin-fernandez2012}
+{Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
+  {Palarea-Albaladejo} J.
+\newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
+  {{Classical}} and Robust Approaches.
+\newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
+\newblock doi:{10.1016/j.csda.2012.02.012}.
+
+\end{thebibliography}
+
 \section*{Supporting information}

### 2021-11-30 cbea_revised.tex

--- 2021-11-30 cbea.tex
@@
 \usepackage[parfill]{parskip}
+\usepackage{float}
 
@@
 Research in human associated microbiomes often involves the analysis of taxonomic count tables generated via high-throughput sequencing.
-It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional.
+It is difficult to apply statistical tools as the data is high-dimensional, sparse, and compositional.
 An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets.
@@
 This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
-Here, we attempt to fill this gap with a new single sample taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature.
-Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
+Here, we attempt to fill this gap with a new single sample taxon enrichment method with using a novel log-ratio formulation based on the competitive null hypothesis commonly used in the enrichment analysis literature.
+Our approach, titled competitive balances for taxonomic enrichment analysis (CBEA), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
 We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
-Herein we demonstrate using both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios.
-Additionally, it provides informative scores that can be inputs to downstream differential abundance and prediction tasks.
+Herein we demonstrate using both real data applications and simulations that CBEA controls for type I error even under high sparsity and high inter-taxa correlation scenarios.
+Additionally, it provides informative scores that can be inputs to downstream analyses such as prediction tasks.
 
@@
 Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability.
-Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given that compared to other 'omics data types microbiome taxonomic data is strictly compositional.
-We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis.
-Our scores can be used for statistical inference, and as inputs to other downstream analyses such as differential abundance and prediction models.
+Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given that compared to other 'omics data types microbiome taxonomic data is compositional.
+We developed a new tool to generate taxon set enrichment scores at the sample level through a novel log-ratio formulation based on the competitive null hypothesis.
+Our scores can be used for statistical inference, and as inputs to other downstream analyses such as prediction models.
 We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.
@@
 
+Second, the total number of reads varies across samples due to limitations of the sequencing instrument \cite{quinn2018b}, where normalization approaches has to be applied to allow for comparison of feature abundances across samples.
+However, microbiome data sets do not follow certain assumptions that enable the usage of methods from similar fields (such as RNA-seq) \cite{quinn2019}.
+For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes that do not differ in expression levels across samples.
+As such, practitioners often rely on total sum normalization to transform count data into relative proportions constrained to sum to one \cite{weiss2017}, which based on empirical studies showed better performance in analysis tasks \cite{mcknight2019}.
+This normalization schema imposes spurious negative correlation between the two
 Second, the number of reads obtained is constrained by the sequencing instrument at an arbitrary limit, and is inconsistent across samples, resulting in a variable number of total read counts per sample.
@@
 We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a multiplicative aggregation method that addresses the downsides of the naive summation-based method presented above \cite{mclaren2019, silverman2017}.
-The resulting method, titled competitive isometric log-ratio (cILR), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing.
+The resulting method, titled competitive isometric log-ratio (CBEA), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing.
 These scores can then act as inputs to differential abundance and predictive modeling tasks downstream.
 
-In the following sections, we provide the formulation of cILR and discuss some statistical properties.
-We illustrate significance testing at the sample level using cILR and evaluate type I error and power under different simulation scenarios and real data applications.
-We assess the informativeness of cILR generated scores, and evaluate how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis.
-We compare the performance of cILR in these respective tasks against standard microbiome taxonomic data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods.
+In the following sections, we provide the formulation of CBEA and discuss some statistical properties.
+We illustrate significance testing at the sample level using CBEA and evaluate type I error and power under different simulation scenarios and real data applications.
+We assess the informativeness of CBEA generated scores, and evaluate how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis.
+We compare the performance of CBEA in these respective tasks against standard microbiome taxonomic data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods.
 An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).
@@
 \section*{Materials and Methods} \label{methods}
-\subsection*{Competitive Isometric Log-ratio (cILR)}
-The cILR method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}.
-Details on the computational implementation of cILR can be found in the supplemental materials.
-The cILR method takes two inputs:
+\subsection*{Competitive Isometric Log-ratio (CBEA)}
+The CBEA method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}.
+Details on the computational implementation of CBEA can be found in the supplemental materials.
+The CBEA method takes two inputs:
 \begin{itemize}
@@
 \end{itemize}
-The cILR method generates one output:  
+The CBEA method generates one output:  
 \begin{itemize} 
@@
 \begin{enumerate}
-    \item \textbf{Compute the cILR statistic}: Let $\mathbf{M}$ be a $n$ by $m$ matrix of cILR scores. Let $\mathbf{M}_{i,k}$ be cILR scores for set $k$ of sample $i$:   
+    \item \textbf{Compute the CBEA statistic}: Let $\mathbf{M}$ be a $n$ by $m$ matrix of CBEA scores. Let $\mathbf{M}_{i,k}$ be CBEA scores for set $k$ of sample $i$:   
     \begin{equation}\label{main_eq}
@@
     \item \textbf{Estimate the empirical null distribution} Enrichment scores represent the test statistic for the $Q_1$ null hypothesis $H_o$ that relative abundances in $\mathbf{X}$ of members of set $k$ are not enriched compared to those not in set $k$.
-Since the distribution of cILR under the null vary depending on data characteristics (Fig~\ref{fig:1}), an empirical null distribution will be estimated from data.
+Since the distribution of CBEA under the null vary depending on data characteristics (Fig~\ref{fig:1}), an empirical null distribution will be estimated from data.
     \begin{itemize}
-        \item \textbf{Compute the cILR statistic on permuted and un-permuted $\mathbf{X}$}.
-Let $\mathbf{X}_{perm}$ be the column permuted relative abundance matrix, and $\mathbf{M}_{perm}$ be the corresponding cILR scores generated from $\mathbf{X}_{perm}$.
-Similarly, we have $\mathbf{M}_{unperm}$ be cILR scores generated from $\mathbf{X}$.
+        \item \textbf{Compute the CBEA statistic on permuted and un-permuted $\mathbf{X}$}.
+Let $\mathbf{X}_{perm}$ be the column permuted relative abundance matrix, and $\mathbf{M}_{perm}$ be the corresponding CBEA scores generated from $\mathbf{X}_{perm}$.
+Similarly, we have $\mathbf{M}_{unperm}$ be CBEA scores generated from $\mathbf{X}$.
         \item \textbf{Estimate correlation-adjusted empirical distribution for each set}.
@@
     \end{itemize}
-    \item \textbf{Calculate finalized cILR scores with respect to the empirical null}.
+    \item \textbf{Calculate finalized CBEA scores with respect to the empirical null}.
 Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution.
@@
 
-\subsection*{Properties of cILR}
-\subsubsection*{cILR and the Isometric Log Ratio Transformation}
-The cILR statistic is a special instance of the isometric log-ratio transformation (ILR) \cite{egozcue2003}.
+\subsection*{Properties of CBEA}
+\subsubsection*{CBEA and the Isometric Log Ratio Transformation}
+The CBEA statistic is a special instance of the isometric log-ratio transformation (ILR) \cite{egozcue2003}.
 The standard ILR is a transformation method to address the negative correlation bias inherent in compositional data by providing an isometry between the $D$-dimensional simplex $\mathbb{S}^D$ and coordinates in the $D-1$ real space $\mathbb{R}^{D-1}$ \cite{egozcue2003,washburne2017}.
@@
 They are usually termed ``compositional balances", and have been leveraged in many recent approaches in variable transformation \cite{washburne2017,silverman2017,morton2017}.
-The cILR formulation in Eq~\eqref{main_eq} is a special case of Eq~\eqref{ilr_standard} defined on a node that splits the taxa into two disjoint sets, one representing the set of interest, the other representing the remaining taxa.
-As such, the cILR transformation inherits the properties of the ILR as a log-ratio method applicable to compositional data sets.
-However, unlike the ILR and its variants \cite{silverman2017, morton2017, washburne2017}, the axes defined by each cILR set are not orthogonal (since the balances are mutually exclusive between sets and do not belong in the same SBP).
-Hence, a correlation can exist between cILR aggregated variables.
-
-\subsubsection*{Statistical Properties of cILR}
-We can perform significance testing on the cILR statistic which corresponds to the null hypothesis that the center of the subcomposition defined by the set is equal to the center of the subcomposition defined by the complement of the set.
+The CBEA formulation in Eq~\eqref{main_eq} is a special case of Eq~\eqref{ilr_standard} defined on a node that splits the taxa into two disjoint sets, one representing the set of interest, the other representing the remaining taxa.
+As such, the CBEA transformation inherits the properties of the ILR as a log-ratio method applicable to compositional data sets.
+However, unlike the ILR and its variants \cite{silverman2017, morton2017, washburne2017}, the axes defined by each CBEA set are not orthogonal (since the balances are mutually exclusive between sets and do not belong in the same SBP).
+Hence, a correlation can exist between CBEA aggregated variables.
+
+\subsubsection*{Statistical Properties of CBEA}
+We can perform significance testing on the CBEA statistic which corresponds to the null hypothesis that the center of the subcomposition defined by the set is equal to the center of the subcomposition defined by the complement of the set.
 This is equivalent to the $Q_1$ competitive null hypothesis in the gene set testing literature \cite{tian2005} where the enrichment of a gene set is defined with respect to genes outside the set.
 
-We can apply prior usage of the ILR statistic in hypothesis testing to cILR by assuming that the null distribution of cILR follows a standard normal distribution \cite{egozcue2005}.
-However, when applying cILR for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses.
+We can apply prior usage of the ILR statistic in hypothesis testing to CBEA by assuming that the null distribution of CBEA follows a standard normal distribution \cite{egozcue2005}.
+However, when applying CBEA for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses.
 Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies.
-As such, to perform significance testing using cILR, we also estimated the null distribution from observed raw cILR variables.
+As such, to perform significance testing using CBEA, we also estimated the null distribution from observed raw CBEA variables.
 
 This assumption is also supported by preliminary simulation studies (detailed below).
-In panel A of Fig~\ref{fig:1}, we simulated microbiome taxonomic count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
+In panel A of Fig~\ref{fig:1}, we simulated microbiome taxonomic count data under the global null across different data features and compute raw CBEA scores and compute kurtosis and skewness.
 It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation.
@@
     \includegraphics[width=\linewidth]{figures/kurtosis_skewness_gof.png}
-    \caption{{\bf Properties of the null distribution of cILR under the global null simulations}.
-Panel \textbf{(B)} presents kurtosis and skewness of cILR scores while panel \textbf{(A)} presents the goodness of fit (as Kolmogorov-Smirnov D statistic) for mixture normal and normal distributions.
+    \caption{{\bf Properties of the null distribution of CBEA under the global null simulations}.
+Panel \textbf{(B)} presents kurtosis and skewness of CBEA scores while panel \textbf{(A)} presents the goodness of fit (as Kolmogorov-Smirnov D statistic) for mixture normal and normal distributions.
 Panel \textbf{(C)} is a density plot of the shape of the null distribution.
@@
 To address this issue, we also evaluated a mixture distribution of two normal components.
-Panel B of Fig~\ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
+Panel B of Fig~\ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on CBEA scores in simulation scenarios under the global null.
 We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.
 
-We performed our empirical null estimation by fitting our distribution of choice and computing relevant parameters on raw cILR scores on taxa-permuted data (equivalent to gene permutation in the gene expression literature).
+We performed our empirical null estimation by fitting our distribution of choice and computing relevant parameters on raw CBEA scores on taxa-permuted data (equivalent to gene permutation in the gene expression literature).
 As such, the null distribution is characterized by scores computed on sets of equal size with randomly drawn taxa.
@@
 We account for this by combining the mean estimate from permuted data with the variance estimate from unpermuted data, where the inter-taxa correlation structure remains undisturbed.
-However, this procedure assumes that the variance of cILR is equal under both the null and alternate hypotheses.
+However, this procedure assumes that the variance of CBEA is equal under both the null and alternate hypotheses.
 
 \subsection*{Evaluation}
-All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/cILR\_analysis}{qpmnguyen/cILR\_analysis}).
+All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).
 
 \subsubsection*{Parametric Simulations}  
-To address the performance of cILR for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
+To address the performance of CBEA for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
 Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
@@
 
-\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of cILR in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}.
+\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of CBEA in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}.
 We utilized this data set following the approach outlined in Calagaro et al.
@@
 
-\noindent \textbf{Differential Abundance Analsysis}: To assess type I error using cILR scores in differential abundance analysis, we utilized the 16S rRNA gene sequencing of stool samples from the Human Microbiome Project \cite{consortium2012, proctor2019}.
+\noindent \textbf{Differential Abundance Analsysis}: To assess type I error using CBEA scores in differential abundance analysis, we utilized the 16S rRNA gene sequencing of stool samples from the Human Microbiome Project \cite{consortium2012, proctor2019}.
 Here, we randomly assign samples a label of case or control, and repeated this process 500 times, assessing all candidate methods at each iteration.
@@
 
-\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank sum test.
+\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the CBEA method with a naive Wilcoxon rank sum test.
 We added a pseudocount of 1 to all values.
 This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample.
-For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test.
+For classification performance, we compared CBEA methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test.
 All three approaches were applied directly on count data (after pseudocount).
@@
 
-\noindent \textbf{Differential Abundance}: Since cILR are sample-level enrichment scores, we performed differential abundance by using a Wilcoxon Rank Sum test and Welch's t-test across case/control status on cILR generated scores.
+\noindent \textbf{Differential Abundance}: Since CBEA are sample-level enrichment scores, we performed differential abundance by using a Wilcoxon Rank Sum test and Welch's t-test across case/control status on CBEA generated scores.
 We added a pseudocount of 1 to all values.
@@
 
-\noindent \textbf{Disease Prediction}: We fit random forest on cILR scores, as well as ssGSEA \cite{hanzelmann2013} and GSVA \cite{barbie2009} similar to single sample enrichment section.
+\noindent \textbf{Disease Prediction}: We fit random forest on CBEA scores, as well as ssGSEA \cite{hanzelmann2013} and GSVA \cite{barbie2009} similar to single sample enrichment section.
 We added a pseudocount of 1 to all values.
-Additionally, we also compared performance using cILR against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.
+Additionally, we also compared performance using CBEA against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.
 
@@
 \subsection*{Enrichment testing at the sample level}
-cILR provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{methods}.
-Here, we present empirical results for this application of cILR assessing type I error, power, and classification capacity.
+CBEA provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{methods}.
+Here, we present empirical results for this application of CBEA assessing type I error, power, and classification capacity.
 
@@
 Panel A and B in Fig~\ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
-We benchmarked the results of the cILR method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
+We benchmarked the results of the CBEA method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
 All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions.
-However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted cILR variants show high levels of inflated type I error, where Wilcoxon test performed the worst.
-On the other hand, adjusted cILR methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.
+However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted CBEA variants show high levels of inflated type I error, where Wilcoxon test performed the worst.
+On the other hand, adjusted CBEA methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.
 
@@
     \includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
-    \caption{Sample-level inference with cILR under parametric simulations (\textbf{(A)} and \textbf{B}), and real data analysis (\textbf{(C)} and \textbf{(D)}).
+    \caption{Sample-level inference with CBEA under parametric simulations (\textbf{(A)} and \textbf{B}), and real data analysis (\textbf{(C)} and \textbf{(D)}).
 In simulation analyses, panel \textbf{(A)} shows type I error rate, and panel \textbf{(B)} shows power for single sample enrichment test for a specified set and was compared against a Wilcoxon rank sum test at $\alpha$ of 0.05.
@@
 Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach.
-Adjusted cILR demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate.
-However, this resulted in lower power for adjusted cILR methods.}
+Adjusted CBEA demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate.
+However, this resulted in lower power for adjusted CBEA methods.}
     \label{fig:2}
@@
 However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B.
-In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un-adjusted ones.
+In situations where there is no inter-taxa correlation, CBEA still outperforms the wilcoxon rank sum test, however adjusted versions of CBEA did not perform as well as un-adjusted ones.
 However, in higher correlation scenarios, the difference in power is much more dramatic.
-At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted cILR was only performing at 50\% power, while unadjusted cILR and wilcoxon rank sum test were able to reach 80\%.
-These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of cILR.
-Most importantly, cILR demonstrate higher power in all scenarios where type I error is properly controlled.
-
-To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set.
-This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated.
+At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted CBEA was only performing at 50\% power, while unadjusted CBEA and wilcoxon rank sum test were able to reach 80\%.
+These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of CBEA.
+Most importantly, CBEA demonstrate higher power in all scenarios where type I error is properly controlled.
+
+To further assess the utility of CBEA in classifying samples with enriched sets, we generated AUC scores for different CBEA scores using true labels of whether a sample has an inflated set.
+This analysis, therefore, assessed the relative ranking of samples using CBEA scores whereby high scores should correspond to samples that are known to be inflated.
 Fig~\ref{fig:3} presents this result.
-We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
-Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
+We compared different variants of CBEA against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
+Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), CBEA scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
 In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
@@
 Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations.
-As expected in Fig~\ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods.
-Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05.
-Power analysis (Fig~\ref{fig:2}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
+As expected in Fig~\ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted CBEA methods.
+Conversely, adjusted CBEA controls for false positives adequately at the correct $\alpha$ level of 0.05.
+Power analysis (Fig~\ref{fig:2}D) showed similar patterns, where unadjusted CBEA methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
 
@@
     \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
-    \caption{Classification performance via AUC of cILR, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}.
+    \caption{Classification performance via AUC of CBEA, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}.
 Performance scores measure whether scores can highly rank samples that are known to have inflated abundance.
@@
 \subsection*{Differential abundance analysis}
-cILR generates sample-specific scores representing the degree of enrichment of a pre-defined set.
+CBEA generates sample-specific scores representing the degree of enrichment of a pre-defined set.
 As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test).
-We compared the performance of this approach with cILR and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
+We compared the performance of this approach with CBEA and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
 
@@
     \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
-For cILR methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by cILR (across different output types and distributional assumptions).
-cILR associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
+For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions).
+CBEA associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
     \label{fig:4}
@@
 All methods control for type I error well across both sparsity and correlation levels, where the estimated rate was consistently around the 0.05 pre-defined threshold.
-Results were similar across all evaluated methods, although in some instances, for example in medium correlation setting ($\rho = 0.2$), the unadjusted cILR resulted in higher type I error, regardless of difference in means test and distribution of choice.
+Results were similar across all evaluated methods, although in some instances, for example in medium correlation setting ($\rho = 0.2$), the unadjusted CBEA resulted in higher type I error, regardless of difference in means test and distribution of choice.
 
 The difference between the methods is more noticeable when evaluating power.
-All cILR associated variants showed much higher power even when the effect size is limited (fold change is 1.5), and there is a noticeable gap in performance between cILR and both DESeq2 and corncob.
+All CBEA associated variants showed much higher power even when the effect size is limited (fold change is 1.5), and there is a noticeable gap in performance between CBEA and both DESeq2 and corncob.
 Surprisingly, this effect is consistent across correlation levels and sparsity, even though we expectedly see performance in power drop as a function of sparsity especially in low effect size settings.
@@
 
-We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of cILR were controlling for type I error at the defined $\alpha$ threshold of 0.05.
+We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of CBEA were controlling for type I error at the defined $\alpha$ threshold of 0.05.
 This is surprising given the consistency of preserving type I error for both corncob and DESeq2 in all simulation evaluations.
@@
     \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
-    \caption{Differential abundance analysis using corncob, DESeq2 and cILR with either Wilcoxon rank sum test or Welch's t-test.
+    \caption{Differential abundance analysis using corncob, DESeq2 and CBEA with either Wilcoxon rank sum test or Welch's t-test.
 Panel \textbf{(A)} shows type I error results as the proportion of significant genera after 500 iterations where case/control status was assigned randomly to each sample.
@@
 Type I error evaluation used stool samples while the true positive rate evaluation used samples from the gingival site.
-Results showed that cILR associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2}
+Results showed that CBEA associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2}
     \label{fig:5}
@@
 \subsection*{Disease Prediction}   
-Since cILR can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models.
-In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating cILR, ssGSEA, and GSVA).
+Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models.
+In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating CBEA, ssGSEA, and GSVA).
 Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.
@@
 As expected, performance across all assessed methods increased with a higher signal-to-noise ratio.
-Both CLR and cILR approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks.
-This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while cILR and CLR did not experience a similar level of impact.
+Both CLR and CBEA approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks.
+This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while CBEA and CLR did not experience a similar level of impact.
 As such, performance gap widens with increasing correlation and sparsity.
@@
 
-In this analysis, cILR unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
+In this analysis, CBEA unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
 This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation.
-However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and cILR/CLR.
+However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and CBEA/CLR.
 Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity.
-Additionally, we did not observe any performance difference between the different variations of cILR.
+Additionally, we did not observe any performance difference between the different variations of CBEA.
 
@@
     \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
-    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
+    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
 Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure.
 Panel \textbf{(B)} presents performance on a classification task with AUC as the evaluation measure.
-cILR approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
+CBEA approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
     \label{fig:6}
@@
 %    \includegraphics[width=\textwidth]{figures/sim_pred_rsq.png}
-%    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on cILR,     %ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data %across sparsity levels, correlation, and signal-to-noise ratio.
+%    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on CBEA,     %ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data %across sparsity levels, correlation, and signal-to-noise ratio.
 Panel \textbf{(A)} and \textbf{(B)} %represent results across different levels of model saturation (proportion of sets associated with the %outcome).
-cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
+CBEA approaches outperformed GSVA and ssGSEA but not standard CLR.}
 %    \label{fig:sim_pred_rsq}
@@
 \subsubsection*{Real data evaluations}
-In addition to parametric simulations, we also assessed the performance of using cILR scores in predictive models with real data sets.
+In addition to parametric simulations, we also assessed the performance of using CBEA scores in predictive models with real data sets.
 Fig~\ref{fig:7} presents results for two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
@@
     \includegraphics[width = \linewidth]{figures/data_prediction_plot.png}
-    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
+    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
 Data sets used span both 16S rRNA gene sequencing (Gevers et al.
@@
 \cite{nielsen2014}).
-cILR performs better than GSVA and ssGSEA but not as well as CLR, with the exception of the whole genome sequencing data set.}
+CBEA performs better than GSVA and ssGSEA but not as well as CLR, with the exception of the whole genome sequencing data set.}
     \label{fig:7}
@@
 
-Similar to simulation experiments, we also fitted a naive random forest model using CILR, ssGSEA, GSVA, or CLR transformed variables as inputs, and use AUC as the performance criteria.
-Results also replicated that of the simulations, where across both data sets cILR and CLR methods provide much better performance than both GSVA or ssGSEA.
-Interestingly, the cILR approach performed better than CLR in the whole genome data set but did not perform as well in the 16S rRNA gene sequencing data set.
-However, these results indicate that cILR generated scores are informative, providing competitive performance when acting as inputs to disease predictive models.
+Similar to simulation experiments, we also fitted a naive random forest model using CBEA, ssGSEA, GSVA, or CLR transformed variables as inputs, and use AUC as the performance criteria.
+Results also replicated that of the simulations, where across both data sets CBEA and CLR methods provide much better performance than both GSVA or ssGSEA.
+Interestingly, the CBEA approach performed better than CLR in the whole genome data set but did not perform as well in the 16S rRNA gene sequencing data set.
+However, these results indicate that CBEA generated scores are informative, providing competitive performance when acting as inputs to disease predictive models.
 Most importantly, performance values are consistent across both simulated and real data sets.
 
-These results demonstrate that cILR generated scores are informative features in disease prediction tasks.
-Simulation results indicate that cILR methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach.
-Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
+These results demonstrate that CBEA generated scores are informative features in disease prediction tasks.
+Simulation results indicate that CBEA methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach.
+Interestingly, however, CBEA methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
 
@@
 
-\subsection*{Inference with cILR}
-The formulation of cILR as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}.
-This allows for conducting inference with cILR even at the sample-level.
-We assessed the usage of cILR in this type of analysis by evaluating type I error and power across both simulation studies and real data applications.
-Most importantly, we demonstrated that our adjusted cILR approach was able to address the issue of variance inflation due to correlation \cite{wu2012} by controlling for type I error at the appropriate $\alpha$ level across different levels of simulated inter-taxa correlation (Fig~\ref{fig:2}) while conversely unadjusted cILR and the naive Wilcoxon rank sum test showed much higher rates of error.
+\subsection*{Inference with CBEA}
+The formulation of CBEA as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}.
+This allows for conducting inference with CBEA even at the sample-level.
+We assessed the usage of CBEA in this type of analysis by evaluating type I error and power across both simulation studies and real data applications.
+Most importantly, we demonstrated that our adjusted CBEA approach was able to address the issue of variance inflation due to correlation \cite{wu2012} by controlling for type I error at the appropriate $\alpha$ level across different levels of simulated inter-taxa correlation (Fig~\ref{fig:2}) while conversely unadjusted CBEA and the naive Wilcoxon rank sum test showed much higher rates of error.
 This is further encouraged in real data analysis where the false discovery rate was around $0.05$ when a collection of true null and true alternate hypotheses were tested.
@@
 The conservativeness of the test attenuates with higher sparsity and correlation, where power was not approaching even 50\% even at the highest effect sizes.
-However, when the degree of such data features are reasonable, cILR will still be able to detect a reasonable proportion of samples with inflated counts.
-
-We also observed that choosing different distributional forms did not alter performance values for cILR.
-This runs contrary to our comparison analysis in Fig~\ref{fig:1}, where we demonstrated that the mixture normal distribution had superior fit compared to the simple normal for raw cILR scores computed under the global null.
+However, when the degree of such data features are reasonable, CBEA will still be able to detect a reasonable proportion of samples with inflated counts.
+
+We also observed that choosing different distributional forms did not alter performance values for CBEA.
+This runs contrary to our comparison analysis in Fig~\ref{fig:1}, where we demonstrated that the mixture normal distribution had superior fit compared to the simple normal for raw CBEA scores computed under the global null.
 We hypothesized that this might be due to the difficulty in fitting mixture distribution to data using the expectation maximization algorithm, as the convergence rate is slow when there is high overlap between the mixtures, resulting in a small mixing coefficient for one of the components \cite{naim2012}.
-As such, in our implementation of cILR, in order to ensure convergence for the estimating procedure we increased the number of iterations while relaxing the tolerance parameter.
+As such, in our implementation of CBEA, in order to ensure convergence for the estimating procedure we increased the number of iterations while relaxing the tolerance parameter.
 Furthermore, there are also possible problems with our adjustment procedure for the mixture distribution that might impact overall fit.
@@
 
-Despite these concerns, empirical results still indicate that cILR can confidently identify samples with inflated counts.
-The conservativeness of the correlation adjustment procedure ensures that significant results can be trusted by practitioners, even if cILR might not be able to exhaustively identify all samples with inflated counts.
+Despite these concerns, empirical results still indicate that CBEA can confidently identify samples with inflated counts.
+The conservativeness of the correlation adjustment procedure ensures that significant results can be trusted by practitioners, even if CBEA might not be able to exhaustively identify all samples with inflated counts.
 In situations where either the data is less sparse (e.g.
 containing a lot of core taxa that are prevalent across all samples), there is less inter-taxa correlation within the set (e.g.
-taxa that do not participate in common pathways but have shared characteristics like pathogenicity), or if the effect size is large, then cILR will still be able to produce reasonable power.
-A practitioner can use cILR to screen for samples for subsequent analysis that might involve significant costs, or perform hypothesis generation using a less stringent criteria alongside a multiple testing adjustment procedure (such as Benjamini-Hochberg \cite{benjamini1995}).
+taxa that do not participate in common pathways but have shared characteristics like pathogenicity), or if the effect size is large, then CBEA will still be able to produce reasonable power.
+A practitioner can use CBEA to screen for samples for subsequent analysis that might involve significant costs, or perform hypothesis generation using a less stringent criteria alongside a multiple testing adjustment procedure (such as Benjamini-Hochberg \cite{benjamini1995}).
 
 \subsection*{Downstream analysis}
-The sample-level enrichment scores generated by the cILR method can be used in downstream analyses commonly performed in microbiome research: differential abundance testing and disease prediction.
+The sample-level enrichment scores generated by the CBEA method can be used in downstream analyses commonly performed in microbiome research: differential abundance testing and disease prediction.
 
 \subsubsection*{Differential abundance analysis}
-For differential abundance testing, we evaluated whether using cILR scores alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes.
-We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014} applied on data where taxa were aggregated using the naive sum method.
+For differential abundance testing, we evaluated whether using CBEA scores alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes.
+We compared CBEA against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014} applied on data where taxa were aggregated using the naive sum method.
 We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}.
@@
 Surprisingly, we found some conflicting results when evaluating comparisons across parametric simulations and real data analysis.
-The performance of cILR was consistent across two evaluation criteria, demonstrating good type I error and respectable power.
-However, corncob and DESeq2 showed opposite effects: in simulation experiments, both methods show good type I error control but low power, while in real data analyses, conversely both corncob and DESeq2 showed inflated type I error but comparable power with respect to cILR methods.
-Despite such discrepancy, results still indicate good performance of cILR scores when used as inputs to downstream differential abundance analysis compared to using aggregated raw counts, even in methods designed to handle that type of data such as corncob and DESeq2.
+The performance of CBEA was consistent across two evaluation criteria, demonstrating good type I error and respectable power.
+However, corncob and DESeq2 showed opposite effects: in simulation experiments, both methods show good type I error control but low power, while in real data analyses, conversely both corncob and DESeq2 showed inflated type I error but comparable power with respect to CBEA methods.
+Despite such discrepancy, results still indicate good performance of CBEA scores when used as inputs to downstream differential abundance analysis compared to using aggregated raw counts, even in methods designed to handle that type of data such as corncob and DESeq2.
 
@@
 
-The fact that the performance of cILR remains consistent across both simulation and real data analysis shows that cILR is invariant to taxon-specific biases.
-Furthermore, our evaluation indicates that even a simple difference in means test when used in combination with cILR scores can preserve type I error while maintaining good power.
-As such, a practioner can use cILR as a pre-processing step prior to performing a differential abundance test.
+The fact that the performance of CBEA remains consistent across both simulation and real data analysis shows that CBEA is invariant to taxon-specific biases.
+Furthermore, our evaluation indicates that even a simple difference in means test when used in combination with CBEA scores can preserve type I error while maintaining good power.
+As such, a practioner can use CBEA as a pre-processing step prior to performing a differential abundance test.
 
 \subsubsection*{Predictive models}
-For disease prediction, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using cILR generated scores as inputs.
-Similar to our inference analysis, we compared cILR against both ssGSEA and GSVA.
-Additionally, we also evaluated cILR with the approach where counts of a set were aggregated using sums and then centered log-ratio transformed.
+For disease prediction, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using CBEA generated scores as inputs.
+Similar to our inference analysis, we compared CBEA against both ssGSEA and GSVA.
+Additionally, we also evaluated CBEA with the approach where counts of a set were aggregated using sums and then centered log-ratio transformed.
 This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}.
-Results indicated that cILR produces good performance values across both real data analysis and simulation scenarios.
-Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
-Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome taxonomic data sets.
-This is consistent with our sample ranking analysis (Fig~\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
-However, cILR did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data.
-
-However, in simulation studies, this performance gap between CLR and cILR decreases with higher sparsity and correlation, especially in low effect saturation scenarios.
+Results indicated that CBEA produces good performance values across both real data analysis and simulation scenarios.
+Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that CBEA scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
+Comparatively, CBEA generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome taxonomic data sets.
+This is consistent with our sample ranking analysis (Fig~\ref{fig:3}), where CBEA scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
+However, CBEA did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data.
+
+However, in simulation studies, this performance gap between CLR and CBEA decreases with higher sparsity and correlation, especially in low effect saturation scenarios.
 Additionally, there are also downsides to applying CLR.
@@
 Second, the procedure still relies on using summation of counts prior to transformation, which means that we still can't compare across sets of different sizes, and any bias might still be propagated \cite{mclaren2019}.
-As such, despite benefits in performance for a naive random forest model, there is still space for using cILR as primary inputs into predictive models.
-
-Similar to other experiments in downstream usage of cILR, performance did not change with different underlying distributions, output types, or correlation status.
+As such, despite benefits in performance for a naive random forest model, there is still space for using CBEA as primary inputs into predictive models.
+
+Similar to other experiments in downstream usage of CBEA, performance did not change with different underlying distributions, output types, or correlation status.
 This is surprising since we expect z-scores to perform better as they are able to capture the direction of an association.
 The fact that this effect persisted even onto our real data analysis suggests that this is not due to a deficiency of our simulation design.
-As such, practitioners who wish to use cILR in predictive models might be suited to use the settings that is the fastest to compute.
-
-Ultimately, results indicate that cILR can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity.
-Even though there exists situations where it might not provide maximum predictive values, the flexibility of cILR in various types of analyses enable even though in some scenarios it might not provide maximum predictive values.
+As such, practitioners who wish to use CBEA in predictive models might be suited to use the settings that is the fastest to compute.
+
+Ultimately, results indicate that CBEA can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity.
+Even though there exists situations where it might not provide maximum predictive values, the flexibility of CBEA in various types of analyses enable even though in some scenarios it might not provide maximum predictive values.
 
 \subsection*{Limitations and future directions} 
-There are various limitations to our evaluation of cILR.
+There are various limitations to our evaluation of CBEA.
 First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome taxonomic data.
@@
 Finally, we assumed that taxa within a set are all equally associated with the outcome.
-This limits our ability to evaluate the performance of cILR when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.
-
-Our evaluation showed various drawbacks of the cILR method.
-First, inference with cILR is limited in being able to exhaustively detect all samples with significant inflated counts for a set in situations where there is a high degree of sparsity and inter-taxa correlation.
-Second, for downstream analyses, cILR might not always perform better than competing methods, especially when being used to generate inputs to predictive models.
+This limits our ability to evaluate the performance of CBEA when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.
+
+Our evaluation showed various drawbacks of the CBEA method.
+First, inference with CBEA is limited in being able to exhaustively detect all samples with significant inflated counts for a set in situations where there is a high degree of sparsity and inter-taxa correlation.
+Second, for downstream analyses, CBEA might not always perform better than competing methods, especially when being used to generate inputs to predictive models.
 We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated with adjusting the mixture normal distribution.
@@
 
-In addition, there are possible extensions cILR can we can consider to provide more flexibility across different data analysis scenarios in data analysis.
-First, cILR did not address the sparsity of microbiome taxonomic data and relies on a pseudocount to ensure log operations are valid.
+In addition, there are possible extensions CBEA can we can consider to provide more flexibility across different data analysis scenarios in data analysis.
+First, CBEA did not address the sparsity of microbiome taxonomic data and relies on a pseudocount to ensure log operations are valid.
 We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
-Second, cILR also treated all taxa within the set as equally contributing to the set.
+Second, CBEA also treated all taxa within the set as equally contributing to the set.
 Incorporation of taxa-specific weights could reduce the influence of outliers, such as rare or highly invariant taxa.
@@
 \label{S1_Fig}
-{\bf Computational performance of cILR.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
+{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
 Evaluation was performed on simulated data sets.
@@
 Evaluation was performed on simulated null data sets of 10,000 samples testing for enrichment of a set of size 50.
-For correlation of 0.5, p-values represent correlation adjusted cILR while for correlation of 0, p-values represent unadjusted cILR.
+For correlation of 0.5, p-values represent correlation adjusted CBEA while for correlation of 0, p-values represent unadjusted CBEA.
 
@@
 \label{S1_File}
-{\bf Supplemental derivations.} Includes additional details on addressing variance inflation due to correlation in cILR, as well as computational performance and p-value distribution of the method.
+{\bf Supplemental derivations.} Includes additional details on addressing variance inflation due to correlation in CBEA, as well as computational performance and p-value distribution of the method.
 

### 2021-11-30 cbea_revised.tex
Updates from Overleaf

@@
 The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host.
-Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
+Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019a}, and obesity \cite{aoun2020}.
 To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
@@
 
-However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019,li2015}.
+However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019a,li2015}.
 First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present.
@@
 
-% \bibliography{tax_agg}{}
-
-\begin{thebibliography}{10}
-
-\bibitem{proctor2019}
-Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
-\newblock The {{Integrative Human Microbiome Project}}.
-\newblock Nature. 2019;569(7758):641--648.
-\newblock doi:{10.1038/s41586-019-1238-8}.
-
-\bibitem{sharma2019}
-Sharma S, Tripathi P.
-\newblock Gut Microbiome and Type 2 Diabetes: Where We Are and Where to Go?
-\newblock The Journal of Nutritional Biochemistry. 2019;63:101--108.
-\newblock doi:{10.1016/j.jnutbio.2018.10.003}.
-
-\bibitem{aoun2020}
-Aoun A, Darwish F, Hamod N.
-\newblock The {{Influence}} of the {{Gut Microbiome}} on {{Obesity}} in
-  {{Adults}} and the {{Role}} of {{Probiotics}}, {{Prebiotics}}, and
-  {{Synbiotics}} for {{Weight Loss}}.
-\newblock Preventive Nutrition and Food Science. 2020;25(2):113--123.
-\newblock doi:{10.3746/pnf.2020.25.2.113}.
-
-\bibitem{cho2012}
-Cho I, Blaser MJ.
-\newblock The Human Microbiome: At the Interface of Health and Disease.
-\newblock Nature Reviews Genetics. 2012;13(4):260--270.
-\newblock doi:{10.1038/nrg3182}.
-
-\bibitem{callahan2016}
-Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
-\newblock {{DADA2}}: {{High}}-Resolution Sample Inference from {{Illumina}}
-  Amplicon Data.
-\newblock Nature Methods. 2016;13(7):581--583.
-\newblock doi:{10.1038/nmeth.3869}.
-
-\bibitem{truong2015}
-Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
-\newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
-\newblock Nature Methods. 2015;12(10):902--903.
-\newblock doi:{10.1038/nmeth.3589}.
-
-\bibitem{li2019}
-Li H.
-\newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
-  {{Metagenomics}}.
-\newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
-  2019. p. 977--550.
-
-\bibitem{li2015}
-Li H.
-\newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
-  Compositional Data Analysis}}.
-\newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
-\newblock doi:{10.1146/annurev-statistics-010814-020351}.
-
-\bibitem{shi2016}
-Shi P, Zhang A, Li H.
-\newblock Regression Analysis for Microbiome Compositional Data.
-\newblock The Annals of Applied Statistics. 2016;10(2):1019--1040.
-\newblock doi:{10.1214/16-AOAS928}.
-
-\bibitem{sankaran2014}
-Sankaran K, Holmes S.
-\newblock {{structSSI}}: {{Simultaneous}} and {{Selective Inference}} for
-  {{Grouped}} or {{Hierarchically Structured Data}}.
-\newblock Journal of statistical software. 2014;59(13):1--21.
-\newblock doi:{10.18637/jss.v059.i13}.
-
-\bibitem{benjamini1995}
-Benjamini Y, Hochberg Y.
-\newblock Controlling the {{False Discovery Rate}}: {{A Practical}} and
-  {{Powerful Approach}} to {{Multiple Testing}}.
-\newblock Journal of the Royal Statistical Society Series B (Methodological).
-  1995;57(1):289--300.
-
-\bibitem{chen2018}
-Chen J, King E, Deek R, Wei Z, Yu Y, Grill D, et~al.
-\newblock An Omnibus Test for Differential Distribution Analysis of Microbiome
-  Sequencing Data.
-\newblock Bioinformatics. 2018;34(4):643--651.
-\newblock doi:{10.1093/bioinformatics/btx650}.
-
-\bibitem{weiss2017}
-Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
-\newblock Normalization and Microbial Differential Abundance Strategies Depend
-  upon Data Characteristics.
-\newblock Microbiome. 2017;5(1).
-\newblock doi:{10.1186/s40168-017-0237-y}.
-
-\bibitem{mcmurdie2014}
-McMurdie PJ, Holmes S.
-\newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
-  Inadmissible}}.
-\newblock PLOS Computational Biology. 2014;10(4):e1003531.
-\newblock doi:{10.1371/journal.pcbi.1003531}.
-
-\bibitem{love2014}
-Love MI, Huber W, Anders S.
-\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
-  Data with {{DESeq2}}.
-\newblock Genome Biology. 2014;15(12):550.
-\newblock doi:{10.1186/s13059-014-0550-8}.
-
-\bibitem{quinn2019}
-Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
-\newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
-\newblock GigaScience. 2019;8(giz107).
-\newblock doi:{10.1093/gigascience/giz107}.
-
-\bibitem{quinn2018b}
-Quinn TP, Erb I, Richardson MF, Crowley TM.
-\newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
-\newblock Bioinformatics. 2018;34(16):2870--2878.
-\newblock doi:{10.1093/bioinformatics/bty175}.
-
-\bibitem{gloor2017}
-Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
-\newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
-  Optional}}.
-\newblock Frontiers in Microbiology. 2017;8.
-\newblock doi:{10.3389/fmicb.2017.02224}.
-
-\bibitem{aitchison1999}
-Aitchison J.
-\newblock A {{Concise Guide}} to {{Compositional Data Analysis}}. 1999; p. 134.
-
-\bibitem{kurtz2015}
-Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
-\newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
-  Ecological Networks}}.
-\newblock PLOS Computational Biology. 2015;11(5):e1004226.
-\newblock doi:{10.1371/journal.pcbi.1004226}.
-
-\bibitem{kaul2017}
-Kaul A, Davidov O, Peddada SD.
-\newblock Structural Zeros in High-Dimensional Data with Applications to
-  Microbiome Studies.
-\newblock Biostatistics. 2017;18(3):422--433.
-\newblock doi:{10.1093/biostatistics/kxw053}.
-
-\bibitem{kaul2017a}
-Kaul A, Mandal S, Davidov O, Peddada SD.
-\newblock Analysis of {{Microbiome Data}} in the {{Presence}} of {{Excess
-  Zeros}}.
-\newblock Frontiers in Microbiology. 2017;8.
-\newblock doi:{10.3389/fmicb.2017.02114}.
-
-\bibitem{khatri2012}
-Khatri P, Sirota M, Butte AJ.
-\newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
-  {{Outstanding Challenges}}.
-\newblock PLOS Computational Biology. 2012;8(2):e1002375.
-\newblock doi:{10.1371/journal.pcbi.1002375}.
-
-\bibitem{goeman2007}
-Goeman JJ, B{\"u}hlmann P.
-\newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
-  Issues.
-\newblock Bioinformatics. 2007;23(8):980--987.
-\newblock doi:{10.1093/bioinformatics/btm051}.
-
-\bibitem{subramanian2005}
-Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
-\newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
-  Interpreting Genome-Wide Expression Profiles.
-\newblock Proceedings of the National Academy of Sciences.
-  2005;102(43):15545--15550.
-\newblock doi:{10.1073/pnas.0506580102}.
-
-\bibitem{ashburner2000}
-Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
-\newblock Gene {{Ontology}}: Tool for the Unification of Biology.
-\newblock Nature genetics. 2000;25(1):25--29.
-\newblock doi:{10.1038/75556}.
-
-\bibitem{hanzelmann2013}
-H{\"a}nzelmann S, Castelo R, Guinney J.
-\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
-  {{RNA}}-{{Seq}} Data.
-\newblock BMC Bioinformatics. 2013;14(1):7.
-\newblock doi:{10.1186/1471-2105-14-7}.
-
-\bibitem{frost2020a}
-Frost HR.
-\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
-  Method for Cell-Specific Gene Set Scoring.
-\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
-\newblock doi:{10.1093/nar/gkaa582}.
-
-\bibitem{mclaren2019}
-McLaren MR, Willis AD, Callahan BJ.
-\newblock Consistent and Correctable Bias in Metagenomic Sequencing
-  Experiments.
-\newblock eLife. 2019;8:e46923.
-\newblock doi:{10.7554/eLife.46923}.
-
-\bibitem{egozcue2005}
-Egozcue JJ, {Pawlowsky-Glahn} V.
-\newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
-  Analysis}}.
-\newblock Mathematical Geology. 2005;37(7):795--828.
-\newblock doi:{10.1007/s11004-005-7381-9}.
-
-\bibitem{chong2020}
-Chong J, Liu P, Zhou G, Xia J.
-\newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
-  Functional, and Meta-Analysis of Microbiome Data.
-\newblock Nature Protocols. 2020;15(3):799--821.
-\newblock doi:{10.1038/s41596-019-0264-1}.
-
-\bibitem{tian2005}
-Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
-\newblock Discovering Statistically Significant Pathways in Expression
-  Profiling Studies.
-\newblock Proceedings of the National Academy of Sciences.
-  2005;102(38):13544--13549.
-\newblock doi:{10.1073/pnas.0506577102}.
-
-\bibitem{egozcue2003}
-Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcelo-Vidal} C.
-\newblock Isometric {{Logratio Transformations}} for {{Compositional Data
-  Analysis}}.
-\newblock Mathematical Geology. 2003; p.~22.
-
-\bibitem{silverman2017}
-Silverman JD, Washburne AD, Mukherjee S, David LA.
-\newblock A Phylogenetic Transform Enhances Analysis of Compositional
-  Microbiota Data.
-\newblock eLife. 2017;6:e21887.
-\newblock doi:{10.7554/eLife.21887}.
-
-\bibitem{barbie2009}
-Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
-\newblock Systematic {{RNA}} Interference Reveals That Oncogenic
-  {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
-\newblock Nature. 2009;462(7269):108--112.
-\newblock doi:{10.1038/nature08460}.
-
-\bibitem{quast2013}
-Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
-\newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
-  Processing and Web-Based Tools.
-\newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
-\newblock doi:{10.1093/nar/gks1219}.
-
-\bibitem{delignette-muller2015}
-{Delignette-Muller} ML, Dutang C.
-\newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
-\newblock Journal of Statistical Software. 2015;64(4):1--34.
-
-\bibitem{benaglia2009}
-Benaglia T, Chauveau D, Hunter DR, Young D.
-\newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
-\newblock Journal of Statistical Software. 2009;32(6):1--29.
-
-\bibitem{washburne2017}
-Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL, Mukherjee S, et~al.
-\newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
-  Associations in Microbiome Datasets.
-\newblock PeerJ. 2017;5:e2969.
-\newblock doi:{10.7717/peerj.2969}.
-
-\bibitem{morton2017}
-Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
-  et~al.
-\newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
-\newblock mSystems. 2017;2(1).
-\newblock doi:{10.1128/mSystems.00162-16}.
-
-\bibitem{efron2004}
-Efron B.
-\newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
-\newblock Journal of the American Statistical Association.
-  2004;99(465):96--104.
-\newblock doi:{10.1198/016214504000000089}.
-
-\bibitem{wu2012}
-Wu D, Smyth GK.
-\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
-  Correlation.
-\newblock Nucleic Acids Research. 2012;40(17):e133.
-\newblock doi:{10.1093/nar/gks461}.
-
-\bibitem{calgaro2020}
-Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
-\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
-  and Metagenomics Applied to Microbiome Data.
-\newblock Genome Biology. 2020;21(1):191.
-\newblock doi:{10.1186/s13059-020-02104-1}.
-
-\bibitem{cario1997}
-Cario MC.
-\newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
-  Distributions}} and {{Correlation Matrix}}. 1997; p.~19.
-
-\bibitem{agresti1998}
-Agresti A, Coull BA.
-\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
-  Estimation}} of {{Binomial Proportions}}.
-\newblock The American Statistician. 1998;52(2):119--126.
-\newblock doi:{10.2307/2685469}.
-
-\bibitem{delong1988}
-DeLong ER, DeLong DM, {Clarke-Pearson} DL.
-\newblock Comparing the Areas under Two or More Correlated Receiver Operating
-  Characteristic Curves: A Nonparametric Approach.
-\newblock Biometrics. 1988;44(3):837--845.
-
-\bibitem{hawinkel2019}
-Hawinkel S, Mattiello F, Bijnens L, Thas O.
-\newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
-  Control the False Discovery Rate.
-\newblock Briefings in Bioinformatics. 2019;20(1):210--221.
-\newblock doi:{10.1093/bib/bbx104}.
-
-\bibitem{breiman2001}
-Breiman L.
-\newblock Random {{Forests}}.
-\newblock Machine Learning. 2001;45(1):5--32.
-\newblock doi:{10.1023/A:1010933404324}.
-
-\bibitem{kuhn2020}
-Kuhn M, Wickham H.
-\newblock Tidymodels: {{Easily}} Install and Load the 'tidymodels' Packages;
-  2020.
-
-\bibitem{xiao2018}
-Xiao J, Chen L, Yu Y, Zhang X, Chen J.
-\newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
-  {{Predictive Modeling}} of {{Microbial Community Data}}.
-\newblock Frontiers in Microbiology. 2018;9.
-\newblock doi:{10.3389/fmicb.2018.03112}.
-
-\bibitem{dong2020}
-Dong M, Li L, Chen M, Kusalik A, Xu W.
-\newblock Predictive Analysis Methods for Human Microbiome Data with
-  Application to {{Parkinson}}'s Disease.
-\newblock PLOS ONE. 2020;15(8):e0237779.
-\newblock doi:{10.1371/journal.pone.0237779}.
-
-\bibitem{pasolli2017}
-Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
-\newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
-\newblock Nature Methods. 2017;14(11):1023--1024.
-\newblock doi:{10.1038/nmeth.4468}.
-
-\bibitem{schiffer2019}
-Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
-\newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
-  through Bioconductor.
-\newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
-
-\bibitem{gonzalez2018}
-Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
-  Ackermann G, et~al.
-\newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
-\newblock Nature Methods. 2018;15(10):796--798.
-\newblock doi:{10.1038/s41592-018-0141-9}.
-
-\bibitem{consortium2012}
-Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
-  et~al.
-\newblock Structure, Function and Diversity of the Healthy Human Microbiome.
-\newblock Nature. 2012;486(7402):207--214.
-\newblock doi:{10.1038/nature11234}.
-
-\bibitem{thurnheer2016}
-Thurnheer T, Bostanci N, Belibasakis GN.
-\newblock Microbial Dynamics during Conversion from Supragingival to
-  Subgingival Biofilms in an in Vitro Model.
-\newblock Molecular Oral Microbiology. 2016;31(2):125--135.
-\newblock doi:{10.1111/omi.12108}.
-
-\bibitem{beghini2019}
-Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
-\newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
-  in the {{New York City Health}} and {{Nutrition Examination Study}}.
-\newblock Annals of Epidemiology. 2019;34:18--25.e3.
-\newblock doi:{10.1016/j.annepidem.2019.03.005}.
-
-\bibitem{matteocalgaro2020}
-Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
-\newblock Zenodo.
-
-\bibitem{nielsen2014}
-Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
-\newblock Identification and Assembly of Genomes and Genetic Elements in
-  Complex Metagenomic Samples without Using Reference Genomes.
-\newblock Nature Biotechnology. 2014;32(8):822--828.
-\newblock doi:{10.1038/nbt.2939}.
-
-\bibitem{gevers2014}
-Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
-  et~al.
-\newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
-  {{Disease}}.
-\newblock Cell Host \& Microbe. 2014;15(3):382--392.
-\newblock doi:{10.1016/j.chom.2014.02.005}.
-
-\bibitem{martin2020}
-Martin BD, Witten D, Willis AD.
-\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
-  Regression.
-\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
-\newblock doi:{10.1214/19-AOAS1283}.
-
-\bibitem{naim2012}
-Naim I, Gildea D.
-\newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
-  {{Unbalanced Mixing Coefficients}}.
-\newblock Proceedings of the 29th International Coference on International
-  Conference on Machine Learning. 2012; p.~8.
-\newblock doi:{10.5555/3042573.3042756}.
-
-\bibitem{wu2016}
-Wu C, Chen J, Kim J, Pan W.
-\newblock An Adaptive Association Test for Microbiome Data.
-\newblock Genome Medicine. 2016;8(1):56.
-\newblock doi:{10.1186/s13073-016-0302-3}.
-
-\bibitem{martin-fernandez2012}
-{Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
-  {Palarea-Albaladejo} J.
-\newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
-  {{Classical}} and Robust Approaches.
-\newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
-\newblock doi:{10.1016/j.csda.2012.02.012}.
-
-\end{thebibliography}
+\bibliography{mac_taxagg}{}
 

### 2021-11-30 placeholder.tex
Updates from Overleaf

--- 2021-11-30 cbea.tex
@@
-% Template for PLoS
-% Version 3.5 March 2018
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-%\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-
-%% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
-
-%% END MACROS SECTION
-
-% Adding packages 
-\usepackage{bm}
-\usepackage[parfill]{parskip}
-
-\begin{document}
-\vspace*{0.2in}
-
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{cILR: Competitive isometric log-ratio for taxonomic enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Quang P. Nguyen\textsuperscript{1,2},
-Anne G. Hoen\textsuperscript{1,2\ddag},
-H. Robert Frost\textsuperscript{1*\ddag\dag},
-\\
-\bigskip
-\textbf{1} Department of Biomedical Data Science, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA
-\\
-\textbf{2} Department of Epidemiology, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA
-\\
-\bigskip
-
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-%\Yinyang These authors contributed equally to this work.
-
-% Additional Equal Contribution Note
-% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-\ddag These authors jointly supervised this work.\\
-\dag Corresponding author
-% Current address notes
-%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
-% \textcurrency b Insert second current address 
-% \textcurrency c Insert third current address
-
-% Deceased author note
-% \dag Deceased
-
-% Group/Consortium Author Note
-%\textpilcrow Membership list can be found in the Acknowledgments section.
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* hildreth.r.frost@dartmouth.edu
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
-Research in human associated microbiomes often involves the analysis of taxonomic count tables generated via high-throughput sequencing.
-It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional.
-An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets.
-Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis.
-Unfortunately, there is a lack of sophisticated set-based analysis methods specific to microbiome taxonomic data, where current practice often employs abundance summation as a technique for aggregation.
-This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
-Here, we attempt to fill this gap with a new single sample taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature.
-Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
-We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
-Herein we demonstrate using both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios.
-Additionally, it provides informative scores that can be inputs to downstream differential abundance and prediction tasks.
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-\section*{Author summary}
-The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing.
-However, microbiome taxonomic data is sparse and high dimensional which prevents the application of standard statistical techniques.
-One approach to address this problem is to perform analyses at the level of taxon sets.
-Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability.
-Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given that compared to other 'omics data types microbiome taxonomic data is strictly compositional.
-We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis.
-Our scores can be used for statistical inference, and as inputs to other downstream analyses such as differential abundance and prediction models.
-We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.
-
-\linenumbers
-
-% Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
-The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host.
-Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
-To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
-amplicon sequencing) or by profiling all the genomic content of the sample (i.e.
-whole-genome shotgun sequencing) \cite{cho2012}.
-Raw sequencing data is then processed through a variety of bioinformatic pipelines \cite{callahan2016, truong2015}, yielding various data products, one of which are taxonomic tables which can be used to study associations between members of the microbiome and an exposure or outcome of interest.
-
-However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019,li2015}.
-First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present.
-For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016}, allowing for simultaneous model fitting and variable selection.
-For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.
-
-Second, the number of reads obtained is constrained by the sequencing instrument at an arbitrary limit, and is inconsistent across samples, resulting in a variable number of total read counts per sample.
-Many normalization methods \cite{weiss2017} have been proposed to address these issues, including cross-applying methods from the gene expression literature \cite{mcmurdie2014}.
-However, these methods rely on assumptions specific to the original bulk RNA-seq data sets such as the presence of housekeeping genes with consistent expression levels \cite{love2014}, which might not be not true in the context of microbiome relative abundance data \cite{quinn2019,quinn2018b}.
-As such, microbiome taxonomic data is strictly compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another.
-Consequently, log-ratio transformations from the compositional data literature are often utilized \cite{aitchison1999}.
-
-Third, the data are highly zero-inflated, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment).
-Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015}, or applying zero-inflated models \cite{chen2018, kaul2017}.
-Newer methods developed recently have focused on understanding the different types of zeros in the data, providing more sophisticated heuristics around when pseudocounts can be utilized \cite{kaul2017a}.
-
-Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}.
-Aggregated sets are less sparse than their constituent elements, and testing on a smaller number of variables reduces the multiple testing burden, thereby increasing power and reproducibility.
-Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations.
-There exists a diverse set of available methods developed to perform such analyses.
-More traditional set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest \cite{goeman2007}.
-Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values.
-The most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size.
-Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples.
-Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample level and are more akin to a transformation.
-This strategy allows for the flexible incorporation of different statistical techniques downstream, such as prediction models, as well as for visualization purposes in ordination plots.
-
-In microbiome research, even though no explicit enrichment analysis is performed, standard practice often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum by simple summation of abundances \cite{mclaren2019}.
-Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, it increases bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}.
-As such, there is a need for microbiome researchers to adopt more robust sample-level set enrichment methods.
-Unfortunately, limited work has been done to extend existing methods to be more specific to microbiome relative abundance data.
-Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}.
-However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used.
-
-Here, we present a novel method that generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}.
-We leverage the concept of the $Q_1$ competitive hypothesis presented in Tian et al.
-\cite{tian2005}, which tests the null that the value of variables within a specific set is equal to the value of measured variables not in the set.
-The competitive null hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables.
-We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a multiplicative aggregation method that addresses the downsides of the naive summation-based method presented above \cite{mclaren2019, silverman2017}.
-The resulting method, titled competitive isometric log-ratio (cILR), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing.
-These scores can then act as inputs to differential abundance and predictive modeling tasks downstream.
-
-In the following sections, we provide the formulation of cILR and discuss some statistical properties.
-We illustrate significance testing at the sample level using cILR and evaluate type I error and power under different simulation scenarios and real data applications.
-We assess the informativeness of cILR generated scores, and evaluate how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis.
-We compare the performance of cILR in these respective tasks against standard microbiome taxonomic data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods.
-An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).
-
-\section*{Materials and Methods} \label{methods}
-\subsection*{Competitive Isometric Log-ratio (cILR)}
-The cILR method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}.
-Details on the computational implementation of cILR can be found in the supplemental materials.
-The cILR method takes two inputs:
-\begin{itemize}
-    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing.
-Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing.
-    \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxon $p$ to $m$ sets of interest.
-These sets can be Linnean taxonomic classifications annotated using databases such as SILVA \cite{quast2013}, or those based on more functionally driven categories such as tropism or ecosystem roles ($A_{i,j} = 1$ indicates that microbe $i$ belongs to set $j$).
-\end{itemize}
-The cILR method generates one output:  
-\begin{itemize} 
-    \item $\mathbf{E}$: $n$ by $m$ matrix indicating the enrichment score of $m$ pre-defined sets identified in $\mathbf{A}$ across $n$ samples. 
-\end{itemize}
-The procedure is as follows:  
-\begin{enumerate}
-    \item \textbf{Compute the cILR statistic}: Let $\mathbf{M}$ be a $n$ by $m$ matrix of cILR scores. Let $\mathbf{M}_{i,k}$ be cILR scores for set $k$ of sample $i$:   
-    \begin{equation}\label{main_eq}
-        \mathbf{M}_{i,k} = \sqrt{\frac{\sum_k A_{ik}(p - \sum_k A_{ik})}{p}} \ln \left( \frac{g(\mathbf{X}_{i,j}|\mathbf{A}_{j,k} = 1)}{g(\mathbf{X}_{i,j}|\mathbf{A}_{j,k} \neq 1))} \right)
-    \end{equation}
-    where $g(.)$ is the geometric mean. This represents the ratio of the geometric mean of the relative abundance of taxa assigned to set $k$ and remainder taxa. 
-    \item \textbf{Estimate the empirical null distribution} Enrichment scores represent the test statistic for the $Q_1$ null hypothesis $H_o$ that relative abundances in $\mathbf{X}$ of members of set $k$ are not enriched compared to those not in set $k$.
-Since the distribution of cILR under the null vary depending on data characteristics (Fig~\ref{fig:1}), an empirical null distribution will be estimated from data.
-    \begin{itemize}
-        \item \textbf{Compute the cILR statistic on permuted and un-permuted $\mathbf{X}$}.
-Let $\mathbf{X}_{perm}$ be the column permuted relative abundance matrix, and $\mathbf{M}_{perm}$ be the corresponding cILR scores generated from $\mathbf{X}_{perm}$.
-Similarly, we have $\mathbf{M}_{unperm}$ be cILR scores generated from $\mathbf{X}$.
-        \item \textbf{Estimate correlation-adjusted empirical distribution for each set}.
-For each set, a fit a parametric distribution to both $\mathbf{M}_{perm}$ and $\mathbf{M}_{unperm}$.
-The location measure estimated from $\mathbf{M}_{perm}$ and the spread measure estimated from $\mathbf{M}_{unperm}$ will be combined as the correlation-adjusted empirical null distribution $\mathbf{P}_{emp}$ for each set.
-Two available options are the normal distribution and the mixture normal distribution.
-For the normal distribution, parameters were estimated using the method of maximum likelihood implemented in the \emph{fitdistr} package \cite{delignette-muller2015}.
-For the mixture normal distribution, parameters were estimated using an expectation-maximization algorithm implemented in the \emph{mixtools} package \cite{benaglia2009}.
-    \end{itemize}
-    \item \textbf{Calculate finalized cILR scores with respect to the empirical null}.
-Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution.
-P-values can be calculated by subtracting $\mathbf{E}$ from 1.
-\end{enumerate}
-
-\subsection*{Properties of cILR}
-\subsubsection*{cILR and the Isometric Log Ratio Transformation}
-The cILR statistic is a special instance of the isometric log-ratio transformation (ILR) \cite{egozcue2003}.
-The standard ILR is a transformation method to address the negative correlation bias inherent in compositional data by providing an isometry between the $D$-dimensional simplex $\mathbb{S}^D$ and coordinates in the $D-1$ real space $\mathbb{R}^{D-1}$ \cite{egozcue2003,washburne2017}.
-This is accomplished by projecting the composition onto a chosen orthonormal basis in $\mathbb{R}$, which can be defined by a sequential binary partition (SBP) of the variables (e.g.
-a rooted phylogenetic tree).
-The ILR transformed variables are the coordinates of nodes within an SBP tree of the variables.
-Without loss of generalizability, in a given SBP with node $i$ splitting variables between sets $R$ and $S$, we have the ILR coordinate $x^{*}_{i}$ as:
-\begin{equation}\label{ilr_standard}
-    x^{*}_i = \sqrt{\frac{rs}{r+s}} \log\left(\frac{g(X_{j|j \in R})}{g(X_{j|j \in S})}\right)
-\end{equation}
-where $r$ and $s$ are the cardinalities of sets $R$ and $S$ respectively, $g(z)$ is the geometric mean, and $X_{j}$ are values of the original predictors with indexes defined by membership in $R$ and $S$.
-The ILR confer many important benefits.
-First, ILR coordinates exist in real space, whereby common statistical methods can be used.
-Second, ILR aggregated variables preserve inter-sample distances before and after aggregation \cite{egozcue2005}.
-Third, ILR variables are not constrained to sum to 0 as that of the centered log-ratio transformation, resulting in a covariance matrix that is not singular \cite{egozcue2003}.
-
-The usage of the ILR statistic is not uncommon in the microbiome literature.
-They are usually termed ``compositional balances", and have been leveraged in many recent approaches in variable transformation \cite{washburne2017,silverman2017,morton2017}.
-The cILR formulation in Eq~\eqref{main_eq} is a special case of Eq~\eqref{ilr_standard} defined on a node that splits the taxa into two disjoint sets, one representing the set of interest, the other representing the remaining taxa.
-As such, the cILR transformation inherits the properties of the ILR as a log-ratio method applicable to compositional data sets.
-However, unlike the ILR and its variants \cite{silverman2017, morton2017, washburne2017}, the axes defined by each cILR set are not orthogonal (since the balances are mutually exclusive between sets and do not belong in the same SBP).
-Hence, a correlation can exist between cILR aggregated variables.
-
-\subsubsection*{Statistical Properties of cILR}
-We can perform significance testing on the cILR statistic which corresponds to the null hypothesis that the center of the subcomposition defined by the set is equal to the center of the subcomposition defined by the complement of the set.
-This is equivalent to the $Q_1$ competitive null hypothesis in the gene set testing literature \cite{tian2005} where the enrichment of a gene set is defined with respect to genes outside the set.
-
-We can apply prior usage of the ILR statistic in hypothesis testing to cILR by assuming that the null distribution of cILR follows a standard normal distribution \cite{egozcue2005}.
-However, when applying cILR for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses.
-Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies.
-As such, to perform significance testing using cILR, we also estimated the null distribution from observed raw cILR variables.
-
-This assumption is also supported by preliminary simulation studies (detailed below).
-In panel A of Fig~\ref{fig:1}, we simulated microbiome taxonomic count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
-It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation.
-Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis).
-The effect is most dramatic under both high inter-taxa correlation and sparsity.
-This heterogeneity further supports the decision to estimate an empirical null distribution, similar to Efron \cite{efron2004}.
-
-\begin{figure} [!h]
-    \centering
-    \includegraphics[width=\linewidth]{figures/kurtosis_skewness_gof.png}
-    \caption{{\bf Properties of the null distribution of cILR under the global null simulations}.
-Panel \textbf{(B)} presents kurtosis and skewness of cILR scores while panel \textbf{(A)} presents the goodness of fit (as Kolmogorov-Smirnov D statistic) for mixture normal and normal distributions.
-Panel \textbf{(C)} is a density plot of the shape of the null distribution.
-Results indicated the necessity of estimating an empirical null and demonstrating that the mixture distribution was the better fit compared to the basic normal.}
-    \label{fig:1}
-\end{figure}
-
-Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null.
-To address this issue, we also evaluated a mixture distribution of two normal components.
-Panel B of Fig~\ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
-We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.
-
-We performed our empirical null estimation by fitting our distribution of choice and computing relevant parameters on raw cILR scores on taxa-permuted data (equivalent to gene permutation in the gene expression literature).
-As such, the null distribution is characterized by scores computed on sets of equal size with randomly drawn taxa.
-However, null distribution based on taxa-permutation is sensitive to inter-taxa correlations within the set \cite{wu2012}.
-Since the permutation procedure does not preserve correlation structures, estimating parameters from empirical scores on permuted data will underestimate the variance inflation due to correlation.
-We account for this by combining the mean estimate from permuted data with the variance estimate from unpermuted data, where the inter-taxa correlation structure remains undisturbed.
-However, this procedure assumes that the variance of cILR is equal under both the null and alternate hypotheses.
-
-\subsection*{Evaluation}
-All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/cILR\_analysis}{qpmnguyen/cILR\_analysis}).
-
-\subsubsection*{Parametric Simulations}  
-To address the performance of cILR for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
-Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
-\begin{equation}
-    \mathbf{X}_{ij} =
-      \begin{cases}
-        0 & \text{with probability $p_j$}\\
-        \mathbf{NB}(\mu_j, \phi_j) & \text{with probability $1 - p_j$}\\
-      \end{cases}       
-\end{equation}
-
-where $\mu_j$ and $\phi_j$ are mean and dispersion parameters, respectively.
-To incorporate a flexible correlation structure into our simulation model, we utilized the NorTA (Normal to Anything) method \cite{cario1997}.
-Given an $n$ by $p$ matrix of values $\mathbf{U}$ sampled from multivariate normal distribution with correlation matrix $\mathbf{\rho}$, we can generate target microbiome count vector $\mathbf{X_{.j}}$ for taxa $j$ following the marginal distribution $\mathbf{NB}$ characterized by the negative binomial cumulative distribution function $\mathbb{F_{\mathbf{NB}}}$:
-\begin{equation}
-    \mathbf{X}_{.j} = \mathbb{F_{\mathbf{NB}}}^{-1}(\Phi_{U_i})
-\end{equation}
-In this instance, for each taxon $j$, we set elements in $\mathbf{U}_{.j}$ to be zero with probability $p_j$ and applied $\mathbf{NB}^{-1}(\mu_j, \phi_j)$ on non-zero elements to generate our final count matrix $\mathbf{X}$.
-To ensure that our simulations match closely to real data, we fitted negative binomial distribution using a maximum likelihood approach (with the \emph{fitdistrplus} package in R \cite{delignette-muller2015}) to non-zero counts for each taxon from 16S rRNA profiling of stool samples from the Human Microbiome Project (HMP).
-We take the median values of the estimated mean and dispersion parameters as the baseline of our simulations.
-For simplicity, we assumed that inter-taxa correlation follows an exchangeable structure
-
-\noindent \textbf{Single Sample Enrichment}: To assess type I error rate and power for enrichment significance testing at the sample level, we simulated data based on the schema above, and assessed enrichment for one focal set.
-Type I error was obtained under the global null as the number of samples where the null hypothesis was rejected at $\alpha = 0.05$ over the total number of samples (which represents the total number of hypotheses tested).
-Power was obtained using the same formulation as type I error rate but under the global alternate.
-We treated type I error and power as estimates of binomial proportions and utilized the Agresti-Couli \cite{agresti1998} formulation to calculate 95\% confidence intervals.
-Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$) and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$).
-For type I error analysis, we also varied the size of the set (50, 100, 150).
-For power analyses, set size was kept constant at 100 but different effect sizes (fold change of 1.5, 2, and 3).
-All sample sizes were set at 10,000.
-
-For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUROC/AUC).
-This is a strategy used in Frost \cite{frost2020a} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
-whether samples with inflated counts are highly ranked using estimated scores).
-DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate.
-Simulation settings for classification performance were identical to power analyses as detailed in the previous paragraph.
-
-\noindent \textbf{Differential Abundance Analysis}: To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition.
-Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null.
-Power is defined similarly, but instead under the global alternate hypothesis.
-Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error.
-A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size.
-Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$).
-Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3).
-Due to the compositional nature of microbiome taxonomic data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets.
-As such, we applied a compensation procedure as described in Hawinkel et al.
-\cite{hawinkel2019} to ensure the validity of simulation results.
-All sample sizes were set at 2,000.
-
-\noindent \textbf{Prediction}: To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}.
-For binary outcomes, we use AUC similar to the classification analyses above.
-For continuous outcomes, we used root mean squared error (RMSE).
-All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020} suite of packages.
-Across both learning tasks, we varied sparsity ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation ($\rho = 0, 0.2, 0.5$).
-Continuous outcomes $Y_{cont}$ were generated as linear combinations of taxa counts.
-\begin{equation}
-    Y_{cont} = f(\mathbf{X}) + \mathbf{\epsilon}
-\end{equation}
-where $\mathbf{\epsilon} \sim N(0, \sigma_{\epsilon}^2)$ and $f(\mathbf{X}) = \beta_0 + \mathbf{X}\mathbf{\beta}$.
-For each simulation, we set $\beta_0$ to be $\frac{6}{\sqrt{10}}$ similar to \cite{xiao2018}.
-The degree of model saturation (the number of non zero $\mathbf{\beta}$ values) were varied between 0.1 and 0.5, and signal to noise ratio (SNR = $\frac{\sigma(f(\mathbf{X}))}{\sigma_{\epsilon}}$) was varied between 1.5, 2, and 3.
-
-For binary outcomes, we generate $Y_{binary}$ as Bernoulli draws with probability $p_{binary}$, where 
-\begin{equation}
-    p_{binary} = \frac{1}{1 + \exp(f(\mathbf{X}) + \mathbf{\epsilon})}
-\end{equation}
-To ensure a balance of classes, we applied the strategy described in Dong et al.
-\cite{dong2020} where the associated $\beta$ values are evenly split between positive and negative associations.
-All data sets generated from prediction tasks have 2,000 samples with 5,000 taxa over 50 sets with a size of 100 taxa per set.
-
-\subsubsection*{Real Datasets}
-
-In addition to simulation analyses, we also evaluated our method using real data sets based on both 16S rRNA gene sequencing and whole-genome sequencing.
-All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.
-
-\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of cILR in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}.
-We utilized this data set following the approach outlined in Calagaro et al.
-\cite{calgaro2020}.
-This data set is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}.
-Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false positive rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment.
-Microbial tropism annotation at the genus level was from Beghini et al.
-\cite{beghini2019} and was downloaded directly from the GitHub repository associated with Calagaro et al.
-\cite{matteocalgaro2020}.
-
-\noindent \textbf{Differential Abundance Analsysis}: To assess type I error using cILR scores in differential abundance analysis, we utilized the 16S rRNA gene sequencing of stool samples from the Human Microbiome Project \cite{consortium2012, proctor2019}.
-Here, we randomly assign samples a label of case or control, and repeated this process 500 times, assessing all candidate methods at each iteration.
-Type I error is then the number of taxa identified as differentially abundant across all tested taxa.
-For the true positive rate, we used the same gingival data set as described above.
-However, instead of testing for aerobic microbes as a group, the true positive rate is the number of aerobic/anaerobic genera identified as differentially abundant across all aerobic or anaerobic genera.
-
-\noindent \textbf{Disease Prediction}: To assess predictive power, we utilized the whole genome sequencing of stool samples of inflammatory bowel disease (IBD) patients from the MetaHIT consortium \cite{nielsen2014}.
-This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease).
-Additionally, we also utilized a similar data set from Gevers et al.
-\cite{gevers2014} which also profiles the gut microbiome of IBD patients and controls but using 16S rRNA gene sequencing.
-This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada.
-Of the 671 samples obtained, 500 samples belong to patients with IBD.
-
-\subsubsection*{Comparison Methods}
-
-\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank sum test.
-We added a pseudocount of 1 to all values.
-This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample.
-For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test.
-All three approaches were applied directly on count data (after pseudocount).
-For GSVA, the Poisson kernel was used.
-
-\noindent \textbf{Differential Abundance}: Since cILR are sample-level enrichment scores, we performed differential abundance by using a Wilcoxon Rank Sum test and Welch's t-test across case/control status on cILR generated scores.
-We added a pseudocount of 1 to all values.
-For comparison, we chose representative state-of-the-art methods in differential abundance analysis, namely DESeq2 \cite{love2014,mcmurdie2014} and corncob \cite{martin2020}.
-For DESeq2, we performed a likelihood ratio test against an intercept only reduced model with dispersion estimated with local fit.
-For corncob, we also performed a likelihood ratio test against an intercept only reduced model without bootstrapping.
-
-\noindent \textbf{Disease Prediction}: We fit random forest on cILR scores, as well as ssGSEA \cite{hanzelmann2013} and GSVA \cite{barbie2009} similar to single sample enrichment section.
-We added a pseudocount of 1 to all values.
-Additionally, we also compared performance using cILR against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.
-
-\section*{Results}
-In this section, we present the performance of our proposed method for three applicable microbiome analysis tasks: sample level enrichment, differential abundance, and disease prediction.
-We obtained these results from both parametric simulations and examples from real data.
-
-\subsection*{Enrichment testing at the sample level}
-cILR provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{methods}.
-Here, we present empirical results for this application of cILR assessing type I error, power, and classification capacity.
-
-\subsubsection*{Simulation studies}
-Panel A and B in Fig~\ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
-We benchmarked the results of the cILR method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
-All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions.
-However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted cILR variants show high levels of inflated type I error, where Wilcoxon test performed the worst.
-On the other hand, adjusted cILR methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
-    \caption{Sample-level inference with cILR under parametric simulations (\textbf{(A)} and \textbf{B}), and real data analysis (\textbf{(C)} and \textbf{(D)}).
-In simulation analyses, panel \textbf{(A)} shows type I error rate, and panel \textbf{(B)} shows power for single sample enrichment test for a specified set and was compared against a Wilcoxon rank sum test at $\alpha$ of 0.05.
-In real data analysis, panel \textbf{(C)} shows the false-positive rate, and panel \textbf{(D)} shows the true positive rate.
-For this analysis, 16S rRNA data from the oral microbiome of the gingival site was used.
-The set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples.
-Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach.
-Adjusted cILR demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate.
-However, this resulted in lower power for adjusted cILR methods.}
-    \label{fig:2}
-\end{figure}
-
-However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B.
-In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un-adjusted ones.
-However, in higher correlation scenarios, the difference in power is much more dramatic.
-At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted cILR was only performing at 50\% power, while unadjusted cILR and wilcoxon rank sum test were able to reach 80\%.
-These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of cILR.
-Most importantly, cILR demonstrate higher power in all scenarios where type I error is properly controlled.
-
-To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set.
-This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated.
-Fig~\ref{fig:3} presents this result.
-We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
-Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
-In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
-Notably, the W-statistic provided the least information for classifying samples with inflated taxa.
-
-\subsubsection*{Real data evaluations}  
-These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{methods}.
-Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations.
-As expected in Fig~\ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods.
-Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05.
-Power analysis (Fig~\ref{fig:2}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
-    \caption{Classification performance via AUC of cILR, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}.
-Performance scores measure whether scores can highly rank samples that are known to have inflated abundance.
-In the gingival data set presented in panel \textbf{(B)}, samples from the supragingival site are assumed to have an inflated abundance of aerobic microbes.
-Error bars are the 95\% DeLong confidence intervals for AUC \cite{delong1988}}
-    \label{fig:3}
-\end{figure}
-
-\subsection*{Differential abundance analysis}
-cILR generates sample-specific scores representing the degree of enrichment of a pre-defined set.
-As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test).
-We compared the performance of this approach with cILR and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
-
-\subsubsection*{Simulation studies}
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
-    \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
-For cILR methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by cILR (across different output types and distributional assumptions).
-cILR associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
-    \label{fig:4}
-\end{figure}
-
-Fig.~\ref{fig:4} present results for simulation studies for both type I error (panel A) and power (panel B) evaluations.
-All methods control for type I error well across both sparsity and correlation levels, where the estimated rate was consistently around the 0.05 pre-defined threshold.
-Results were similar across all evaluated methods, although in some instances, for example in medium correlation setting ($\rho = 0.2$), the unadjusted cILR resulted in higher type I error, regardless of difference in means test and distribution of choice.
-
-The difference between the methods is more noticeable when evaluating power.
-All cILR associated variants showed much higher power even when the effect size is limited (fold change is 1.5), and there is a noticeable gap in performance between cILR and both DESeq2 and corncob.
-Surprisingly, this effect is consistent across correlation levels and sparsity, even though we expectedly see performance in power drop as a function of sparsity especially in low effect size settings.
-
-\subsubsection*{Real data evaluations}
-In addition to simulation studies, we also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP (Fig~\ref{fig:5}).
-For type I error evaluations, we use stool samples and randomly assign them with case/control status and calculated type I error as the proportion of genera identified as significantly different.
-For true positive rate evaluations, we use the gingival data set as detailed in the previous section, and calculated the true positive rate as the proportion of genera labeled as either anaerobic or aerobic that were found to be significant.
-
-We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of cILR were controlling for type I error at the defined $\alpha$ threshold of 0.05.
-This is surprising given the consistency of preserving type I error for both corncob and DESeq2 in all simulation evaluations.
-
-In true positive experiments with data from the gingival site, estimated rates were more similar across the different methods.
-As expected, using the Wilcoxon rank sum test resulted in lower true positive rate compared to remaining methods, but the difference was not noticable.
-This is also surprising given that in simulation studies, both corncob and DESeq2 showed markedly lower power across all effect sizes.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
-    \caption{Differential abundance analysis using corncob, DESeq2 and cILR with either Wilcoxon rank sum test or Welch's t-test.
-Panel \textbf{(A)} shows type I error results as the proportion of significant genera after 500 iterations where case/control status was assigned randomly to each sample.
-Panel \textbf{(B)} shows true positive rate results as the proportion of significant genera who are either obligate anaerobes or aerobes.
-Both evaluations use 16S rRNA gene sequencing data from HMP.
-Type I error evaluation used stool samples while the true positive rate evaluation used samples from the gingival site.
-Results showed that cILR associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2}
-    \label{fig:5}
-\end{figure}
-
-\subsection*{Disease Prediction}   
-Since cILR can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models.
-In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating cILR, ssGSEA, and GSVA).
-Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.
-\subsubsection*{Simulation studies}
-Fig~\ref{fig:6} shows results for simulation studies as detailed in the \nameref{methods} section.
-Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome.
-As expected, performance across all assessed methods increased with a higher signal-to-noise ratio.
-Both CLR and cILR approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks.
-This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while cILR and CLR did not experience a similar level of impact.
-As such, performance gap widens with increasing correlation and sparsity.
-Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.
-
-In this analysis, cILR unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
-This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation.
-However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and cILR/CLR.
-Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity.
-Additionally, we did not observe any performance difference between the different variations of cILR.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
-    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
-Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure.
-Panel \textbf{(B)} presents performance on a classification task with AUC as the evaluation measure.
-cILR approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
-    \label{fig:6}
-\end{figure}
-
-%\begin{figure}[!ht]
-%    \centering
-%    \includegraphics[width=\textwidth]{figures/sim_pred_rsq.png}
-%    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on cILR,     %ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data %across sparsity levels, correlation, and signal-to-noise ratio.
-Panel \textbf{(A)} and \textbf{(B)} %represent results across different levels of model saturation (proportion of sets associated with the %outcome).
-cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
-%    \label{fig:sim_pred_rsq}
-%\end{figure}
-
-\subsubsection*{Real data evaluations}
-In addition to parametric simulations, we also assessed the performance of using cILR scores in predictive models with real data sets.
-Fig~\ref{fig:7} presents results for two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
-The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al.
-\cite{gevers2014} data set uses 16S rRNA gene sequencing, while the Nielsen et al \cite{nielsen2014} data set uses whole genome shotgun sequencing.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \linewidth]{figures/data_prediction_plot.png}
-    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
-Data sets used span both 16S rRNA gene sequencing (Gevers et al.
-\cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al.
-\cite{nielsen2014}).
-cILR performs better than GSVA and ssGSEA but not as well as CLR, with the exception of the whole genome sequencing data set.}
-    \label{fig:7}
-\end{figure}
-
-Similar to simulation experiments, we also fitted a naive random forest model using CILR, ssGSEA, GSVA, or CLR transformed variables as inputs, and use AUC as the performance criteria.
-Results also replicated that of the simulations, where across both data sets cILR and CLR methods provide much better performance than both GSVA or ssGSEA.
-Interestingly, the cILR approach performed better than CLR in the whole genome data set but did not perform as well in the 16S rRNA gene sequencing data set.
-However, these results indicate that cILR generated scores are informative, providing competitive performance when acting as inputs to disease predictive models.
-Most importantly, performance values are consistent across both simulated and real data sets.
-
-These results demonstrate that cILR generated scores are informative features in disease prediction tasks.
-Simulation results indicate that cILR methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach.
-Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
-
-\section*{Discussion}
-
-\subsection*{Inference with cILR}
-The formulation of cILR as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}.
-This allows for conducting inference with cILR even at the sample-level.
-We assessed the usage of cILR in this type of analysis by evaluating type I error and power across both simulation studies and real data applications.
-Most importantly, we demonstrated that our adjusted cILR approach was able to address the issue of variance inflation due to correlation \cite{wu2012} by controlling for type I error at the appropriate $\alpha$ level across different levels of simulated inter-taxa correlation (Fig~\ref{fig:2}) while conversely unadjusted cILR and the naive Wilcoxon rank sum test showed much higher rates of error.
-This is further encouraged in real data analysis where the false discovery rate was around $0.05$ when a collection of true null and true alternate hypotheses were tested.
-Unfortunately the trade-off of good type I error control is lower power.
-The conservativeness of the test attenuates with higher sparsity and correlation, where power was not approaching even 50\% even at the highest effect sizes.
-However, when the degree of such data features are reasonable, cILR will still be able to detect a reasonable proportion of samples with inflated counts.
-
-We also observed that choosing different distributional forms did not alter performance values for cILR.
-This runs contrary to our comparison analysis in Fig~\ref{fig:1}, where we demonstrated that the mixture normal distribution had superior fit compared to the simple normal for raw cILR scores computed under the global null.
-We hypothesized that this might be due to the difficulty in fitting mixture distribution to data using the expectation maximization algorithm, as the convergence rate is slow when there is high overlap between the mixtures, resulting in a small mixing coefficient for one of the components \cite{naim2012}.
-As such, in our implementation of cILR, in order to ensure convergence for the estimating procedure we increased the number of iterations while relaxing the tolerance parameter.
-Furthermore, there are also possible problems with our adjustment procedure for the mixture distribution that might impact overall fit.
-In order to combine the scale and location estimate of two mixture distributions, we fixed the overall mean, standard deviation, mixing coefficient and component-wise means and used an optimization procedure to find the component-wise variances.
-However, this means that we have one equation for the overall variance but two possible parameters to estimate.
-As such, there is no guaranteed unique solution to component-wise variances.
-We hypothesized that the instability and degeneracy in component-wise variance estimates might impact the fidelity of estimates at the tails of the distribution, thereby affecting inference.
-
-Despite these concerns, empirical results still indicate that cILR can confidently identify samples with inflated counts.
-The conservativeness of the correlation adjustment procedure ensures that significant results can be trusted by practitioners, even if cILR might not be able to exhaustively identify all samples with inflated counts.
-In situations where either the data is less sparse (e.g.
-containing a lot of core taxa that are prevalent across all samples), there is less inter-taxa correlation within the set (e.g.
-taxa that do not participate in common pathways but have shared characteristics like pathogenicity), or if the effect size is large, then cILR will still be able to produce reasonable power.
-A practitioner can use cILR to screen for samples for subsequent analysis that might involve significant costs, or perform hypothesis generation using a less stringent criteria alongside a multiple testing adjustment procedure (such as Benjamini-Hochberg \cite{benjamini1995}).
-
-\subsection*{Downstream analysis}
-The sample-level enrichment scores generated by the cILR method can be used in downstream analyses commonly performed in microbiome research: differential abundance testing and disease prediction.
-
-\subsubsection*{Differential abundance analysis}
-For differential abundance testing, we evaluated whether using cILR scores alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes.
-We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014} applied on data where taxa were aggregated using the naive sum method.
-We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}.
-Conversely, corncob is a newer method developed specifically for microbiome taxonomic data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.
-
-Surprisingly, we found some conflicting results when evaluating comparisons across parametric simulations and real data analysis.
-The performance of cILR was consistent across two evaluation criteria, demonstrating good type I error and respectable power.
-However, corncob and DESeq2 showed opposite effects: in simulation experiments, both methods show good type I error control but low power, while in real data analyses, conversely both corncob and DESeq2 showed inflated type I error but comparable power with respect to cILR methods.
-Despite such discrepancy, results still indicate good performance of cILR scores when used as inputs to downstream differential abundance analysis compared to using aggregated raw counts, even in methods designed to handle that type of data such as corncob and DESeq2.
-
-We hypothesized that the above behavior can arise due to issues with performing inference in the presence taxon-specific protocol biases \cite{mclaren2019}.
-According to McLaren et al., the observed relative abundance of taxa is different than the true relative abundance due to protocol bias, and importantly this bias is specific to each taxon \cite{mclaren2019}.
-This is especially true in the context of sum-based aggregations, where the resulting bias of the aggregated taxon are dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}).
-Conceptually, this means that measurement error for a taxon aggregate is different across samples as relative abundance of contributing changes, leading to issues when attempting to perform inference.
-As such, we expect methods like corncob or DESeq2 when performed on such aggregates to have inflated type I error compared to our taxon-ratio based approach.
-
-The bias model also helps explain differences in performance of DESeq2 and corncob in simulation analyses compared to real data.
-Our simulation protocol does not explicitly include bias in our formulation, and all taxa were generated from the same underlying distribution with similar variances across all samples (the only difference is in the mean value where a taxa is expected to have inflated counts).
-As such, we do not expect our simulated taxa to have any taxon-specific biases, which is not the case in real data settings.
-Therefore, we can expect DESeq2 and corncob to retain their expected type I error control in simulation analyses compared to real data.
-It is still surprising to see lower power for both methods in simulation analyses, which might be due to the fact that the evaluation protocol only considers default settings for both methods and did not attempt to optimize performance.
-
-The fact that the performance of cILR remains consistent across both simulation and real data analysis shows that cILR is invariant to taxon-specific biases.
-Furthermore, our evaluation indicates that even a simple difference in means test when used in combination with cILR scores can preserve type I error while maintaining good power.
-As such, a practioner can use cILR as a pre-processing step prior to performing a differential abundance test.
-
-\subsubsection*{Predictive models}
-For disease prediction, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using cILR generated scores as inputs.
-Similar to our inference analysis, we compared cILR against both ssGSEA and GSVA.
-Additionally, we also evaluated cILR with the approach where counts of a set were aggregated using sums and then centered log-ratio transformed.
-This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}.
-Results indicated that cILR produces good performance values across both real data analysis and simulation scenarios.
-Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
-Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome taxonomic data sets.
-This is consistent with our sample ranking analysis (Fig~\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
-However, cILR did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data.
-
-However, in simulation studies, this performance gap between CLR and cILR decreases with higher sparsity and correlation, especially in low effect saturation scenarios.
-Additionally, there are also downsides to applying CLR.
-First, the singular covariance matrix of CLR transformed variables is singular due to a sum to zero constraint \cite{gloor2017}, preventing the proper usage of approaches that rely on matrix decomposition.
-Second, the procedure still relies on using summation of counts prior to transformation, which means that we still can't compare across sets of different sizes, and any bias might still be propagated \cite{mclaren2019}.
-As such, despite benefits in performance for a naive random forest model, there is still space for using cILR as primary inputs into predictive models.
-
-Similar to other experiments in downstream usage of cILR, performance did not change with different underlying distributions, output types, or correlation status.
-This is surprising since we expect z-scores to perform better as they are able to capture the direction of an association.
-The fact that this effect persisted even onto our real data analysis suggests that this is not due to a deficiency of our simulation design.
-As such, practitioners who wish to use cILR in predictive models might be suited to use the settings that is the fastest to compute.
-
-Ultimately, results indicate that cILR can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity.
-Even though there exists situations where it might not provide maximum predictive values, the flexibility of cILR in various types of analyses enable even though in some scenarios it might not provide maximum predictive values.
-
-\subsection*{Limitations and future directions} 
-There are various limitations to our evaluation of cILR.
-First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome taxonomic data.
-There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation of prior studies.
-Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is not perfect.
-This is because the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut.
-As such, results from power analyses using this data set is only relative between the comparison methods.
-Finally, we assumed that taxa within a set are all equally associated with the outcome.
-This limits our ability to evaluate the performance of cILR when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.
-
-Our evaluation showed various drawbacks of the cILR method.
-First, inference with cILR is limited in being able to exhaustively detect all samples with significant inflated counts for a set in situations where there is a high degree of sparsity and inter-taxa correlation.
-Second, for downstream analyses, cILR might not always perform better than competing methods, especially when being used to generate inputs to predictive models.
-We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated with adjusting the mixture normal distribution.
-As such, we hope to refine the null distribution estimating procedure by either choosing a better distributional form, or to further constrain the optimization procedure of the mixture normal distribution by fixing the third and fourth moments.
-
-In addition, there are possible extensions cILR can we can consider to provide more flexibility across different data analysis scenarios in data analysis.
-First, cILR did not address the sparsity of microbiome taxonomic data and relies on a pseudocount to ensure log operations are valid.
-We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
-Second, cILR also treated all taxa within the set as equally contributing to the set.
-Incorporation of taxa-specific weights could reduce the influence of outliers, such as rare or highly invariant taxa.
-Finally, curating sets based on \emph{apriori} characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.
-
-\section*{Conclusion}
-Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomics data sets.
-However, limited work has been done developing set based methods specifically for microbiome relative abundance data.
-We introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level.
-We demonstrated that our method can control for type I error for significance testing at the sample level, while generated scores are also valid inputs in downstream analyses, including disease prediction and differential abundance.
-
-\section*{Acknowledgments}
-The authors thank Becky Lebeaux, Modupe Coker, Erika Dade, Jie Zhou, and Weston Viles for insightful comments and suggestions that greatly improved the paper.
-
-\nolinenumbers
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-% \bibliography{tax_agg}{}
-
 \begin{thebibliography}{10}
@@
 \end{thebibliography}
-
-\section*{Supporting information}
-
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Fig.}
-\label{S1_Fig}
-{\bf Computational performance of cILR.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
-Evaluation was performed on simulated data sets.
-For sample size analysis, only 10 sets were evaluated.
-For taxa set analysis, sample size was fixed at 1,000.
-Across all evaluations, the size of each taxa set was also fixed at 50.
-
-\paragraph*{S2 Fig.}
-\label{S2_Fig}
-{\bf Distribution of p-values.} Q-Q plot of 10,000 p-values compared against a uniform distribution bounded between 0 and 1.
-Evaluation was performed on simulated null data sets of 10,000 samples testing for enrichment of a set of size 50.
-For correlation of 0.5, p-values represent correlation adjusted cILR while for correlation of 0, p-values represent unadjusted cILR.
-
-\paragraph*{S1 File.}
-\label{S1_File}
-{\bf Supplemental derivations.} Includes additional details on addressing variance inflation due to correlation in cILR, as well as computational performance and p-value distribution of the method.
-
-\end{document}

### 2021-12-02 cbea_revised.tex
Updates from Overleaf

@@
 However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019a,li2015}.
-First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present.
-For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016}, allowing for simultaneous model fitting and variable selection.
-For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.
+The data is sparse, high-dimensional, and likely compositional \cite{gloor2017, li2019a, li2015}.
+Even though the these problems are challenging, a very approachable solution is to to use set-based analysis methods, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}.
+Aggregated variables can be less sparse, and testing on a smaller number of features can reduce the multiple-testing burden.
+As such, gene set testing methods have been shown to increase power and reproducibility of genomic analyses.
+Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more biologically informative interpretations.
 
-Second, the total number of reads varies across samples due to limitations of the sequencing instrument \cite{quinn2018b}, where normalization approaches has to be applied to allow for comparison of feature abundances across samples.
-However, microbiome data sets do not follow certain assumptions that enable the usage of methods from similar fields (such as RNA-seq) \cite{quinn2019}.
-For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes that do not differ in expression levels across samples.
-As such, practitioners often rely on total sum normalization to transform count data into relative proportions constrained to sum to one \cite{weiss2017}, which based on empirical studies showed better performance in analysis tasks \cite{mcknight2019}.
-This normalization schema imposes spurious negative correlation between the two
-Second, the number of reads obtained is constrained by the sequencing instrument at an arbitrary limit, and is inconsistent across samples, resulting in a variable number of total read counts per sample.
-Many normalization methods \cite{weiss2017} have been proposed to address these issues, including cross-applying methods from the gene expression literature \cite{mcmurdie2014}.
-However, these methods rely on assumptions specific to the original bulk RNA-seq data sets such as the presence of housekeeping genes with consistent expression levels \cite{love2014}, which might not be not true in the context of microbiome relative abundance data \cite{quinn2019,quinn2018b}.
-As such, microbiome taxonomic data is strictly compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another.
-Consequently, log-ratio transformations from the compositional data literature are often utilized \cite{aitchison1999}.
+There exists a diverse set of methods already developed in this field.
+Traditional set testing methods utilize the hypergeometric distribution to test for the overrepresentation of a gene set using a candidate list of genes screened based on a marginal model \cite{goeman2007}.
+Unfortunately, these approaches are sensitive to the differential expression test as well as the chosen threshold where genes can be part of the candidate list.
+Aggregate score methods, which are generally more preferred \cite{irizarry2009}, instead assigns a score for each gene set based on gene-specific statistics such as z-scores or fold change.
+Of these approaches, methods such as GSEA \cite{subramanian2005} performs a test for each gene set at the population level, summarizing information across all samples.
+Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020}, generate enrichment scores at the sample level and are more akin to a transformation.
+In addition to being able to screen for enriched sets per sample, this strategy also allows for the flexible incorporation of different downstream analyses, such as fitting prediction models, or performing dimension reduction.
 
-Third, the data are highly zero-inflated, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment).
-Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015}, or applying zero-inflated models \cite{chen2018, kaul2017}.
+%First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present.
+For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016a}, allowing for simultaneous model fitting and variable selection.
+For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018c} to address hypothesis testing burden.
+
+%Second, the total number of reads varies across samples due to limitations of the sequencing instrument \cite{quinn2018}.
+
+%Third, the data are highly sparse, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment).
+Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015a}, or applying zero-inflated models \cite{chen2018c, kaul2017}.
 Newer methods developed recently have focused on understanding the different types of zeros in the data, providing more sophisticated heuristics around when pseudocounts can be utilized \cite{kaul2017a}.
 
-Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}.
-Aggregated sets are less sparse than their constituent elements, and testing on a smaller number of variables reduces the multiple testing burden, thereby increasing power and reproducibility.
-Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations.
-There exists a diverse set of available methods developed to perform such analyses.
-More traditional set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest \cite{goeman2007}.
-Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values.
-The most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size.
-Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples.
-Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample level and are more akin to a transformation.
-This strategy allows for the flexible incorporation of different statistical techniques downstream, such as prediction models, as well as for visualization purposes in ordination plots.
+In microbiome research, even though no explicit enrichment analysis is performed, researchers often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum.
+However, despite this interest, there is limited research done to extend existing set-based methods to microbiome relative abundance data.
+Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}.
+However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used and p-value threshold.
+One of the primary challenges for adapting gene set analysis to the microbiome context is the compositional nature of the data.
+Sequencing technologies constrain the total number of reads, and samples are expected to have the same number of reads instead of DNA content \cite{quinn2019, quinn2018}.
+However, different samples still yield arbitrarily different total read counts \cite{morton2019, gloor2017}, suggesting the use of normalization methods to allow for proper comparison of feature abundances across samples.
+However, microbiome data sets do not follow certain assumptions that enable the cross-application of methods from similar fields (such as RNA-seq) \cite{quinn2019, quinn2018}.
+For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes with constant expression levels across samples.
+As such, practitioners often rely on total sum normalization to transform count data into relative proportions that sum to one \cite{weiss2017}.
+Empirical studies have supported this decision, showing increased performance of analysis tasks under this normalization schema \cite{mcknight2019}.
+Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another.
+Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1999} are motivated to address this issue.
 
-In microbiome research, even though no explicit enrichment analysis is performed, standard practice often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum by simple summation of abundances \cite{mclaren2019}.
-Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, it increases bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}.
-As such, there is a need for microbiome researchers to adopt more robust sample-level set enrichment methods.
-Unfortunately, limited work has been done to extend existing methods to be more specific to microbiome relative abundance data.
-Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}.
-However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used.
+Standard practice for aggregating variables in microbiome research uses element-wise summations \cite{mclaren2019}.
+Unfortunately, there are disadvantages to using this approach (referred to as amalgamations in the CoDA literature).
+First, inter-sample Aitchison distances computed on original parts are not preserved after amalgamation \cite{egozcue2005}.
+This means that cluster analyses might show different results depending on the level of amalgamation and differs from the those computed from original variables.
+Second, amalgamations do not allow for comparison between sets of different sizes within the same experimental condition since larger sets will have larger means and variances.
+Third, considering that each taxa has specific measurement biases \cite{mclaren2019}, an amalgamation based approach would make the bias of the amalgamated variable dependent on the relative abundance of the its constituents.
+In other words, if taxon $1$ has abundance $A_1$ and bias $B_1$, while taxon $2$ has abundance $A_2$ and bias $B_2$, then the bias of the aggregate variable (for example, a genera) is $(A_1B_1 + A_2B_2)/(A_1 + A_2)$ (see Appendix 1.
+from McLaren et al.
+\cite{mclaren2019}).
+This means that bias invariant approaches (such as analyses of ratios) would no longer be invariant when applied to amalgamated variables as bias now varies across samples.
+The alternative would be to multiply the proportions rather than to sum them \cite{egozcue2005}.
 
-Here, we present a novel method that generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}.
+Here, we present a taxon-set testing method for microbiome relative abundance data that addresses the aforementioned issues.
+Our approach generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020}.
 We leverage the concept of the $Q_1$ competitive hypothesis presented in Tian et al.
-\cite{tian2005}, which tests the null that the value of variables within a specific set is equal to the value of measured variables not in the set.
-The competitive null hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables.
-We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a multiplicative aggregation method that addresses the downsides of the naive summation-based method presented above \cite{mclaren2019, silverman2017}.
-The resulting method, titled competitive isometric log-ratio (CBEA), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing.
-These scores can then act as inputs to differential abundance and predictive modeling tasks downstream.
+\cite{tian2005} to formulate the enrichment of a set as the compositional balance \cite{rivera-pinto} of taxa within the set and remainder taxa using multiplication as the method of aggregating proportions \cite{egozcue2005}.
+This well-defined null hypothesis allows us to perform significance testing with interpretable results through estimating the empirical distribution of our statistic under the null that can also account for variance inflation due to inter-taxa correlation \cite{wu2012}.
 
-In the following sections, we provide the formulation of CBEA and discuss some statistical properties.
-We illustrate significance testing at the sample level using CBEA and evaluate type I error and power under different simulation scenarios and real data applications.
-We assess the informativeness of CBEA generated scores, and evaluate how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis.
-We compare the performance of CBEA in these respective tasks against standard microbiome taxonomic data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods.
-An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).
+In the following sections, we present our approach titled competitive balances for taxonomic erichment analysis (CBEA).
+First, we presented the step-by-step formulation of CBEA and discuss its statistical properties.
+Second, we detailed our evaluation strategy using both real data and parametric simulations, and the methods we're comparing against.
+Third, we presented results on enrichment testing using CBEA for single samples as well as at the population level.
+Fourth, we showed the performance of CBEA in downstream disease prediction.
+Finally, we discussed our results and the limitations of our method.
+An R package implementation of CBEA can be found on GitHub (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/CBEA}).
 
 \section*{Materials and Methods} \label{methods}
-\subsection*{Competitive Isometric Log-ratio (CBEA)}
+\subsection*{Competitive balances for taxonomic enrichment analysis (CBEA)}
 The CBEA method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}.
@@
 \subsection*{Properties of CBEA}
-\subsubsection*{CBEA and the Isometric Log Ratio Transformation}
+\subsubsection*{CBEA and balances of groups of parts}
 The CBEA statistic is a special instance of the isometric log-ratio transformation (ILR) \cite{egozcue2003}.
@@
 
-\subsubsection*{Statistical Properties of CBEA}
+\subsubsection*{Estimating the null distribution}
 We can perform significance testing on the CBEA statistic which corresponds to the null hypothesis that the center of the subcomposition defined by the set is equal to the center of the subcomposition defined by the complement of the set.

### 2021-12-03 cbea_revised.tex
New updates

@@
 The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host.
-Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019a}, and obesity \cite{aoun2020}.
+Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
 To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
@@
 \subsection*{Competitive balances for taxonomic enrichment analysis (CBEA)}
-The CBEA method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}.
+The CBEA method generates sample-specific enrichment scores for microbial sets using products of proportions \cite{egozcue2003}.
 Details on the computational implementation of CBEA can be found in the supplemental materials.
@@
 Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing.
+CBEA does not accept $X$ matrices with zeroes since it invalidates the log-ratio transformation.
+Users can generate a dense matrix $X$ using a method of choice, however the default mode for CBEA would be add a pseudocount of 1 if zeroes are detected in the matrix.
     \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxon $p$ to $m$ sets of interest.
@@
 \subsubsection*{CBEA and balances of groups of parts}
-The CBEA statistic is a special instance of the isometric log-ratio transformation (ILR) \cite{egozcue2003}.
-The standard ILR is a transformation method to address the negative correlation bias inherent in compositional data by providing an isometry between the $D$-dimensional simplex $\mathbb{S}^D$ and coordinates in the $D-1$ real space $\mathbb{R}^{D-1}$ \cite{egozcue2003,washburne2017}.
-This is accomplished by projecting the composition onto a chosen orthonormal basis in $\mathbb{R}$, which can be defined by a sequential binary partition (SBP) of the variables (e.g.
-a rooted phylogenetic tree).
-The ILR transformed variables are the coordinates of nodes within an SBP tree of the variables.
-Without loss of generalizability, in a given SBP with node $i$ splitting variables between sets $R$ and $S$, we have the ILR coordinate $x^{*}_{i}$ as:
+The CBEA statistic is based on the multiplication-based aggregation approach used to calculate balances between groups of parts \cite{egozcue2005}.
+These balances are computed using the isometric log ratio (ILR) transformation \cite{egozcue2003} formula.
+For a given balance $i$ splitting variables across sets $R$ and and $S$, we have the balance coordinate $x^{*}_{i}$ as:
 \begin{equation}\label{ilr_standard}
@@
 where $r$ and $s$ are the cardinalities of sets $R$ and $S$ respectively, $g(z)$ is the geometric mean, and $X_{j}$ are values of the original predictors with indexes defined by membership in $R$ and $S$.
-The ILR confer many important benefits.
-First, ILR coordinates exist in real space, whereby common statistical methods can be used.
-Second, ILR aggregated variables preserve inter-sample distances before and after aggregation \cite{egozcue2005}.
-Third, ILR variables are not constrained to sum to 0 as that of the centered log-ratio transformation, resulting in a covariance matrix that is not singular \cite{egozcue2003}.
 
-The usage of the ILR statistic is not uncommon in the microbiome literature.
-They are usually termed ``compositional balances", and have been leveraged in many recent approaches in variable transformation \cite{washburne2017,silverman2017,morton2017}.
-The CBEA formulation in Eq~\eqref{main_eq} is a special case of Eq~\eqref{ilr_standard} defined on a node that splits the taxa into two disjoint sets, one representing the set of interest, the other representing the remaining taxa.
-As such, the CBEA transformation inherits the properties of the ILR as a log-ratio method applicable to compositional data sets.
-However, unlike the ILR and its variants \cite{silverman2017, morton2017, washburne2017}, the axes defined by each CBEA set are not orthogonal (since the balances are mutually exclusive between sets and do not belong in the same SBP).
-Hence, a correlation can exist between CBEA aggregated variables.
+CBEA belongs to a set of methods that seeks to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto, silverman2017, morton2017}.
+Unlike methods such as PhILR \cite{silverman2017}, CBEA does not provide a sequential binary partition that forms the basis for ILR procedure \cite{egozcue2003} and is therefore not a subclass of ILR.
+A similar method to CBEA would be phylofactor \cite{washburne2017}.
+However, instead of performing an optimization procedure to identify interesting balances, CBEA constructs balances apriori using pre-defined sets, and formulates the enrichment of a set as the scaled log-ratio between the center of the subcomposition represented by microbes within the set and the center of the subcomposition represented by remainder taxa.
+This formulation aligns with the $Q_1$ null hypothesis from the gene set testing literature \cite{tian2005}.
 
 \subsubsection*{Estimating the null distribution}
-We can perform significance testing on the CBEA statistic which corresponds to the null hypothesis that the center of the subcomposition defined by the set is equal to the center of the subcomposition defined by the complement of the set.
-This is equivalent to the $Q_1$ competitive null hypothesis in the gene set testing literature \cite{tian2005} where the enrichment of a gene set is defined with respect to genes outside the set.
-
-We can apply prior usage of the ILR statistic in hypothesis testing to CBEA by assuming that the null distribution of CBEA follows a standard normal distribution \cite{egozcue2005}.
+We can assume that the CBEA statistic, similar to other log-ratio based transforms, follows a normal distribution \cite{egozcue2003, aitchison1980}.
 However, when applying CBEA for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses.
@@
 As such, the null distribution is characterized by scores computed on sets of equal size with randomly drawn taxa.
+
 However, null distribution based on taxa-permutation is sensitive to inter-taxa correlations within the set \cite{wu2012}.
@@
 
-\subsection*{Evaluation}
+\section*{Evaluation}
 All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).
 
-\subsubsection*{Parametric Simulations}  
+\subsection*{Simulation Analysis}
+\subsubsection*{Data generation model}  
 To address the performance of CBEA for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
@@
 
-\noindent \textbf{Single Sample Enrichment}: To assess type I error rate and power for enrichment significance testing at the sample level, we simulated data based on the schema above, and assessed enrichment for one focal set.
+\subsubsection*{Simulation scenarios for enrichment analysis at the sample level}
+To assess type I error rate and power for enrichment significance testing at the sample level, we simulated data based on the schema above, and assessed enrichment for one focal set.
 Type I error was obtained under the global null as the number of samples where the null hypothesis was rejected at $\alpha = 0.05$ over the total number of samples (which represents the total number of hypotheses tested).
@@
 
-For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUROC/AUC).
+For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUC).
 This is a strategy used in Frost \cite{frost2020a} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
@@
 
-\noindent \textbf{Differential Abundance Analysis}: To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition.
+\subsubsection*{Simulation scenarios for enrichment analysis at the population level}
+To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition.
 Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null.
@@
 
-\noindent \textbf{Prediction}: To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}.
+\subsubsection*{Simulation scenarios for downstream prediction}
+To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}.
 For binary outcomes, we use AUC similar to the classification analyses above.
@@
 
-\subsubsection*{Real Datasets}
-
+\subsection*{Evaluation using real data}
 In addition to simulation analyses, we also evaluated our method using real data sets based on both 16S rRNA gene sequencing and whole-genome sequencing.
-All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.
+We follow the benchmark set by Geistlinger et al.
+\cite{geistlinger} All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.
 
@@
 
-\bibliography{mac_taxagg}{}
+\bibliography{references}{}
 

### 2021-12-03 cbea_revised.tex

@@
 
-CBEA belongs to a set of methods that seeks to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto, silverman2017, morton2017}.
+CBEA belongs to a set of methods that seeks to leverage compositional balances for the analysis of microbiome data \cite{washburne2017a, rivera-pinto, silverman2017, morton2017}.
 Unlike methods such as PhILR \cite{silverman2017}, CBEA does not provide a sequential binary partition that forms the basis for ILR procedure \cite{egozcue2003} and is therefore not a subclass of ILR.
-A similar method to CBEA would be phylofactor \cite{washburne2017}.
+A similar method to CBEA would be phylofactor \cite{washburne2017a}.
 However, instead of performing an optimization procedure to identify interesting balances, CBEA constructs balances apriori using pre-defined sets, and formulates the enrichment of a set as the scaled log-ratio between the center of the subcomposition represented by microbes within the set and the center of the subcomposition represented by remainder taxa.

### 2021-12-03 cbea_revised.tex

@@
 Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another.
-Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1999} are motivated to address this issue.
+Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1994} are motivated to address this issue.
 

### 2021-12-03 cbea_revised.tex

@@
 For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUC).
-This is a strategy used in Frost \cite{frost2020a} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
+This is a strategy used in Frost \cite{frost2020} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
 whether samples with inflated counts are highly ranked using estimated scores).
@@
 We follow the benchmark set by Geistlinger et al.
-\cite{geistlinger} All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.
+\cite{geistlinger2021} All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.
 
@@
 First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome taxonomic data.
-There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation of prior studies.
+There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016a} have been used in the evaluation of prior studies.
 Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is not perfect.

### 2021-12-03 cbea_revised.tex

@@
 We leverage the concept of the $Q_1$ competitive hypothesis presented in Tian et al.
-\cite{tian2005} to formulate the enrichment of a set as the compositional balance \cite{rivera-pinto} of taxa within the set and remainder taxa using multiplication as the method of aggregating proportions \cite{egozcue2005}.
+\cite{tian2005} to formulate the enrichment of a set as the compositional balance \cite{rivera-pinto2018} of taxa within the set and remainder taxa using multiplication as the method of aggregating proportions \cite{egozcue2005}.
 This well-defined null hypothesis allows us to perform significance testing with interpretable results through estimating the empirical distribution of our statistic under the null that can also account for variance inflation due to inter-taxa correlation \cite{wu2012}.
@@
 
-CBEA belongs to a set of methods that seeks to leverage compositional balances for the analysis of microbiome data \cite{washburne2017a, rivera-pinto, silverman2017, morton2017}.
+CBEA belongs to a set of methods that seeks to leverage compositional balances for the analysis of microbiome data \cite{washburne2017a, rivera-pinto2018, silverman2017, morton2017}.
 Unlike methods such as PhILR \cite{silverman2017}, CBEA does not provide a sequential binary partition that forms the basis for ILR procedure \cite{egozcue2003} and is therefore not a subclass of ILR.
-A similar method to CBEA would be phylofactor \cite{washburne2017a}.
+A similar method to CBEA would be phylofactor \cite{washburne2017}.
 However, instead of performing an optimization procedure to identify interesting balances, CBEA constructs balances apriori using pre-defined sets, and formulates the enrichment of a set as the scaled log-ratio between the center of the subcomposition represented by microbes within the set and the center of the subcomposition represented by remainder taxa.

### 2021-12-03 cbea_revised.tex
Updates from Overleaf

@@
 
-CBEA belongs to a set of methods that seeks to leverage compositional balances for the analysis of microbiome data \cite{washburne2017a, rivera-pinto2018, silverman2017, morton2017}.
+CBEA belongs to a set of methods that seeks to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto2018, silverman2017, morton2017}.
 Unlike methods such as PhILR \cite{silverman2017}, CBEA does not provide a sequential binary partition that forms the basis for ILR procedure \cite{egozcue2003} and is therefore not a subclass of ILR.

### 2021-12-03 cbea_revised.tex
Updates from Overleaf

@@
 For continuous outcomes, we used root mean squared error (RMSE).
-All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020} suite of packages.
+All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020b} suite of packages.
 Across both learning tasks, we varied sparsity ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation ($\rho = 0, 0.2, 0.5$).
@@
 
-\bibliography{references}{}
+%\bibliography{references}{}
+\begin{thebibliography}{10}
+
+\bibitem{proctor2019}
+Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
+\newblock The {{Integrative Human Microbiome Project}}.
+\newblock Nature. 2019;569(7758):641--648.
+\newblock doi:{10.1038/s41586-019-1238-8}.
+
+\bibitem{sharma2019a}
+Sharma S, Tripathi P.
+\newblock Gut Microbiome and Type 2 Diabetes: Where We Are and Where to Go?
+\newblock The Journal of Nutritional Biochemistry. 2019;63:101--108.
+\newblock doi:{10.1016/j.jnutbio.2018.10.003}.
+
+\bibitem{aoun2020}
+Aoun A, Darwish F, Hamod N.
+\newblock The {{Influence}} of the {{Gut Microbiome}} on {{Obesity}} in
+  {{Adults}} and the {{Role}} of {{Probiotics}}, {{Prebiotics}}, and
+  {{Synbiotics}} for {{Weight Loss}}.
+\newblock Preventive Nutrition and Food Science. 2020;25(2):113--123.
+\newblock doi:{10.3746/pnf.2020.25.2.113}.
+
+\bibitem{cho2012}
+Cho I, Blaser MJ.
+\newblock The Human Microbiome: At the Interface of Health and Disease.
+\newblock Nature Reviews Genetics. 2012;13(4):260--270.
+\newblock doi:{10.1038/nrg3182}.
+
+\bibitem{callahan2016}
+Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
+\newblock {{DADA2}}: High-Resolution Sample Inference from {{Illumina}}
+  Amplicon Data.
+\newblock Nature Methods. 2016;13(7):581--583.
+\newblock doi:{10.1038/nmeth.3869}.
+
+\bibitem{truong2015}
+Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
+\newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
+\newblock Nature Methods. 2015;12(10):902--903.
+\newblock doi:{10.1038/nmeth.3589}.
+
+\bibitem{li2019a}
+Li H.
+\newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
+  {{Metagenomics}}.
+\newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
+  2019. p. 977--550.
+
+\bibitem{li2015}
+Li H.
+\newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
+  Compositional Data Analysis}}.
+\newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
+\newblock doi:{10.1146/annurev-statistics-010814-020351}.
+
+\bibitem{gloor2017}
+Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
+\newblock Microbiome {{Datasets Are Compositional}}: And {{This Is Not
+  Optional}}.
+\newblock Frontiers in Microbiology. 2017;8.
+\newblock doi:{10.3389/fmicb.2017.02224}.
+
+\bibitem{khatri2012}
+Khatri P, Sirota M, Butte AJ.
+\newblock Ten {{Years}} of {{Pathway Analysis}}: Current {{Approaches}} and
+  {{Outstanding Challenges}}.
+\newblock PLOS Computational Biology. 2012;8(2):e1002375.
+\newblock doi:{10.1371/journal.pcbi.1002375}.
+
+\bibitem{goeman2007}
+Goeman JJ, B{\"u}hlmann P.
+\newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
+  Issues.
+\newblock Bioinformatics. 2007;23(8):980--987.
+\newblock doi:{10.1093/bioinformatics/btm051}.
+
+\bibitem{subramanian2005}
+Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
+\newblock Gene Set Enrichment Analysis: A Knowledge-Based Approach for
+  Interpreting Genome-Wide Expression Profiles.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(43):15545--15550.
+\newblock doi:{10.1073/pnas.0506580102}.
+
+\bibitem{ashburner2000}
+Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
+\newblock Gene {{Ontology}}: Tool for the Unification of Biology.
+\newblock Nature genetics. 2000;25(1):25--29.
+\newblock doi:{10.1038/75556}.
+
+\bibitem{irizarry2009}
+Irizarry RA, Wang C, Zhou Y, Speed TP.
+\newblock Gene {{Set Enrichment Analysis Made Simple}}.
+\newblock Statistical methods in medical research. 2009;18(6):565--575.
+\newblock doi:{10.1177/0962280209351908}.
+
+\bibitem{hanzelmann2013}
+H{\"a}nzelmann S, Castelo R, Guinney J.
+\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
+  {{RNA}}-{{Seq}} Data.
+\newblock BMC Bioinformatics. 2013;14(1):7.
+\newblock doi:{10.1186/1471-2105-14-7}.
+
+\bibitem{frost2020}
+Frost HR.
+\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
+  Method for Cell-Specific Gene Set Scoring.
+\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
+\newblock doi:{10.1093/nar/gkaa582}.
+
+\bibitem{chong2020}
+Chong J, Liu P, Zhou G, Xia J.
+\newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
+  Functional, and Meta-Analysis of Microbiome Data.
+\newblock Nature Protocols. 2020;15(3):799--821.
+\newblock doi:{10.1038/s41596-019-0264-1}.
+
+\bibitem{quinn2019}
+Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
+\newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
+\newblock GigaScience. 2019;8(giz107).
+\newblock doi:{10.1093/gigascience/giz107}.
+
+\bibitem{quinn2018}
+Quinn TP, Erb I, Richardson MF, Crowley TM.
+\newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
+\newblock Bioinformatics. 2018;34(16):2870--2878.
+\newblock doi:{10.1093/bioinformatics/bty175}.
+
+\bibitem{morton2019}
+Morton JT, Marotz C, Washburne A, Silverman J, Zaramela LS, Edlund A, et~al.
+\newblock Establishing Microbial Composition Measurement Standards with
+  Reference Frames.
+\newblock Nature Communications. 2019;10(1).
+\newblock doi:{10.1038/s41467-019-10656-5}.
+
+\bibitem{love2014}
+Love MI, Huber W, Anders S.
+\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
+  Data with {{DESeq2}}.
+\newblock Genome Biology. 2014;15(12):550.
+\newblock doi:{10.1186/s13059-014-0550-8}.
+
+\bibitem{weiss2017}
+Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
+\newblock Normalization and Microbial Differential Abundance Strategies Depend
+  upon Data Characteristics.
+\newblock Microbiome. 2017;5(1).
+\newblock doi:{10.1186/s40168-017-0237-y}.
+
+\bibitem{mcknight2019}
+McKnight DT, Huerlimann R, Bower DS, Schwarzkopf L, Alford RA, Zenger KR.
+\newblock Methods for Normalizing Microbiome Data: An Ecological Perspective.
+\newblock Methods in Ecology and Evolution. 2019;10(3):389--400.
+\newblock doi:{10.1111/2041-210X.13115}.
+
+\bibitem{aitchison1994}
+Aitchison J.
+\newblock Principles of Compositional Data Analysis.
+\newblock Lecture Notes-Monograph Series. 1994; p. 73--81.
+
+\bibitem{mclaren2019}
+McLaren MR, Willis AD, Callahan BJ.
+\newblock Consistent and Correctable Bias in Metagenomic Sequencing
+  Experiments.
+\newblock eLife. 2019;8:e46923.
+\newblock doi:{10.7554/eLife.46923}.
+
+\bibitem{egozcue2005}
+Egozcue JJ, {Pawlowsky-Glahn} V.
+\newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2005;37(7):795--828.
+\newblock doi:{10.1007/s11004-005-7381-9}.
+
+\bibitem{tian2005}
+Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
+\newblock Discovering Statistically Significant Pathways in Expression
+  Profiling Studies.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(38):13544--13549.
+\newblock doi:{10.1073/pnas.0506577102}.
+
+\bibitem{rivera-pinto2018}
+{Rivera-Pinto} J, Egozcue JJ, {Pawlowsky-Glahn} V, Paredes R, {Noguera-Julian}
+  M, Calle ML.
+\newblock Balances: A {{New Perspective}} for {{Microbiome Analysis}}.
+\newblock mSystems. 2018;3(4):e00053--18.
+\newblock doi:{10.1128/mSystems.00053-18}.
+
+\bibitem{wu2012}
+Wu D, Smyth GK.
+\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
+  Correlation.
+\newblock Nucleic Acids Research. 2012;40(17):e133.
+\newblock doi:{10.1093/nar/gks461}.
+
+\bibitem{egozcue2003}
+Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcel{\'o}-Vidal} C.
+\newblock Isometric {{Logratio Transformations}} for {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2003;35(3):279--300.
+\newblock doi:{10.1023/A:1023818214614}.
+
+\bibitem{quast2013}
+Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
+\newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
+  Processing and Web-Based Tools.
+\newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
+\newblock doi:{10.1093/nar/gks1219}.
+
+\bibitem{delignette-muller2015}
+{Delignette-Muller} ML, Dutang C.
+\newblock Fitdistrplus: An {{R}} Package for Fitting Distributions.
+\newblock Journal of Statistical Software. 2015;64(4):1--34.
+
+\bibitem{benaglia2009}
+Benaglia T, Chauveau D, Hunter DR, Young D.
+\newblock Mixtools: An {{R}} Package for Analyzing Finite Mixture Models.
+\newblock Journal of Statistical Software. 2009;32(6):1--29.
+
+\bibitem{washburne2017}
+Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL.
+\newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
+  Associations in Microbiome Datasets.
+\newblock PeerJ. 2017; p.~26.
+
+\bibitem{silverman2017}
+Silverman JD, Washburne AD, Mukherjee S, David LA.
+\newblock A Phylogenetic Transform Enhances Analysis of Compositional
+  Microbiota Data.
+\newblock eLife. 2017;6:e21887.
+\newblock doi:{10.7554/eLife.21887}.
+
+\bibitem{morton2017}
+Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
+  et~al.
+\newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
+\newblock mSystems. 2017;2(1):e00162--16.
+\newblock doi:{10.1128/mSystems.00162-16}.
+
+\bibitem{aitchison1980}
+Aitchison J, Shen SM.
+\newblock Logistic-Normal Distributions:Some Properties and Uses.
+\newblock Biometrika. 1980;67(2):261--272.
+\newblock doi:{10.1093/biomet/67.2.261}.
+
+\bibitem{efron2004}
+Efron B.
+\newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
+\newblock Journal of the American Statistical Association.
+  2004;99(465):96--104.
+\newblock doi:{10.1198/016214504000000089}.
+
+\bibitem{calgaro2020}
+Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
+\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
+  and Metagenomics Applied to Microbiome Data.
+\newblock Genome Biology. 2020;21(1):191.
+\newblock doi:{10.1186/s13059-020-02104-1}.
+
+\bibitem{cario1997}
+Cario MC.
+\newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
+  Distributions}} and {{Correlation Matrix}}; 1997.
+
+\bibitem{agresti1998}
+Agresti A, Coull BA.
+\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
+  Estimation}} of {{Binomial Proportions}}.
+\newblock The American Statistician. 1998;52(2):119--126.
+\newblock doi:{10.2307/2685469}.
+
+\bibitem{delong1988}
+DeLong ER, DeLong DM, {Clarke-Pearson} DL.
+\newblock Comparing the Areas under Two or More Correlated Receiver Operating
+  Characteristic Curves: A Nonparametric Approach.
+\newblock Biometrics. 1988;44(3):837--845.
+
+\bibitem{hawinkel2019}
+Hawinkel S, Mattiello F, Bijnens L, Thas O.
+\newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
+  Control the False Discovery Rate.
+\newblock Briefings in Bioinformatics. 2019;20(1):210--221.
+\newblock doi:{10.1093/bib/bbx104}.
+
+\bibitem{breiman2001}
+Breiman L.
+\newblock Random {{Forests}}.
+\newblock Machine Learning. 2001;45(1):5--32.
+\newblock doi:{10.1023/A:1010933404324}.
+
+\bibitem{kuhn2020b}
+Kuhn M, Wickham H.
+\newblock {{Tidymodels}}: a collection of packages for modeling and machine learning using tidyverse        principles; 2021.
+\newblock url:{https://www.tidymodels.org}
+
+\bibitem{xiao2018}
+Xiao J, Chen L, Yu Y, Zhang X, Chen J.
+\newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
+  {{Predictive Modeling}} of {{Microbial Community Data}}.
+\newblock Frontiers in Microbiology. 2018;9.
+\newblock doi:{10.3389/fmicb.2018.03112}.
+
+\bibitem{dong2020}
+Dong M, Li L, Chen M, Kusalik A, Xu W.
+\newblock Predictive Analysis Methods for Human Microbiome Data with
+  Application to {{Parkinson}}'s Disease.
+\newblock PLOS ONE. 2020;15(8):e0237779.
+\newblock doi:{10.1371/journal.pone.0237779}.
+
+\bibitem{geistlinger2021}
+Geistlinger L, Csaba G, Santarelli M, Ramos M, Schiffer L, Turaga N, et~al.
+\newblock Toward a Gold Standard for Benchmarking Gene Set Enrichment Analysis.
+\newblock Briefings in bioinformatics. 2021;22(1):545--556.
+
+\bibitem{pasolli2017}
+Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
+\newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
+\newblock Nature Methods. 2017;14(11):1023--1024.
+\newblock doi:{10.1038/nmeth.4468}.
+
+\bibitem{schiffer2019}
+Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
+\newblock {{HMP16SData}}: Efficient Access to the Human Microbiome Project
+  through Bioconductor.
+\newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
+
+\bibitem{gonzalez2018}
+Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
+  Ackermann G, et~al.
+\newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
+\newblock Nature Methods. 2018;15(10):796--798.
+\newblock doi:{10.1038/s41592-018-0141-9}.
+
+\bibitem{consortium2012}
+Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
+  et~al.
+\newblock Structure, Function and Diversity of the Healthy Human Microbiome.
+\newblock Nature. 2012;486(7402):207--214.
+\newblock doi:{10.1038/nature11234}.
+
+\bibitem{thurnheer2016}
+Thurnheer T, Bostanci N, Belibasakis GN.
+\newblock Microbial Dynamics during Conversion from Supragingival to
+  Subgingival Biofilms in an in Vitro Model.
+\newblock Molecular Oral Microbiology. 2016;31(2):125--135.
+\newblock doi:{10.1111/omi.12108}.
+
+\bibitem{beghini2019}
+Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
+\newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
+  in the {{New York City Health}} and {{Nutrition Examination Study}}.
+\newblock Annals of Epidemiology. 2019;34:18--25.e3.
+\newblock doi:{10.1016/j.annepidem.2019.03.005}.
+
+\bibitem{matteocalgaro2020}
+Calgaro M. Mcalgaro93/Sc2meta: Paper {{Release}}; 2020.
+\newblock Zenodo.
+
+\bibitem{nielsen2014}
+Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
+\newblock Identification and Assembly of Genomes and Genetic Elements in
+  Complex Metagenomic Samples without Using Reference Genomes.
+\newblock Nature Biotechnology. 2014;32(8):822--828.
+\newblock doi:{10.1038/nbt.2939}.
+
+\bibitem{gevers2014}
+Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
+  et~al.
+\newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
+  {{Disease}}.
+\newblock Cell Host \& Microbe. 2014;15(3):382--392.
+\newblock doi:{10.1016/j.chom.2014.02.005}.
+
+\bibitem{barbie2009}
+Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
+\newblock Systematic {{RNA}} Interference Reveals That Oncogenic
+  {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
+\newblock Nature. 2009;462(7269):108--112.
+\newblock doi:{10.1038/nature08460}.
+
+\bibitem{mcmurdie2014}
+McMurdie PJ, Holmes S.
+\newblock Waste {{Not}}, {{Want Not}}: Why {{Rarefying Microbiome Data Is
+  Inadmissible}}.
+\newblock PLOS Computational Biology. 2014;10(4):e1003531.
+\newblock doi:{10.1371/journal.pcbi.1003531}.
+
+\bibitem{martin2020}
+Martin BD, Witten D, Willis AD.
+\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
+  Regression.
+\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
+\newblock doi:{10.1214/19-AOAS1283}.
+
+\bibitem{naim2012}
+Naim I, Gildea D.
+\newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
+  {{Unbalanced Mixing Coefficients}}.
+\newblock Proceedings of the 29th International Coference on International
+  Conference on Machine Learning. 2012; p.~8.
+\newblock doi:{10.5555/3042573.3042756}.
+
+\bibitem{benjamini1995}
+Benjamini Y, Hochberg Y.
+\newblock Controlling the {{False Discovery Rate}}: A {{Practical}} and
+  {{Powerful Approach}} to {{Multiple Testing}}.
+\newblock Journal of the Royal Statistical Society Series B (Methodological).
+  1995;57(1):289--300.
+
+\bibitem{wu2016a}
+Wu C, Chen J, Kim J, Pan W.
+\newblock An Adaptive Association Test for Microbiome Data.
+\newblock Genome Medicine. 2016;8(1):56.
+\newblock doi:{10.1186/s13073-016-0302-3}.
+
+\bibitem{martin-fernandez2012}
+{Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
+  {Palarea-Albaladejo} J.
+\newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
+  Classical and Robust Approaches.
+\newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
+\newblock doi:{10.1016/j.csda.2012.02.012}.
+
+\bibitem{kaul2017a}
+Kaul A, Davidov O, Peddada SD.
+\newblock Structural Zeros in High-Dimensional Data with Applications to
+  Microbiome Studies.
+\newblock Biostatistics. 2017;18(3):422--433.
+\newblock doi:{10.1093/biostatistics/kxw053}.
+
+\end{thebibliography}
 

### 2022-01-26 cbea_revised.tex
Updates from Overleaf

@@
 \usepackage{float}
+
+% Adding overline macro 
+\newcommand{\overbar}[1]{\mkern 2mu\overline{\mkern-2mu#1\mkern-20mu}\mkern 20mu}
 
@@
 As such, gene set testing methods have been shown to increase power and reproducibility of genomic analyses.
-Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more biologically informative interpretations.
+Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analysis also allows for more biologically informative interpretations.
 
 There exists a diverse set of methods already developed in this field.
-Traditional set testing methods utilize the hypergeometric distribution to test for the overrepresentation of a gene set using a candidate list of genes screened based on a marginal model \cite{goeman2007}.
-Unfortunately, these approaches are sensitive to the differential expression test as well as the chosen threshold where genes can be part of the candidate list.
+Traditional methods utilize the hypergeometric distribution to test for the overrepresentation of a gene set using a candidate list of genes screened based on a marginal model \cite{goeman2007}.
+Unfortunately, these approaches are sensitive to the differential expression test as well as the chosen threshold when trying to select genes for the candidate list.
 Aggregate score methods, which are generally more preferred \cite{irizarry2009}, instead assigns a score for each gene set based on gene-specific statistics such as z-scores or fold change.
@@
 
-In microbiome research, even though no explicit enrichment analysis is performed, researchers often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum.
+In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate taxa to higher Linnean classification levels such as genus, family, or phylum.
 However, despite this interest, there is limited research done to extend existing set-based methods to microbiome relative abundance data.
@@
 The CBEA method generates sample-specific enrichment scores for microbial sets using products of proportions \cite{egozcue2003}.
-Details on the computational implementation of CBEA can be found in the supplemental materials.
+Details on the computational implementation of CBEA can be found in the Supplementary Materials.
 The CBEA method takes two inputs:
@@
 The effect is most dramatic under both high inter-taxa correlation and sparsity.
-This heterogeneity further supports the decision to estimate an empirical null distribution, similar to Efron \cite{efron2004}.
+This heterogeneity further supports the decision to estimate an empirical null distribution, as suggested by Efron \cite{efron2004}.
 
@@
 Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null.
-To address this issue, we also evaluated a mixture distribution of two normal components.
+To address this issue, we also evaluated a two-component normal mixture distribution.
 Panel B of Fig~\ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on CBEA scores in simulation scenarios under the global null.
@@
 
-However, null distribution based on taxa-permutation is sensitive to inter-taxa correlations within the set \cite{wu2012}.
-Since the permutation procedure does not preserve correlation structures, estimating parameters from empirical scores on permuted data will underestimate the variance inflation due to correlation.
-We account for this by combining the mean estimate from permuted data with the variance estimate from unpermuted data, where the inter-taxa correlation structure remains undisturbed.
-However, this procedure assumes that the variance of CBEA is equal under both the null and alternate hypotheses.
+\subsubsection*{Variance inflation due to inter-taxa correlation}  
+When taxa within a set are highly correlated, the variance of the sample mean of taxon-wise statistics is inflated.
+Without loss of generalizability, for a set of taxa with taxa-specific statistics $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:
+\begin{equation} \label{vareq:1}
+    Var(\bar{x}) = \frac{1}{m^2}\left(\sum_{i = 1}(\sigma_i^2) + \sum_{i < j}\rho_{ij}\sigma_i\sigma_j\right)
+\end{equation}
+where $\sigma_i$ is the standard deviation of taxon $i$ while $\rho_{ij}$ is the correlation between $i$ and $j$.
+The second term of (\ref{vareq:1}) is the correlation dependent variance component, which goes to 0 if there is no correlation.
+The CBEA statistic follows a similar pattern.
+Since the geometric mean of a set of variables is equivalent to the exponential of the arithmetic mean of their logarithms, we can re-write CBEA score for a set $k$ with size $K$ as follows:
+\begin{equation}\label{vareq:2}
+    M_{i,k} = \sqrt{\frac{K(p - K)}{K + (p - K)}} \left( \overbar{\log{X_{i,j|j \in K}}} - \overbar{\log{X_{i,j|j \notin K}}} \right)   
+\end{equation}
+where $p$ is the overall number of taxa, $j$ is the index of a taxa and $K$ is the set of indices of taxa in set $k$.
+The CBEA statistic then looks similar to a t-statistic for difference in means of log-transformed proportions.
+As such, the pooled variance of cILR is dependent on the variance inflation of both mean components $\overbar{\log{X_{i,j|j \in K}}}$ and $\overbar{\log{X_{i,j|j \notin K}}}$.
+The result of this variance inflation is inflated type I error since highly correlated sets are also detected as significantly enriched.
+
+However, as Wu et al.
+\cite{wu2012} showed, performing column permutation to estimate the null distribution of a competitive test statistic doesn't allow for adequate capture of this variance inflation factor since the permutation procedure disrupts the natural correlation structure of the original variables.
+It is important to address this problem since there is strong inter-taxa correlation within the microbiome \cite{kurtz2015a}.
+Our strategy for addressing issue is location (or mean) estimate from the column permuted raw score matrix with the spread (or variance) estimate from the original un-permuted scores.
+This still allows us to leverage the null generated via column permutation while using the proper variance estimate taken from scores where the correlation structure has not been disrupted.
+As such, this procedure assumes that the variance of the test statistic under the alternate hypothesis is the same as that of the null.
+Details of the computational implementation to this estimation process can be found in the Supplementary Materials.
+
+However, set-based analysis is an exploratory approach that can help generate functionally informative hypotheses, and as such users might not want strict type I error control in favor of higher power.
+This is especially true for competitive hypotheses, where its stricter formulation compared to the self-contained approach implies that the test naturally has lower power \cite{goeman2007, ackerman2009}.
+Furthermore, sets that are highly correlated compared to background can be biologically relevant.
+Therefore, CBEA provides an option for users to specify whether correlation adjustment is desired.
 
 \section*{Evaluation}
+We based our evaluation strategy on gene set testing benchmarking standards set by Geistlinger et al.
+\cite{geistlinger2021} and utilized the same approaches whenever possible.
+All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.
 All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).
 
-\subsection*{Simulation Analysis}
-\subsubsection*{Data generation model}  
-To address the performance of CBEA for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
-Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
-\begin{equation}
-    \mathbf{X}_{ij} =
-      \begin{cases}
-        0 & \text{with probability $p_j$}\\
-        \mathbf{NB}(\mu_j, \phi_j) & \text{with probability $1 - p_j$}\\
-      \end{cases}       
-\end{equation}
-
-where $\mu_j$ and $\phi_j$ are mean and dispersion parameters, respectively.
-To incorporate a flexible correlation structure into our simulation model, we utilized the NorTA (Normal to Anything) method \cite{cario1997}.
-Given an $n$ by $p$ matrix of values $\mathbf{U}$ sampled from multivariate normal distribution with correlation matrix $\mathbf{\rho}$, we can generate target microbiome count vector $\mathbf{X_{.j}}$ for taxa $j$ following the marginal distribution $\mathbf{NB}$ characterized by the negative binomial cumulative distribution function $\mathbb{F_{\mathbf{NB}}}$:
-\begin{equation}
-    \mathbf{X}_{.j} = \mathbb{F_{\mathbf{NB}}}^{-1}(\Phi_{U_i})
-\end{equation}
-In this instance, for each taxon $j$, we set elements in $\mathbf{U}_{.j}$ to be zero with probability $p_j$ and applied $\mathbf{NB}^{-1}(\mu_j, \phi_j)$ on non-zero elements to generate our final count matrix $\mathbf{X}$.
-To ensure that our simulations match closely to real data, we fitted negative binomial distribution using a maximum likelihood approach (with the \emph{fitdistrplus} package in R \cite{delignette-muller2015}) to non-zero counts for each taxon from 16S rRNA profiling of stool samples from the Human Microbiome Project (HMP).
-We take the median values of the estimated mean and dispersion parameters as the baseline of our simulations.
-For simplicity, we assumed that inter-taxa correlation follows an exchangeable structure
-
-\subsubsection*{Simulation scenarios for enrichment analysis at the sample level}
-To assess type I error rate and power for enrichment significance testing at the sample level, we simulated data based on the schema above, and assessed enrichment for one focal set.
-Type I error was obtained under the global null as the number of samples where the null hypothesis was rejected at $\alpha = 0.05$ over the total number of samples (which represents the total number of hypotheses tested).
-Power was obtained using the same formulation as type I error rate but under the global alternate.
-We treated type I error and power as estimates of binomial proportions and utilized the Agresti-Couli \cite{agresti1998} formulation to calculate 95\% confidence intervals.
-Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$) and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$).
-For type I error analysis, we also varied the size of the set (50, 100, 150).
-For power analyses, set size was kept constant at 100 but different effect sizes (fold change of 1.5, 2, and 3).
-All sample sizes were set at 10,000.
-
-For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUC).
-This is a strategy used in Frost \cite{frost2020} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
-whether samples with inflated counts are highly ranked using estimated scores).
+\subsection*{Statistical significance}
+We evaluate the inference procedure of CBEA compared to alternate methods using two approaches: randomly sampled taxa sets and sample label permutation.
+These analyses were performed on the 16S rRNA gene sequencing of the oral microbiome from the Human Microbiome Project \cite{consortium2012, proctor2019}.
+This data set contains 369 samples split into two subsites: supragingival and subgingival.
+
+\subsubsection*{Sample-level inference} 
+Due to CBEA's self-contained null hypothesis, we can perform inference at the sample level for the enrichment of a set.
+We evaluated this application by generating one random taxa of different sizes $S \in \{20, 50, 100, 150, 200\}$ across 500 iterations.
+Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample sets of randomly sampled taxa should not be significantly enriched compared to the remainder background taxa.
+For this evaluation, we estimated type I error as the fraction of samples that detects our random set as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.
+Additionally, this analysis also demonstrate whether CBEA is sensitive to different set sizes.
+
+\subsubsection*{Population-level inference} 
+We can perform enrichment testing at the population level by generating corresponding sample level CBEA scores and performing a two-sample test such as Welch's t-test.
+In order to evaluate CBEA under this context, we generated CBEA scores of sets representing genus-level annotation in above gingival data set \cite{consortium2012, proctor2019} and applied a t-test to test for enrichment (similar to GSVA \cite{hanzelmann2013}) across a randomly generated variable indicating case/control status (repeated 500 times).
+Type I error is estimated as the fraction of sets per iteration found to be significantly enriched with confidence bands computed from the standard error across all iterations.
+In addition, we also performed a random set analysis assessing, where we generated 100 sets of different set sizes $S \in \{20, 50, 100, 150, 200\}$ and evaluated the fraction of genera that were found to be differentially abundant across the original labels (supragingival versus subgingival subsite).
+95\% confidence intervals were computed using the Agresti-Couli approach \cite{agresti1998}.
+
+\subsection*{Phenotype relevance}
+We want to evaluate whether sets found to be significantly enriched by CBEA are relevant to the research question.
+To perform this assessment, we relied on the gingival data set mentioned prior \cite{consortium2012, proctor2019}.
+This data set was chosen due to its clear biological interpretation that can serve as the ground truth.
+Specifically, we expect aerobic microbes to be enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}.
+Genus-level annotations for microbial metabolism from Beghini et al.
+\cite{beghini2019} and was obtained from the GitHub repository associated with Calagaro et al.
+\cite{matteocalgaro2020}.
+For sample-level inference, we assessed power as the fraction of supragingival samples where aerobic microbes are significantly enriched.
+For population-level inference, power is the fraction of sets representing genus level taxonomic assignments were significant across subsite labels.
+
+In addition to statistical power, we also assessed phenotype relevance through evaluating whether highly ranked sets based on CBEA scores are more likely to be enriched according to ground truth.
+This is represented by the area under the receiving operator curve (AUROC/AUC) scores computed on CBEA scores against true labels (similar to strategy employed by VAM \cite{frost2020}).
 DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate.
-Simulation settings for classification performance were identical to power analyses as detailed in the previous paragraph.
-
-\subsubsection*{Simulation scenarios for enrichment analysis at the population level}
-To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition.
-Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null.
-Power is defined similarly, but instead under the global alternate hypothesis.
-Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error.
-A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size.
-Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$).
-Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3).
-Due to the compositional nature of microbiome taxonomic data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets.
-As such, we applied a compensation procedure as described in Hawinkel et al.
-\cite{hawinkel2019} to ensure the validity of simulation results.
-All sample sizes were set at 2,000.
-
-\subsubsection*{Simulation scenarios for downstream prediction}
-To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}.
-For binary outcomes, we use AUC similar to the classification analyses above.
-For continuous outcomes, we used root mean squared error (RMSE).
-All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020b} suite of packages.
-Across both learning tasks, we varied sparsity ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation ($\rho = 0, 0.2, 0.5$).
-Continuous outcomes $Y_{cont}$ were generated as linear combinations of taxa counts.
-\begin{equation}
-    Y_{cont} = f(\mathbf{X}) + \mathbf{\epsilon}
-\end{equation}
-where $\mathbf{\epsilon} \sim N(0, \sigma_{\epsilon}^2)$ and $f(\mathbf{X}) = \beta_0 + \mathbf{X}\mathbf{\beta}$.
-For each simulation, we set $\beta_0$ to be $\frac{6}{\sqrt{10}}$ similar to \cite{xiao2018}.
-The degree of model saturation (the number of non zero $\mathbf{\beta}$ values) were varied between 0.1 and 0.5, and signal to noise ratio (SNR = $\frac{\sigma(f(\mathbf{X}))}{\sigma_{\epsilon}}$) was varied between 1.5, 2, and 3.
-
-For binary outcomes, we generate $Y_{binary}$ as Bernoulli draws with probability $p_{binary}$, where 
-\begin{equation}
-    p_{binary} = \frac{1}{1 + \exp(f(\mathbf{X}) + \mathbf{\epsilon})}
-\end{equation}
-To ensure a balance of classes, we applied the strategy described in Dong et al.
-\cite{dong2020} where the associated $\beta$ values are evenly split between positive and negative associations.
-All data sets generated from prediction tasks have 2,000 samples with 5,000 taxa over 50 sets with a size of 100 taxa per set.
-
-\subsection*{Evaluation using real data}
-In addition to simulation analyses, we also evaluated our method using real data sets based on both 16S rRNA gene sequencing and whole-genome sequencing.
-We follow the benchmark set by Geistlinger et al.
-\cite{geistlinger2021} All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.
-
-\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of CBEA in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}.
-We utilized this data set following the approach outlined in Calagaro et al.
-\cite{calgaro2020}.
-This data set is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}.
-Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false positive rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment.
-Microbial tropism annotation at the genus level was from Beghini et al.
-\cite{beghini2019} and was downloaded directly from the GitHub repository associated with Calagaro et al.
-\cite{matteocalgaro2020}.
-
-\noindent \textbf{Differential Abundance Analsysis}: To assess type I error using CBEA scores in differential abundance analysis, we utilized the 16S rRNA gene sequencing of stool samples from the Human Microbiome Project \cite{consortium2012, proctor2019}.
-Here, we randomly assign samples a label of case or control, and repeated this process 500 times, assessing all candidate methods at each iteration.
-Type I error is then the number of taxa identified as differentially abundant across all tested taxa.
-For the true positive rate, we used the same gingival data set as described above.
-However, instead of testing for aerobic microbes as a group, the true positive rate is the number of aerobic/anaerobic genera identified as differentially abundant across all aerobic or anaerobic genera.
-
-\noindent \textbf{Disease Prediction}: To assess predictive power, we utilized the whole genome sequencing of stool samples of inflammatory bowel disease (IBD) patients from the MetaHIT consortium \cite{nielsen2014}.
+
+\subsection*{Disease Prediction}
+CBEA scores can also be used downstream analyses such as disease prediction tasks. We utilized two data sets for this evaluation:
+
+\begin{enumerate}
+\item Whole genome sequencing of stool samples from inflammatory bowel disease (IBD) patients in the MetaHIT consortium \cite{nielsen2014}.
 This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease).
-Additionally, we also utilized a similar data set from Gevers et al.
-\cite{gevers2014} which also profiles the gut microbiome of IBD patients and controls but using 16S rRNA gene sequencing.
+
+\item 16S rRNA gene sequencing of stool samples from IBD patients in the pediatric RISK cohort \cite{gevers2014}.
 This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada.
 Of the 671 samples obtained, 500 samples belong to patients with IBD.
-
-\subsubsection*{Comparison Methods}
-
-\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the CBEA method with a naive Wilcoxon rank sum test.
-We added a pseudocount of 1 to all values.
-This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample.
-For classification performance, we compared CBEA methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test.
-All three approaches were applied directly on count data (after pseudocount).
-For GSVA, the Poisson kernel was used.
-
-\noindent \textbf{Differential Abundance}: Since CBEA are sample-level enrichment scores, we performed differential abundance by using a Wilcoxon Rank Sum test and Welch's t-test across case/control status on CBEA generated scores.
-We added a pseudocount of 1 to all values.
-For comparison, we chose representative state-of-the-art methods in differential abundance analysis, namely DESeq2 \cite{love2014,mcmurdie2014} and corncob \cite{martin2020}.
-For DESeq2, we performed a likelihood ratio test against an intercept only reduced model with dispersion estimated with local fit.
-For corncob, we also performed a likelihood ratio test against an intercept only reduced model without bootstrapping.
-
-\noindent \textbf{Disease Prediction}: We fit random forest on CBEA scores, as well as ssGSEA \cite{hanzelmann2013} and GSVA \cite{barbie2009} similar to single sample enrichment section.
-We added a pseudocount of 1 to all values.
-Additionally, we also compared performance using CBEA against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.
+\end{enumerate}
+
+We evaluate disease prediction performance by fitting a random forest model \cite{breiman2001} using inputs as CBEA scores to classify samples of patients with IBD and healthy controls.
+Random forest was chosen as a baseline learner due to its flexibility as an out-of-the-box model that is easy to fit without requiring complex hyperparameter tuning.
+
+\subsection*{Comparison Methods} 
+We benchmarked the statistical properties of CBEA against existing baseline approaches.
+For sample-level inference analyses, utilized the Wilcoxon rank-sum test, which non-parametrically tests the difference in mean counts between taxa from a pre-defined set and its remainder similar to CBEA.
+For assessments at the population level, we compared CBEA against performing a standard test for differential abundance with set-level features generated via element-wise summations instead.
+We chose DESeq2 \cite{love2014} and corncob \cite{martin2020} since they represent both methods extrapolated from RNA-seq \cite{mcmurdie2014} and those developed specifically for microbiome data.
+
+Since disease prediction models and rankings-based phenotype relevance analyses seek to evaluate the informativeness of CBEA scores instead of relying on computing p-values, we compared performance against other single sample based approaches from the gene set testing literature, specifically ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}.
+Additionally, for evaluating prediction, we also compared performance against a standard analysis plan where inputs are count-aggregated sets with the centered log-ratio (CLR) transformation.
 
 \section*{Results}
-In this section, we present the performance of our proposed method for three applicable microbiome analysis tasks: sample level enrichment, differential abundance, and disease prediction.
-We obtained these results from both parametric simulations and examples from real data.
-
-\subsection*{Enrichment testing at the sample level}
+In this section, we present results for evaluating statistical significance, phenotype relevance, and predictive performance.
+In addition to real data, we also evaluated models based on parametric simulations.
+Results for this section can be found in Supplemental Materials.
+
+\subsection*{Sample level inference}
 CBEA provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{methods}.
@@
 
-\subsubsection*{Simulation studies}
-Panel A and B in Fig~\ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
-We benchmarked the results of the CBEA method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
-All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions.
-However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted CBEA variants show high levels of inflated type I error, where Wilcoxon test performed the worst.
-On the other hand, adjusted CBEA methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.
-
 \begin{figure}[!h]
     \centering
-    \includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
-    \caption{Sample-level inference with CBEA under parametric simulations (\textbf{(A)} and \textbf{B}), and real data analysis (\textbf{(C)} and \textbf{(D)}).
-In simulation analyses, panel \textbf{(A)} shows type I error rate, and panel \textbf{(B)} shows power for single sample enrichment test for a specified set and was compared against a Wilcoxon rank sum test at $\alpha$ of 0.05.
-In real data analysis, panel \textbf{(C)} shows the false-positive rate, and panel \textbf{(D)} shows the true positive rate.
-For this analysis, 16S rRNA data from the oral microbiome of the gingival site was used.
-The set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples.
-Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach.
-Adjusted CBEA demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate.
-However, this resulted in lower power for adjusted CBEA methods.}
-    \label{fig:2}
+    \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
+    \caption{Classification performance via AUC of CBEA, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}.
+Performance scores measure whether scores can highly rank samples that are known to have inflated abundance.
+In the gingival data set presented in panel \textbf{(B)}, samples from the supragingival site are assumed to have an inflated abundance of aerobic microbes.
+Error bars are the 95\% DeLong confidence intervals for AUC \cite{delong1988}}
+    \label{fig:3}
 \end{figure}
@@
 Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.
+
 \subsubsection*{Simulation studies}
@@
 
-These results demonstrate that CBEA generated scores are informative features in disease prediction tasks.
+These results demonstrate for given set-based features, CBEA can generate scores can could be informative in disease prediction tasks.
 Simulation results indicate that CBEA methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach.
@@
 The formulation of CBEA as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}.
-This allows for conducting inference with CBEA even at the sample-level.
+Since CBEA generate sample-specific scores, this allows users to identify significantly enriched taxa even at the at the sample-level.
 We assessed the usage of CBEA in this type of analysis by evaluating type I error and power across both simulation studies and real data applications.
@@
 \subsubsection*{Predictive models}
-For disease prediction, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using CBEA generated scores as inputs.
+We evaluated whether CBEA can be used to generate set-based features for disease prediction models.
+Here, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using CBEA generated scores as inputs.
 Similar to our inference analysis, we compared CBEA against both ssGSEA and GSVA.
-Additionally, we also evaluated CBEA with the approach where counts of a set were aggregated using sums and then centered log-ratio transformed.
+Additionally, we also evaluated CBEA with the approach where counts of a set were aggregated using sums and applied the centered log-ratio transformation (CLR).
 This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}.
-Results indicated that CBEA produces good performance values across both real data analysis and simulation scenarios.
+Results showed that CBEA generate scores perform well  across both real data settings and simulation scenarios.
 Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that CBEA scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
@@
 \newblock {{Tidymodels}}: a collection of packages for modeling and machine learning using tidyverse        principles; 2021.
-\newblock url:{https://www.tidymodels.org}
 

### 2022-01-26 saving_texts.tex

--- 2021-11-30 cbea.tex
@@
-% Template for PLoS
-% Version 3.5 March 2018
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-%\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-
-%% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
-
-%% END MACROS SECTION
-
-% Adding packages 
-\usepackage{bm}
-\usepackage[parfill]{parskip}
-
-\begin{document}
-\vspace*{0.2in}
-
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{cILR: Competitive isometric log-ratio for taxonomic enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Quang P. Nguyen\textsuperscript{1,2},
-Anne G. Hoen\textsuperscript{1,2\ddag},
-H. Robert Frost\textsuperscript{1*\ddag\dag},
-\\
-\bigskip
-\textbf{1} Department of Biomedical Data Science, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA
-\\
-\textbf{2} Department of Epidemiology, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA
-\\
-\bigskip
-
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-%\Yinyang These authors contributed equally to this work.
-
-% Additional Equal Contribution Note
-% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-\ddag These authors jointly supervised this work.\\
-\dag Corresponding author
-% Current address notes
-%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
-% \textcurrency b Insert second current address 
-% \textcurrency c Insert third current address
-
-% Deceased author note
-% \dag Deceased
-
-% Group/Consortium Author Note
-%\textpilcrow Membership list can be found in the Acknowledgments section.
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* hildreth.r.frost@dartmouth.edu
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
-Research in human associated microbiomes often involves the analysis of taxonomic count tables generated via high-throughput sequencing.
-It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional.
-An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets.
-Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis.
-Unfortunately, there is a lack of sophisticated set-based analysis methods specific to microbiome taxonomic data, where current practice often employs abundance summation as a technique for aggregation.
-This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
-Here, we attempt to fill this gap with a new single sample taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature.
-Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
-We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
-Herein we demonstrate using both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios.
-Additionally, it provides informative scores that can be inputs to downstream differential abundance and prediction tasks.
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-\section*{Author summary}
-The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing.
-However, microbiome taxonomic data is sparse and high dimensional which prevents the application of standard statistical techniques.
-One approach to address this problem is to perform analyses at the level of taxon sets.
-Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability.
-Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given that compared to other 'omics data types microbiome taxonomic data is strictly compositional.
-We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis.
-Our scores can be used for statistical inference, and as inputs to other downstream analyses such as differential abundance and prediction models.
-We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.
-
-\linenumbers
-
-% Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
-The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host.
-Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}.
-To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e.
-amplicon sequencing) or by profiling all the genomic content of the sample (i.e.
-whole-genome shotgun sequencing) \cite{cho2012}.
-Raw sequencing data is then processed through a variety of bioinformatic pipelines \cite{callahan2016, truong2015}, yielding various data products, one of which are taxonomic tables which can be used to study associations between members of the microbiome and an exposure or outcome of interest.
-
-However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019,li2015}.
-First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present.
-For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016}, allowing for simultaneous model fitting and variable selection.
-For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.
-
-Second, the number of reads obtained is constrained by the sequencing instrument at an arbitrary limit, and is inconsistent across samples, resulting in a variable number of total read counts per sample.
-Many normalization methods \cite{weiss2017} have been proposed to address these issues, including cross-applying methods from the gene expression literature \cite{mcmurdie2014}.
-However, these methods rely on assumptions specific to the original bulk RNA-seq data sets such as the presence of housekeeping genes with consistent expression levels \cite{love2014}, which might not be not true in the context of microbiome relative abundance data \cite{quinn2019,quinn2018b}.
-As such, microbiome taxonomic data is strictly compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another.
-Consequently, log-ratio transformations from the compositional data literature are often utilized \cite{aitchison1999}.
-
-Third, the data are highly zero-inflated, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment).
-Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015}, or applying zero-inflated models \cite{chen2018, kaul2017}.
-Newer methods developed recently have focused on understanding the different types of zeros in the data, providing more sophisticated heuristics around when pseudocounts can be utilized \cite{kaul2017a}.
-
-Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}.
-Aggregated sets are less sparse than their constituent elements, and testing on a smaller number of variables reduces the multiple testing burden, thereby increasing power and reproducibility.
-Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations.
-There exists a diverse set of available methods developed to perform such analyses.
-More traditional set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest \cite{goeman2007}.
-Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values.
-The most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size.
-Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples.
-Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample level and are more akin to a transformation.
-This strategy allows for the flexible incorporation of different statistical techniques downstream, such as prediction models, as well as for visualization purposes in ordination plots.
-
-In microbiome research, even though no explicit enrichment analysis is performed, standard practice often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum by simple summation of abundances \cite{mclaren2019}.
-Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, it increases bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}.
-As such, there is a need for microbiome researchers to adopt more robust sample-level set enrichment methods.
-Unfortunately, limited work has been done to extend existing methods to be more specific to microbiome relative abundance data.
-Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}.
-However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used.
-
-Here, we present a novel method that generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}.
-We leverage the concept of the $Q_1$ competitive hypothesis presented in Tian et al.
-\cite{tian2005}, which tests the null that the value of variables within a specific set is equal to the value of measured variables not in the set.
-The competitive null hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables.
-We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a multiplicative aggregation method that addresses the downsides of the naive summation-based method presented above \cite{mclaren2019, silverman2017}.
-The resulting method, titled competitive isometric log-ratio (cILR), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing.
-These scores can then act as inputs to differential abundance and predictive modeling tasks downstream.
-
-In the following sections, we provide the formulation of cILR and discuss some statistical properties.
-We illustrate significance testing at the sample level using cILR and evaluate type I error and power under different simulation scenarios and real data applications.
-We assess the informativeness of cILR generated scores, and evaluate how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis.
-We compare the performance of cILR in these respective tasks against standard microbiome taxonomic data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods.
-An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).
-
-\section*{Materials and Methods} \label{methods}
-\subsection*{Competitive Isometric Log-ratio (cILR)}
-The cILR method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}.
-Details on the computational implementation of cILR can be found in the supplemental materials.
-The cILR method takes two inputs:
-\begin{itemize}
-    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing.
-Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing.
-    \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxon $p$ to $m$ sets of interest.
-These sets can be Linnean taxonomic classifications annotated using databases such as SILVA \cite{quast2013}, or those based on more functionally driven categories such as tropism or ecosystem roles ($A_{i,j} = 1$ indicates that microbe $i$ belongs to set $j$).
-\end{itemize}
-The cILR method generates one output:  
-\begin{itemize} 
-    \item $\mathbf{E}$: $n$ by $m$ matrix indicating the enrichment score of $m$ pre-defined sets identified in $\mathbf{A}$ across $n$ samples. 
-\end{itemize}
-The procedure is as follows:  
-\begin{enumerate}
-    \item \textbf{Compute the cILR statistic}: Let $\mathbf{M}$ be a $n$ by $m$ matrix of cILR scores. Let $\mathbf{M}_{i,k}$ be cILR scores for set $k$ of sample $i$:   
-    \begin{equation}\label{main_eq}
-        \mathbf{M}_{i,k} = \sqrt{\frac{\sum_k A_{ik}(p - \sum_k A_{ik})}{p}} \ln \left( \frac{g(\mathbf{X}_{i,j}|\mathbf{A}_{j,k} = 1)}{g(\mathbf{X}_{i,j}|\mathbf{A}_{j,k} \neq 1))} \right)
-    \end{equation}
-    where $g(.)$ is the geometric mean. This represents the ratio of the geometric mean of the relative abundance of taxa assigned to set $k$ and remainder taxa. 
-    \item \textbf{Estimate the empirical null distribution} Enrichment scores represent the test statistic for the $Q_1$ null hypothesis $H_o$ that relative abundances in $\mathbf{X}$ of members of set $k$ are not enriched compared to those not in set $k$.
-Since the distribution of cILR under the null vary depending on data characteristics (Fig~\ref{fig:1}), an empirical null distribution will be estimated from data.
-    \begin{itemize}
-        \item \textbf{Compute the cILR statistic on permuted and un-permuted $\mathbf{X}$}.
-Let $\mathbf{X}_{perm}$ be the column permuted relative abundance matrix, and $\mathbf{M}_{perm}$ be the corresponding cILR scores generated from $\mathbf{X}_{perm}$.
-Similarly, we have $\mathbf{M}_{unperm}$ be cILR scores generated from $\mathbf{X}$.
-        \item \textbf{Estimate correlation-adjusted empirical distribution for each set}.
-For each set, a fit a parametric distribution to both $\mathbf{M}_{perm}$ and $\mathbf{M}_{unperm}$.
-The location measure estimated from $\mathbf{M}_{perm}$ and the spread measure estimated from $\mathbf{M}_{unperm}$ will be combined as the correlation-adjusted empirical null distribution $\mathbf{P}_{emp}$ for each set.
-Two available options are the normal distribution and the mixture normal distribution.
-For the normal distribution, parameters were estimated using the method of maximum likelihood implemented in the \emph{fitdistr} package \cite{delignette-muller2015}.
-For the mixture normal distribution, parameters were estimated using an expectation-maximization algorithm implemented in the \emph{mixtools} package \cite{benaglia2009}.
-    \end{itemize}
-    \item \textbf{Calculate finalized cILR scores with respect to the empirical null}.
-Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution.
-P-values can be calculated by subtracting $\mathbf{E}$ from 1.
-\end{enumerate}
-
-\subsection*{Properties of cILR}
-\subsubsection*{cILR and the Isometric Log Ratio Transformation}
-The cILR statistic is a special instance of the isometric log-ratio transformation (ILR) \cite{egozcue2003}.
-The standard ILR is a transformation method to address the negative correlation bias inherent in compositional data by providing an isometry between the $D$-dimensional simplex $\mathbb{S}^D$ and coordinates in the $D-1$ real space $\mathbb{R}^{D-1}$ \cite{egozcue2003,washburne2017}.
-This is accomplished by projecting the composition onto a chosen orthonormal basis in $\mathbb{R}$, which can be defined by a sequential binary partition (SBP) of the variables (e.g.
-a rooted phylogenetic tree).
-The ILR transformed variables are the coordinates of nodes within an SBP tree of the variables.
-Without loss of generalizability, in a given SBP with node $i$ splitting variables between sets $R$ and $S$, we have the ILR coordinate $x^{*}_{i}$ as:
-\begin{equation}\label{ilr_standard}
-    x^{*}_i = \sqrt{\frac{rs}{r+s}} \log\left(\frac{g(X_{j|j \in R})}{g(X_{j|j \in S})}\right)
-\end{equation}
-where $r$ and $s$ are the cardinalities of sets $R$ and $S$ respectively, $g(z)$ is the geometric mean, and $X_{j}$ are values of the original predictors with indexes defined by membership in $R$ and $S$.
-The ILR confer many important benefits.
-First, ILR coordinates exist in real space, whereby common statistical methods can be used.
-Second, ILR aggregated variables preserve inter-sample distances before and after aggregation \cite{egozcue2005}.
-Third, ILR variables are not constrained to sum to 0 as that of the centered log-ratio transformation, resulting in a covariance matrix that is not singular \cite{egozcue2003}.
-
-The usage of the ILR statistic is not uncommon in the microbiome literature.
-They are usually termed ``compositional balances", and have been leveraged in many recent approaches in variable transformation \cite{washburne2017,silverman2017,morton2017}.
-The cILR formulation in Eq~\eqref{main_eq} is a special case of Eq~\eqref{ilr_standard} defined on a node that splits the taxa into two disjoint sets, one representing the set of interest, the other representing the remaining taxa.
-As such, the cILR transformation inherits the properties of the ILR as a log-ratio method applicable to compositional data sets.
-However, unlike the ILR and its variants \cite{silverman2017, morton2017, washburne2017}, the axes defined by each cILR set are not orthogonal (since the balances are mutually exclusive between sets and do not belong in the same SBP).
-Hence, a correlation can exist between cILR aggregated variables.
-
-\subsubsection*{Statistical Properties of cILR}
-We can perform significance testing on the cILR statistic which corresponds to the null hypothesis that the center of the subcomposition defined by the set is equal to the center of the subcomposition defined by the complement of the set.
-This is equivalent to the $Q_1$ competitive null hypothesis in the gene set testing literature \cite{tian2005} where the enrichment of a gene set is defined with respect to genes outside the set.
-
-We can apply prior usage of the ILR statistic in hypothesis testing to cILR by assuming that the null distribution of cILR follows a standard normal distribution \cite{egozcue2005}.
-However, when applying cILR for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses.
-Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies.
-As such, to perform significance testing using cILR, we also estimated the null distribution from observed raw cILR variables.
-
-This assumption is also supported by preliminary simulation studies (detailed below).
-In panel A of Fig~\ref{fig:1}, we simulated microbiome taxonomic count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness.
-It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation.
-Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis).
-The effect is most dramatic under both high inter-taxa correlation and sparsity.
-This heterogeneity further supports the decision to estimate an empirical null distribution, similar to Efron \cite{efron2004}.
-
-\begin{figure} [!h]
-    \centering
-    \includegraphics[width=\linewidth]{figures/kurtosis_skewness_gof.png}
-    \caption{{\bf Properties of the null distribution of cILR under the global null simulations}.
-Panel \textbf{(B)} presents kurtosis and skewness of cILR scores while panel \textbf{(A)} presents the goodness of fit (as Kolmogorov-Smirnov D statistic) for mixture normal and normal distributions.
-Panel \textbf{(C)} is a density plot of the shape of the null distribution.
-Results indicated the necessity of estimating an empirical null and demonstrating that the mixture distribution was the better fit compared to the basic normal.}
-    \label{fig:1}
-\end{figure}
-
-Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null.
-To address this issue, we also evaluated a mixture distribution of two normal components.
-Panel B of Fig~\ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null.
-We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.
-
-We performed our empirical null estimation by fitting our distribution of choice and computing relevant parameters on raw cILR scores on taxa-permuted data (equivalent to gene permutation in the gene expression literature).
-As such, the null distribution is characterized by scores computed on sets of equal size with randomly drawn taxa.
-However, null distribution based on taxa-permutation is sensitive to inter-taxa correlations within the set \cite{wu2012}.
-Since the permutation procedure does not preserve correlation structures, estimating parameters from empirical scores on permuted data will underestimate the variance inflation due to correlation.
-We account for this by combining the mean estimate from permuted data with the variance estimate from unpermuted data, where the inter-taxa correlation structure remains undisturbed.
-However, this procedure assumes that the variance of cILR is equal under both the null and alternate hypotheses.
-
-\subsection*{Evaluation}
-All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/cILR\_analysis}{qpmnguyen/cILR\_analysis}).
-
-\subsubsection*{Parametric Simulations}  
-To address the performance of cILR for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
+\subsection*{Simulation Analysis}
+\subsubsection*{Data generation model}  
+To address the performance of CBEA for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
 Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
@@
 
-For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUROC/AUC).
-This is a strategy used in Frost \cite{frost2020a} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
+For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUC).
+This is a strategy used in Frost \cite{frost2020} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
 whether samples with inflated counts are highly ranked using estimated scores).
@@
 
-\noindent \textbf{Prediction}: To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}.
+\subsubsection*{Simulation scenarios for downstream prediction}
+To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}.
 For binary outcomes, we use AUC similar to the classification analyses above.
 For continuous outcomes, we used root mean squared error (RMSE).
-All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020} suite of packages.
+All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020b} suite of packages.
 Across both learning tasks, we varied sparsity ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation ($\rho = 0, 0.2, 0.5$).
@@
 
-\subsubsection*{Real Datasets}
-
-In addition to simulation analyses, we also evaluated our method using real data sets based on both 16S rRNA gene sequencing and whole-genome sequencing.
-All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.
-
-\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of cILR in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}.
-We utilized this data set following the approach outlined in Calagaro et al.
-\cite{calgaro2020}.
-This data set is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}.
-Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false positive rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment.
-Microbial tropism annotation at the genus level was from Beghini et al.
-\cite{beghini2019} and was downloaded directly from the GitHub repository associated with Calagaro et al.
-\cite{matteocalgaro2020}.
-
-\noindent \textbf{Differential Abundance Analsysis}: To assess type I error using cILR scores in differential abundance analysis, we utilized the 16S rRNA gene sequencing of stool samples from the Human Microbiome Project \cite{consortium2012, proctor2019}.
-Here, we randomly assign samples a label of case or control, and repeated this process 500 times, assessing all candidate methods at each iteration.
-Type I error is then the number of taxa identified as differentially abundant across all tested taxa.
-For the true positive rate, we used the same gingival data set as described above.
-However, instead of testing for aerobic microbes as a group, the true positive rate is the number of aerobic/anaerobic genera identified as differentially abundant across all aerobic or anaerobic genera.
-
-\noindent \textbf{Disease Prediction}: To assess predictive power, we utilized the whole genome sequencing of stool samples of inflammatory bowel disease (IBD) patients from the MetaHIT consortium \cite{nielsen2014}.
-This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease).
-Additionally, we also utilized a similar data set from Gevers et al.
-\cite{gevers2014} which also profiles the gut microbiome of IBD patients and controls but using 16S rRNA gene sequencing.
-This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada.
-Of the 671 samples obtained, 500 samples belong to patients with IBD.
-
-\subsubsection*{Comparison Methods}
-
-\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank sum test.
-We added a pseudocount of 1 to all values.
-This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample.
-For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test.
-All three approaches were applied directly on count data (after pseudocount).
-For GSVA, the Poisson kernel was used.
-
-\noindent \textbf{Differential Abundance}: Since cILR are sample-level enrichment scores, we performed differential abundance by using a Wilcoxon Rank Sum test and Welch's t-test across case/control status on cILR generated scores.
-We added a pseudocount of 1 to all values.
-For comparison, we chose representative state-of-the-art methods in differential abundance analysis, namely DESeq2 \cite{love2014,mcmurdie2014} and corncob \cite{martin2020}.
-For DESeq2, we performed a likelihood ratio test against an intercept only reduced model with dispersion estimated with local fit.
-For corncob, we also performed a likelihood ratio test against an intercept only reduced model without bootstrapping.
-
-\noindent \textbf{Disease Prediction}: We fit random forest on cILR scores, as well as ssGSEA \cite{hanzelmann2013} and GSVA \cite{barbie2009} similar to single sample enrichment section.
-We added a pseudocount of 1 to all values.
-Additionally, we also compared performance using cILR against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.
-
-\section*{Results}
-In this section, we present the performance of our proposed method for three applicable microbiome analysis tasks: sample level enrichment, differential abundance, and disease prediction.
-We obtained these results from both parametric simulations and examples from real data.
-
-\subsection*{Enrichment testing at the sample level}
-cILR provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{methods}.
-Here, we present empirical results for this application of cILR assessing type I error, power, and classification capacity.
-
-\subsubsection*{Simulation studies}
+\section{Results}
+\subsubsection*{Simulation studies for single sample}
 Panel A and B in Fig~\ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
-We benchmarked the results of the cILR method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
+We benchmarked the results of the CBEA method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
 All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions.
-However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted cILR variants show high levels of inflated type I error, where Wilcoxon test performed the worst.
-On the other hand, adjusted cILR methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.
+However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted CBEA variants show high levels of inflated type I error, where Wilcoxon test performed the worst.
+On the other hand, adjusted CBEA methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.
 
@@
 Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach.
-Adjusted cILR demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate.
-However, this resulted in lower power for adjusted cILR methods.}
+Adjusted CBEA demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate.
+However, this resulted in lower power for adjusted CBEA methods.}
     \label{fig:2}
 \end{figure}
-
-However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B.
-In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un-adjusted ones.
-However, in higher correlation scenarios, the difference in power is much more dramatic.
-At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted cILR was only performing at 50\% power, while unadjusted cILR and wilcoxon rank sum test were able to reach 80\%.
-These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of cILR.
-Most importantly, cILR demonstrate higher power in all scenarios where type I error is properly controlled.
-
-To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set.
-This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated.
-Fig~\ref{fig:3} presents this result.
-We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
-Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
-In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
-Notably, the W-statistic provided the least information for classifying samples with inflated taxa.
-
-\subsubsection*{Real data evaluations}  
-These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{methods}.
-Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations.
-As expected in Fig~\ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods.
-Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05.
-Power analysis (Fig~\ref{fig:2}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
-    \caption{Classification performance via AUC of cILR, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}.
-Performance scores measure whether scores can highly rank samples that are known to have inflated abundance.
-In the gingival data set presented in panel \textbf{(B)}, samples from the supragingival site are assumed to have an inflated abundance of aerobic microbes.
-Error bars are the 95\% DeLong confidence intervals for AUC \cite{delong1988}}
-    \label{fig:3}
-\end{figure}
-
-\subsection*{Differential abundance analysis}
-cILR generates sample-specific scores representing the degree of enrichment of a pre-defined set.
-As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test).
-We compared the performance of this approach with cILR and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
-
-\subsubsection*{Simulation studies}
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
-    \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
-For cILR methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by cILR (across different output types and distributional assumptions).
-cILR associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
-    \label{fig:4}
-\end{figure}
-
-Fig.~\ref{fig:4} present results for simulation studies for both type I error (panel A) and power (panel B) evaluations.
-All methods control for type I error well across both sparsity and correlation levels, where the estimated rate was consistently around the 0.05 pre-defined threshold.
-Results were similar across all evaluated methods, although in some instances, for example in medium correlation setting ($\rho = 0.2$), the unadjusted cILR resulted in higher type I error, regardless of difference in means test and distribution of choice.
-
-The difference between the methods is more noticeable when evaluating power.
-All cILR associated variants showed much higher power even when the effect size is limited (fold change is 1.5), and there is a noticeable gap in performance between cILR and both DESeq2 and corncob.
-Surprisingly, this effect is consistent across correlation levels and sparsity, even though we expectedly see performance in power drop as a function of sparsity especially in low effect size settings.
-
-\subsubsection*{Real data evaluations}
-In addition to simulation studies, we also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP (Fig~\ref{fig:5}).
-For type I error evaluations, we use stool samples and randomly assign them with case/control status and calculated type I error as the proportion of genera identified as significantly different.
-For true positive rate evaluations, we use the gingival data set as detailed in the previous section, and calculated the true positive rate as the proportion of genera labeled as either anaerobic or aerobic that were found to be significant.
-
-We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of cILR were controlling for type I error at the defined $\alpha$ threshold of 0.05.
-This is surprising given the consistency of preserving type I error for both corncob and DESeq2 in all simulation evaluations.
-
-In true positive experiments with data from the gingival site, estimated rates were more similar across the different methods.
-As expected, using the Wilcoxon rank sum test resulted in lower true positive rate compared to remaining methods, but the difference was not noticable.
-This is also surprising given that in simulation studies, both corncob and DESeq2 showed markedly lower power across all effect sizes.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
-    \caption{Differential abundance analysis using corncob, DESeq2 and cILR with either Wilcoxon rank sum test or Welch's t-test.
-Panel \textbf{(A)} shows type I error results as the proportion of significant genera after 500 iterations where case/control status was assigned randomly to each sample.
-Panel \textbf{(B)} shows true positive rate results as the proportion of significant genera who are either obligate anaerobes or aerobes.
-Both evaluations use 16S rRNA gene sequencing data from HMP.
-Type I error evaluation used stool samples while the true positive rate evaluation used samples from the gingival site.
-Results showed that cILR associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2}
-    \label{fig:5}
-\end{figure}
-
-\subsection*{Disease Prediction}   
-Since cILR can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models.
-In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating cILR, ssGSEA, and GSVA).
-Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.
-\subsubsection*{Simulation studies}
-Fig~\ref{fig:6} shows results for simulation studies as detailed in the \nameref{methods} section.
-Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome.
-As expected, performance across all assessed methods increased with a higher signal-to-noise ratio.
-Both CLR and cILR approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks.
-This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while cILR and CLR did not experience a similar level of impact.
-As such, performance gap widens with increasing correlation and sparsity.
-Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.
-
-In this analysis, cILR unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
-This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation.
-However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and cILR/CLR.
-Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity.
-Additionally, we did not observe any performance difference between the different variations of cILR.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
-    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
-Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure.
-Panel \textbf{(B)} presents performance on a classification task with AUC as the evaluation measure.
-cILR approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
-    \label{fig:6}
-\end{figure}
-
-%\begin{figure}[!ht]
-%    \centering
-%    \includegraphics[width=\textwidth]{figures/sim_pred_rsq.png}
-%    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on cILR,     %ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data %across sparsity levels, correlation, and signal-to-noise ratio.
-Panel \textbf{(A)} and \textbf{(B)} %represent results across different levels of model saturation (proportion of sets associated with the %outcome).
-cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
-%    \label{fig:sim_pred_rsq}
-%\end{figure}
-
-\subsubsection*{Real data evaluations}
-In addition to parametric simulations, we also assessed the performance of using cILR scores in predictive models with real data sets.
-Fig~\ref{fig:7} presents results for two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
-The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al.
-\cite{gevers2014} data set uses 16S rRNA gene sequencing, while the Nielsen et al \cite{nielsen2014} data set uses whole genome shotgun sequencing.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \linewidth]{figures/data_prediction_plot.png}
-    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
-Data sets used span both 16S rRNA gene sequencing (Gevers et al.
-\cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al.
-\cite{nielsen2014}).
-cILR performs better than GSVA and ssGSEA but not as well as CLR, with the exception of the whole genome sequencing data set.}
-    \label{fig:7}
-\end{figure}
-
-Similar to simulation experiments, we also fitted a naive random forest model using CILR, ssGSEA, GSVA, or CLR transformed variables as inputs, and use AUC as the performance criteria.
-Results also replicated that of the simulations, where across both data sets cILR and CLR methods provide much better performance than both GSVA or ssGSEA.
-Interestingly, the cILR approach performed better than CLR in the whole genome data set but did not perform as well in the 16S rRNA gene sequencing data set.
-However, these results indicate that cILR generated scores are informative, providing competitive performance when acting as inputs to disease predictive models.
-Most importantly, performance values are consistent across both simulated and real data sets.
-
-These results demonstrate that cILR generated scores are informative features in disease prediction tasks.
-Simulation results indicate that cILR methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach.
-Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
-
-\section*{Discussion}
-
-\subsection*{Inference with cILR}
-The formulation of cILR as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}.
-This allows for conducting inference with cILR even at the sample-level.
-We assessed the usage of cILR in this type of analysis by evaluating type I error and power across both simulation studies and real data applications.
-Most importantly, we demonstrated that our adjusted cILR approach was able to address the issue of variance inflation due to correlation \cite{wu2012} by controlling for type I error at the appropriate $\alpha$ level across different levels of simulated inter-taxa correlation (Fig~\ref{fig:2}) while conversely unadjusted cILR and the naive Wilcoxon rank sum test showed much higher rates of error.
-This is further encouraged in real data analysis where the false discovery rate was around $0.05$ when a collection of true null and true alternate hypotheses were tested.
-Unfortunately the trade-off of good type I error control is lower power.
-The conservativeness of the test attenuates with higher sparsity and correlation, where power was not approaching even 50\% even at the highest effect sizes.
-However, when the degree of such data features are reasonable, cILR will still be able to detect a reasonable proportion of samples with inflated counts.
-
-We also observed that choosing different distributional forms did not alter performance values for cILR.
-This runs contrary to our comparison analysis in Fig~\ref{fig:1}, where we demonstrated that the mixture normal distribution had superior fit compared to the simple normal for raw cILR scores computed under the global null.
-We hypothesized that this might be due to the difficulty in fitting mixture distribution to data using the expectation maximization algorithm, as the convergence rate is slow when there is high overlap between the mixtures, resulting in a small mixing coefficient for one of the components \cite{naim2012}.
-As such, in our implementation of cILR, in order to ensure convergence for the estimating procedure we increased the number of iterations while relaxing the tolerance parameter.
-Furthermore, there are also possible problems with our adjustment procedure for the mixture distribution that might impact overall fit.
-In order to combine the scale and location estimate of two mixture distributions, we fixed the overall mean, standard deviation, mixing coefficient and component-wise means and used an optimization procedure to find the component-wise variances.
-However, this means that we have one equation for the overall variance but two possible parameters to estimate.
-As such, there is no guaranteed unique solution to component-wise variances.
-We hypothesized that the instability and degeneracy in component-wise variance estimates might impact the fidelity of estimates at the tails of the distribution, thereby affecting inference.
-
-Despite these concerns, empirical results still indicate that cILR can confidently identify samples with inflated counts.
-The conservativeness of the correlation adjustment procedure ensures that significant results can be trusted by practitioners, even if cILR might not be able to exhaustively identify all samples with inflated counts.
-In situations where either the data is less sparse (e.g.
-containing a lot of core taxa that are prevalent across all samples), there is less inter-taxa correlation within the set (e.g.
-taxa that do not participate in common pathways but have shared characteristics like pathogenicity), or if the effect size is large, then cILR will still be able to produce reasonable power.
-A practitioner can use cILR to screen for samples for subsequent analysis that might involve significant costs, or perform hypothesis generation using a less stringent criteria alongside a multiple testing adjustment procedure (such as Benjamini-Hochberg \cite{benjamini1995}).
-
-\subsection*{Downstream analysis}
-The sample-level enrichment scores generated by the cILR method can be used in downstream analyses commonly performed in microbiome research: differential abundance testing and disease prediction.
-
-\subsubsection*{Differential abundance analysis}
-For differential abundance testing, we evaluated whether using cILR scores alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes.
-We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014} applied on data where taxa were aggregated using the naive sum method.
-We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}.
-Conversely, corncob is a newer method developed specifically for microbiome taxonomic data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.
-
-Surprisingly, we found some conflicting results when evaluating comparisons across parametric simulations and real data analysis.
-The performance of cILR was consistent across two evaluation criteria, demonstrating good type I error and respectable power.
-However, corncob and DESeq2 showed opposite effects: in simulation experiments, both methods show good type I error control but low power, while in real data analyses, conversely both corncob and DESeq2 showed inflated type I error but comparable power with respect to cILR methods.
-Despite such discrepancy, results still indicate good performance of cILR scores when used as inputs to downstream differential abundance analysis compared to using aggregated raw counts, even in methods designed to handle that type of data such as corncob and DESeq2.
-
-We hypothesized that the above behavior can arise due to issues with performing inference in the presence taxon-specific protocol biases \cite{mclaren2019}.
-According to McLaren et al., the observed relative abundance of taxa is different than the true relative abundance due to protocol bias, and importantly this bias is specific to each taxon \cite{mclaren2019}.
-This is especially true in the context of sum-based aggregations, where the resulting bias of the aggregated taxon are dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}).
-Conceptually, this means that measurement error for a taxon aggregate is different across samples as relative abundance of contributing changes, leading to issues when attempting to perform inference.
-As such, we expect methods like corncob or DESeq2 when performed on such aggregates to have inflated type I error compared to our taxon-ratio based approach.
-
-The bias model also helps explain differences in performance of DESeq2 and corncob in simulation analyses compared to real data.
-Our simulation protocol does not explicitly include bias in our formulation, and all taxa were generated from the same underlying distribution with similar variances across all samples (the only difference is in the mean value where a taxa is expected to have inflated counts).
-As such, we do not expect our simulated taxa to have any taxon-specific biases, which is not the case in real data settings.
-Therefore, we can expect DESeq2 and corncob to retain their expected type I error control in simulation analyses compared to real data.
-It is still surprising to see lower power for both methods in simulation analyses, which might be due to the fact that the evaluation protocol only considers default settings for both methods and did not attempt to optimize performance.
-
-The fact that the performance of cILR remains consistent across both simulation and real data analysis shows that cILR is invariant to taxon-specific biases.
-Furthermore, our evaluation indicates that even a simple difference in means test when used in combination with cILR scores can preserve type I error while maintaining good power.
-As such, a practioner can use cILR as a pre-processing step prior to performing a differential abundance test.
-
-\subsubsection*{Predictive models}
-For disease prediction, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using cILR generated scores as inputs.
-Similar to our inference analysis, we compared cILR against both ssGSEA and GSVA.
-Additionally, we also evaluated cILR with the approach where counts of a set were aggregated using sums and then centered log-ratio transformed.
-This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}.
-Results indicated that cILR produces good performance values across both real data analysis and simulation scenarios.
-Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
-Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome taxonomic data sets.
-This is consistent with our sample ranking analysis (Fig~\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
-However, cILR did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data.
-
-However, in simulation studies, this performance gap between CLR and cILR decreases with higher sparsity and correlation, especially in low effect saturation scenarios.
-Additionally, there are also downsides to applying CLR.
-First, the singular covariance matrix of CLR transformed variables is singular due to a sum to zero constraint \cite{gloor2017}, preventing the proper usage of approaches that rely on matrix decomposition.
-Second, the procedure still relies on using summation of counts prior to transformation, which means that we still can't compare across sets of different sizes, and any bias might still be propagated \cite{mclaren2019}.
-As such, despite benefits in performance for a naive random forest model, there is still space for using cILR as primary inputs into predictive models.
-
-Similar to other experiments in downstream usage of cILR, performance did not change with different underlying distributions, output types, or correlation status.
-This is surprising since we expect z-scores to perform better as they are able to capture the direction of an association.
-The fact that this effect persisted even onto our real data analysis suggests that this is not due to a deficiency of our simulation design.
-As such, practitioners who wish to use cILR in predictive models might be suited to use the settings that is the fastest to compute.
-
-Ultimately, results indicate that cILR can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity.
-Even though there exists situations where it might not provide maximum predictive values, the flexibility of cILR in various types of analyses enable even though in some scenarios it might not provide maximum predictive values.
-
-\subsection*{Limitations and future directions} 
-There are various limitations to our evaluation of cILR.
-First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome taxonomic data.
-There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation of prior studies.
-Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is not perfect.
-This is because the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut.
-As such, results from power analyses using this data set is only relative between the comparison methods.
-Finally, we assumed that taxa within a set are all equally associated with the outcome.
-This limits our ability to evaluate the performance of cILR when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.
-
-Our evaluation showed various drawbacks of the cILR method.
-First, inference with cILR is limited in being able to exhaustively detect all samples with significant inflated counts for a set in situations where there is a high degree of sparsity and inter-taxa correlation.
-Second, for downstream analyses, cILR might not always perform better than competing methods, especially when being used to generate inputs to predictive models.
-We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated with adjusting the mixture normal distribution.
-As such, we hope to refine the null distribution estimating procedure by either choosing a better distributional form, or to further constrain the optimization procedure of the mixture normal distribution by fixing the third and fourth moments.
-
-In addition, there are possible extensions cILR can we can consider to provide more flexibility across different data analysis scenarios in data analysis.
-First, cILR did not address the sparsity of microbiome taxonomic data and relies on a pseudocount to ensure log operations are valid.
-We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
-Second, cILR also treated all taxa within the set as equally contributing to the set.
-Incorporation of taxa-specific weights could reduce the influence of outliers, such as rare or highly invariant taxa.
-Finally, curating sets based on \emph{apriori} characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.
-
-\section*{Conclusion}
-Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomics data sets.
-However, limited work has been done developing set based methods specifically for microbiome relative abundance data.
-We introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level.
-We demonstrated that our method can control for type I error for significance testing at the sample level, while generated scores are also valid inputs in downstream analyses, including disease prediction and differential abundance.
-
-\section*{Acknowledgments}
-The authors thank Becky Lebeaux, Modupe Coker, Erika Dade, Jie Zhou, and Weston Viles for insightful comments and suggestions that greatly improved the paper.
-
-\nolinenumbers
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-% \bibliography{tax_agg}{}
-
-\begin{thebibliography}{10}
-
-\bibitem{proctor2019}
-Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
-\newblock The {{Integrative Human Microbiome Project}}.
-\newblock Nature. 2019;569(7758):641--648.
-\newblock doi:{10.1038/s41586-019-1238-8}.
-
-\bibitem{sharma2019}
-Sharma S, Tripathi P.
-\newblock Gut Microbiome and Type 2 Diabetes: Where We Are and Where to Go?
-\newblock The Journal of Nutritional Biochemistry. 2019;63:101--108.
-\newblock doi:{10.1016/j.jnutbio.2018.10.003}.
-
-\bibitem{aoun2020}
-Aoun A, Darwish F, Hamod N.
-\newblock The {{Influence}} of the {{Gut Microbiome}} on {{Obesity}} in
-  {{Adults}} and the {{Role}} of {{Probiotics}}, {{Prebiotics}}, and
-  {{Synbiotics}} for {{Weight Loss}}.
-\newblock Preventive Nutrition and Food Science. 2020;25(2):113--123.
-\newblock doi:{10.3746/pnf.2020.25.2.113}.
-
-\bibitem{cho2012}
-Cho I, Blaser MJ.
-\newblock The Human Microbiome: At the Interface of Health and Disease.
-\newblock Nature Reviews Genetics. 2012;13(4):260--270.
-\newblock doi:{10.1038/nrg3182}.
-
-\bibitem{callahan2016}
-Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
-\newblock {{DADA2}}: {{High}}-Resolution Sample Inference from {{Illumina}}
-  Amplicon Data.
-\newblock Nature Methods. 2016;13(7):581--583.
-\newblock doi:{10.1038/nmeth.3869}.
-
-\bibitem{truong2015}
-Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
-\newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
-\newblock Nature Methods. 2015;12(10):902--903.
-\newblock doi:{10.1038/nmeth.3589}.
-
-\bibitem{li2019}
-Li H.
-\newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
-  {{Metagenomics}}.
-\newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
-  2019. p. 977--550.
-
-\bibitem{li2015}
-Li H.
-\newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
-  Compositional Data Analysis}}.
-\newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
-\newblock doi:{10.1146/annurev-statistics-010814-020351}.
-
-\bibitem{shi2016}
-Shi P, Zhang A, Li H.
-\newblock Regression Analysis for Microbiome Compositional Data.
-\newblock The Annals of Applied Statistics. 2016;10(2):1019--1040.
-\newblock doi:{10.1214/16-AOAS928}.
-
-\bibitem{sankaran2014}
-Sankaran K, Holmes S.
-\newblock {{structSSI}}: {{Simultaneous}} and {{Selective Inference}} for
-  {{Grouped}} or {{Hierarchically Structured Data}}.
-\newblock Journal of statistical software. 2014;59(13):1--21.
-\newblock doi:{10.18637/jss.v059.i13}.
-
-\bibitem{benjamini1995}
-Benjamini Y, Hochberg Y.
-\newblock Controlling the {{False Discovery Rate}}: {{A Practical}} and
-  {{Powerful Approach}} to {{Multiple Testing}}.
-\newblock Journal of the Royal Statistical Society Series B (Methodological).
-  1995;57(1):289--300.
-
-\bibitem{chen2018}
-Chen J, King E, Deek R, Wei Z, Yu Y, Grill D, et~al.
-\newblock An Omnibus Test for Differential Distribution Analysis of Microbiome
-  Sequencing Data.
-\newblock Bioinformatics. 2018;34(4):643--651.
-\newblock doi:{10.1093/bioinformatics/btx650}.
-
-\bibitem{weiss2017}
-Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
-\newblock Normalization and Microbial Differential Abundance Strategies Depend
-  upon Data Characteristics.
-\newblock Microbiome. 2017;5(1).
-\newblock doi:{10.1186/s40168-017-0237-y}.
-
-\bibitem{mcmurdie2014}
-McMurdie PJ, Holmes S.
-\newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
-  Inadmissible}}.
-\newblock PLOS Computational Biology. 2014;10(4):e1003531.
-\newblock doi:{10.1371/journal.pcbi.1003531}.
-
-\bibitem{love2014}
-Love MI, Huber W, Anders S.
-\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
-  Data with {{DESeq2}}.
-\newblock Genome Biology. 2014;15(12):550.
-\newblock doi:{10.1186/s13059-014-0550-8}.
-
-\bibitem{quinn2019}
-Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
-\newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
-\newblock GigaScience. 2019;8(giz107).
-\newblock doi:{10.1093/gigascience/giz107}.
-
-\bibitem{quinn2018b}
-Quinn TP, Erb I, Richardson MF, Crowley TM.
-\newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
-\newblock Bioinformatics. 2018;34(16):2870--2878.
-\newblock doi:{10.1093/bioinformatics/bty175}.
-
-\bibitem{gloor2017}
-Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
-\newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
-  Optional}}.
-\newblock Frontiers in Microbiology. 2017;8.
-\newblock doi:{10.3389/fmicb.2017.02224}.
-
-\bibitem{aitchison1999}
-Aitchison J.
-\newblock A {{Concise Guide}} to {{Compositional Data Analysis}}. 1999; p. 134.
-
-\bibitem{kurtz2015}
-Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
-\newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
-  Ecological Networks}}.
-\newblock PLOS Computational Biology. 2015;11(5):e1004226.
-\newblock doi:{10.1371/journal.pcbi.1004226}.
-
-\bibitem{kaul2017}
-Kaul A, Davidov O, Peddada SD.
-\newblock Structural Zeros in High-Dimensional Data with Applications to
-  Microbiome Studies.
-\newblock Biostatistics. 2017;18(3):422--433.
-\newblock doi:{10.1093/biostatistics/kxw053}.
-
-\bibitem{kaul2017a}
-Kaul A, Mandal S, Davidov O, Peddada SD.
-\newblock Analysis of {{Microbiome Data}} in the {{Presence}} of {{Excess
-  Zeros}}.
-\newblock Frontiers in Microbiology. 2017;8.
-\newblock doi:{10.3389/fmicb.2017.02114}.
-
-\bibitem{khatri2012}
-Khatri P, Sirota M, Butte AJ.
-\newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
-  {{Outstanding Challenges}}.
-\newblock PLOS Computational Biology. 2012;8(2):e1002375.
-\newblock doi:{10.1371/journal.pcbi.1002375}.
-
-\bibitem{goeman2007}
-Goeman JJ, B{\"u}hlmann P.
-\newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
-  Issues.
-\newblock Bioinformatics. 2007;23(8):980--987.
-\newblock doi:{10.1093/bioinformatics/btm051}.
-
-\bibitem{subramanian2005}
-Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
-\newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
-  Interpreting Genome-Wide Expression Profiles.
-\newblock Proceedings of the National Academy of Sciences.
-  2005;102(43):15545--15550.
-\newblock doi:{10.1073/pnas.0506580102}.
-
-\bibitem{ashburner2000}
-Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
-\newblock Gene {{Ontology}}: Tool for the Unification of Biology.
-\newblock Nature genetics. 2000;25(1):25--29.
-\newblock doi:{10.1038/75556}.
-
-\bibitem{hanzelmann2013}
-H{\"a}nzelmann S, Castelo R, Guinney J.
-\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
-  {{RNA}}-{{Seq}} Data.
-\newblock BMC Bioinformatics. 2013;14(1):7.
-\newblock doi:{10.1186/1471-2105-14-7}.
-
-\bibitem{frost2020a}
-Frost HR.
-\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
-  Method for Cell-Specific Gene Set Scoring.
-\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
-\newblock doi:{10.1093/nar/gkaa582}.
-
-\bibitem{mclaren2019}
-McLaren MR, Willis AD, Callahan BJ.
-\newblock Consistent and Correctable Bias in Metagenomic Sequencing
-  Experiments.
-\newblock eLife. 2019;8:e46923.
-\newblock doi:{10.7554/eLife.46923}.
-
-\bibitem{egozcue2005}
-Egozcue JJ, {Pawlowsky-Glahn} V.
-\newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
-  Analysis}}.
-\newblock Mathematical Geology. 2005;37(7):795--828.
-\newblock doi:{10.1007/s11004-005-7381-9}.
-
-\bibitem{chong2020}
-Chong J, Liu P, Zhou G, Xia J.
-\newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
-  Functional, and Meta-Analysis of Microbiome Data.
-\newblock Nature Protocols. 2020;15(3):799--821.
-\newblock doi:{10.1038/s41596-019-0264-1}.
-
-\bibitem{tian2005}
-Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
-\newblock Discovering Statistically Significant Pathways in Expression
-  Profiling Studies.
-\newblock Proceedings of the National Academy of Sciences.
-  2005;102(38):13544--13549.
-\newblock doi:{10.1073/pnas.0506577102}.
-
-\bibitem{egozcue2003}
-Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcelo-Vidal} C.
-\newblock Isometric {{Logratio Transformations}} for {{Compositional Data
-  Analysis}}.
-\newblock Mathematical Geology. 2003; p.~22.
-
-\bibitem{silverman2017}
-Silverman JD, Washburne AD, Mukherjee S, David LA.
-\newblock A Phylogenetic Transform Enhances Analysis of Compositional
-  Microbiota Data.
-\newblock eLife. 2017;6:e21887.
-\newblock doi:{10.7554/eLife.21887}.
-
-\bibitem{barbie2009}
-Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
-\newblock Systematic {{RNA}} Interference Reveals That Oncogenic
-  {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
-\newblock Nature. 2009;462(7269):108--112.
-\newblock doi:{10.1038/nature08460}.
-
-\bibitem{quast2013}
-Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
-\newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
-  Processing and Web-Based Tools.
-\newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
-\newblock doi:{10.1093/nar/gks1219}.
-
-\bibitem{delignette-muller2015}
-{Delignette-Muller} ML, Dutang C.
-\newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
-\newblock Journal of Statistical Software. 2015;64(4):1--34.
-
-\bibitem{benaglia2009}
-Benaglia T, Chauveau D, Hunter DR, Young D.
-\newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
-\newblock Journal of Statistical Software. 2009;32(6):1--29.
-
-\bibitem{washburne2017}
-Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL, Mukherjee S, et~al.
-\newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
-  Associations in Microbiome Datasets.
-\newblock PeerJ. 2017;5:e2969.
-\newblock doi:{10.7717/peerj.2969}.
-
-\bibitem{morton2017}
-Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
-  et~al.
-\newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
-\newblock mSystems. 2017;2(1).
-\newblock doi:{10.1128/mSystems.00162-16}.
-
-\bibitem{efron2004}
-Efron B.
-\newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
-\newblock Journal of the American Statistical Association.
-  2004;99(465):96--104.
-\newblock doi:{10.1198/016214504000000089}.
-
-\bibitem{wu2012}
-Wu D, Smyth GK.
-\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
-  Correlation.
-\newblock Nucleic Acids Research. 2012;40(17):e133.
-\newblock doi:{10.1093/nar/gks461}.
-
-\bibitem{calgaro2020}
-Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
-\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
-  and Metagenomics Applied to Microbiome Data.
-\newblock Genome Biology. 2020;21(1):191.
-\newblock doi:{10.1186/s13059-020-02104-1}.
-
-\bibitem{cario1997}
-Cario MC.
-\newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
-  Distributions}} and {{Correlation Matrix}}. 1997; p.~19.
-
-\bibitem{agresti1998}
-Agresti A, Coull BA.
-\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
-  Estimation}} of {{Binomial Proportions}}.
-\newblock The American Statistician. 1998;52(2):119--126.
-\newblock doi:{10.2307/2685469}.
-
-\bibitem{delong1988}
-DeLong ER, DeLong DM, {Clarke-Pearson} DL.
-\newblock Comparing the Areas under Two or More Correlated Receiver Operating
-  Characteristic Curves: A Nonparametric Approach.
-\newblock Biometrics. 1988;44(3):837--845.
-
-\bibitem{hawinkel2019}
-Hawinkel S, Mattiello F, Bijnens L, Thas O.
-\newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
-  Control the False Discovery Rate.
-\newblock Briefings in Bioinformatics. 2019;20(1):210--221.
-\newblock doi:{10.1093/bib/bbx104}.
-
-\bibitem{breiman2001}
-Breiman L.
-\newblock Random {{Forests}}.
-\newblock Machine Learning. 2001;45(1):5--32.
-\newblock doi:{10.1023/A:1010933404324}.
-
-\bibitem{kuhn2020}
-Kuhn M, Wickham H.
-\newblock Tidymodels: {{Easily}} Install and Load the 'tidymodels' Packages;
-  2020.
-
-\bibitem{xiao2018}
-Xiao J, Chen L, Yu Y, Zhang X, Chen J.
-\newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
-  {{Predictive Modeling}} of {{Microbial Community Data}}.
-\newblock Frontiers in Microbiology. 2018;9.
-\newblock doi:{10.3389/fmicb.2018.03112}.
-
-\bibitem{dong2020}
-Dong M, Li L, Chen M, Kusalik A, Xu W.
-\newblock Predictive Analysis Methods for Human Microbiome Data with
-  Application to {{Parkinson}}'s Disease.
-\newblock PLOS ONE. 2020;15(8):e0237779.
-\newblock doi:{10.1371/journal.pone.0237779}.
-
-\bibitem{pasolli2017}
-Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
-\newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
-\newblock Nature Methods. 2017;14(11):1023--1024.
-\newblock doi:{10.1038/nmeth.4468}.
-
-\bibitem{schiffer2019}
-Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
-\newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
-  through Bioconductor.
-\newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
-
-\bibitem{gonzalez2018}
-Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
-  Ackermann G, et~al.
-\newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
-\newblock Nature Methods. 2018;15(10):796--798.
-\newblock doi:{10.1038/s41592-018-0141-9}.
-
-\bibitem{consortium2012}
-Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
-  et~al.
-\newblock Structure, Function and Diversity of the Healthy Human Microbiome.
-\newblock Nature. 2012;486(7402):207--214.
-\newblock doi:{10.1038/nature11234}.
-
-\bibitem{thurnheer2016}
-Thurnheer T, Bostanci N, Belibasakis GN.
-\newblock Microbial Dynamics during Conversion from Supragingival to
-  Subgingival Biofilms in an in Vitro Model.
-\newblock Molecular Oral Microbiology. 2016;31(2):125--135.
-\newblock doi:{10.1111/omi.12108}.
-
-\bibitem{beghini2019}
-Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
-\newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
-  in the {{New York City Health}} and {{Nutrition Examination Study}}.
-\newblock Annals of Epidemiology. 2019;34:18--25.e3.
-\newblock doi:{10.1016/j.annepidem.2019.03.005}.
-
-\bibitem{matteocalgaro2020}
-Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
-\newblock Zenodo.
-
-\bibitem{nielsen2014}
-Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
-\newblock Identification and Assembly of Genomes and Genetic Elements in
-  Complex Metagenomic Samples without Using Reference Genomes.
-\newblock Nature Biotechnology. 2014;32(8):822--828.
-\newblock doi:{10.1038/nbt.2939}.
-
-\bibitem{gevers2014}
-Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
-  et~al.
-\newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
-  {{Disease}}.
-\newblock Cell Host \& Microbe. 2014;15(3):382--392.
-\newblock doi:{10.1016/j.chom.2014.02.005}.
-
-\bibitem{martin2020}
-Martin BD, Witten D, Willis AD.
-\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
-  Regression.
-\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
-\newblock doi:{10.1214/19-AOAS1283}.
-
-\bibitem{naim2012}
-Naim I, Gildea D.
-\newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
-  {{Unbalanced Mixing Coefficients}}.
-\newblock Proceedings of the 29th International Coference on International
-  Conference on Machine Learning. 2012; p.~8.
-\newblock doi:{10.5555/3042573.3042756}.
-
-\bibitem{wu2016}
-Wu C, Chen J, Kim J, Pan W.
-\newblock An Adaptive Association Test for Microbiome Data.
-\newblock Genome Medicine. 2016;8(1):56.
-\newblock doi:{10.1186/s13073-016-0302-3}.
-
-\bibitem{martin-fernandez2012}
-{Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
-  {Palarea-Albaladejo} J.
-\newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
-  {{Classical}} and Robust Approaches.
-\newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
-\newblock doi:{10.1016/j.csda.2012.02.012}.
-
-\end{thebibliography}
-
-\section*{Supporting information}
-
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Fig.}
-\label{S1_Fig}
-{\bf Computational performance of cILR.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
-Evaluation was performed on simulated data sets.
-For sample size analysis, only 10 sets were evaluated.
-For taxa set analysis, sample size was fixed at 1,000.
-Across all evaluations, the size of each taxa set was also fixed at 50.
-
-\paragraph*{S2 Fig.}
-\label{S2_Fig}
-{\bf Distribution of p-values.} Q-Q plot of 10,000 p-values compared against a uniform distribution bounded between 0 and 1.
-Evaluation was performed on simulated null data sets of 10,000 samples testing for enrichment of a set of size 50.
-For correlation of 0.5, p-values represent correlation adjusted cILR while for correlation of 0, p-values represent unadjusted cILR.
-
-\paragraph*{S1 File.}
-\label{S1_File}
-{\bf Supplemental derivations.} Includes additional details on addressing variance inflation due to correlation in cILR, as well as computational performance and p-value distribution of the method.
-
-\end{document}

### 2022-01-26 cbea_revised.tex

@@
 
-\subsection*{Sample level inference}
+\subsection*{Statistical Significance}
+\subsubsection*{Inference at the sample level}
 CBEA provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{methods}.
@@
     \centering
-    \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
-    \caption{Classification performance via AUC of CBEA, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}.
-Performance scores measure whether scores can highly rank samples that are known to have inflated abundance.
-In the gingival data set presented in panel \textbf{(B)}, samples from the supragingival site are assumed to have an inflated abundance of aerobic microbes.
-Error bars are the 95\% DeLong confidence intervals for AUC \cite{delong1988}}
-    \label{fig:3}
+    \includegraphics[width = \textwidth]{figures/data_ss_fdr_new.png}
+    \caption{Type I error for sample-level inference with CBEA under different parametric assumptions compared against a Wilcoxon rank-sum test at a nominal $\alpha$ of 0.05. } 
+    \label{fig:2}
 \end{figure}

### 2022-01-26 cbea_revised.tex

--- 2022-01-26 cbea_revised.tex
@@
 In addition to statistical power, we also assessed phenotype relevance through evaluating whether highly ranked sets based on CBEA scores are more likely to be enriched according to ground truth.
-This is represented by the area under the receiving operator curve (AUROC/AUC) scores computed on CBEA scores against true labels (similar to strategy employed by VAM \cite{frost2020}).
+This is represented by the area under the receiving operator curve (AUROC/AUC) scores computed on CBEA scores against true labels (similar approach was used to evaluate VAM \cite{frost2020}).
 DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate.
@@
 In this section, we present results for evaluating statistical significance, phenotype relevance, and predictive performance.
-In addition to real data, we also evaluated models based on parametric simulations.
-Results for this section can be found in Supplemental Materials.
-
-\subsection*{Sample level inference}
-CBEA provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{methods}.
-Here, we present empirical results for this application of CBEA assessing type I error, power, and classification capacity.
+In addition to real data, we also evaluated models based on parametric simulations, where results can be found in the Supplemental Materials.
+
+\subsection*{Statistical Significance}
+\subsubsection*{Inference at the sample level}
+CBEA provides significance testing at the sample level through a self-contained competitive null hypothesis.
+Generating random sets approximate the global null setting where within each sample, sets generated by randomly sampling taxa should not be significantly more enriched than remainder taxa.
 
@@
     \centering
-    \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
-    \caption{Classification performance via AUC of CBEA, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}.
-Performance scores measure whether scores can highly rank samples that are known to have inflated abundance.
-In the gingival data set presented in panel \textbf{(B)}, samples from the supragingival site are assumed to have an inflated abundance of aerobic microbes.
-Error bars are the 95\% DeLong confidence intervals for AUC \cite{delong1988}}
-    \label{fig:3}
+    \includegraphics[width = \textwidth]{figures/data_ss_fdr_new.png}
+    \caption{Random taxa set analyses for inference at the sample level of CBEA under different parametric assumptions compared against a Wilcoxon rank-sum test.
+Type I error (\emph{y}-axis) was evaluated by generating random sets of different sizes (\emph{x}-axis) (500 replications per size) and computing the fraction of samples where the set was found to be significantly enriched at $\alpha = 0.05$.
+Confidence bands represent the mean type I error $\pm$ sample standard error computed across 500 replications of the experiment.}
+    \label{fig:2}
 \end{figure}
+
+Fig~\ref{fig:2} demonstrates type I error of sample-level inference evaluated using the random set approach.
+The Wilcoxon rank sum test and unadjusted CBEA under mixture normal assumption demonstrated good type I error control at the appropriate $\alpha$ level.
+This fits with our expectations since the mixture normal distribution has much better fit than the normal distribution especially at the tails (Fig~\ref{fig:1}).
+However, other variants of CBEA demonstrated inflated type I error, especially correlation adjusted variants compared to their unadjusted counter parts.
+Encouragingly, all methods demonstrate consistent performance across all set sizes, with a slight increase in type I error at the highest levels.
+
+\subsection*{Phenotype Relevance}.
 

### 2022-01-27 cbea_revised.tex

@@
 Type I error (\emph{y}-axis) was evaluated by generating random sets of different sizes (\emph{x}-axis) (500 replications per size) and computing the fraction of samples where the set was found to be significantly enriched at $\alpha = 0.05$.
-Confidence bands represent the mean type I error $\pm$ sample standard error computed across 500 replications of the experiment.}
+Error bars represent the mean type I error $\pm$ sample standard error computed across 500 replications of the experiment.}
     \label{fig:2}
@@
 
-\subsection*{Phenotype Relevance}.
-
+\subsection*{Phenotype Relevance}
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width = \textwidth]{figures/data_ss_pwr_new.png}
+    \caption{Statistical power (\textbf{A}) and score rankings (\textbf{B}) to assess phenotype relevance.
+Power (\emph{x}-axis) was estimated as the overall fraction of aerobic microbes found to be enriched in supragingival samples at $\alpha = 0.05$.
+95\% confidence intervals were computed using the Agresti-Couli approach for binomial proportions.
+Score rankings were evaluated by comparing computed scores against true values using AUCROC (\emph{x}-axis).
+DeLong 95 \% confidence intervals for AUROC were computed.}
+    \label{fig:3}
+\end{figure}
+
+Unsurprisingly, our approaches for type I 
 However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B.

### 2022-01-27 supplemental.tex

@@
 
-\title{Supplementary Materials for cILR: Taxonomic enrichment analysis with isometric log ratios}
+\title{Supplementary Materials for CBEA: Competitive Balances for Taxonomic Enrichment Analysis}
 \begin{document}
@@
 \section{Addressing variance inflation due to correlation}
-In order to perform inference with competitive isometric log ratio (cILR), we estimated the null distribution empirically following Efron et al.
-\cite{efron2004}.
+
+In order to perform inference with CBEA, we estimated the null distribution empirically.
+This can be done either non-parametrically by constructing a null distribution through computing scores on multiple permutations of the data, or parametrically via estimating parameters of a distribution using the same permuted scores.
+This means that our null distribution for a given set is equivalent to scores computed for sets of similar sizes but containing randomly chosen taxa.
 We chose two distributional forms for the null: the normal distribution and a two-component mixture normal distribution.
-Subsequently, we estimated the parameters of the respective distributions from a column (or variable) permuted raw score matrix.
-This means that our null distribution for a given set is equivalent to scores computed for sets of similar sizes but containing randomly chosen taxa.
 For the normal distribution, we estimated parameters using the maximum likelihood using the \emph{fitdistrplus} package \cite{delignette-muller2015}.
@@
 
-However, when variables are correlated, the variance of the sample mean of the variables is inflated \cite{wu2012}.
-Without loss of generalizability, for a set of variables $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:
-\begin{equation} \label{eq:1}
-    Var(\bar{x}) = \frac{1}{m^2}\left(\sum_{i = 1}(\sigma_i^2) + \sum_{i < j}\rho_{ij}\sigma_i\sigma_j\right)
-\end{equation}
-where $\sigma_i$ is the standard deviation of variable $i$ while $\rho_{ij}$ is the correlation between $i$ and $j$.
-The second term of (\ref{eq:1}) is the correlation dependent variance component, which goes to 0 if there is no correlation.
-Wu et al.
-\cite{wu2012} showed that a null distribution estimated from a column permuted matrix does not account for this variance inflation, since the permutation procedure disrupts the natural correlation structure of the original variables.
-
-The cILR statistic follows a similar pattern.
-Since the geometric mean of a set of variables is equivalent to the exponential of the arithmetic mean of their logarithms, we can re-write cILR for a set $k$ with size $\kappa$ as follows:
-\begin{equation}\label{eq:2}
-    cILR_k = \sqrt{\frac{\kappa(p - \kappa)}{\kappa + (p - \kappa)}} \left( \overbar{\log{X_{j|j \in K}}} - \overbar{\log{X_{j|j \notin K}}} \right)   
-\end{equation}
-where $p$ is the overall number of taxa, $j$ is the index of a taxa and $K$ is the set of indices of taxa in set $k$.
-The cILR statistic then looks similar to a t-statistic for difference in means of log-transformed proportions.
-As such, the pooled variance of cILR is dependent on the variance inflation of both mean components $\overbar{\log{X_{j|j \in K}}}$ and $\overbar{\log{X_{j|j \notin K}}}$.
-
-It is important to address this problem since there is strong inter-taxa correlation within the microbiome \cite{kurtz2015}.
-Even though Wu et al.
-proposed a correlation adjusted t-statistic for the CAMERA method \cite{wu2012}, it is difficult to adapt such modifications to the cILR statistic.
-This is because CAMERA requires an estimation procedure for pair-wise correlations of the $\log{X_j}$, as well as making the assumption that the marginal variance of each $\log{X_j}$ is equivalent to each other.
-
-Instead, our strategy is to combine the location (or mean) estimate from the column permuted raw score matrix with the spread (or variance) estimate from the original un-permuted scores.
+An advantage of estimating the null using a parametric approach is the ability to correct for variance inflation due to inter-taxa correlation within the set compared to overall background correlation \cite{wu2012}.
+CBEA addresses this issue by combining the location (or mean) estimate from the column permuted raw score matrix with the spread (or variance) estimate from the original un-permuted scores.
 This still allows us to leverage the null generated via column permutation while using the proper variance estimate taken from scores where the correlation structure has not been disrupted.
@@
 \end{equation}
+
 $\lambda'_1$, $\lambda'_2$, $\mu'_1$, $\mu'_2$, and $M'$ are component-wise mixing coefficients, component-wise means, and overall mean of the mixture distribution estimated from column-permuted scores while $SD$ is the overall standard deviation of the mixture distribution estimated from unpermuted scores.
 We solve for this optimization problem using a quasi Newton method with box constraints (L-BFGS-B) with the default finite-difference approximation of the gradient, as implemented in the \emph{optim} function in R.
+
+There are also different variations to this approach.
+We can choose to vary both $\sigma_1$ and $\sigma_2$ or to keep either $\sigma_1$ or $\sigma_2$ constant and varying the remaining component.
+We can assume that null distribution is a two-component mixture distribution where there is one major component with smaller mean (representing the bulk of the distribution centered at the permuted mean) and one minor component with higher mean (representing the inflated right tail).
+Under this assumption, we can modify the optimization problem to only estimate the variance parameter of the smaller component (i.e.
+withouth loss of generalizability keeping $\sigma_1$ constant where $\lambda'_1 > \lambda'_2$).
+This allows for the optimization procedure to more properly capture the right tail distribution rather than increasing weight on the left tail of the distribution which impacts the computation of p-values for a one-sided test.
+Empirical experiments (data not shown) done on simulated data and random set analyses suggest that this adjustment improves performance of the adjusted CBEA under mixture-normal assumption.
+In the R implementation of CBEA, users can control this behaviour by specifying the
 
@@
 \newpage
-\bibliography{tax_agg}{}
+\bibliography{mac_taxagg}{}
 \bibliographystyle{plos2015}

### 2022-01-27 cbea_revised.tex

@@
 
-Unsurprisingly, our approaches for type I 
-However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B.
-In situations where there is no inter-taxa correlation, CBEA still outperforms the wilcoxon rank sum test, however adjusted versions of CBEA did not perform as well as un-adjusted ones.
-However, in higher correlation scenarios, the difference in power is much more dramatic.
-At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted CBEA was only performing at 50\% power, while unadjusted CBEA and wilcoxon rank sum test were able to reach 80\%.
-These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of CBEA.
-Most importantly, CBEA demonstrate higher power in all scenarios where type I error is properly controlled.
-
-To further assess the utility of CBEA in classifying samples with enriched sets, we generated AUC scores for different CBEA scores using true labels of whether a sample has an inflated set.
-This analysis, therefore, assessed the relative ranking of samples using CBEA scores whereby high scores should correspond to samples that are known to be inflated.
-Fig~\ref{fig:3} presents this result.
-We compared different variants of CBEA against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
-Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), CBEA scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
-In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
-Notably, the W-statistic provided the least information for classifying samples with inflated taxa.
-
-\subsubsection*{Real data evaluations}  
-These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{methods}.
-Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations.
-As expected in Fig~\ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted CBEA methods.
-Conversely, adjusted CBEA controls for false positives adequately at the correct $\alpha$ level of 0.05.
-Power analysis (Fig~\ref{fig:2}D) showed similar patterns, where unadjusted CBEA methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
-    \caption{Classification performance via AUC of CBEA, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}.
-Performance scores measure whether scores can highly rank samples that are known to have inflated abundance.
-In the gingival data set presented in panel \textbf{(B)}, samples from the supragingival site are assumed to have an inflated abundance of aerobic microbes.
-Error bars are the 95\% DeLong confidence intervals for AUC \cite{delong1988}}
-    \label{fig:3}
-\end{figure}
+Fig~\ref{fig:3} shows results where we assess the phenotype relevance of CBEA results.
+For power analyses, we observed that adjusted CBEA approaches demonstrate much lower power compared to the Wilcoxon rank-sum test and unadjusted variants.
+This is surprising given the fact that in statistical significance analyses, the adjusted CBEA approach provides inflated type I error, especially if the normal distribution assumption was chosen.
+Here, we did not see any differences in performance between distribution choices.
+
+We also evaluate phenotype relevance by assessing whether enriched sets according to ground truth are preferentially ranked higher using CBEA scores.
+This aspect is captured through computing AUROC by comparing computed enrichment scores and true labels.
+Consistent with prior approaches, adjusting for correlation did not improve performance, where the AUROC values are equivalent to using the Wilcoxon Rank sum statistic at around 0.5.
+Unadjusted methods were much better at ranking true enriched sets, however the mean AUROC values are lower than alternate single sample enrichment methods (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}) even though this difference is not significant due to overlapping confidence intervals.
+Unfortunately, evaluating power and score rankings using real data does not allow for observing performance trends when effect sizes, sparsity, and inter-taxa correlation are varied.
+In equivalent simulation analyses, we observed the opposite effect, where
 

### 2022-01-27 saving_texts.tex

@@
 \end{figure}
+
+\subsection{Power analyses}
+However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B.
+In situations where there is no inter-taxa correlation, CBEA still outperforms the wilcoxon rank sum test, however adjusted versions of CBEA did not perform as well as un-adjusted ones.
+However, in higher correlation scenarios, the difference in power is much more dramatic.
+At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted CBEA was only performing at 50\% power, while unadjusted CBEA and wilcoxon rank sum test were able to reach 80\%.
+These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of CBEA.
+Most importantly, CBEA demonstrate higher power in all scenarios where type I error is properly controlled.
+
+To further assess the utility of CBEA in classifying samples with enriched sets, we generated AUC scores for different CBEA scores using true labels of whether a sample has an inflated set.
+This analysis, therefore, assessed the relative ranking of samples using CBEA scores whereby high scores should correspond to samples that are known to be inflated.
+Fig~\ref{fig:3} presents this result.
+We compared different variants of CBEA against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
+Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), CBEA scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
+In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
+Notably, the W-statistic provided the least information for classifying samples with inflated taxa.
+
+\subsubsection*{Real data evaluations}  
+These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{methods}.
+Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations.
+As expected in Fig~\ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted CBEA methods.
+Conversely, adjusted CBEA controls for false positives adequately at the correct $\alpha$ level of 0.05.
+Power analysis (Fig~\ref{fig:2}D) showed similar patterns, where unadjusted CBEA methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.

### 2022-01-27 saving_texts.tex
Updates from Overleaf

@@
 Power analysis (Fig~\ref{fig:2}D) showed similar patterns, where unadjusted CBEA methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
+
+\subsection*{Differential abundance analysis}
+CBEA generates sample-specific scores representing the degree of enrichment of a pre-defined set.
+As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test).
+We compared the performance of this approach with CBEA and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
+
+\subsubsection*{Simulation studies}
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
+    \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
+For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions).
+CBEA associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
+    \label{fig:4}
+\end{figure}
+
+Fig.~\ref{fig:4} present results for simulation studies for both type I error (panel A) and power (panel B) evaluations.
+All methods control for type I error well across both sparsity and correlation levels, where the estimated rate was consistently around the 0.05 pre-defined threshold.
+Results were similar across all evaluated methods, although in some instances, for example in medium correlation setting ($\rho = 0.2$), the unadjusted CBEA resulted in higher type I error, regardless of difference in means test and distribution of choice.
+
+The difference between the methods is more noticeable when evaluating power.
+All CBEA associated variants showed much higher power even when the effect size is limited (fold change is 1.5), and there is a noticeable gap in performance between CBEA and both DESeq2 and corncob.
+Surprisingly, this effect is consistent across correlation levels and sparsity, even though we expectedly see performance in power drop as a function of sparsity especially in low effect size settings.
+
+\subsubsection*{Real data evaluations}
+In addition to simulation studies, we also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP (Fig~\ref{fig:5}).
+For type I error evaluations, we use stool samples and randomly assign them with case/control status and calculated type I error as the proportion of genera identified as significantly different.
+For true positive rate evaluations, we use the gingival data set as detailed in the previous section, and calculated the true positive rate as the proportion of genera labeled as either anaerobic or aerobic that were found to be significant.
+
+We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of CBEA were controlling for type I error at the defined $\alpha$ threshold of 0.05.
+This is surprising given the consistency of preserving type I error for both corncob and DESeq2 in all simulation evaluations.
+
+In true positive experiments with data from the gingival site, estimated rates were more similar across the different methods.
+As expected, using the Wilcoxon rank sum test resulted in lower true positive rate compared to remaining methods, but the difference was not noticable.
+This is also surprising given that in simulation studies, both corncob and DESeq2 showed markedly lower power across all effect sizes.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
+    \caption{Differential abundance analysis using corncob, DESeq2 and CBEA with either Wilcoxon rank sum test or Welch's t-test.
+Panel \textbf{(A)} shows type I error results as the proportion of significant genera after 500 iterations where case/control status was assigned randomly to each sample.
+Panel \textbf{(B)} shows true positive rate results as the proportion of significant genera who are either obligate anaerobes or aerobes.
+Both evaluations use 16S rRNA gene sequencing data from HMP.
+Type I error evaluation used stool samples while the true positive rate evaluation used samples from the gingival site.
+Results showed that CBEA associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2}
+    \label{fig:5}
+\end{figure}

### 2022-01-27 cbea_revised.tex

@@
 
+\subsubsection*{Inference at the population level}
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width = \textwidth]{figures/data_diffab_fdr.eps}
+    \caption{Statistical power (\textbf{A}) and score rankings (\textbf{B}) to assess phenotype relevance.
+Power (\emph{x}-axis) was estimated as the overall fraction of aerobic microbes found to be enriched in supragingival samples at $\alpha = 0.05$.
+95\% confidence intervals were computed using the Agresti-Couli approach for binomial proportions.
+Score rankings were evaluated by comparing computed scores against true values using AUCROC (\emph{x}-axis).
+DeLong 95 \% confidence intervals for AUROC were computed.}
+    \label{fig:3}
+\end{figure}
+
 \subsection*{Phenotype Relevance}
@@
 DeLong 95 \% confidence intervals for AUROC were computed.}
-    \label{fig:3}
+    \label{fig:4}
 \end{figure}
@@
 Unadjusted methods were much better at ranking true enriched sets, however the mean AUROC values are lower than alternate single sample enrichment methods (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}) even though this difference is not significant due to overlapping confidence intervals.
+
 Unfortunately, evaluating power and score rankings using real data does not allow for observing performance trends when effect sizes, sparsity, and inter-taxa correlation are varied.
 In equivalent simulation analyses, we observed the opposite effect, where
-
-\subsection*{Differential abundance analysis}
-CBEA generates sample-specific scores representing the degree of enrichment of a pre-defined set.
-As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test).
-We compared the performance of this approach with CBEA and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
-
-\subsubsection*{Simulation studies}
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
-    \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
-For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions).
-CBEA associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
-    \label{fig:4}
-\end{figure}
-
-Fig.~\ref{fig:4} present results for simulation studies for both type I error (panel A) and power (panel B) evaluations.
-All methods control for type I error well across both sparsity and correlation levels, where the estimated rate was consistently around the 0.05 pre-defined threshold.
-Results were similar across all evaluated methods, although in some instances, for example in medium correlation setting ($\rho = 0.2$), the unadjusted CBEA resulted in higher type I error, regardless of difference in means test and distribution of choice.
-
-The difference between the methods is more noticeable when evaluating power.
-All CBEA associated variants showed much higher power even when the effect size is limited (fold change is 1.5), and there is a noticeable gap in performance between CBEA and both DESeq2 and corncob.
-Surprisingly, this effect is consistent across correlation levels and sparsity, even though we expectedly see performance in power drop as a function of sparsity especially in low effect size settings.
-
-\subsubsection*{Real data evaluations}
-In addition to simulation studies, we also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP (Fig~\ref{fig:5}).
-For type I error evaluations, we use stool samples and randomly assign them with case/control status and calculated type I error as the proportion of genera identified as significantly different.
-For true positive rate evaluations, we use the gingival data set as detailed in the previous section, and calculated the true positive rate as the proportion of genera labeled as either anaerobic or aerobic that were found to be significant.
-
-We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of CBEA were controlling for type I error at the defined $\alpha$ threshold of 0.05.
-This is surprising given the consistency of preserving type I error for both corncob and DESeq2 in all simulation evaluations.
-
-In true positive experiments with data from the gingival site, estimated rates were more similar across the different methods.
-As expected, using the Wilcoxon rank sum test resulted in lower true positive rate compared to remaining methods, but the difference was not noticable.
-This is also surprising given that in simulation studies, both corncob and DESeq2 showed markedly lower power across all effect sizes.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
-    \caption{Differential abundance analysis using corncob, DESeq2 and CBEA with either Wilcoxon rank sum test or Welch's t-test.
-Panel \textbf{(A)} shows type I error results as the proportion of significant genera after 500 iterations where case/control status was assigned randomly to each sample.
-Panel \textbf{(B)} shows true positive rate results as the proportion of significant genera who are either obligate anaerobes or aerobes.
-Both evaluations use 16S rRNA gene sequencing data from HMP.
-Type I error evaluation used stool samples while the true positive rate evaluation used samples from the gingival site.
-Results showed that CBEA associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2}
-    \label{fig:5}
-\end{figure}
 

### 2022-02-01 cbea_revised.tex
Updates from Overleaf

@@
 % Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
+\section*{Introduction} \label{introduction}
 The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host.
@@
 
-\section*{Evaluation}
+\section*{Evaluation} \label{evaluation}
 We based our evaluation strategy on gene set testing benchmarking standards set by Geistlinger et al.
@@
 This data set contains 369 samples split into two subsites: supragingival and subgingival.
+We processed this data set by removing all samples with total read counts less than 1000 and OTUs whose presence (at least 1 count) is in 10\% of samples or less.
 
@@
 This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease).
+We processed this data by removing all samples with less than 1,000 total read counts as well as any OTU who was present (at least 1 count) in 10\% of samples or less.
+Prior to model fitting, we back-transformed relative abundances into count data (to align the format with our 16S rRNA gene sequencing data set) using provided total number of reads aligned to MetaPhlan marker genes (per sample).
 
@@
 Of the 671 samples obtained, 500 samples belong to patients with IBD.
+We processed this data set by removing all samples with less than 1,000 total read counts as well as any OTU who was present (at least 1 count) in 10\% of samples or less.
 \end{enumerate}
@@
 We evaluate disease prediction performance by fitting a random forest model \cite{breiman2001} using inputs as CBEA scores to classify samples of patients with IBD and healthy controls.
-Random forest was chosen as a baseline learner due to its flexibility as an out-of-the-box model that is easy to fit without requiring complex hyperparameter tuning.
+Random forest was chosen as a baseline learner due to its flexibility as an out-of-the-box model that is easy to fit.
+In this instance we evaluated predictive performance of a default random forest model (without hyperparameter tuning) AUROC after 10-fold cross validation.
+Additionally, we utilized SMOTE to correct for class imbalances \cite{chawla2002}.
+Implementation was done using the \emph{tidymodels} suite of packages \cite{kuhn2020b}.
 
@@
 
+Similar to other single sample approaches to gene set testing such as GSVA \cite{hanzelmann2013}, we can perform inference at the population level by utilizing a two-sample difference in means test.
+Here, we evaluate using CBEA scores generated under different settings with Welch's t-test in a supervised manner to assess whether a set is enriched across case/control status.
+
 \begin{figure}[!h]
@@
     \includegraphics[width = \textwidth]{figures/data_diffab_fdr.eps}
-    \caption{Statistical power (\textbf{A}) and score rankings (\textbf{B}) to assess phenotype relevance.
-Power (\emph{x}-axis) was estimated as the overall fraction of aerobic microbes found to be enriched in supragingival samples at $\alpha = 0.05$.
-95\% confidence intervals were computed using the Agresti-Couli approach for binomial proportions.
-Score rankings were evaluated by comparing computed scores against true values using AUCROC (\emph{x}-axis).
-DeLong 95 \% confidence intervals for AUROC were computed.}
+    \caption{Random sample label (\textbf{A}) and random set (\textbf{B}) analyses for population level inference.
+(\textbf{A}) Type I error (\emph{x}-axis) was estimated as the overall fraction of sets found to be enriched $\alpha = 0.05$ using randomly generated sample labels (500 permutations).
+Error bars represent the mean type I error $\pm$ sample standard error.
+(\textbf{B}) Proportion of significant sets (\emph{y}-axis) using 100 randomly generated sets of different set sizes (\emph{x}-axis).
+Confidence intervals computed using Agresti-Couli method for binomial proportions.}
     \label{fig:3}
@@
 
+Fig~\ref{fig:3} shows results for this scenario using both random sample label and random set evaluations.
+The random sample label approach (Fig~\ref{fig:3}A) provides a controlled setting where we can estimate type I error rate controlled at $\alpha = 0.05$.
+Across all replications, CBEA methods were able to control for type I error at the nominal threshold of 0.05, with CBEA raw scores being the most performant.
+Neither output types, correlation adjustment, nor distributional assumption improved performance values.
+Surprisingly, DESeq2 and corncob both exhibit significantly inflated type I error.
+
+We also assessed the impact of set-size on the inference procedure by testing for enrichment using the original sample labels but with randomly sampled sets of different sizes.
+Overall we observed very similar values across CBEA as well as corncob and DESeq2, suggesting that no individual method is systematically identifying too many significant sets.
+Additionally, similar to analogous analyses at the sample level, no approach were significantly sensitive to set-size.
+
 \subsection*{Phenotype Relevance} 
-
+\subsubsection*{Inference at the sample level}
 \begin{figure}[!h]
@@
     \caption{Statistical power (\textbf{A}) and score rankings (\textbf{B}) to assess phenotype relevance.
-Power (\emph{x}-axis) was estimated as the overall fraction of aerobic microbes found to be enriched in supragingival samples at $\alpha = 0.05$.
+(\textbf{A}) Power (\emph{x}-axis) was estimated as the overall fraction of aerobic microbes found to be enriched in supragingival samples at $\alpha = 0.05$.
 95\% confidence intervals were computed using the Agresti-Couli approach for binomial proportions.
-Score rankings were evaluated by comparing computed scores against true values using AUCROC (\emph{x}-axis).
+(\textbf{B}) Score rankings were evaluated by comparing computed scores against true values using AUROC (\emph{x}-axis).
 DeLong 95 \% confidence intervals for AUROC were computed.}
@@
 
-Fig~\ref{fig:3} shows results where we assess the phenotype relevance of CBEA results.
-For power analyses, we observed that adjusted CBEA approaches demonstrate much lower power compared to the Wilcoxon rank-sum test and unadjusted variants.
+In Fig~\ref{fig:4}, we evaluate whether sets found to be significant by CBEA are relevant to the phenotype of interest.
+We leveraged the gingival data set as stated in \nameref{evaluation} section where we know beforehand that aerobic microbes are more likely to be enriched in supragingival subsite samples and vice versa.
+
+We estimated statistical power using this data set as the fraction of supragingival samples where the set representing aerobic microbes were significantly enriched.
+We observed that adjusted CBEA approaches demonstrate much lower power compared to the Wilcoxon rank-sum test and unadjusted variants.
 This is surprising given the fact that in statistical significance analyses, the adjusted CBEA approach provides inflated type I error, especially if the normal distribution assumption was chosen.
@@
 
-We also evaluate phenotype relevance by assessing whether enriched sets according to ground truth are preferentially ranked higher using CBEA scores.
-This aspect is captured through computing AUROC by comparing computed enrichment scores and true labels.
+We also evaluated phenotype relevance by assessing whether enriched sets according to ground truth are preferentially ranked higher using assigned continuous scores (instead of performing a hypothesis test).
+This aspect is captured through computing AUROC values comparing computed enrichment scores and true labels.
 Consistent with prior approaches, adjusting for correlation did not improve performance, where the AUROC values are equivalent to using the Wilcoxon Rank sum statistic at around 0.5.
@@
 
+\subsubsection*{Inference at the population level}
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width = \textwidth]{figures/data_diffab_pwr.png}
+    \caption{Statistical power to assess phenotype relevance of inference tasks at the population level.
+Power (\emph{x}-axis) was estimated as the overall fraction of sets representing genera that are aerobic or anaerobic microbes found to be differentially enriched across sample type (supragingival or subgingival).
+95\% confidence intervals were computed using the Agresti-Couli approach for binomial proportions.}
+    \label{fig:5}
+\end{figure}
+
+We also assessed statistical power for population level inference scenarios using a similar approach.
+Here, enrichment scores for sets representing all identified genera were computed, and estimated power as the fraction of sets found to be differentially enriched across sample site labels (supragingival or subgingival).
+We compared these results against performing a differential abundance test of genus level features generated via sum-based approaches.
+Results are shown in Fig~\ref{fig:5}.
+Some CBEA variants, such as CDF outputs for the mixture normal distributional assumption, did not correctly detect as many significant sets as DESeq2 or corncob despite very close performance values.
+Using raw CBEA scores was best approach, however it did not exceed values obtained from DESeq2 and corncob.
+
 \subsection*{Disease Prediction}   
 Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models.
-In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating CBEA, ssGSEA, and GSVA).
+In this section we assessed the predictive performance of a standard baseline random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating CBEA, ssGSEA, and GSVA).
 Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.
-
-\subsubsection*{Simulation studies}
-Fig~\ref{fig:6} shows results for simulation studies as detailed in the \nameref{methods} section.
-Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome.
-As expected, performance across all assessed methods increased with a higher signal-to-noise ratio.
-Both CLR and CBEA approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks.
-This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while CBEA and CLR did not experience a similar level of impact.
-As such, performance gap widens with increasing correlation and sparsity.
-Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.
-
-In this analysis, CBEA unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
-This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation.
-However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and CBEA/CLR.
-Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity.
-Additionally, we did not observe any performance difference between the different variations of CBEA.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
-    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
-Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure.
-Panel \textbf{(B)} presents performance on a classification task with AUC as the evaluation measure.
-CBEA approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
-    \label{fig:6}
-\end{figure}
 
@@
 
-\subsubsection*{Real data evaluations}
-In addition to parametric simulations, we also assessed the performance of using CBEA scores in predictive models with real data sets.
-Fig~\ref{fig:7} presents results for two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
+We fit our model to two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
 The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al.
@@
 CBEA performs better than GSVA and ssGSEA but not as well as CLR, with the exception of the whole genome sequencing data set.}
-    \label{fig:7}
+    \label{fig:6}
 \end{figure}
 
-Similar to simulation experiments, we also fitted a naive random forest model using CBEA, ssGSEA, GSVA, or CLR transformed variables as inputs, and use AUC as the performance criteria.
-Results also replicated that of the simulations, where across both data sets CBEA and CLR methods provide much better performance than both GSVA or ssGSEA.
+Fig~\ref{fig:6} illustrates results from using CBEA, ssGSEA, GSVA, or CLR transformed variables as inputs to our model, with AUROC as the performance criteria.
+Results also replicated that of the simulations (Supplementary materials), where across both data sets CBEA and CLR methods provide much better performance than both GSVA or ssGSEA.
 Interestingly, the CBEA approach performed better than CLR in the whole genome data set but did not perform as well in the 16S rRNA gene sequencing data set.
@@
 \subsection*{Inference with CBEA} 
+CBEA is a microbiome-specific approach to generate sample specific enrichment scores for taxonomic sets defined apriori.
 The formulation of CBEA as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}.
-Since CBEA generate sample-specific scores, this allows users to identify significantly enriched taxa even at the at the sample-level.
-We assessed the usage of CBEA in this type of analysis by evaluating type I error and power across both simulation studies and real data applications.
-Most importantly, we demonstrated that our adjusted CBEA approach was able to address the issue of variance inflation due to correlation \cite{wu2012} by controlling for type I error at the appropriate $\alpha$ level across different levels of simulated inter-taxa correlation (Fig~\ref{fig:2}) while conversely unadjusted CBEA and the naive Wilcoxon rank sum test showed much higher rates of error.
-This is further encouraged in real data analysis where the false discovery rate was around $0.05$ when a collection of true null and true alternate hypotheses were tested.
-Unfortunately the trade-off of good type I error control is lower power.
-The conservativeness of the test attenuates with higher sparsity and correlation, where power was not approaching even 50\% even at the highest effect sizes.
-However, when the degree of such data features are reasonable, CBEA will still be able to detect a reasonable proportion of samples with inflated counts.
-
-We also observed that choosing different distributional forms did not alter performance values for CBEA.
-This runs contrary to our comparison analysis in Fig~\ref{fig:1}, where we demonstrated that the mixture normal distribution had superior fit compared to the simple normal for raw CBEA scores computed under the global null.
-We hypothesized that this might be due to the difficulty in fitting mixture distribution to data using the expectation maximization algorithm, as the convergence rate is slow when there is high overlap between the mixtures, resulting in a small mixing coefficient for one of the components \cite{naim2012}.
-As such, in our implementation of CBEA, in order to ensure convergence for the estimating procedure we increased the number of iterations while relaxing the tolerance parameter.
-Furthermore, there are also possible problems with our adjustment procedure for the mixture distribution that might impact overall fit.
-In order to combine the scale and location estimate of two mixture distributions, we fixed the overall mean, standard deviation, mixing coefficient and component-wise means and used an optimization procedure to find the component-wise variances.
-However, this means that we have one equation for the overall variance but two possible parameters to estimate.
-As such, there is no guaranteed unique solution to component-wise variances.
-We hypothesized that the instability and degeneracy in component-wise variance estimates might impact the fidelity of estimates at the tails of the distribution, thereby affecting inference.
-
-Despite these concerns, empirical results still indicate that CBEA can confidently identify samples with inflated counts.
-The conservativeness of the correlation adjustment procedure ensures that significant results can be trusted by practitioners, even if CBEA might not be able to exhaustively identify all samples with inflated counts.
-In situations where either the data is less sparse (e.g.
-containing a lot of core taxa that are prevalent across all samples), there is less inter-taxa correlation within the set (e.g.
-taxa that do not participate in common pathways but have shared characteristics like pathogenicity), or if the effect size is large, then CBEA will still be able to produce reasonable power.
-A practitioner can use CBEA to screen for samples for subsequent analysis that might involve significant costs, or perform hypothesis generation using a less stringent criteria alongside a multiple testing adjustment procedure (such as Benjamini-Hochberg \cite{benjamini1995}).
-
-\subsection*{Downstream analysis}
-The sample-level enrichment scores generated by the CBEA method can be used in downstream analyses commonly performed in microbiome research: differential abundance testing and disease prediction.
-
-\subsubsection*{Differential abundance analysis}
-For differential abundance testing, we evaluated whether using CBEA scores alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes.
-We compared CBEA against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014} applied on data where taxa were aggregated using the naive sum method.
+Since this null hypothesis is self-contained per sample, this allows users perform enrichment testing at the sample level.
+Additionally, in combination with a difference in means test, CBEA can also test for enrichment at the population level across case/control status similar to GSVA \cite{hanzelmann2013}.
+
+For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}) while also demonstrating respectable power (Fig~\ref{fig:4}).
+This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), where the mixture normal displayed superior fit to the null distribution.
+Unfortunately, other variants of CBEA did not demonstrate good type I error control.
+Interestingly, while the adjusted methods showed inflated type I error and low power in real data evaluations (Fig~\ref{fig:2}), they performed well in simulation studies (\nameref{S1_Fig}, \nameref{S3_Fig}).
+
+For population-level inference task, CBEA also performed very well.
+Under the permutation global null, representing genera abundance using CBEA scores in combination with Welch's t-test controls for type I error the correct $\alpha$ threshold while also showing respectable power.
+Since population level enrichment test is equivalent to differential abundance using set-based features, we compared CBEA approach against using element-wise summations with corncob \cite{martin2020} and DESeq2 \cite{love2014} to test for set-level differential abundance.
 We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}.
-Conversely, corncob is a newer method developed specifically for microbiome taxonomic data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.
-
-Surprisingly, we found some conflicting results when evaluating comparisons across parametric simulations and real data analysis.
-The performance of CBEA was consistent across two evaluation criteria, demonstrating good type I error and respectable power.
-However, corncob and DESeq2 showed opposite effects: in simulation experiments, both methods show good type I error control but low power, while in real data analyses, conversely both corncob and DESeq2 showed inflated type I error but comparable power with respect to CBEA methods.
-Despite such discrepancy, results still indicate good performance of CBEA scores when used as inputs to downstream differential abundance analysis compared to using aggregated raw counts, even in methods designed to handle that type of data such as corncob and DESeq2.
-
-We hypothesized that the above behavior can arise due to issues with performing inference in the presence taxon-specific protocol biases \cite{mclaren2019}.
-According to McLaren et al., the observed relative abundance of taxa is different than the true relative abundance due to protocol bias, and importantly this bias is specific to each taxon \cite{mclaren2019}.
-This is especially true in the context of sum-based aggregations, where the resulting bias of the aggregated taxon are dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}).
+Alternately, corncob is a newer method developed specifically for microbiome taxonomic data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation.
+We observed that using this approach resulted in an inflated type I error compared to all variants of CBEA (Fig~\ref{fig:2}), yet did not improve power (Fig~\ref{fig:4}).
+Results for CBEA approaches were replicated in simulation analyses, however for corncob and DESeq2 we observed an opposite effect: in simulation experiments, both methods show good type I error control but low power (\nameref{S2_Fig}, \nameref{S4_Fig}).
+
+We hypothesized that the discrepancy between simulation and real data evaluations can can arise due to differences in our assumptions regarding the data generating process that informed our simulation schema.
+One important consideration, as described in the \nameref{introduction} section, are taxon-specific biases that distort the observed relative abundance of taxa compred to true values \cite{mclaren2019}.
+In the context of sum-based aggregations, the resulting bias of the aggregated taxon are dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}).
 Conceptually, this means that measurement error for a taxon aggregate is different across samples as relative abundance of contributing changes, leading to issues when attempting to perform inference.
-As such, we expect methods like corncob or DESeq2 when performed on such aggregates to have inflated type I error compared to our taxon-ratio based approach.
-
-The bias model also helps explain differences in performance of DESeq2 and corncob in simulation analyses compared to real data.
-Our simulation protocol does not explicitly include bias in our formulation, and all taxa were generated from the same underlying distribution with similar variances across all samples (the only difference is in the mean value where a taxa is expected to have inflated counts).
-As such, we do not expect our simulated taxa to have any taxon-specific biases, which is not the case in real data settings.
-Therefore, we can expect DESeq2 and corncob to retain their expected type I error control in simulation analyses compared to real data.
+As such, we expect methods like corncob or DESeq2 when performed on such sum aggregates in the presence of taxon-specific biases to have inflated type I error compared to our taxon-ratio based approach.
+Conversely, in simulation studies, we did not explicitly include bias in our formulation and all taxa were generated from the same negative binomial distribution with similar variances across all samples (the only difference is when an enriched taxa is expected to have inflated means).
+As such, DESeq2 and corncob controls well for type I error in simulation scenarios where the effect of taxon-specific biases is absent.
 It is still surprising to see lower power for both methods in simulation analyses, which might be due to the fact that the evaluation protocol only considers default settings for both methods and did not attempt to optimize performance.
-
-The fact that the performance of CBEA remains consistent across both simulation and real data analysis shows that CBEA is invariant to taxon-specific biases.
-Furthermore, our evaluation indicates that even a simple difference in means test when used in combination with CBEA scores can preserve type I error while maintaining good power.
-As such, a practioner can use CBEA as a pre-processing step prior to performing a differential abundance test.
-
-\subsubsection*{Predictive models}
-We evaluated whether CBEA can be used to generate set-based features for disease prediction models.
-Here, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using CBEA generated scores as inputs.
+We also suspect that
+
+\subsection*{Downstream analysis using predictive models}
+The sample-level enrichment scores generated by the CBEA method can be used in downstream analyses such as disease prediction.
+Here, we evaluated whether CBEA can be used to generate set-based features for disease prediction models.
+
+We fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using CBEA generated scores as inputs.
 Similar to our inference analysis, we compared CBEA against both ssGSEA and GSVA.
@@
 
-Similar to other experiments in downstream usage of CBEA, performance did not change with different underlying distributions, output types, or correlation status.
-This is surprising since we expect z-scores to perform better as they are able to capture the direction of an association.
-The fact that this effect persisted even onto our real data analysis suggests that this is not due to a deficiency of our simulation design.
-As such, practitioners who wish to use CBEA in predictive models might be suited to use the settings that is the fastest to compute.
-
-Ultimately, results indicate that CBEA can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity.
-Even though there exists situations where it might not provide maximum predictive values, the flexibility of CBEA in various types of analyses enable even though in some scenarios it might not provide maximum predictive values.
-
 \subsection*{Limitations and future directions}
-There are various limitations to our evaluation of CBEA.
+
+These above results demonstrate the applicability of CBEA under different data analysis scenarios.
+If researchers are interested in perform inference, they can decide between an unsupervised sample level approach (i.e.
+screen samples for enrichment of certain characteristics) or a supervised population level approach (i.e.
+identifying characteristics that are differentially abundant across case/control status).
+
+For the unsupervised approach, utilizing the unadjusted CBEA with the mixture normal distribution provides a good initial starting point.
+In the case where the screening procedure only wants to detect mean-inflated taxon sets (instead of additionally detecting taxon sets with increased correlation), users can apply the adjusted approach with the normal distribution, which can be effective at conserving type I error even at high correlation scenarios.
+However, the trade off for this adjustment is power that decreases with increasing correlation.
+For the supervised analysis, all CBEA variants control well for type I error and provides adequate statistical power.
+However, using raw CBEA scores with difference-in-means test such as Welch's t-test is preferable since it requires the least amount of computing resource (no estimation process) while still outperforming using sum-based approach with a standard differential abundance test.
+
+Beyond inference, CBEA scores are flexible and can be useful for downstream analysis.
+We demonstrated that for a given number of set-based features, CBEA can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity.
+This is especially true for whole genome sequencing data sets, where CBEA outperfrorms the standard approach of applying a centered log-ratio transformation.
+Even though we only evaluated prediction models in this manuscript, researchers can benchmark their own usage of CBEA for other downstream tasks such as sample ordination.
+
+However, there are various limitations to our evaluation of CBEA.
 First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome taxonomic data.
 There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016a} have been used in the evaluation of prior studies.
-Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is not perfect.
-This is because the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut.
-As such, results from power analyses using this data set is only relative between the comparison methods.
+Second, we were not able to evaluate the phenotypical relevance of enrichment results as in Geistlinger et al.
+\cite{geistlinger2021} due to limited consistent annotations for microbiome signatures in health and disease, especially those that are experimentally verified (and not just from differential abundance studies).
+We attempted to perform this evaluation by leveraging the gingival data set similar to \cite{calgaro2020}.
+However, we acknowledge that this is not a perfect solution, since oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut.
+As such, results from power analyses using this data set is only relative between the comparison methods instead of treated as absolute measures of power or phenotype relevance.
 Finally, we assumed that taxa within a set are all equally associated with the outcome.
@@
 
-Our evaluation showed various drawbacks of the CBEA method.
-First, inference with CBEA is limited in being able to exhaustively detect all samples with significant inflated counts for a set in situations where there is a high degree of sparsity and inter-taxa correlation.
+Our evaluation also showed various drawbacks of the CBEA method itself.
+First, inference with CBEA at the sample level is limited, and can be affected by inter-taxa correlation if users wish to only detect mean-inflated sets, especially when inter-taxa correlation within the set is high.
 Second, for downstream analyses, CBEA might not always perform better than competing methods, especially when being used to generate inputs to predictive models.
-We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated with adjusting the mixture normal distribution.
+We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated the estimation procedure associated with adjusting the mixture normal distribution.
 As such, we hope to refine the null distribution estimating procedure by either choosing a better distributional form, or to further constrain the optimization procedure of the mixture normal distribution by fixing the third and fourth moments.
 
-In addition, there are possible extensions CBEA can we can consider to provide more flexibility across different data analysis scenarios in data analysis.
-First, CBEA did not address the sparsity of microbiome taxonomic data and relies on a pseudocount to ensure log operations are valid.
-We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
+In addition, CBEA itself did not consider other aspects of microbiome data that can merit further extension.
+First, across all analyses, we relied on adding a pseudocount to ensure log operations are valid.
+Users can directly addressing this by incorporating model-based zero correction methods prior to modelling such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
+However, in our simulation studies, sparsity seems to not have a significant impact on the overall performance of our approach.
 Second, CBEA also treated all taxa within the set as equally contributing to the set.
-Incorporation of taxa-specific weights could reduce the influence of outliers, such as rare or highly invariant taxa.
-Finally, curating sets based on \emph{apriori} characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.
+Incorporation of taxa-specific weights (similar to PhILR \cite{silverman2017} could reduce the influence of outliers, such as rare or highly invariant taxa.
+Finally, even though for a given set of apriori annotations CBEA can generate useful summary scores, such values are limited in their utility if the annotation themselves are not meaningful.
+As such, curating and validating sets (similar to MSigDB \cite{subramanian2005}) based on physiological or genomic characteristics of microbes \cite{weissman2021} or their association with human disease (in beta BugSigDB \url{https://bugsigdb.org/Main_Page}) can allow for incorporating functional insights into microbiome-outcome analyses.
 
@@
 
-%\bibliography{references}{}
-\begin{thebibliography}{10}
-
-\bibitem{proctor2019}
-Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
-\newblock The {{Integrative Human Microbiome Project}}.
-\newblock Nature. 2019;569(7758):641--648.
-\newblock doi:{10.1038/s41586-019-1238-8}.
-
-\bibitem{sharma2019a}
-Sharma S, Tripathi P.
-\newblock Gut Microbiome and Type 2 Diabetes: Where We Are and Where to Go?
-\newblock The Journal of Nutritional Biochemistry. 2019;63:101--108.
-\newblock doi:{10.1016/j.jnutbio.2018.10.003}.
-
-\bibitem{aoun2020}
-Aoun A, Darwish F, Hamod N.
-\newblock The {{Influence}} of the {{Gut Microbiome}} on {{Obesity}} in
-  {{Adults}} and the {{Role}} of {{Probiotics}}, {{Prebiotics}}, and
-  {{Synbiotics}} for {{Weight Loss}}.
-\newblock Preventive Nutrition and Food Science. 2020;25(2):113--123.
-\newblock doi:{10.3746/pnf.2020.25.2.113}.
-
-\bibitem{cho2012}
-Cho I, Blaser MJ.
-\newblock The Human Microbiome: At the Interface of Health and Disease.
-\newblock Nature Reviews Genetics. 2012;13(4):260--270.
-\newblock doi:{10.1038/nrg3182}.
-
-\bibitem{callahan2016}
-Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
-\newblock {{DADA2}}: High-Resolution Sample Inference from {{Illumina}}
-  Amplicon Data.
-\newblock Nature Methods. 2016;13(7):581--583.
-\newblock doi:{10.1038/nmeth.3869}.
-
-\bibitem{truong2015}
-Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
-\newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
-\newblock Nature Methods. 2015;12(10):902--903.
-\newblock doi:{10.1038/nmeth.3589}.
-
-\bibitem{li2019a}
-Li H.
-\newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
-  {{Metagenomics}}.
-\newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
-  2019. p. 977--550.
-
-\bibitem{li2015}
-Li H.
-\newblock Microbiome, {{Metagenomics}}, and {{High}}-{{Dimensional
-  Compositional Data Analysis}}.
-\newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
-\newblock doi:{10.1146/annurev-statistics-010814-020351}.
-
-\bibitem{gloor2017}
-Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
-\newblock Microbiome {{Datasets Are Compositional}}: And {{This Is Not
-  Optional}}.
-\newblock Frontiers in Microbiology. 2017;8.
-\newblock doi:{10.3389/fmicb.2017.02224}.
-
-\bibitem{khatri2012}
-Khatri P, Sirota M, Butte AJ.
-\newblock Ten {{Years}} of {{Pathway Analysis}}: Current {{Approaches}} and
-  {{Outstanding Challenges}}.
-\newblock PLOS Computational Biology. 2012;8(2):e1002375.
-\newblock doi:{10.1371/journal.pcbi.1002375}.
-
-\bibitem{goeman2007}
-Goeman JJ, B{\"u}hlmann P.
-\newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
-  Issues.
-\newblock Bioinformatics. 2007;23(8):980--987.
-\newblock doi:{10.1093/bioinformatics/btm051}.
-
-\bibitem{subramanian2005}
-Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
-\newblock Gene Set Enrichment Analysis: A Knowledge-Based Approach for
-  Interpreting Genome-Wide Expression Profiles.
-\newblock Proceedings of the National Academy of Sciences.
-  2005;102(43):15545--15550.
-\newblock doi:{10.1073/pnas.0506580102}.
-
-\bibitem{ashburner2000}
-Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
-\newblock Gene {{Ontology}}: Tool for the Unification of Biology.
-\newblock Nature genetics. 2000;25(1):25--29.
-\newblock doi:{10.1038/75556}.
-
-\bibitem{irizarry2009}
-Irizarry RA, Wang C, Zhou Y, Speed TP.
-\newblock Gene {{Set Enrichment Analysis Made Simple}}.
-\newblock Statistical methods in medical research. 2009;18(6):565--575.
-\newblock doi:{10.1177/0962280209351908}.
-
-\bibitem{hanzelmann2013}
-H{\"a}nzelmann S, Castelo R, Guinney J.
-\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and
-  {{RNA}}-{{Seq}} Data.
-\newblock BMC Bioinformatics. 2013;14(1):7.
-\newblock doi:{10.1186/1471-2105-14-7}.
-
-\bibitem{frost2020}
-Frost HR.
-\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
-  Method for Cell-Specific Gene Set Scoring.
-\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
-\newblock doi:{10.1093/nar/gkaa582}.
-
-\bibitem{chong2020}
-Chong J, Liu P, Zhou G, Xia J.
-\newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
-  Functional, and Meta-Analysis of Microbiome Data.
-\newblock Nature Protocols. 2020;15(3):799--821.
-\newblock doi:{10.1038/s41596-019-0264-1}.
-
-\bibitem{quinn2019}
-Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
-\newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
-\newblock GigaScience. 2019;8(giz107).
-\newblock doi:{10.1093/gigascience/giz107}.
-
-\bibitem{quinn2018}
-Quinn TP, Erb I, Richardson MF, Crowley TM.
-\newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
-\newblock Bioinformatics. 2018;34(16):2870--2878.
-\newblock doi:{10.1093/bioinformatics/bty175}.
-
-\bibitem{morton2019}
-Morton JT, Marotz C, Washburne A, Silverman J, Zaramela LS, Edlund A, et~al.
-\newblock Establishing Microbial Composition Measurement Standards with
-  Reference Frames.
-\newblock Nature Communications. 2019;10(1).
-\newblock doi:{10.1038/s41467-019-10656-5}.
-
-\bibitem{love2014}
-Love MI, Huber W, Anders S.
-\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA}}-Seq
-  Data with {{DESeq2}}.
-\newblock Genome Biology. 2014;15(12):550.
-\newblock doi:{10.1186/s13059-014-0550-8}.
-
-\bibitem{weiss2017}
-Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
-\newblock Normalization and Microbial Differential Abundance Strategies Depend
-  upon Data Characteristics.
-\newblock Microbiome. 2017;5(1).
-\newblock doi:{10.1186/s40168-017-0237-y}.
-
-\bibitem{mcknight2019}
-McKnight DT, Huerlimann R, Bower DS, Schwarzkopf L, Alford RA, Zenger KR.
-\newblock Methods for Normalizing Microbiome Data: An Ecological Perspective.
-\newblock Methods in Ecology and Evolution. 2019;10(3):389--400.
-\newblock doi:{10.1111/2041-210X.13115}.
-
-\bibitem{aitchison1994}
-Aitchison J.
-\newblock Principles of Compositional Data Analysis.
-\newblock Lecture Notes-Monograph Series. 1994; p. 73--81.
-
-\bibitem{mclaren2019}
-McLaren MR, Willis AD, Callahan BJ.
-\newblock Consistent and Correctable Bias in Metagenomic Sequencing
-  Experiments.
-\newblock eLife. 2019;8:e46923.
-\newblock doi:{10.7554/eLife.46923}.
-
-\bibitem{egozcue2005}
-Egozcue JJ, {Pawlowsky-Glahn} V.
-\newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
-  Analysis}}.
-\newblock Mathematical Geology. 2005;37(7):795--828.
-\newblock doi:{10.1007/s11004-005-7381-9}.
-
-\bibitem{tian2005}
-Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
-\newblock Discovering Statistically Significant Pathways in Expression
-  Profiling Studies.
-\newblock Proceedings of the National Academy of Sciences.
-  2005;102(38):13544--13549.
-\newblock doi:{10.1073/pnas.0506577102}.
-
-\bibitem{rivera-pinto2018}
-{Rivera-Pinto} J, Egozcue JJ, {Pawlowsky-Glahn} V, Paredes R, {Noguera-Julian}
-  M, Calle ML.
-\newblock Balances: A {{New Perspective}} for {{Microbiome Analysis}}.
-\newblock mSystems. 2018;3(4):e00053--18.
-\newblock doi:{10.1128/mSystems.00053-18}.
-
-\bibitem{wu2012}
-Wu D, Smyth GK.
-\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
-  Correlation.
-\newblock Nucleic Acids Research. 2012;40(17):e133.
-\newblock doi:{10.1093/nar/gks461}.
-
-\bibitem{egozcue2003}
-Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcel{\'o}-Vidal} C.
-\newblock Isometric {{Logratio Transformations}} for {{Compositional Data
-  Analysis}}.
-\newblock Mathematical Geology. 2003;35(3):279--300.
-\newblock doi:{10.1023/A:1023818214614}.
-
-\bibitem{quast2013}
-Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
-\newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
-  Processing and Web-Based Tools.
-\newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
-\newblock doi:{10.1093/nar/gks1219}.
-
-\bibitem{delignette-muller2015}
-{Delignette-Muller} ML, Dutang C.
-\newblock Fitdistrplus: An {{R}} Package for Fitting Distributions.
-\newblock Journal of Statistical Software. 2015;64(4):1--34.
-
-\bibitem{benaglia2009}
-Benaglia T, Chauveau D, Hunter DR, Young D.
-\newblock Mixtools: An {{R}} Package for Analyzing Finite Mixture Models.
-\newblock Journal of Statistical Software. 2009;32(6):1--29.
-
-\bibitem{washburne2017}
-Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL.
-\newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
-  Associations in Microbiome Datasets.
-\newblock PeerJ. 2017; p.~26.
-
-\bibitem{silverman2017}
-Silverman JD, Washburne AD, Mukherjee S, David LA.
-\newblock A Phylogenetic Transform Enhances Analysis of Compositional
-  Microbiota Data.
-\newblock eLife. 2017;6:e21887.
-\newblock doi:{10.7554/eLife.21887}.
-
-\bibitem{morton2017}
-Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
-  et~al.
-\newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
-\newblock mSystems. 2017;2(1):e00162--16.
-\newblock doi:{10.1128/mSystems.00162-16}.
-
-\bibitem{aitchison1980}
-Aitchison J, Shen SM.
-\newblock Logistic-Normal Distributions:Some Properties and Uses.
-\newblock Biometrika. 1980;67(2):261--272.
-\newblock doi:{10.1093/biomet/67.2.261}.
-
-\bibitem{efron2004}
-Efron B.
-\newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
-\newblock Journal of the American Statistical Association.
-  2004;99(465):96--104.
-\newblock doi:{10.1198/016214504000000089}.
-
-\bibitem{calgaro2020}
-Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
-\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA}}-Seq,
-  and Metagenomics Applied to Microbiome Data.
-\newblock Genome Biology. 2020;21(1):191.
-\newblock doi:{10.1186/s13059-020-02104-1}.
-
-\bibitem{cario1997}
-Cario MC.
-\newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
-  Distributions}} and {{Correlation Matrix}}; 1997.
-
-\bibitem{agresti1998}
-Agresti A, Coull BA.
-\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
-  Estimation}} of {{Binomial Proportions}}.
-\newblock The American Statistician. 1998;52(2):119--126.
-\newblock doi:{10.2307/2685469}.
-
-\bibitem{delong1988}
-DeLong ER, DeLong DM, {Clarke-Pearson} DL.
-\newblock Comparing the Areas under Two or More Correlated Receiver Operating
-  Characteristic Curves: A Nonparametric Approach.
-\newblock Biometrics. 1988;44(3):837--845.
-
-\bibitem{hawinkel2019}
-Hawinkel S, Mattiello F, Bijnens L, Thas O.
-\newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
-  Control the False Discovery Rate.
-\newblock Briefings in Bioinformatics. 2019;20(1):210--221.
-\newblock doi:{10.1093/bib/bbx104}.
-
-\bibitem{breiman2001}
-Breiman L.
-\newblock Random {{Forests}}.
-\newblock Machine Learning. 2001;45(1):5--32.
-\newblock doi:{10.1023/A:1010933404324}.
-
-\bibitem{kuhn2020b}
-Kuhn M, Wickham H.
-\newblock {{Tidymodels}}: a collection of packages for modeling and machine learning using tidyverse        principles; 2021.
-
-\bibitem{xiao2018}
-Xiao J, Chen L, Yu Y, Zhang X, Chen J.
-\newblock A {{Phylogeny}}-{{Regularized Sparse Regression Model}} for
-  {{Predictive Modeling}} of {{Microbial Community Data}}.
-\newblock Frontiers in Microbiology. 2018;9.
-\newblock doi:{10.3389/fmicb.2018.03112}.
-
-\bibitem{dong2020}
-Dong M, Li L, Chen M, Kusalik A, Xu W.
-\newblock Predictive Analysis Methods for Human Microbiome Data with
-  Application to {{Parkinson}}'s Disease.
-\newblock PLOS ONE. 2020;15(8):e0237779.
-\newblock doi:{10.1371/journal.pone.0237779}.
-
-\bibitem{geistlinger2021}
-Geistlinger L, Csaba G, Santarelli M, Ramos M, Schiffer L, Turaga N, et~al.
-\newblock Toward a Gold Standard for Benchmarking Gene Set Enrichment Analysis.
-\newblock Briefings in bioinformatics. 2021;22(1):545--556.
-
-\bibitem{pasolli2017}
-Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
-\newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
-\newblock Nature Methods. 2017;14(11):1023--1024.
-\newblock doi:{10.1038/nmeth.4468}.
-
-\bibitem{schiffer2019}
-Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
-\newblock {{HMP16SData}}: Efficient Access to the Human Microbiome Project
-  through Bioconductor.
-\newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
-
-\bibitem{gonzalez2018}
-Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
-  Ackermann G, et~al.
-\newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
-\newblock Nature Methods. 2018;15(10):796--798.
-\newblock doi:{10.1038/s41592-018-0141-9}.
-
-\bibitem{consortium2012}
-Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
-  et~al.
-\newblock Structure, Function and Diversity of the Healthy Human Microbiome.
-\newblock Nature. 2012;486(7402):207--214.
-\newblock doi:{10.1038/nature11234}.
-
-\bibitem{thurnheer2016}
-Thurnheer T, Bostanci N, Belibasakis GN.
-\newblock Microbial Dynamics during Conversion from Supragingival to
-  Subgingival Biofilms in an in Vitro Model.
-\newblock Molecular Oral Microbiology. 2016;31(2):125--135.
-\newblock doi:{10.1111/omi.12108}.
-
-\bibitem{beghini2019}
-Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
-\newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
-  in the {{New York City Health}} and {{Nutrition Examination Study}}.
-\newblock Annals of Epidemiology. 2019;34:18--25.e3.
-\newblock doi:{10.1016/j.annepidem.2019.03.005}.
-
-\bibitem{matteocalgaro2020}
-Calgaro M. Mcalgaro93/Sc2meta: Paper {{Release}}; 2020.
-\newblock Zenodo.
-
-\bibitem{nielsen2014}
-Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
-\newblock Identification and Assembly of Genomes and Genetic Elements in
-  Complex Metagenomic Samples without Using Reference Genomes.
-\newblock Nature Biotechnology. 2014;32(8):822--828.
-\newblock doi:{10.1038/nbt.2939}.
-
-\bibitem{gevers2014}
-Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
-  et~al.
-\newblock The {{Treatment}}-{{Naive Microbiome}} in {{New}}-{{Onset Crohn}}'s
-  {{Disease}}.
-\newblock Cell Host \& Microbe. 2014;15(3):382--392.
-\newblock doi:{10.1016/j.chom.2014.02.005}.
-
-\bibitem{barbie2009}
-Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
-\newblock Systematic {{RNA}} Interference Reveals That Oncogenic
-  {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
-\newblock Nature. 2009;462(7269):108--112.
-\newblock doi:{10.1038/nature08460}.
-
-\bibitem{mcmurdie2014}
-McMurdie PJ, Holmes S.
-\newblock Waste {{Not}}, {{Want Not}}: Why {{Rarefying Microbiome Data Is
-  Inadmissible}}.
-\newblock PLOS Computational Biology. 2014;10(4):e1003531.
-\newblock doi:{10.1371/journal.pcbi.1003531}.
-
-\bibitem{martin2020}
-Martin BD, Witten D, Willis AD.
-\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
-  Regression.
-\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
-\newblock doi:{10.1214/19-AOAS1283}.
-
-\bibitem{naim2012}
-Naim I, Gildea D.
-\newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
-  {{Unbalanced Mixing Coefficients}}.
-\newblock Proceedings of the 29th International Coference on International
-  Conference on Machine Learning. 2012; p.~8.
-\newblock doi:{10.5555/3042573.3042756}.
-
-\bibitem{benjamini1995}
-Benjamini Y, Hochberg Y.
-\newblock Controlling the {{False Discovery Rate}}: A {{Practical}} and
-  {{Powerful Approach}} to {{Multiple Testing}}.
-\newblock Journal of the Royal Statistical Society Series B (Methodological).
-  1995;57(1):289--300.
-
-\bibitem{wu2016a}
-Wu C, Chen J, Kim J, Pan W.
-\newblock An Adaptive Association Test for Microbiome Data.
-\newblock Genome Medicine. 2016;8(1):56.
-\newblock doi:{10.1186/s13073-016-0302-3}.
-
-\bibitem{martin-fernandez2012}
-{Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
-  {Palarea-Albaladejo} J.
-\newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
-  Classical and Robust Approaches.
-\newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
-\newblock doi:{10.1016/j.csda.2012.02.012}.
-
-\bibitem{kaul2017a}
-Kaul A, Davidov O, Peddada SD.
-\newblock Structural Zeros in High-Dimensional Data with Applications to
-  Microbiome Studies.
-\newblock Biostatistics. 2017;18(3):422--433.
-\newblock doi:{10.1093/biostatistics/kxw053}.
-
-\end{thebibliography}
+\bibliography{mac_taxagg}{}
+%\begin{thebibliography}{10}
+%\end{thebibliography}
 
@@
 
+\paragraph*{S3 Fig.}
+\label{S3_Fig}
+{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
+Evaluation was performed on simulated data sets.
+For sample size analysis, only 10 sets were evaluated.
+For taxa set analysis, sample size was fixed at 1,000.
+Across all evaluations, the size of each taxa set was also fixed at 50.
+
+\paragraph*{S4 Fig.}
+\label{S4_Fig}
+{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
+Evaluation was performed on simulated data sets.
+For sample size analysis, only 10 sets were evaluated.
+For taxa set analysis, sample size was fixed at 1,000.
+Across all evaluations, the size of each taxa set was also fixed at 50.
+
+\paragraph*{S5 Fig.}
+\label{S5_Fig}
+{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
+Evaluation was performed on simulated data sets.
+For sample size analysis, only 10 sets were evaluated.
+For taxa set analysis, sample size was fixed at 1,000.
+Across all evaluations, the size of each taxa set was also fixed at 50.
+
+\paragraph*{S6 Fig.}
+\label{S6_Fig}
+{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
+Evaluation was performed on simulated data sets.
+For sample size analysis, only 10 sets were evaluated.
+For taxa set analysis, sample size was fixed at 1,000.
+Across all evaluations, the size of each taxa set was also fixed at 50.
+
+\paragraph*{S7 Fig.}
+\label{S7_Fig}
+{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
+Evaluation was performed on simulated data sets.
+For sample size analysis, only 10 sets were evaluated.
+For taxa set analysis, sample size was fixed at 1,000.
+Across all evaluations, the size of each taxa set was also fixed at 50.
+
 \paragraph*{S1 File.}

### 2022-02-01 saving_texts.tex

@@
 \end{figure}
+
+\subsection{Prediction analysis}
+\subsubsection*{Simulation studies}
+Fig~\ref{fig:6} shows results for simulation studies as detailed in the \nameref{methods} section.
+Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome.
+As expected, performance across all assessed methods increased with a higher signal-to-noise ratio.
+Both CLR and CBEA approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks.
+This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while CBEA and CLR did not experience a similar level of impact.
+As such, performance gap widens with increasing correlation and sparsity.
+Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.
+
+In this analysis, CBEA unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
+This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation.
+However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and CBEA/CLR.
+Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity.
+Additionally, we did not observe any performance difference between the different variations of CBEA.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
+    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
+Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure.
+Panel \textbf{(B)} presents performance on a classification task with AUC as the evaluation measure.
+CBEA approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
+    \label{fig:6}
+\end{figure}

### 2022-02-01 supplemental.tex
Updates from Overleaf

@@
 \newcommand{\overbar}[1]{\mkern 2mu\overline{\mkern-2mu#1\mkern-20mu}\mkern 20mu}
+\renewcommand{\thefigure}{S\arabic{figure}}
 
@@
 Under this assumption, we can modify the optimization problem to only estimate the variance parameter of the smaller component (i.e.
-withouth loss of generalizability keeping $\sigma_1$ constant where $\lambda'_1 > \lambda'_2$).
+without loss of generalizability keeping $\sigma_1$ constant where $\lambda'_1 > \lambda'_2$).
 This allows for the optimization procedure to more properly capture the right tail distribution rather than increasing weight on the left tail of the distribution which impacts the computation of p-values for a one-sided test.
 Empirical experiments (data not shown) done on simulated data and random set analyses suggest that this adjustment improves performance of the adjusted CBEA under mixture-normal assumption.
-In the R implementation of CBEA, users can control this behaviour by specifying the
+In the R implementation of CBEA, users can control this behaviour by specifying the \emph{fix\_comp} parameter as part of the \emph{control} argument.
+
+\section{Numerical Simulations}
+\subsection{Design}
+To address the performance of CBEA for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
+Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
+\begin{equation}
+    \mathbf{X}_{ij} =
+      \begin{cases}
+        0 & \text{with probability $p_j$}\\
+        \mathbf{NB}(\mu_j, \phi_j) & \text{with probability $1 - p_j$}\\
+      \end{cases}       
+\end{equation}
+
+where $\mu_j$ and $\phi_j$ are mean and dispersion parameters, respectively.
+To incorporate a flexible correlation structure into our simulation model, we utilized the NorTA (Normal to Anything) method \cite{cario1997}.
+Given an $n$ by $p$ matrix of values $\mathbf{U}$ sampled from multivariate normal distribution with correlation matrix $\mathbf{\rho}$, we can generate target microbiome count vector $\mathbf{X_{.j}}$ for taxa $j$ following the marginal distribution $\mathbf{NB}$ characterized by the negative binomial cumulative distribution function $\mathbb{F_{\mathbf{NB}}}$:
+\begin{equation}
+    \mathbf{X}_{.j} = \mathbb{F_{\mathbf{NB}}}^{-1}(\Phi_{U_i})
+\end{equation}
+In this instance, for each taxon $j$, we set elements in $\mathbf{U}_{.j}$ to be zero with probability $p_j$ and applied $\mathbf{NB}^{-1}(\mu_j, \phi_j)$ on non-zero elements to generate our final count matrix $\mathbf{X}$.
+To ensure that our simulations match closely to real data, we fitted negative binomial distribution using a maximum likelihood approach (with the \emph{fitdistrplus} package in R \cite{delignette-muller2015}) to non-zero counts for each taxon from 16S rRNA profiling of stool samples from the Human Microbiome Project (HMP).
+We take the median values of the estimated mean and dispersion parameters as the baseline of our simulations.
+For simplicity, we assumed that inter-taxa correlation follows an exchangeable structure
+
+\subsubsection{Simulation scenarios for enrichment analysis at the sample level}
+To assess type I error rate and power for enrichment significance testing at the sample level, we simulated data based on the schema above, and assessed enrichment for one focal set.
+Type I error was obtained under the global null as the number of samples where the null hypothesis was rejected at $\alpha = 0.05$ over the total number of samples (which represents the total number of hypotheses tested).
+Power was obtained using the same formulation as type I error rate but under the global alternate.
+We treated type I error and power as estimates of binomial proportions and utilized the Agresti-Couli \cite{agresti1998} formulation to calculate 95\% confidence intervals.
+Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$) and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$).
+For type I error analysis, we also varied the size of the set (50, 100, 150).
+For power analyses, set size was kept constant at 100 but different effect sizes (fold change of 1.5, 2, and 3).
+All sample sizes were set at 10,000.
+
+For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUC).
+This is a strategy used in Frost \cite{frost2020} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
+whether samples with inflated counts are highly ranked using estimated scores).
+DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate.
+Simulation settings for classification performance were identical to power analyses as detailed in the previous paragraph.
+
+\subsubsection{Simulation scenarios for enrichment analysis at the population level}
+To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition.
+Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null.
+Power is defined similarly, but instead under the global alternate hypothesis.
+Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error.
+A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size.
+Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$).
+Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3).
+Due to the compositional nature of microbiome taxonomic data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets.
+As such, we applied a compensation procedure as described in Hawinkel et al.
+\cite{hawinkel2019} to ensure the validity of simulation results.
+All sample sizes were set at 2,000.
+
+\subsubsection{Simulation scenarios for downstream prediction}
+To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}.
+For binary outcomes, we use AUC similar to the classification analyses above.
+For continuous outcomes, we used root mean squared error (RMSE).
+All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020b} suite of packages.
+Across both learning tasks, we varied sparsity ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation ($\rho = 0, 0.2, 0.5$).
+Continuous outcomes $Y_{cont}$ were generated as linear combinations of taxa counts.
+\begin{equation}
+    Y_{cont} = f(\mathbf{X}) + \mathbf{\epsilon}
+\end{equation}
+where $\mathbf{\epsilon} \sim N(0, \sigma_{\epsilon}^2)$ and $f(\mathbf{X}) = \beta_0 + \mathbf{X}\mathbf{\beta}$.
+For each simulation, we set $\beta_0$ to be $\frac{6}{\sqrt{10}}$ similar to \cite{xiao2018}.
+The degree of model saturation (the number of non zero $\mathbf{\beta}$ values) were varied between 0.1 and 0.5, and signal to noise ratio (SNR = $\frac{\sigma(f(\mathbf{X}))}{\sigma_{\epsilon}}$) was varied between 1.5, 2, and 3.
+
+For binary outcomes, we generate $Y_{binary}$ as Bernoulli draws with probability $p_{binary}$, where 
+\begin{equation}
+    p_{binary} = \frac{1}{1 + \exp(f(\mathbf{X}) + \mathbf{\epsilon})}
+\end{equation}
+To ensure a balance of classes, we applied the strategy described in Dong et al.
+\cite{dong2020} where the associated $\beta$ values are evenly split between positive and negative associations.
+All data sets generated from prediction tasks have 2,000 samples with 5,000 taxa over 50 sets with a size of 100 taxa per set.
+
+\subsection{Results} 
+\subsubsection{Statistical Inference}
+Panel A and B in Fig~\ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
+We benchmarked the results of the CBEA method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
+All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions.
+However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted CBEA variants show high levels of inflated type I error, where Wilcoxon test performed the worst.
+On the other hand, adjusted CBEA methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{figures/sim_ss_fdr.png}
+    \caption{Simulation results for type I error evaluation for CBEA sample-level inference.
+Type I error rate (\emph{y} axis) was estimated for each approach across data sparsity levels (\emph{x} axis) across different set sizes (horizontal) and inter-taxa correlation within the set (vertical).
+We compared variatns of CBEA against a Wilcoxon rank sum test at $\alpha$ of 0.05.
+For each scenario, a data set of 10,000 samples (equivalent to 10,000 hypotheses) was utilized.
+Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach}
+    \label{fig:s1}
+\end{figure}
+
+CBEA generates sample-specific scores representing the degree of enrichment of a pre-defined set.
+As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test).
+We compared the performance of this approach with CBEA and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
+    \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
+For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions).
+CBEA associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
+    \label{fig:s2}
+\end{figure}
+
+Fig.~\ref{fig:4} present results for simulation studies for both type I error (panel A) and power (panel B) evaluations.
+All methods control for type I error well across both sparsity and correlation levels, where the estimated rate was consistently around the 0.05 pre-defined threshold.
+Results were similar across all evaluated methods, although in some instances, for example in medium correlation setting ($\rho = 0.2$), the unadjusted CBEA resulted in higher type I error, regardless of difference in means test and distribution of choice.
+
+\subsubsection{Phenotype relevance}.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width = \textwidth]{figures/sim_ss_pwr.png}
+    \caption{Simulation results for phenotype relevance valuation for CBEA sample-level inference.
+(\textbf{(A)}) demonstrate statistical power (\emph{y} axis) across different data sparsity levels (\emph{x} axis) and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
+For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions).
+CBEA associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
+    \label{fig:s3}
+\end{figure}
+
+However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B.
+In situations where there is no inter-taxa correlation, CBEA still outperforms the wilcoxon rank sum test, however adjusted versions of CBEA did not perform as well as un-adjusted ones.
+However, in higher correlation scenarios, the difference in power is much more dramatic.
+At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted CBEA was only performing at 50\% power, while unadjusted CBEA and wilcoxon rank sum test were able to reach 80\%.
+These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of CBEA.
+Most importantly, CBEA demonstrate higher power in all scenarios where type I error is properly controlled.
+
+To further assess the utility of CBEA in classifying samples with enriched sets, we generated AUC scores for different CBEA scores using true labels of whether a sample has an inflated set.
+This analysis, therefore, assessed the relative ranking of samples using CBEA scores whereby high scores should correspond to samples that are known to be inflated.
+Fig~\ref{fig:3} presents this result.
+We compared different variants of CBEA against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
+Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), CBEA scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
+In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
+Notably, the W-statistic provided the least information for classifying samples with inflated taxa.
+
+The difference between the methods is more noticeable when evaluating power.
+All CBEA associated variants showed much higher power even when the effect size is limited (fold change is 1.5), and there is a noticeable gap in performance between CBEA and both DESeq2 and corncob.
+Surprisingly, this effect is consistent across correlation levels and sparsity, even though we expectedly see performance in power drop as a function of sparsity especially in low effect size settings.
+
+\subsubsection{Prediction analysis}
+Fig~\ref{fig:6} shows results for simulation studies as detailed in the \nameref{methods} section.
+Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome.
+As expected, performance across all assessed methods increased with a higher signal-to-noise ratio.
+Both CLR and CBEA approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks.
+This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while CBEA and CLR did not experience a similar level of impact.
+As such, performance gap widens with increasing correlation and sparsity.
+Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.
+
+In this analysis, CBEA unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
+This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation.
+However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and CBEA/CLR.
+Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity.
+Additionally, we did not observe any performance difference between the different variations of CBEA.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
+    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
+Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure.
+Panel \textbf{(B)} presents performance on a classification task with AUC as the evaluation measure.
+CBEA approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
+    \label{fig:6}
+\end{figure}
 

### 2022-02-01 cbea_revised.tex

@@
 Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability.
-Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given that compared to other 'omics data types microbiome taxonomic data is compositional.
+Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given its unique features compared to other 'omics data types.
 We developed a new tool to generate taxon set enrichment scores at the sample level through a novel log-ratio formulation based on the competitive null hypothesis.
-Our scores can be used for statistical inference, and as inputs to other downstream analyses such as prediction models.
+Our scores can be used for statistical inference at both the sample and population levels, as well as inputs to other downstream analyses such as prediction models.
 We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.
@@
 As such, practitioners often rely on total sum normalization to transform count data into relative proportions that sum to one \cite{weiss2017}.
-Empirical studies have supported this decision, showing increased performance of analysis tasks under this normalization schema \cite{mcknight2019}.
+Some studies have provided emprical performance vealuations supporting this normalization schema \cite{mcknight2019}.
 Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another.
@@
 
-Standard practice for aggregating variables in microbiome research uses element-wise summations \cite{mclaren2019}.
-Unfortunately, there are disadvantages to using this approach (referred to as amalgamations in the CoDA literature).
+Unfortunately, the standard practice for aggregating variables using element-wise summations (referred to as amalgamations in the CoDA literature), does not adequately address the compositional issue \cite{mclaren2019}.
 First, inter-sample Aitchison distances computed on original parts are not preserved after amalgamation \cite{egozcue2005}.
@@
 \begin{itemize}
-    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing.
+    \item $\mathbf{X}$: $n$ by $p$ matrix of positive propotions for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing.
 Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing.
 CBEA does not accept $X$ matrices with zeroes since it invalidates the log-ratio transformation.
-Users can generate a dense matrix $X$ using a method of choice, however the default mode for CBEA would be add a pseudocount of 1 if zeroes are detected in the matrix.
+Users can generate a dense matrix $X$ using a method of choice, however the default mode for CBEA would be add a pseudocount of $10^{-5}$ if zeroes are detected in the matrix.
     \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxon $p$ to $m$ sets of interest.
@@
 Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution.
-P-values can be calculated by subtracting $\mathbf{E}$ from 1.
+Raw p-values can be calculated by subtracting $\mathbf{E}$ from 1.
 \end{enumerate}
@@
 However, when applying CBEA for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses.
-Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies.
+Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic directly (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies.
 As such, to perform significance testing using CBEA, we also estimated the null distribution from observed raw CBEA variables.
@@
 The CBEA statistic then looks similar to a t-statistic for difference in means of log-transformed proportions.
-As such, the pooled variance of cILR is dependent on the variance inflation of both mean components $\overbar{\log{X_{i,j|j \in K}}}$ and $\overbar{\log{X_{i,j|j \notin K}}}$.
+As such, the pooled variance of CBEA is dependent on the variance inflation of both mean components $\overbar{\log{X_{i,j|j \in K}}}$ and $\overbar{\log{X_{i,j|j \notin K}}}$.
 The result of this variance inflation is inflated type I error since highly correlated sets are also detected as significantly enriched.
@@
 All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).
+Additional packages used to support this analysis includes: \emph{tidyverse} \cite{wickham2019}, \emph{pROC} \cite{robin2011}, \emph{phyloseq} \cite{mcmurdie2014}, \emph{mia} \cite{ernst2021}, \emph{targets} \cite{landau2021}.
 
@@
 We evaluated this application by generating one random taxa of different sizes $S \in \{20, 50, 100, 150, 200\}$ across 500 iterations.
-Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample sets of randomly sampled taxa should not be significantly enriched compared to the remainder background taxa.
+Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample sets of randomly drawn taxa should not be significantly enriched compared to the remainder background taxa.
 For this evaluation, we estimated type I error as the fraction of samples that detects our random set as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.
@@
 Type I error is estimated as the fraction of sets per iteration found to be significantly enriched with confidence bands computed from the standard error across all iterations.
-In addition, we also performed a random set analysis assessing, where we generated 100 sets of different set sizes $S \in \{20, 50, 100, 150, 200\}$ and evaluated the fraction of genera that were found to be differentially abundant across the original labels (supragingival versus subgingival subsite).
+In addition, we also performed a random set analysis assessment, where we generated 100 sets of different set sizes $S \in \{20, 50, 100, 150, 200\}$ and evaluated the fraction of genera that were found to be differentially abundant across the original labels (supragingival versus subgingival subsite).
 95\% confidence intervals were computed using the Agresti-Couli approach \cite{agresti1998}.
@@
 For sample-level inference, we assessed power as the fraction of supragingival samples where aerobic microbes are significantly enriched.
-For population-level inference, power is the fraction of sets representing genus level taxonomic assignments were significant across subsite labels.
+For population-level inference, power is the fraction of sets representing genus level taxonomic assignments that were significant across subsite labels.
 
@@
 This is represented by the area under the receiving operator curve (AUROC/AUC) scores computed on CBEA scores against true labels (similar approach was used to evaluate VAM \cite{frost2020}).
-DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate.
+DeLong 95\% confidence intervals for AUROC \cite{delong1988} were obtained for each estimate.
 
@@
 This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease).
-We processed this data by removing all samples with less than 1,000 total read counts as well as any OTU who was present (at least 1 count) in 10\% of samples or less.
+We processed this data by removing all samples with less than 1,000 total read counts as well as any OTU who was present (with non-zero proportions) in 10\% of samples or less.
 Prior to model fitting, we back-transformed relative abundances into count data (to align the format with our 16S rRNA gene sequencing data set) using provided total number of reads aligned to MetaPhlan marker genes (per sample).
@@
 For assessments at the population level, we compared CBEA against performing a standard test for differential abundance with set-level features generated via element-wise summations instead.
-We chose DESeq2 \cite{love2014} and corncob \cite{martin2020} since they represent both methods extrapolated from RNA-seq \cite{mcmurdie2014} and those developed specifically for microbiome data.
+We chose DESeq2 \cite{love2014} and corncob \cite{martin2020} because they represent both methods extrapolated from RNA-seq \cite{mcmurdie2014} and those developed specifically for microbiome data.
 
@@
 Type I error (\emph{y}-axis) was evaluated by generating random sets of different sizes (\emph{x}-axis) (500 replications per size) and computing the fraction of samples where the set was found to be significantly enriched at $\alpha = 0.05$.
-Error bars represent the mean type I error $\pm$ sample standard error computed across 500 replications of the experiment.}
+Error bars represent the mean type I error $\pm$ sample standard error computed across 500 replications of the experiment.
+Only the unadjusted CBEA with the mixture normal distribution and the Wilcoxon rank sum test were able to control for type I error at 0.05.
+All approaches are invariant to set sizes.}
     \label{fig:2}
@@
 Encouragingly, all methods demonstrate consistent performance across all set sizes, with a slight increase in type I error at the highest levels.
+
+Interestingly, simulation results (\nameref{S1_Fig}) showed an opposite pattern.
+Adjusted approaches were good at controlling for type I error, especially under the low inter-taxa correlation values within the set (similar to generating random sets where the natural correlation structure is disrupted).
+In these simulations, unadjusted approaches and the Wilcoxon rank sum test had significant type I error inflation with increasing correlation.
+All approaches seems to be invariant to the level of data sparsity.
 
@@
 (\textbf{B}) Proportion of significant sets (\emph{y}-axis) using 100 randomly generated sets of different set sizes (\emph{x}-axis).
-Confidence intervals computed using Agresti-Couli method for binomial proportions.}
+Confidence intervals computed using Agresti-Couli method for binomial proportions.
+For sample label permutation (\textbf{A}), all CBEA approaches were able to control for type I error but not for corncob and DESeq2.
+For random set analyses (\textbf{B}), all approaches demonstrate similar rate of accepting significant sets and were invariant to overall set size.}
     \label{fig:3}
@@
 
-We also assessed the impact of set-size on the inference procedure by testing for enrichment using the original sample labels but with randomly sampled sets of different sizes.
+We also assessed the impact of set-size on the inference procedure by testing for enrichment using the original sample labels but with randomly sampled sets of different sizes (Fig~\ref{fig:3}B).
 Overall we observed very similar values across CBEA as well as corncob and DESeq2, suggesting that no individual method is systematically identifying too many significant sets.
-Additionally, similar to analogous analyses at the sample level, no approach were significantly sensitive to set-size.
+Additionally, similar to analogous analyses at the sample level, no approach were significantly sensitive to changes in set sizes.
 
@@
 We observed that adjusted CBEA approaches demonstrate much lower power compared to the Wilcoxon rank-sum test and unadjusted variants.
-This is surprising given the fact that in statistical significance analyses, the adjusted CBEA approach provides inflated type I error, especially if the normal distribution assumption was chosen.
-Here, we did not see any differences in performance between distribution choices.
+This is surprising given the fact that in statistical significance analyses, the adjusted CBEA approach provides inflated type I error, especially if the normal distribution assumption was chosen, which indicates a mismatch in estimating the null distribution since a high type I error did not result in increased power.
 
@@
 
-Unfortunately, evaluating power and score rankings using real data does not allow for observing performance trends when effect sizes, sparsity, and inter-taxa correlation are varied.
-In equivalent simulation analyses, we observed the opposite effect, where
+The above results were replicated in simulation studies where we observed that adjusted approaches was very conservative and demonstrated significantly lower power (\nameref{S3_Fig}) with increasing correlation even at the highest evaluated effect sizes.
+When assessing score rankings, the performance of CBEA was closer to ssGSEA and GSVA compared to real data evaluations, however all single sample approaches were much better than using the W statistic from the Wilcoxon Rank Sum test.
 
@@
 \subsection*{Disease Prediction}   
-Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models.
-In this section we assessed the predictive performance of a standard baseline random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating CBEA, ssGSEA, and GSVA).
+Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:4}), we want to assess whether they can also act as useful inputs to predictive models.
+In this section we assessed the predictive performance of a standard baseline random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (CBEA, ssGSEA, and GSVA).
 Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.
@@
     \centering
-    \includegraphics[width = \linewidth]{figures/data_prediction_plot.png}
+    \includegraphics[width = \linewidth]{figures/data_pred.png}
     \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
@@
 
-Fig~\ref{fig:6} illustrates results from using CBEA, ssGSEA, GSVA, or CLR transformed variables as inputs to our model, with AUROC as the performance criteria.
-Results also replicated that of the simulations (Supplementary materials), where across both data sets CBEA and CLR methods provide much better performance than both GSVA or ssGSEA.
-Interestingly, the CBEA approach performed better than CLR in the whole genome data set but did not perform as well in the 16S rRNA gene sequencing data set.
-However, these results indicate that CBEA generated scores are informative, providing competitive performance when acting as inputs to disease predictive models.
-Most importantly, performance values are consistent across both simulated and real data sets.
-
-These results demonstrate for given set-based features, CBEA can generate scores can could be informative in disease prediction tasks.
-Simulation results indicate that CBEA methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach.
-Interestingly, however, CBEA methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
+Fig~\ref{fig:6} illustrates the performance of our model with AUROC as the evaluation criteria.
+In the 16S rRNA data set, the best performing CBEA variant (CDF values computed from an unadjusted mixture normal distribution) outperforms both GSVA and ssGSEA but not the standard CLR approach.
+Interestingly, in the whole genome sequencing data set, CBEA outperforms CLR, but were more similar in performance to GSVA.
+However, due to large confidence intervals, no method were significantly out-performing other approaches evaluated.
+As such, these results indicated that CBEA generated scores are can be informative, providing competitive performance when acting as inputs to disease predictive models.
+Simulation studies (\nameref{S5_Fig}) showed similar results, however CBEA was more consistently underperforming compared to CLR across all scenarios.
+Interestingly, the performance gap decreases with increasing sparsity levels.
 
@@
 This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), where the mixture normal displayed superior fit to the null distribution.
-Unfortunately, other variants of CBEA did not demonstrate good type I error control.
-Interestingly, while the adjusted methods showed inflated type I error and low power in real data evaluations (Fig~\ref{fig:2}), they performed well in simulation studies (\nameref{S1_Fig}, \nameref{S3_Fig}).
+Unfortunately, other variants of CBEA demonstrated neither good type I error control nor power.
+Interestingly, while the adjusted methods showed poor performance in real data evaluations (Fig~\ref{fig:2}), in simulation studies (\nameref{S1_Fig}, \nameref{S3_Fig}) these approaches were able to control for type I error well with the trade-off of much lower power.
 
 For population-level inference task, CBEA also performed very well.
-Under the permutation global null, representing genera abundance using CBEA scores in combination with Welch's t-test controls for type I error the correct $\alpha$ threshold while also showing respectable power.
+Under the permutation global null, representing genera abundance using CBEA scores in combination with Welch's t-test controls for type I error the correct $\alpha$ threshold while also keeping respectable power.
 Since population level enrichment test is equivalent to differential abundance using set-based features, we compared CBEA approach against using element-wise summations with corncob \cite{martin2020} and DESeq2 \cite{love2014} to test for set-level differential abundance.
@@
 We hypothesized that the discrepancy between simulation and real data evaluations can can arise due to differences in our assumptions regarding the data generating process that informed our simulation schema.
-One important consideration, as described in the \nameref{introduction} section, are taxon-specific biases that distort the observed relative abundance of taxa compred to true values \cite{mclaren2019}.
+For the non-zero component of each taxa, we sampled from the same negative binomial distribution where designated enriched taxa were generated with inflated means (but the same dispersion).
+These marginals were simulated to account for block exchangable correlation within the enriched set only.
+This might affect our results in two ways: First, our simulation scenario ensures that any designated non-enriched taxa is identical to each other.
+This is not the case for real data, where our null scenarios involves randomly sampled sets that might by chance all have taxa with inflated means compared to remainder taxa.
+This is represented in \nameref{S7_Fig}, where the distribution of type I error across our 500 replications are right skewed for underperforming CBEA variants, indicating that these approaches are much more sensitive compared to the Wilcoxon rank sum test or unadjusted CBEA with mixture normal distribution.
+Second, as described in the \nameref{introduction} section, we did not consider taxon-specific biases that distort the observed relative abundance of taxa compred to true values \cite{mclaren2019}.
 In the context of sum-based aggregations, the resulting bias of the aggregated taxon are dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}).
@@
 As such, we expect methods like corncob or DESeq2 when performed on such sum aggregates in the presence of taxon-specific biases to have inflated type I error compared to our taxon-ratio based approach.
-Conversely, in simulation studies, we did not explicitly include bias in our formulation and all taxa were generated from the same negative binomial distribution with similar variances across all samples (the only difference is when an enriched taxa is expected to have inflated means).
-As such, DESeq2 and corncob controls well for type I error in simulation scenarios where the effect of taxon-specific biases is absent.
-It is still surprising to see lower power for both methods in simulation analyses, which might be due to the fact that the evaluation protocol only considers default settings for both methods and did not attempt to optimize performance.
-We also suspect that
+Conversely, in simulation studies, where taxon-specific biases are absent, corncob and DESeq2 should perform better.
 
@@
 The sample-level enrichment scores generated by the CBEA method can be used in downstream analyses such as disease prediction.
-Here, we evaluated whether CBEA can be used to generate set-based features for disease prediction models.
+We evaluated whether CBEA can be used to generate set-based features for disease prediction models.
 
@@
 Comparatively, CBEA generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome taxonomic data sets.
-This is consistent with our sample ranking analysis (Fig~\ref{fig:3}), where CBEA scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
+This is consistent with our sample ranking analysis (Fig~\ref{fig:4}), where CBEA scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
 However, CBEA did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data.
-
-However, in simulation studies, this performance gap between CLR and CBEA decreases with higher sparsity and correlation, especially in low effect saturation scenarios.
-Additionally, there are also downsides to applying CLR.
-First, the singular covariance matrix of CLR transformed variables is singular due to a sum to zero constraint \cite{gloor2017}, preventing the proper usage of approaches that rely on matrix decomposition.
-Second, the procedure still relies on using summation of counts prior to transformation, which means that we still can't compare across sets of different sizes, and any bias might still be propagated \cite{mclaren2019}.
-As such, despite benefits in performance for a naive random forest model, there is still space for using CBEA as primary inputs into predictive models.
+Fortunately, in simulation studies, this performance gap between CLR and CBEA decreases with higher sparsity and correlation, especially in low effect saturation scenarios.
 
@@
 identifying characteristics that are differentially abundant across case/control status).
-
 For the unsupervised approach, utilizing the unadjusted CBEA with the mixture normal distribution provides a good initial starting point.
-In the case where the screening procedure only wants to detect mean-inflated taxon sets (instead of additionally detecting taxon sets with increased correlation), users can apply the adjusted approach with the normal distribution, which can be effective at conserving type I error even at high correlation scenarios.
+In the case where researchers only wants to screen samples with mean-inflated taxon sets (instead of additionally detecting taxon sets with increased correlation), they can apply the adjusted approach, which can be effective at conserving type I error even at high correlation scenarios.
 However, the trade off for this adjustment is power that decreases with increasing correlation.
@@
 We demonstrated that for a given number of set-based features, CBEA can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity.
-This is especially true for whole genome sequencing data sets, where CBEA outperfrorms the standard approach of applying a centered log-ratio transformation.
+This is especially true for whole genome sequencing data sets, where CBEA outperfrorms the standard approach of applying a CLR transformation.
+Researchers might find CBEA useful under situations of high sparsity and inter-taxa correlation, or if the property of a singular covariance matrix (a byproduct of the CLR transformation \cite{gloor2017}) is undesired.
 Even though we only evaluated prediction models in this manuscript, researchers can benchmark their own usage of CBEA for other downstream tasks such as sample ordination.
@@
 Our evaluation also showed various drawbacks of the CBEA method itself.
-First, inference with CBEA at the sample level is limited, and can be affected by inter-taxa correlation if users wish to only detect mean-inflated sets, especially when inter-taxa correlation within the set is high.
+First, inference with CBEA at the sample level is limited, and can be affected by inter-taxa correlation if users wish to only detect mean-inflated sets.
 Second, for downstream analyses, CBEA might not always perform better than competing methods, especially when being used to generate inputs to predictive models.
@@
 
-In addition, CBEA itself did not consider other aspects of microbiome data that can merit further extension.
+In addition, CBEA itself did not consider other aspects of microbiome data.
 First, across all analyses, we relied on adding a pseudocount to ensure log operations are valid.
@@
 Second, CBEA also treated all taxa within the set as equally contributing to the set.
-Incorporation of taxa-specific weights (similar to PhILR \cite{silverman2017} could reduce the influence of outliers, such as rare or highly invariant taxa.
+Incorporation of taxa-specific weights (similar to PhILR \cite{silverman2017}) could reduce the influence of outliers, such as rare or highly invariant taxa.
 Finally, even though for a given set of apriori annotations CBEA can generate useful summary scores, such values are limited in their utility if the annotation themselves are not meaningful.
@@
 \label{S1_Fig}
-{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
-Evaluation was performed on simulated data sets.
-For sample size analysis, only 10 sets were evaluated.
-For taxa set analysis, sample size was fixed at 1,000.
-Across all evaluations, the size of each taxa set was also fixed at 50.
+{\bf Simulation results for type I error evaluation for CBEA sample-level inference.} Placeholder text.
 
@@
 \label{S2_Fig}
-{\bf Distribution of p-values.} Q-Q plot of 10,000 p-values compared against a uniform distribution bounded between 0 and 1.
-Evaluation was performed on simulated null data sets of 10,000 samples testing for enrichment of a set of size 50.
-For correlation of 0.5, p-values represent correlation adjusted CBEA while for correlation of 0, p-values represent unadjusted CBEA.
+{\bf Simulation results for type I error evaluation for CBEA population-level inference.} Placeholder.
 
@@
 \label{S3_Fig}
-{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
-Evaluation was performed on simulated data sets.
-For sample size analysis, only 10 sets were evaluated.
-For taxa set analysis, sample size was fixed at 1,000.
-Across all evaluations, the size of each taxa set was also fixed at 50.
+{\bf Simulation results for phenotype relevance evaluation for CBEA sample-level inference.} Placeholder text.
 
@@
 \label{S4_Fig}
-{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
-Evaluation was performed on simulated data sets.
-For sample size analysis, only 10 sets were evaluated.
-For taxa set analysis, sample size was fixed at 1,000.
-Across all evaluations, the size of each taxa set was also fixed at 50.
+{\bf Simulation results for phenotype relevance evaluation for CBEA population-level inference.} Placeholder text.
 
@@
 \label{S5_Fig}
-{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
-Evaluation was performed on simulated data sets.
-For sample size analysis, only 10 sets were evaluated.
-For taxa set analysis, sample size was fixed at 1,000.
-Across all evaluations, the size of each taxa set was also fixed at 50.
+{\bf Simulation results for predictive pefromance evaluation for CBEA.} Placeholder text.
 
@@
 \label{S6_Fig}
-{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
-Evaluation was performed on simulated data sets.
-For sample size analysis, only 10 sets were evaluated.
-For taxa set analysis, sample size was fixed at 1,000.
-Across all evaluations, the size of each taxa set was also fixed at 50.
+{\bf Computational performance of CBEA.} Placeholder text.
 
@@
 \label{S7_Fig}
-{\bf Computational performance of CBEA.} Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
-Evaluation was performed on simulated data sets.
-For sample size analysis, only 10 sets were evaluated.
-For taxa set analysis, sample size was fixed at 1,000.
-Across all evaluations, the size of each taxa set was also fixed at 50.
+{\bf Distribution of type I error values across all replications in real data random set evaluations for CBEA inference at the sample-level.} Placeholder text.
 
@@
 \label{S1_File}
-{\bf Supplemental derivations.} Includes additional details on addressing variance inflation due to correlation in CBEA, as well as computational performance and p-value distribution of the method.
-
+{\bf Supplemental derivations.} Includes additional details on addressing variance inflation due to correlation in CBEA, simulation analyses, and run-time performance.
 \end{document}

### 2022-02-01 cbea_revised.tex
Updates from Overleaf
New version and finalized

@@
 Additionally, we utilized SMOTE to correct for class imbalances \cite{chawla2002}.
-Implementation was done using the \emph{tidymodels} suite of packages \cite{kuhn2020b}.
+Implementation was done using the \emph{tidymodels} suite of packages \cite{kuhn2020}.
 
@@
 
-\bibliography{mac_taxagg}{}
-%\begin{thebibliography}{10}
-%\end{thebibliography}
+%\bibliography{references}{}
+\begin{thebibliography}{10}
+
+\bibitem{proctor2019}
+Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
+\newblock The {{Integrative Human Microbiome Project}}.
+\newblock Nature. 2019;569(7758):641--648.
+\newblock doi:{10.1038/s41586-019-1238-8}.
+
+\bibitem{sharma2019a}
+Sharma S, Tripathi P.
+\newblock Gut Microbiome and Type 2 Diabetes: Where We Are and Where to Go?
+\newblock The Journal of Nutritional Biochemistry. 2019;63:101--108.
+\newblock doi:{10.1016/j.jnutbio.2018.10.003}.
+
+\bibitem{aoun2020}
+Aoun A, Darwish F, Hamod N.
+\newblock The {{Influence}} of the {{Gut Microbiome}} on {{Obesity}} in
+  {{Adults}} and the {{Role}} of {{Probiotics}}, {{Prebiotics}}, and
+  {{Synbiotics}} for {{Weight Loss}}.
+\newblock Preventive Nutrition and Food Science. 2020;25(2):113--123.
+\newblock doi:{10.3746/pnf.2020.25.2.113}.
+
+\bibitem{cho2012}
+Cho I, Blaser MJ.
+\newblock The Human Microbiome: At the Interface of Health and Disease.
+\newblock Nature Reviews Genetics. 2012;13(4):260--270.
+\newblock doi:{10.1038/nrg3182}.
+
+\bibitem{callahan2016}
+Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
+\newblock {{DADA2}}: {{High-resolution}} Sample Inference from {{Illumina}}
+  Amplicon Data.
+\newblock Nature Methods. 2016;13(7):581--583.
+\newblock doi:{10.1038/nmeth.3869}.
+
+\bibitem{truong2015}
+Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
+\newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
+\newblock Nature Methods. 2015;12(10):902--903.
+\newblock doi:{10.1038/nmeth.3589}.
+
+\bibitem{li2019a}
+Li H.
+\newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
+  {{Metagenomics}}.
+\newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
+  2019. p. 977--550.
+
+\bibitem{li2015}
+Li H.
+\newblock Microbiome, {{Metagenomics}}, and {{High-Dimensional Compositional
+  Data Analysis}}.
+\newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
+\newblock doi:{10.1146/annurev-statistics-010814-020351}.
+
+\bibitem{gloor2017}
+Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
+\newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
+  Optional}}.
+\newblock Frontiers in Microbiology. 2017;8.
+\newblock doi:{10.3389/fmicb.2017.02224}.
+
+\bibitem{khatri2012}
+Khatri P, Sirota M, Butte AJ.
+\newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
+  {{Outstanding Challenges}}.
+\newblock PLOS Computational Biology. 2012;8(2):e1002375.
+\newblock doi:{10.1371/journal.pcbi.1002375}.
+
+\bibitem{goeman2007}
+Goeman JJ, B{\"u}hlmann P.
+\newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
+  Issues.
+\newblock Bioinformatics. 2007;23(8):980--987.
+\newblock doi:{10.1093/bioinformatics/btm051}.
+
+\bibitem{subramanian2005}
+Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
+\newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
+  Interpreting Genome-Wide Expression Profiles.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(43):15545--15550.
+\newblock doi:{10.1073/pnas.0506580102}.
+
+\bibitem{ashburner2000}
+Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
+\newblock Gene {{Ontology}}: Tool for the Unification of Biology.
+\newblock Nature genetics. 2000;25(1):25--29.
+\newblock doi:{10.1038/75556}.
+
+\bibitem{irizarry2009}
+Irizarry RA, Wang C, Zhou Y, Speed TP.
+\newblock Gene {{Set Enrichment Analysis Made Simple}}.
+\newblock Statistical methods in medical research. 2009;18(6):565--575.
+\newblock doi:{10.1177/0962280209351908}.
+
+\bibitem{hanzelmann2013}
+H{\"a}nzelmann S, Castelo R, Guinney J.
+\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and {{RNA-Seq}}
+  Data.
+\newblock BMC Bioinformatics. 2013;14(1):7.
+\newblock doi:{10.1186/1471-2105-14-7}.
+
+\bibitem{frost2020}
+Frost HR.
+\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
+  Method for Cell-Specific Gene Set Scoring.
+\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
+\newblock doi:{10.1093/nar/gkaa582}.
+
+\bibitem{chong2020}
+Chong J, Liu P, Zhou G, Xia J.
+\newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
+  Functional, and Meta-Analysis of Microbiome Data.
+\newblock Nature Protocols. 2020;15(3):799--821.
+\newblock doi:{10.1038/s41596-019-0264-1}.
+
+\bibitem{quinn2019}
+Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
+\newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
+\newblock GigaScience. 2019;8(giz107).
+\newblock doi:{10.1093/gigascience/giz107}.
+
+\bibitem{quinn2018}
+Quinn TP, Erb I, Richardson MF, Crowley TM.
+\newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
+\newblock Bioinformatics. 2018;34(16):2870--2878.
+\newblock doi:{10.1093/bioinformatics/bty175}.
+
+\bibitem{morton2019}
+Morton JT, Marotz C, Washburne A, Silverman J, Zaramela LS, Edlund A, et~al.
+\newblock Establishing Microbial Composition Measurement Standards with
+  Reference Frames.
+\newblock Nature Communications. 2019;10(1).
+\newblock doi:{10.1038/s41467-019-10656-5}.
+
+\bibitem{love2014}
+Love MI, Huber W, Anders S.
+\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA-seq}}
+  Data with {{DESeq2}}.
+\newblock Genome Biology. 2014;15(12):550.
+\newblock doi:{10.1186/s13059-014-0550-8}.
+
+\bibitem{weiss2017}
+Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
+\newblock Normalization and Microbial Differential Abundance Strategies Depend
+  upon Data Characteristics.
+\newblock Microbiome. 2017;5(1).
+\newblock doi:{10.1186/s40168-017-0237-y}.
+
+\bibitem{mcknight2019}
+McKnight DT, Huerlimann R, Bower DS, Schwarzkopf L, Alford RA, Zenger KR.
+\newblock Methods for Normalizing Microbiome Data: {{An}} Ecological
+  Perspective.
+\newblock Methods in Ecology and Evolution. 2019;10(3):389--400.
+\newblock doi:{10.1111/2041-210X.13115}.
+
+\bibitem{aitchison1994}
+Aitchison J.
+\newblock Principles of Compositional Data Analysis.
+\newblock Lecture Notes-Monograph Series. 1994; p. 73--81.
+
+\bibitem{mclaren2019}
+McLaren MR, Willis AD, Callahan BJ.
+\newblock Consistent and Correctable Bias in Metagenomic Sequencing
+  Experiments.
+\newblock eLife. 2019;8:e46923.
+\newblock doi:{10.7554/eLife.46923}.
+
+\bibitem{egozcue2005}
+Egozcue JJ, {Pawlowsky-Glahn} V.
+\newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2005;37(7):795--828.
+\newblock doi:{10.1007/s11004-005-7381-9}.
+
+\bibitem{tian2005}
+Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
+\newblock Discovering Statistically Significant Pathways in Expression
+  Profiling Studies.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(38):13544--13549.
+\newblock doi:{10.1073/pnas.0506577102}.
+
+\bibitem{rivera-pinto2018}
+{Rivera-Pinto} J, Egozcue JJ, {Pawlowsky-Glahn} V, Paredes R, {Noguera-Julian}
+  M, Calle ML.
+\newblock Balances: A {{New Perspective}} for {{Microbiome Analysis}}.
+\newblock mSystems. 2018;3(4):e00053--18.
+\newblock doi:{10.1128/mSystems.00053-18}.
+
+\bibitem{wu2012}
+Wu D, Smyth GK.
+\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
+  Correlation.
+\newblock Nucleic Acids Research. 2012;40(17):e133.
+\newblock doi:{10.1093/nar/gks461}.
+
+\bibitem{egozcue2003}
+Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcel{\'o}-Vidal} C.
+\newblock Isometric {{Logratio Transformations}} for {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2003;35(3):279--300.
+\newblock doi:{10.1023/A:1023818214614}.
+
+\bibitem{quast2013}
+Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
+\newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
+  Processing and Web-Based Tools.
+\newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
+\newblock doi:{10.1093/nar/gks1219}.
+
+\bibitem{delignette-muller2015}
+{Delignette-Muller} ML, Dutang C.
+\newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
+\newblock Journal of Statistical Software. 2015;64(4):1--34.
+
+\bibitem{benaglia2009}
+Benaglia T, Chauveau D, Hunter DR, Young D.
+\newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
+\newblock Journal of Statistical Software. 2009;32(6):1--29.
+
+\bibitem{washburne2017}
+Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL.
+\newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
+  Associations in Microbiome Datasets.
+\newblock PeerJ. 2017; p.~26.
+
+\bibitem{silverman2017}
+Silverman JD, Washburne AD, Mukherjee S, David LA.
+\newblock A Phylogenetic Transform Enhances Analysis of Compositional
+  Microbiota Data.
+\newblock eLife. 2017;6:e21887.
+\newblock doi:{10.7554/eLife.21887}.
+
+\bibitem{morton2017}
+Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
+  et~al.
+\newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
+\newblock mSystems. 2017;2(1):e00162--16.
+\newblock doi:{10.1128/mSystems.00162-16}.
+
+\bibitem{aitchison1980}
+Aitchison J, Shen SM.
+\newblock Logistic-Normal Distributions:{{Some}} Properties and Uses.
+\newblock Biometrika. 1980;67(2):261--272.
+\newblock doi:{10.1093/biomet/67.2.261}.
+
+\bibitem{efron2004}
+Efron B.
+\newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
+\newblock Journal of the American Statistical Association.
+  2004;99(465):96--104.
+\newblock doi:{10.1198/016214504000000089}.
+
+\bibitem{kurtz2015a}
+Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
+\newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
+  Ecological Networks}}.
+\newblock PLOS Computational Biology. 2015;11(5):e1004226.
+\newblock doi:{10.1371/journal.pcbi.1004226}.
+
+\bibitem{ackerman2009}
+Ackermann M, Strimmer K.
+\newblock A General Modular Framework for Gene Set Enrichment Analysis.
+\newblock BMC bioinformatics. 2009;10(1):1--20.
+
+\bibitem{geistlinger2021}
+Geistlinger L, Csaba G, Santarelli M, Ramos M, Schiffer L, Turaga N, et~al.
+\newblock Toward a Gold Standard for Benchmarking Gene Set Enrichment Analysis.
+\newblock Briefings in bioinformatics. 2021;22(1):545--556.
+
+\bibitem{pasolli2017}
+Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
+\newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
+\newblock Nature Methods. 2017;14(11):1023--1024.
+\newblock doi:{10.1038/nmeth.4468}.
+
+\bibitem{schiffer2019}
+Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
+\newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
+  through Bioconductor.
+\newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
+
+\bibitem{gonzalez2018}
+Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
+  Ackermann G, et~al.
+\newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
+\newblock Nature Methods. 2018;15(10):796--798.
+\newblock doi:{10.1038/s41592-018-0141-9}.
+
+\bibitem{wickham2019}
+Wickham H, Averick M, Bryan J, Chang W, McGowan LD, Fran{\c c}ois R, et~al.
+\newblock Welcome to the Tidyverse.
+\newblock Journal of Open Source Software. 2019;4(43):1686.
+\newblock doi:{10.21105/joss.01686}.
+
+\bibitem{robin2011}
+Robin X, Turck N, Hainard A, Tiberti N, Lisacek F, Sanchez JC, et~al.
+\newblock {{pROC}}: An Open-Source Package for {{R}} and {{S}}+ to Analyze and
+  Compare {{ROC}} Curves.
+\newblock BMC Bioinformatics. 2011;12:77.
+
+\bibitem{mcmurdie2014}
+McMurdie PJ, Holmes S.
+\newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
+  Inadmissible}}.
+\newblock PLOS Computational Biology. 2014;10(4):e1003531.
+\newblock doi:{10.1371/journal.pcbi.1003531}.
+
+\bibitem{ernst2021}
+Ernst FGM, Shetty SA, Borman T, Lahti L. Mia: {{Microbiome}} Analysis; 2021.
+
+\bibitem{landau2021}
+Landau WM.
+\newblock The Targets {{R}} Package: A Dynamic {{Make-like}} Function-Oriented
+  Pipeline Toolkit for Reproducibility and High-Performance Computing.
+\newblock Journal of Open Source Software. 2021;6(57):2959.
+
+\bibitem{consortium2012}
+Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
+  et~al.
+\newblock Structure, Function and Diversity of the Healthy Human Microbiome.
+\newblock Nature. 2012;486(7402):207--214.
+\newblock doi:{10.1038/nature11234}.
+
+\bibitem{agresti1998}
+Agresti A, Coull BA.
+\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
+  Estimation}} of {{Binomial Proportions}}.
+\newblock The American Statistician. 1998;52(2):119--126.
+\newblock doi:{10.2307/2685469}.
+
+\bibitem{thurnheer2016}
+Thurnheer T, Bostanci N, Belibasakis GN.
+\newblock Microbial Dynamics during Conversion from Supragingival to
+  Subgingival Biofilms in an in Vitro Model.
+\newblock Molecular Oral Microbiology. 2016;31(2):125--135.
+\newblock doi:{10.1111/omi.12108}.
+
+\bibitem{beghini2019}
+Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
+\newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
+  in the {{New York City Health}} and {{Nutrition Examination Study}}.
+\newblock Annals of Epidemiology. 2019;34:18--25.e3.
+\newblock doi:{10.1016/j.annepidem.2019.03.005}.
+
+\bibitem{matteocalgaro2020}
+Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
+\newblock Zenodo.
+
+\bibitem{delong1988}
+DeLong ER, DeLong DM, {Clarke-Pearson} DL.
+\newblock Comparing the Areas under Two or More Correlated Receiver Operating
+  Characteristic Curves: A Nonparametric Approach.
+\newblock Biometrics. 1988;44(3):837--845.
+
+\bibitem{nielsen2014}
+Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
+\newblock Identification and Assembly of Genomes and Genetic Elements in
+  Complex Metagenomic Samples without Using Reference Genomes.
+\newblock Nature Biotechnology. 2014;32(8):822--828.
+\newblock doi:{10.1038/nbt.2939}.
+
+\bibitem{gevers2014}
+Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
+  et~al.
+\newblock The {{Treatment-Naive Microbiome}} in {{New-Onset Crohn}}'s
+  {{Disease}}.
+\newblock Cell Host \& Microbe. 2014;15(3):382--392.
+\newblock doi:{10.1016/j.chom.2014.02.005}.
+
+\bibitem{breiman2001}
+Breiman L.
+\newblock Random {{Forests}}.
+\newblock Machine Learning. 2001;45(1):5--32.
+\newblock doi:{10.1023/A:1010933404324}.
+
+\bibitem{chawla2002}
+Chawla NV, Bowyer KW, Hall LO, Kegelmeyer WP.
+\newblock {{SMOTE}}: {{Synthetic Minority Over-sampling Technique}}.
+\newblock Journal of Artificial Intelligence Research. 2002;16:321--357.
+\newblock doi:{10.1613/jair.953}.
+
+\bibitem{kuhn2020}
+Kuhn M, Wickham H. Tidymodels: A Collection of Packages for Modeling and
+  Machine Learning Using Tidyverse Principles.; 2020.
+
+\bibitem{martin2020}
+Martin BD, Witten D, Willis AD.
+\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
+  Regression.
+\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
+\newblock doi:{10.1214/19-AOAS1283}.
+
+\bibitem{barbie2009}
+Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
+\newblock Systematic {{RNA}} Interference Reveals That Oncogenic
+  {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
+\newblock Nature. 2009;462(7269):108--112.
+\newblock doi:{10.1038/nature08460}.
+
+\bibitem{calgaro2020}
+Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
+\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA-seq}},
+  and Metagenomics Applied to Microbiome Data.
+\newblock Genome Biology. 2020;21(1):191.
+\newblock doi:{10.1186/s13059-020-02104-1}.
+
+\bibitem{wu2016a}
+Wu C, Chen J, Kim J, Pan W.
+\newblock An Adaptive Association Test for Microbiome Data.
+\newblock Genome Medicine. 2016;8(1):56.
+\newblock doi:{10.1186/s13073-016-0302-3}.
+
+\bibitem{martin-fernandez2012}
+{Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
+  {Palarea-Albaladejo} J.
+\newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
+  {{Classical}} and Robust Approaches.
+\newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
+\newblock doi:{10.1016/j.csda.2012.02.012}.
+
+\bibitem{kaul2017a}
+Kaul A, Davidov O, Peddada SD.
+\newblock Structural Zeros in High-Dimensional Data with Applications to
+  Microbiome Studies.
+\newblock Biostatistics. 2017;18(3):422--433.
+\newblock doi:{10.1093/biostatistics/kxw053}.
+
+\bibitem{weissman2021}
+Weissman JL, Dogra S, Javadi K, Bolten S, Flint R, Davati C, et~al.
+\newblock Exploring the Functional Composition of the Human Microbiome Using a
+  Hand-Curated Microbial Trait Database.
+\newblock BMC Bioinformatics. 2021;22(1):306.
+\newblock doi:{10.1186/s12859-021-04216-2}.
+
+\end{thebibliography}
 

### 2022-02-14 supplemental.tex

@@
 \subsection{Design}
-To address the performance of CBEA for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
+Even though real data evaluations provide good estimates for performance of CBEA under typical analysis tasks, it does not allow for understanding of the behavior of the model under different effect sizes, correlations, and sparsity.
+As such, we also perform parametric numerical simulations by generating microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}.
 Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
@@
 To ensure that our simulations match closely to real data, we fitted negative binomial distribution using a maximum likelihood approach (with the \emph{fitdistrplus} package in R \cite{delignette-muller2015}) to non-zero counts for each taxon from 16S rRNA profiling of stool samples from the Human Microbiome Project (HMP).
-We take the median values of the estimated mean and dispersion parameters as the baseline of our simulations.
-For simplicity, we assumed that inter-taxa correlation follows an exchangeable structure
+We take the median values of the estimated mean ($\mu_j$) and dispersion parameters ($\phi_j$) as the baseline of our simulations.
+For simplicity, we assumed that inter-taxa correlation follows an exchangeable structure with correlation equals to $\rho$.
 
@@
 
-For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUC).
+For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUROC).
 This is a strategy used in Frost \cite{frost2020} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e.
 whether samples with inflated counts are highly ranked using estimated scores).
-DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate.
+DeLong 95\% confidence intervals for AUROC \cite{delong1988} were obtained for each estimate.
 Simulation settings for classification performance were identical to power analyses as detailed in the previous paragraph.
@@
 \subsubsection{Simulation scenarios for enrichment analysis at the population level}
-To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition.
-Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null.
+To assess type I error rate and power for inferece at the population level, we simulated data based on the schema above, and assessed the enrichment of 50 sets (with 100 taxa per set) across 10 replicates per simulation condition.
+Type I error is calculated as the number of enriched sets over the total number of sets for each simulation under the global null.
 Power is defined similarly, but instead under the global alternate hypothesis.
 Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error.
-A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size.
 Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$).
-Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3).
+For power analyse, we defined an enriched set as a set where all taxa within a set have inflated means of the same effect size.
+Half of the sets are defined as enriched across case/control status with varying effect sizes (fold change of 1.5, 2, and 3).
 Due to the compositional nature of microbiome taxonomic data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets.
@@
 To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}.
-For binary outcomes, we use AUC similar to the classification analyses above.
+For binary outcomes, we use AUROC similar to the classification analyses above.
 For continuous outcomes, we used root mean squared error (RMSE).
-All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020b} suite of packages.
+All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020} suite of packages.
 Across both learning tasks, we varied sparsity ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation ($\rho = 0, 0.2, 0.5$).
@@
 \subsubsection{Statistical Inference}
-Panel A and B in Fig~\ref{fig:2} demonstrate type I error and power respectively across different simulation conditions.
-We benchmarked the results of the CBEA method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement.
+Fig~\ref{fig:s1} demonstrate type I error evaluations for sample-level inference with CBEA compared to the Wilcoxon rank sum test, which uses a rank-based statistic to compare the mean count difference between taxa in the set its complement.
 All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions.
@@
 On the other hand, adjusted CBEA methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.
+This is opposite from our real data evaluations, where adjusted CBEA demonstrated inflated type I error under random set evalutions.
 
@@
     \includegraphics[width=\textwidth]{figures/sim_ss_fdr.png}
-    \caption{Simulation results for type I error evaluation for CBEA sample-level inference.
+    \caption{\textbf{Simulation results for type I error evaluation for CBEA sample-level inference}.
 Type I error rate (\emph{y} axis) was estimated for each approach across data sparsity levels (\emph{x} axis) across different set sizes (horizontal) and inter-taxa correlation within the set (vertical).
@@
 
-CBEA generates sample-specific scores representing the degree of enrichment of a pre-defined set.
-As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test).
+We also assessed the ability to perform inference at the population level using CBEA similar to GSVA \cite{hanzelmann2013}.
+Here, we test for enrichment of sets across case/control status by generating CBEA scores and performing Welch's t-test as a difference in means test.
 We compared the performance of this approach with CBEA and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
 
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
-    \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
+%\begin{figure}[!h]
+%    \centering
+%    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
+%    \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
 For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions).
 CBEA associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
-    \label{fig:s2}
-\end{figure}
+%    \label{fig:s2}
+%\end{figure}
 
-Fig.~\ref{fig:4} present results for simulation studies for both type I error (panel A) and power (panel B) evaluations.
+%Fig.~\ref{fig:s2} present results for simulation studies for both type I error (panel A) and power (panel B) evaluations.
 All methods control for type I error well across both sparsity and correlation levels, where the estimated rate was consistently around the 0.05 pre-defined threshold.
@@
 
-\subsubsection{Phenotype relevance}.
+\subsubsection{Phenotype relevance}
 
@@
     \includegraphics[width = \textwidth]{figures/sim_ss_pwr.png}
-    \caption{Simulation results for phenotype relevance valuation for CBEA sample-level inference.
-(\textbf{(A)}) demonstrate statistical power (\emph{y} axis) across different data sparsity levels (\emph{x} axis) and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
+    \caption{\textbf{Simulation results for phenotype relevance valuation for CBEA sample-level inference}.
+\textbf{(A)} demonstrate statistical power (\emph{y}-axis) across different data sparsity levels (\emph{x}-axis) and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
 For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions).
@@
 
-However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B.
-In situations where there is no inter-taxa correlation, CBEA still outperforms the wilcoxon rank sum test, however adjusted versions of CBEA did not perform as well as un-adjusted ones.
-However, in higher correlation scenarios, the difference in power is much more dramatic.
-At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted CBEA was only performing at 50\% power, while unadjusted CBEA and wilcoxon rank sum test were able to reach 80\%.
-These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of CBEA.
-Most importantly, CBEA demonstrate higher power in all scenarios where type I error is properly controlled.
-
-To further assess the utility of CBEA in classifying samples with enriched sets, we generated AUC scores for different CBEA scores using true labels of whether a sample has an inflated set.
-This analysis, therefore, assessed the relative ranking of samples using CBEA scores whereby high scores should correspond to samples that are known to be inflated.
-Fig~\ref{fig:3} presents this result.
-We compared different variants of CBEA against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test.
-Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), CBEA scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios.
-In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2).
-Notably, the W-statistic provided the least information for classifying samples with inflated taxa.
-
-The difference between the methods is more noticeable when evaluating power.
-All CBEA associated variants showed much higher power even when the effect size is limited (fold change is 1.5), and there is a noticeable gap in performance between CBEA and both DESeq2 and corncob.
-Surprisingly, this effect is consistent across correlation levels and sparsity, even though we expectedly see performance in power drop as a function of sparsity especially in low effect size settings.
+We assessed phenotype relevance similar to the main manuscript by assessing statistical power and score rankings via AUROC.
+Results for this analysis is shown in Fig~\ref{fig:s3}.
+For statistical power (panel A), under low-correlation settings, all CBEA approaches demonstrate similar power, with the unadjusted methods being slightly more performant at low effect sizes.
+Notably, all CBEA variants outperformed the Wilcoxon rank sum test.
+However, as correlation increases, the adjusted CBEA variants showed much lower power, congruent with the perspective that the adjustment process is conservative and trades off power for type I error control.
+Even at the highest effect size (fold change of 3 in means) adjusted CBEA does not approach 0.8.
+For score rankings (panel B), all methods are close together in performance, with the Wilcoxon W statistic being the worst performer.
+These results are similar to that of our real data evaluations.
+Notably, performance values were not affected by sparsity and increases to near perfect prediction at higher effect sizes.
 
 \subsubsection{Prediction analysis}
-Fig~\ref{fig:6} shows results for simulation studies as detailed in the \nameref{methods} section.
+Fig~\ref{fig:s5} shows results for simulation studies as detailed in the \nameref{methods} section.
 Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome.
@@
 Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure.
-Panel \textbf{(B)} presents performance on a classification task with AUC as the evaluation measure.
+Panel \textbf{(B)} presents performance on a classification task with AUROC as the evaluation measure.
 CBEA approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
-    \label{fig:6}
+    \label{fig:s5}
 \end{figure}
@@
 \section{Performance evaluation} 
-We implemented cILR in the package \emph{teaR} on GitHub with \emph{Rcpp} implementations for geometric mean computations.
+We implemented CBEA in the package \emph{CBEA} on GitHub (\url{https://www.github.com/qpmnguyen/CBEA}).
 We evaluated computational time using the \emph{bench} package in R.
@@
 Across all evaluations, the size of each taxa set was also fixed at 50.}
-    \label{fig:s1}
+    \label{fig:s6}
 \end{figure}
@@
 For correlation of 0.5, p-values represent correlation adjusted cILR while for correlation of 0, p-values represent unadjusted cILR.}
-    \label{fig:s2}
+    \label{fig:s7}
 \end{figure}
@@
 \newpage
-\bibliography{mac_taxagg}{}
+\bibliography{references}{}
 \bibliographystyle{plos2015}

### 2022-02-14 supplemental.tex
Updates from Overleaf

@@
 \cite{hawinkel2019} to ensure the validity of simulation results.
-All sample sizes were set at 2,000.
+All sample sizes were set at 500.
 
@@
 For each scenario, a data set of 10,000 samples (equivalent to 10,000 hypotheses) was utilized.
-Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach}
+Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach.}
     \label{fig:s1}
@@
 We compared the performance of this approach with CBEA and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.
-
-%\begin{figure}[!h]
-%    \centering
-%    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
-%    \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
-For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions).
-CBEA associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.}
-%    \label{fig:s2}
-%\end{figure}
-
-%Fig.~\ref{fig:s2} present results for simulation studies for both type I error (panel A) and power (panel B) evaluations.
-All methods control for type I error well across both sparsity and correlation levels, where the estimated rate was consistently around the 0.05 pre-defined threshold.
-Results were similar across all evaluated methods, although in some instances, for example in medium correlation setting ($\rho = 0.2$), the unadjusted CBEA resulted in higher type I error, regardless of difference in means test and distribution of choice.
+Fig.~\ref{fig:s2} present results for simulation studies for both type I error evaluations.
+All methods were able to control for type I error across both sparsity and correlation levels, where medium level sparsity ($p = 0.4$) and correlation ($\rho = 0.2$) showed the strongest performance.
+In these scenarios, the CDF values of CBEA generated under the adjusted mixture normal distribution performed the best.
+This is different than our real data evaluations, where corncob and DESeq2 showed increased type I error.
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{figures/sim_diffab_fdr.png}
+    \caption{\textbf{Simulation results for type I error evaluation for CBEA population-level inference}.
+Type I error (\emph{y}-axis) was estimated as the average proportion of sets with significant enrichment at 0.05 across 10 replications per simulation condition under the global null.
+Error bars were estimated using standard errors computed across 10 replicated data sets.
+Performance was evaluated across different sparsity (\emph{x}-axis) and inter-taxa correlation levels.
+For CBEA methods, enrichment analysis was performed using a Welch's t-test across case/control status with single sample scores representing set-based features generated by CBEA (across different output types and distributional assumptions).
+For corncob and DESeq2, set-based features were constructed using element-wise summations.}
+    \label{fig:s2}
+\end{figure}
 
@@
 Notably, performance values were not affected by sparsity and increases to near perfect prediction at higher effect sizes.
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width = \textwidth]{figures/sim_diffab_pwr.png}
+    \caption{\textbf{Simulation results for phenotype relevance evaluation for CBEA population-level inference}.
+Power (\emph{y}-axis) was estimated as the average proportion of sets correctly identified as significantly enriched (at 0.05) across 10 replications per simulation condition under the global null.
+Error bars were estimated using standard errors computed across 10 replicated data sets.
+Performance was evaluated across different sparsity (\emph{x}-axis) and inter-taxa correlation levels.
+For CBEA methods, enrichment analysis was performed using a Welch's t-test across case/control status with single sample scores representing set-based features generated by CBEA (across different output types and distributional assumptions).
+For corncob and DESeq2, set-based features were constructed using element-wise summations.}
+    \label{fig:s4}
+\end{figure}
+
+For population-level analyses, we assessed phenotype relevance as statistical power to detect sets that were simulated to be significantly enriched.
+Fig~\ref{fig:s4} showed these results.
+As expected, power decreases with increasing sparsity, where the effect was attenuated at lower effect sizes.
+Correlation did impact power, however the difference was not notable.
+Interestingly, at lower effect sizes both DESeq2 and corncob has comparable power with CBEA, however as effect size increases, the difference in performance values became more stark.
+This is different than our real data evaluations, where power was more comparable (with slight advantage to corncob and DESeq2).
 

### 2022-02-15 saving_texts.tex

@@
 \end{figure}
+
+\section{Distribution of p-values}
+We also demonstrate the validity our inference procedure by assessing the distribution of p-values.
+We used simulated data sets as described in the main manuscript, under different data sparsity levels (0, 0.6) and inter-taxa correlation within the set ($\rho$ = 0, 0.5).
+We evaluated p-values of the correlation adjusted method for $\rho = 0.5$ case and the p-values of the unadjusted method for $\rho = 0$ case.
+
+In general, sparsity has no effect on the distribution of p-values, where it remains uniform at zero correlation values.
+However, for higher correlation values, p-values deviate more from uniform under the global null, with the mixture normal distribution being further away from uniform compared to the normal distribution.
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/pval_distr.png}
+    \caption{Q-Q plot of 10,000 p-values compared against a uniform distribution bounded between 0 and 1.
+Evaluation was performed on simulated null data sets of 10,000 samples testing for enrichment of a set of size 50.
+For correlation of 0.5, p-values represent correlation adjusted cILR while for correlation of 0, p-values represent unadjusted cILR.}
+    \label{fig:s7}
+\end{figure}

### 2022-02-15 supplemental.tex

@@
 We evaluated computational time using the \emph{bench} package in R.
-Sample size and number of sets were varied across different evaluation conditions.
-All set sizes were kept constant at 50.
+We applied CBEA to a standard data set generated using our simulation model consisting of 40 sets (of size 20 each) and 500 samples.
+Benchmark was performed on a single core using a node on the computing cluster (Specifications: Intel Xeon E5-2690 (2.6GHz) with 4GB of RAM).
 
-\begin{figure}[!ht]
+\begin{figure}[!h]
     \centering
     \includegraphics[width=\textwidth]{figures/performance.png}
-    \caption{Computational time (in seconds) as a function of sample size (left panel) and number of taxa sets evaluated (right panel).
-Evaluation was performed on simulated data sets.
-For sample size analysis, only 10 sets were evaluated.
-For taxa set analysis, sample size was fixed at 1,000.
-Across all evaluations, the size of each taxa set was also fixed at 50.}
+    \caption{\textbf{Runtime performance}.
+Overall runtime of CBEA under different parameters for a data set of 500 samples, 800 taxa (40 sets of size 20 each).
+This data set was generated via simulations.}
     \label{fig:s6}
@@
 
-In general, increasing sample size and number of evaluated sets both increase computational time.
-The mixture normal distribution takes longer to compute compared to the normal distribution, and adjusted methods take longer but was most apparent when combined with the mixture distribution.
-Overall, a reasonable data set of 1,000 samples and 100 sets (5,000 taxa) would take around 200 seconds (or approximately 3 minutes) for the longest approach (correlation adjusted mixture normal distribution).
-Benchmark was performed on a single core using a personal Windows laptop (Specifications: Intel Core i7-9750H (2.6-4.5 GHz - 12MB Cache) with 16GB of RAM)
+In general, using the normal distribution is the fastest approach regardless of the total number of permutations performed (2 minutes).
+However, using the mixture normal distribution increased the runtime many folds, especially for the adjusted approach (highest was 5.53 hours).
+This time scales with the number of sets evaluated, as well as the number of samples.
+We believe this is due to the procedure used to estimate parameters of the mixture normal distrubiton as implemented in the \emph{mixtools} package.
+The default parameters used in CBEA also increased the runtime in order to reduce convergence issues.
 
-\section{Distribution of p-values}
-We also demonstrate the validity our inference procedure by assessing the distribution of p-values.
-We used simulated data sets as described in the main manuscript, under different data sparsity levels (0, 0.6) and inter-taxa correlation within the set ($\rho$ = 0, 0.5).
-We evaluated p-values of the correlation adjusted method for $\rho = 0.5$ case and the p-values of the unadjusted method for $\rho = 0$ case.
+In order to reduce runtime, users can attempt the following: Since CBEA fits parametric distributions over permuted values of all samples within a data set (i.e.
+for $N = 100$ and $10$ permutations, the fitting procedure will attempt to estimate parameters from a vector of size $1000$ per set), if the sample size is high users can reduce the number of permutations.
+Additionally, CBEA also implements a procedure to parallelize computation across sets, which might be applicable to situations where there are a lot of sets.
+Finally, a lot of CBEA approaches work well without parametric fit, so users can use the non-parametric approaches.
 
-In general, sparsity has no effect on the distribution of p-values, where it remains uniform at zero correlation values.
-However, for higher correlation values, p-values deviate more from uniform under the global null, with the mixture normal distribution being further away from uniform compared to the normal distribution.
-
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[width=\textwidth]{figures/pval_distr.png}
-    \caption{Q-Q plot of 10,000 p-values compared against a uniform distribution bounded between 0 and 1.
-Evaluation was performed on simulated null data sets of 10,000 samples testing for enrichment of a set of size 50.
-For correlation of 0.5, p-values represent correlation adjusted cILR while for correlation of 0, p-values represent unadjusted cILR.}
-    \label{fig:s7}
-\end{figure}
-
-\newpage
+\clearpage
 \bibliography{references}{}

### 2022-02-15 supplemental.tex
Updates from Overleaf

@@
 \subsubsection{Prediction analysis}
-Fig~\ref{fig:s5} shows results for simulation studies as detailed in the \nameref{methods} section.
-Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome.
-As expected, performance across all assessed methods increased with a higher signal-to-noise ratio.
-Both CLR and CBEA approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks.
-This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while CBEA and CLR did not experience a similar level of impact.
-As such, performance gap widens with increasing correlation and sparsity.
-Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.
-
-In this analysis, CBEA unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}.
-This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation.
-However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and CBEA/CLR.
-Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity.
-Additionally, we did not observe any performance difference between the different variations of CBEA.
 
@@
     \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
-    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
-Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure.
-Panel \textbf{(B)} presents performance on a classification task with AUROC as the evaluation measure.
-CBEA approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
+    \caption{\textbf{Simulation results for predictive performance evaluation for CBEA}.
+Predictive performance of a random forest model (with no hyperparameter tuning) trained on set-based features as inputs.
+Methods to generate these features include CBEA, ssGSEA, GSVA, and the CLR transformation applied on sum-aggregated sets.
+Simulation data was generated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
+Panel \textbf{(A)} presents performance on a regression task using RMSE (root mean squared error) as the evaluation measure.
+Panel \textbf{(B)} presents performance on a classification task with AUROC as the evaluation measure}
     \label{fig:s5}
@@
 
-\section{Performance evaluation} 
+Fig~\ref{fig:s5} shows results for simulation studies as detailed in the \nameref{methods} section.
+Panel A presents results for a regression learning task with a continuous outcome.
+We observed no difference in performance values across different evaluation models.
+Overall, prediction error did not change across sparsity levels, and decreases with increasing signal-to-noise ratio (SNR).
+However, at higher correlation levels, the pattern was more erratic.
+For example, when $\rho$ is set at $0.5$, higher SNR decreases performance only at low sparsity $p = 0.2$, but had the expected pattern medium sparsity $p = 0.4$, where higher SNR correlates with improved performance.
+This effect seems to be specific to low effect saturation scenarios (only a limited number of taxa sets are associated with the outcome).
+Interestingly, higher sparsity levels produces better performance.
+
+Panel B represent results for a classification task with a binary outcome with AUROC as the evaluation criteria.
+Here we observed similar results as that of our real data evaluations, where using CLR transfromed data produces more predictive models across all scenarios.
+Conversely, GSVA and ssGSEA were consistently under performing when compared to CBEA and CLR.
+Interestingly, the degree of difference varies across sparsity and inter-taxa correlations.
+We noticed that increasing sparsity and correlation decreases the gap in performance between CBEA and CLR, while increasing the gap in performance between CBEA/CLR and GSVA/ssGSEA.
+As such, we can hypothesize that both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity.
+Finally, effect saturation did not change model rankings, but did decrease overall performance.
+
+\section{Runtime} 
 We implemented CBEA in the package \emph{CBEA} on GitHub (\url{https://www.github.com/qpmnguyen/CBEA}).
@@
 In order to reduce runtime, users can attempt the following: Since CBEA fits parametric distributions over permuted values of all samples within a data set (i.e.
-for $N = 100$ and $10$ permutations, the fitting procedure will attempt to estimate parameters from a vector of size $1000$ per set), if the sample size is high users can reduce the number of permutations.
-Additionally, CBEA also implements a procedure to parallelize computation across sets, which might be applicable to situations where there are a lot of sets.
-Finally, a lot of CBEA approaches work well without parametric fit, so users can use the non-parametric approaches.
+for $N = 100$ and $10$ permutations, the fitting procedure will attempt to estimate parameters from a vector of size $1000$ for each set), if the sample size is high users can reduce the number of permutations.
+Additionally, CBEA also implements a procedure to parallelize computation across sets, which might be applicable to situations where there are a lot of sets to evaluate.
+Finally, a lot of CBEA approaches work well without parametric fit, so users can use the non-parametric approaches like the permutation test or using raw CBEA scores.
 

### 2022-02-15 cbea_revised.tex

--- 2022-02-01 cbea_revised.tex
@@
 This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
-Here, we attempt to fill this gap with a new single sample taxon enrichment method with using a novel log-ratio formulation based on the competitive null hypothesis commonly used in the enrichment analysis literature.
+Here, we attempt to fill this gap with a new single sample taxon enrichment method that uses a novel log-ratio formulation based on the competitive null hypothesis commonly used in the enrichment analysis literature.
 Our approach, titled competitive balances for taxonomic enrichment analysis (CBEA), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
@@
 Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability.
-Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given its unique features compared to other 'omics data types.
+Unfortunately, there is limited interest in developing new set-based tools tailored for microbiome taxonomic data given its unique features compared to other 'omics data types.
 We developed a new tool to generate taxon set enrichment scores at the sample level through a novel log-ratio formulation based on the competitive null hypothesis.
@@
 The data is sparse, high-dimensional, and likely compositional \cite{gloor2017, li2019a, li2015}.
-Even though the these problems are challenging, a very approachable solution is to to use set-based analysis methods, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}.
+Even though these problems are challenging, a very approachable solution is to use set-based analysis methods, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}.
 Aggregated variables can be less sparse, and testing on a smaller number of features can reduce the multiple-testing burden.
@@
 In the following sections, we present our approach titled competitive balances for taxonomic erichment analysis (CBEA).
-First, we presented the step-by-step formulation of CBEA and discuss its statistical properties.
-Second, we detailed our evaluation strategy using both real data and parametric simulations, and the methods we're comparing against.
-Third, we presented results on enrichment testing using CBEA for single samples as well as at the population level.
-Fourth, we showed the performance of CBEA in downstream disease prediction.
-Finally, we discussed our results and the limitations of our method.
+First, we present the step-by-step formulation of CBEA and discuss its statistical properties.
+Second, we detail our evaluation strategy using both real data and parametric simulations, and the methods we're comparing against.
+Third, we present results on enrichment testing using CBEA for single samples as well as at the population level.
+Fourth, we show the performance of CBEA in downstream disease prediction.
+Finally, we discuss our results and the limitations of our method.
 An R package implementation of CBEA can be found on GitHub (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/CBEA}).
@@
 \begin{enumerate}
-    \item \textbf{Compute the CBEA statistic}: Let $\mathbf{M}$ be a $n$ by $m$ matrix of CBEA scores. Let $\mathbf{M}_{i,k}$ be CBEA scores for set $k$ of sample $i$:   
+    \item \textbf{Compute the CBEA statistic}: Let $\mathbf{M}$ be a $n$ by $m$ matrix of CBEA scores. Let $\mathbf{M}_{i,k}$ be CBEA score for set $k$ and sample $i$:   
     \begin{equation}\label{main_eq}
@@
 These balances are computed using the isometric log ratio (ILR) transformation \cite{egozcue2003} formula.
-For a given balance $i$ splitting variables across sets $R$ and and $S$, we have the balance coordinate $x^{*}_{i}$ as:
+For a given balance $i$ splitting variables across sets $R$ and $S$, we have the balance coordinate $x^{*}_{i}$ as:
 \begin{equation}\label{ilr_standard}
@@
 
-CBEA belongs to a set of methods that seeks to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto2018, silverman2017, morton2017}.
+CBEA belongs to a set of methods that seek to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto2018, silverman2017, morton2017}.
 Unlike methods such as PhILR \cite{silverman2017}, CBEA does not provide a sequential binary partition that forms the basis for ILR procedure \cite{egozcue2003} and is therefore not a subclass of ILR.
@@
 \end{equation}
-where $\sigma_i$ is the standard deviation of taxon $i$ while $\rho_{ij}$ is the correlation between $i$ and $j$.
+where $\sigma_i$ is the standard deviation of taxon $i$ and $\rho_{ij}$ is the correlation between $i$ and $j$.
 The second term of (\ref{vareq:1}) is the correlation dependent variance component, which goes to 0 if there is no correlation.
@@
 It is important to address this problem since there is strong inter-taxa correlation within the microbiome \cite{kurtz2015a}.
-Our strategy for addressing issue is location (or mean) estimate from the column permuted raw score matrix with the spread (or variance) estimate from the original un-permuted scores.
-This still allows us to leverage the null generated via column permutation while using the proper variance estimate taken from scores where the correlation structure has not been disrupted.
+Our strategy for addressing this issue is to use the location (or mean) estimate from the column permuted raw score matrix with the spread (or variance) estimate taken from the original un-permuted scores.
+This still allows us to leverage the null distribution generated via column permutation while using the proper variance estimate taken from scores where the correlation structure has not been disrupted.
 As such, this procedure assumes that the variance of the test statistic under the alternate hypothesis is the same as that of the null.
@@
 Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample sets of randomly drawn taxa should not be significantly enriched compared to the remainder background taxa.
-For this evaluation, we estimated type I error as the fraction of samples that detects our random set as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.
-Additionally, this analysis also demonstrate whether CBEA is sensitive to different set sizes.
+For this evaluation, we estimated type I error as the fraction of samples where our random set is detected as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.
+Additionally, this analysis also tests whether CBEA is sensitive to different set sizes.
 
@@
 We want to evaluate whether sets found to be significantly enriched by CBEA are relevant to the research question.
-To perform this assessment, we relied on the gingival data set mentioned prior \cite{consortium2012, proctor2019}.
-This data set was chosen due to its clear biological interpretation that can serve as the ground truth.
+To perform this assessment, we relied on the gingival data set mentioned above \cite{consortium2012, proctor2019}.
+This data set was chosen because its clear biological interpretation can serve as the ground truth.
 Specifically, we expect aerobic microbes to be enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}.
 Genus-level annotations for microbial metabolism from Beghini et al.
-\cite{beghini2019} and was obtained from the GitHub repository associated with Calagaro et al.
+\cite{beghini2019} were obtained from the GitHub repository associated with Calagaro et al.
 \cite{matteocalgaro2020}.
@@
 
-In addition to statistical power, we also assessed phenotype relevance through evaluating whether highly ranked sets based on CBEA scores are more likely to be enriched according to ground truth.
+In addition to statistical power, we also assessed phenotype relevance through evaluating whether highly ranked sets based on CBEA scores are more likely to be enriched according to the ground truth.
 This is represented by the area under the receiving operator curve (AUROC/AUC) scores computed on CBEA scores against true labels (similar approach was used to evaluate VAM \cite{frost2020}).
@@
 \subsection*{Disease Prediction}
-CBEA scores can also be used downstream analyses such as disease prediction tasks. We utilized two data sets for this evaluation:
+CBEA scores can also be used for downstream analyses such as disease prediction tasks. We utilized two data sets for this evaluation:
 
@@
 This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease).
-We processed this data by removing all samples with less than 1,000 total read counts as well as any OTU who was present (with non-zero proportions) in 10\% of samples or less.
-Prior to model fitting, we back-transformed relative abundances into count data (to align the format with our 16S rRNA gene sequencing data set) using provided total number of reads aligned to MetaPhlan marker genes (per sample).
+We processed this data by removing all samples with less than 1,000 total read counts as well as any OTU that was present (with non-zero proportions) in 10\% of the samples or less.
+Prior to model fitting, we back-transformed relative abundances into count data (to align the format with our 16S rRNA gene sequencing data set) using the provided total number of reads aligned to MetaPhlan marker genes (per sample).
 
@@
 Of the 671 samples obtained, 500 samples belong to patients with IBD.
-We processed this data set by removing all samples with less than 1,000 total read counts as well as any OTU who was present (at least 1 count) in 10\% of samples or less.
+We processed this data set by removing all samples with less than 1,000 total read counts as well as any OTU that was present (at least 1 count) in 10\% of the samples or less.
 \end{enumerate}
 
-We evaluate disease prediction performance by fitting a random forest model \cite{breiman2001} using inputs as CBEA scores to classify samples of patients with IBD and healthy controls.
+We evaluate disease prediction performance by fitting a random forest model \cite{breiman2001} using as inputs CBEA scores to classify samples of patients with IBD and healthy controls.
 Random forest was chosen as a baseline learner due to its flexibility as an out-of-the-box model that is easy to fit.
@@
 The Wilcoxon rank sum test and unadjusted CBEA under mixture normal assumption demonstrated good type I error control at the appropriate $\alpha$ level.
-This fits with our expectations since the mixture normal distribution has much better fit than the normal distribution especially at the tails (Fig~\ref{fig:1}).
+This fits with our expectations since the mixture normal distribution has a much better fit than the normal distribution especially at the tails of the empirical distribution (Fig~\ref{fig:1}).
 However, other variants of CBEA demonstrated inflated type I error, especially correlation adjusted variants compared to their unadjusted counter parts.
@@
 Overall we observed very similar values across CBEA as well as corncob and DESeq2, suggesting that no individual method is systematically identifying too many significant sets.
-Additionally, similar to analogous analyses at the sample level, no approach were significantly sensitive to changes in set sizes.
+Additionally, similar to analogous analyses at the sample level, no approach was significantly sensitive to changes in set sizes.
 
@@
 
-The above results were replicated in simulation studies where we observed that adjusted approaches was very conservative and demonstrated significantly lower power (\nameref{S3_Fig}) with increasing correlation even at the highest evaluated effect sizes.
+The above results were replicated in simulation studies where we observed that adjusted approaches were very conservative and demonstrated significantly lower power (\nameref{S3_Fig}) with increasing correlation even at the highest evaluated effect sizes.
 When assessing score rankings, the performance of CBEA was closer to ssGSEA and GSVA compared to real data evaluations, however all single sample approaches were much better than using the W statistic from the Wilcoxon Rank Sum test.
@@
 We also assessed statistical power for population level inference scenarios using a similar approach.
-Here, enrichment scores for sets representing all identified genera were computed, and estimated power as the fraction of sets found to be differentially enriched across sample site labels (supragingival or subgingival).
+Here, enrichment scores for sets representing all identified genera were computed, and power was estimated as the fraction of sets found to be differentially enriched across sample site labels (supragingival or subgingival).
 We compared these results against performing a differential abundance test of genus level features generated via sum-based approaches.
@@
 In the 16S rRNA data set, the best performing CBEA variant (CDF values computed from an unadjusted mixture normal distribution) outperforms both GSVA and ssGSEA but not the standard CLR approach.
-Interestingly, in the whole genome sequencing data set, CBEA outperforms CLR, but were more similar in performance to GSVA.
-However, due to large confidence intervals, no method were significantly out-performing other approaches evaluated.
-As such, these results indicated that CBEA generated scores are can be informative, providing competitive performance when acting as inputs to disease predictive models.
-Simulation studies (\nameref{S5_Fig}) showed similar results, however CBEA was more consistently underperforming compared to CLR across all scenarios.
-Interestingly, the performance gap decreases with increasing sparsity levels.
+Interestingly, in the whole genome sequencing data set, CBEA outperforms CLR, but was similar in performance to GSVA.
+However, due to large confidence intervals, no method significantly out-performed other evaluated approaches.
+As such, these results indicate that, for a given pre-determined collection of sets, CBEA generated scores are can be informative and provide competitive performance when acting as inputs to disease predictive models.
+Simulation studies (\nameref{S5_Fig}) showed similar results, however CBEA more consistently underperformed compared to CLR across all scenarios.
+Interestingly, the performance gap decreases with increasing sparsity levels and correlation.
 
@@
 Interestingly, while the adjusted methods showed poor performance in real data evaluations (Fig~\ref{fig:2}), in simulation studies (\nameref{S1_Fig}, \nameref{S3_Fig}) these approaches were able to control for type I error well with the trade-off of much lower power.
-
-For population-level inference task, CBEA also performed very well.
-Under the permutation global null, representing genera abundance using CBEA scores in combination with Welch's t-test controls for type I error the correct $\alpha$ threshold while also keeping respectable power.
-Since population level enrichment test is equivalent to differential abundance using set-based features, we compared CBEA approach against using element-wise summations with corncob \cite{martin2020} and DESeq2 \cite{love2014} to test for set-level differential abundance.
+For the population-level inference task, CBEA also performed very well.
+Under the permutation global null, representing genera abundance using CBEA scores in combination with Welch's t-test controls for type I error at the correct $\alpha$ threshold while also keeping respectable power.
+Since the population level enrichment test is equivalent to a differential abundance test using set-based features, we compared the CBEA approach against using element-wise summations with corncob \cite{martin2020} and DESeq2 \cite{love2014} to test for set-level differential abundance.
 We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}.
@@
 These marginals were simulated to account for block exchangable correlation within the enriched set only.
-This might affect our results in two ways: First, our simulation scenario ensures that any designated non-enriched taxa is identical to each other.
-This is not the case for real data, where our null scenarios involves randomly sampled sets that might by chance all have taxa with inflated means compared to remainder taxa.
-This is represented in \nameref{S7_Fig}, where the distribution of type I error across our 500 replications are right skewed for underperforming CBEA variants, indicating that these approaches are much more sensitive compared to the Wilcoxon rank sum test or unadjusted CBEA with mixture normal distribution.
-Second, as described in the \nameref{introduction} section, we did not consider taxon-specific biases that distort the observed relative abundance of taxa compred to true values \cite{mclaren2019}.
+This might affect our results in two ways.
+First, our simulation scenario ensures that all designated non-enriched taxa is identical to each other.
+This is not the case for real data, where our null scenario involves randomly sampled sets that might by chance all have taxa with inflated means compared to remainder taxa.
+This is represented in \nameref{S7_Fig}, where the distribution of type I error across 500 replications are right skewed for underperforming CBEA variants, indicating that these approaches are much more sensitive compared to the Wilcoxon rank sum test or unadjusted CBEA with mixture normal distribution.
+Second, as described in the \nameref{introduction} section, we did not consider taxon-specific biases that distort the observed relative abundance of taxa compared to true values \cite{mclaren2019}.
 In the context of sum-based aggregations, the resulting bias of the aggregated taxon are dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}).
 Conceptually, this means that measurement error for a taxon aggregate is different across samples as relative abundance of contributing changes, leading to issues when attempting to perform inference.
-As such, we expect methods like corncob or DESeq2 when performed on such sum aggregates in the presence of taxon-specific biases to have inflated type I error compared to our taxon-ratio based approach.
-Conversely, in simulation studies, where taxon-specific biases are absent, corncob and DESeq2 should perform better.
+As such, we expect methods like corncob or DESeq2 when performed on such sum aggregates in the presence of taxon-specific biases to have inflated type I error compared to our multiplication based approach.
+This also explains why conversely in simulation studies, where taxon-specific biases are absent, corncob and DESeq2 performed better.
 
@@
 
-We fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using CBEA generated scores as inputs.
+We fit a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using CBEA generated scores as inputs.
 Similar to our inference analysis, we compared CBEA against both ssGSEA and GSVA.
@@
 For the unsupervised approach, utilizing the unadjusted CBEA with the mixture normal distribution provides a good initial starting point.
-In the case where researchers only wants to screen samples with mean-inflated taxon sets (instead of additionally detecting taxon sets with increased correlation), they can apply the adjusted approach, which can be effective at conserving type I error even at high correlation scenarios.
-However, the trade off for this adjustment is power that decreases with increasing correlation.
-For the supervised analysis, all CBEA variants control well for type I error and provides adequate statistical power.
-However, using raw CBEA scores with difference-in-means test such as Welch's t-test is preferable since it requires the least amount of computing resource (no estimation process) while still outperforming using sum-based approach with a standard differential abundance test.
+In the case where researchers only want to screen samples with mean-inflated taxon sets (instead of additionally detecting taxon sets with increased correlation), they can apply the adjusted approach, which can be effective at conserving type I error even for high correlation scenarios.
+However, the trade off for this adjustment is power, which decreases with increasing correlation.
+For the supervised analysis, all CBEA variants control for type I error and provide adequate statistical power.
+However, using raw CBEA scores with difference-in-means test such as Welch's t-test is preferable since is the least computationally expensive (no estimation process) while still outperforming the use of a sum-based approach with a standard differential abundance test.
 
@@
 There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016a} have been used in the evaluation of prior studies.
-Second, we were not able to evaluate the phenotypical relevance of enrichment results as in Geistlinger et al.
+More recent studies have suggested utilizing the hierarchical multinomial logistic normal distribution to model microbiome data sets \cite{morton2021}.
+As such, there is space to evaluate and adapt CBEA to these different distributional assumptions that underlie the data generating process.
+Second, we were not able to evaluate the phenotype relevance of enrichment results as in Geistlinger et al.
 \cite{geistlinger2021} due to limited consistent annotations for microbiome signatures in health and disease, especially those that are experimentally verified (and not just from differential abundance studies).
 We attempted to perform this evaluation by leveraging the gingival data set similar to \cite{calgaro2020}.
-However, we acknowledge that this is not a perfect solution, since oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut.
-As such, results from power analyses using this data set is only relative between the comparison methods instead of treated as absolute measures of power or phenotype relevance.
+However, we acknowledge that this is not a perfect solution, since the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut.
+As such, results from power analyses using this data set is only relative between the comparison methods and cannot be treated as absolute measures of power or phenotype relevance.
+Third, fitting the mixture normal distribution to raw CBEA scores using the expectation-maximization algorithm is difficult, as the convergence rate is slow when there is high overlap between the mixtures, resulting in small mixing coefficient for one of the components and increased runtime (\nameref{S6_Fig}) \cite{naim2012}.
+In our implementation, we attempted to account for this by increasing the maximum number of iterations and relaxed the tolerance threshold.
 Finally, we assumed that taxa within a set are all equally associated with the outcome.
@@
 First, across all analyses, we relied on adding a pseudocount to ensure log operations are valid.
-Users can directly addressing this by incorporating model-based zero correction methods prior to modelling such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
+Users can directly address this by incorporating model-based zero correction methods prior to modelling such as in \cite{martin-fernandez2012} or \cite{kaul2017a}.
 However, in our simulation studies, sparsity seems to not have a significant impact on the overall performance of our approach.
@@
 Incorporation of taxa-specific weights (similar to PhILR \cite{silverman2017}) could reduce the influence of outliers, such as rare or highly invariant taxa.
-Finally, even though for a given set of apriori annotations CBEA can generate useful summary scores, such values are limited in their utility if the annotation themselves are not meaningful.
+Finally, even though for a given set of apriori annotations CBEA can generate useful summary scores, such values are limited in their utility if the annotations themselves are not meaningful.
 As such, curating and validating sets (similar to MSigDB \cite{subramanian2005}) based on physiological or genomic characteristics of microbes \cite{weissman2021} or their association with human disease (in beta BugSigDB \url{https://bugsigdb.org/Main_Page}) can allow for incorporating functional insights into microbiome-outcome analyses.
@@
 The authors thank Becky Lebeaux, Modupe Coker, Erika Dade, Jie Zhou, and Weston Viles for insightful comments and suggestions that greatly improved the paper.
+This work is supported by the National Institutes of Health (grants NCI R21CA253408, NIGMS P20GM130454, NCI P30CA023108, and NLM R01LM012723).
 
@@
 
-\bibliography{mac_taxagg}{}
-%\begin{thebibliography}{10}
-%\end{thebibliography}
+\bibliography{references}{}
 
@@
 \label{S1_Fig}
-{\bf Simulation results for type I error evaluation for CBEA sample-level inference.} Placeholder text.
+{\bf Simulation results for type I error evaluation for CBEA sample-level inference.} Type I error rate (\emph{y} axis) was estimated for each approach across data sparsity levels (\emph{x} axis) across different set sizes (horizontal) and inter-taxa correlation within the set (vertical).
+We compared variatns of CBEA against a Wilcoxon rank sum test at $\alpha$ of 0.05.
+For each scenario, a data set of 10,000 samples (equivalent to 10,000 hypotheses) was utilized.
+Confidence bounds were obtained using Agresti-Couli approach.
 
@@
 \label{S2_Fig}
-{\bf Simulation results for type I error evaluation for CBEA population-level inference.} Placeholder.
+{\bf Simulation results for type I error evaluation for CBEA population-level inference.} Type I error (\emph{y}-axis) was estimated as the average proportion of sets with significant enrichment at 0.05 across 10 replications per simulation condition under the global null.
+Error bars were estimated using standard errors computed across 10 replicated data sets.
+Performance was evaluated across different sparsity (\emph{x}-axis) and inter-taxa correlation levels.
+For CBEA methods, enrichment analysis was performed using a Welch's t-test across case/control status with single sample scores representing set-based features generated by CBEA (across different output types and distributional assumptions).
+For corncob and DESeq2, set-based features were constructed using element-wise summations.
 
@@
 \label{S3_Fig}
-{\bf Simulation results for phenotype relevance evaluation for CBEA sample-level inference.} Placeholder text.
+{\bf Simulation results for phenotype relevance evaluation for CBEA sample-level inference.} \textbf{(A)} demonstrate statistical power (\emph{y}-axis) across different data sparsity levels (\emph{x}-axis) and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
+For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions).
+CBEA associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.
 
@@
 \label{S4_Fig}
-{\bf Simulation results for phenotype relevance evaluation for CBEA population-level inference.} Placeholder text.
+{\bf Simulation results for phenotype relevance evaluation for CBEA population-level inference.} Power (\emph{y}-axis) was estimated as the average proportion of sets correctly identified as significantly enriched (at 0.05) across 10 replications per simulation condition under the global null.
+Error bars were estimated using standard errors computed across 10 replicated data sets.
+Performance was evaluated across different sparsity (\emph{x}-axis) and inter-taxa correlation levels.
+For CBEA methods, enrichment analysis was performed using a Welch's t-test across case/control status with single sample scores representing set-based features generated by CBEA (across different output types and distributional assumptions).
+For corncob and DESeq2, set-based features were constructed using element-wise summations.
 
@@
 \label{S5_Fig}
-{\bf Simulation results for predictive pefromance evaluation for CBEA.} Placeholder text.
+{\bf Simulation results for predictive pefromance evaluation for CBEA.} Predictive performance of a random forest model (with no hyperparameter tuning) trained on set-based features as inputs.
+Methods to generate these features include CBEA, ssGSEA, GSVA, and the CLR transformation applied on sum-aggregated sets.
+Simulation data was generated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio.
+Panel \textbf{(A)} presents performance on a regression task using RMSE (root mean squared error) as the evaluation measure.
+Panel \textbf{(B)} presents performance on a classification task with AUROC as the evaluation measure.
 
@@
 \label{S6_Fig}
-{\bf Computational performance of CBEA.} Placeholder text.
+{\bf Runtime performance}. Overall runtime of CBEA under different parameters for a data set of 500 samples, 800 taxa (40 sets of size 20 each). This data set was generated via simulations.
 

### 2022-02-15 cbea_revised.tex
Updates from Overleaf

@@
 There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016a} have been used in the evaluation of prior studies.
-More recent studies have suggested utilizing the hierarchical multinomial logistic normal distribution to model microbiome data sets \cite{morton2021}.
+More recent studies have suggested utilizing the hierarchical multinomial logistic normal distribution to model microbiome data sets \cite{morton2021, ma2021a}.
 As such, there is space to evaluate and adapt CBEA to these different distributional assumptions that underlie the data generating process.

### 2022-02-15 cbea_revised.tex
Final manuscript

@@
 
-However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019a,li2015}.
+However, there are unique challenges in the analysis of these taxonomic count tables \cite{li2019a,li2015}.
 The data is sparse, high-dimensional, and likely compositional \cite{gloor2017, li2019a, li2015}.
@@
 As such, gene set testing methods have been shown to increase power and reproducibility of genomic analyses.
-Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analysis also allows for more biologically informative interpretations.
-
-There exists a diverse set of methods already developed in this field.
+Furthermore, through the usage of functionally informative sets defined \emph{a priori} based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analysis also allows for more biologically informative interpretations.
+
+A diverse set of methods have already been developed in this field.
 Traditional methods utilize the hypergeometric distribution to test for the overrepresentation of a gene set using a candidate list of genes screened based on a marginal model \cite{goeman2007}.
 Unfortunately, these approaches are sensitive to the differential expression test as well as the chosen threshold when trying to select genes for the candidate list.
-Aggregate score methods, which are generally more preferred \cite{irizarry2009}, instead assigns a score for each gene set based on gene-specific statistics such as z-scores or fold change.
-Of these approaches, methods such as GSEA \cite{subramanian2005} performs a test for each gene set at the population level, summarizing information across all samples.
+Aggregate score methods, which are generally preferred \cite{irizarry2009}, instead assign a score for each gene set based on gene-specific statistics such as z-scores or fold change.
+Of these approaches, methods such as GSEA \cite{subramanian2005} perform a test for each gene set at the population level, summarizing information across all samples.
 Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020}, generate enrichment scores at the sample level and are more akin to a transformation.
@@
 
-In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate taxa to higher Linnean classification levels such as genus, family, or phylum.
-However, despite this interest, there is limited research done to extend existing set-based methods to microbiome relative abundance data.
+In microbiome research, even when no explicit enrichment analysis is performed, researchers often aggregate taxa to higher Linnean classification levels such as genus, family, or phylum.
+However, there is limited research done to extend existing set-based methods to microbiome relative abundance data.
 Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}.
@@
 Some studies have provided emprical performance vealuations supporting this normalization schema \cite{mcknight2019}.
-Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another.
+Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxon can only be interpreted relative to another.
 Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1994} are motivated to address this issue.
@@
 First, we present the step-by-step formulation of CBEA and discuss its statistical properties.
-Second, we detail our evaluation strategy using both real data and parametric simulations, and the methods we're comparing against.
+Second, we detail our evaluation strategy using both real data and parametric simulations, and the methods we are comparing against.
 Third, we present results on enrichment testing using CBEA for single samples as well as at the population level.
@@
 \begin{itemize}
-    \item $\mathbf{X}$: $n$ by $p$ matrix of positive propotions for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing.
+    \item $\mathbf{X}$: $n$ by $p$ matrix of positive proportions for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing.
 Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing.
 CBEA does not accept $X$ matrices with zeroes since it invalidates the log-ratio transformation.
-Users can generate a dense matrix $X$ using a method of choice, however the default mode for CBEA would be add a pseudocount of $10^{-5}$ if zeroes are detected in the matrix.
+Users can generate a dense matrix $X$ using a method of choice, however by default mode CBEA will add a pseudocount of $10^{-5}$ if zeroes are detected in the matrix.
     \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxon $p$ to $m$ sets of interest.
@@
 \subsection*{Properties of CBEA}
-\subsubsection*{CBEA and balances of groups of parts}
+\subsubsection*{CBEA and balances between groups of parts}
 The CBEA statistic is based on the multiplication-based aggregation approach used to calculate balances between groups of parts \cite{egozcue2005}.
@@
 CBEA belongs to a set of methods that seek to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto2018, silverman2017, morton2017}.
-Unlike methods such as PhILR \cite{silverman2017}, CBEA does not provide a sequential binary partition that forms the basis for ILR procedure \cite{egozcue2003} and is therefore not a subclass of ILR.
+Unlike methods such as PhILR \cite{silverman2017}, CBEA does not present an orthonormal basis for the complete ILR transformation (such as a a sequential binary partition) \cite{egozcue2003}.
+Therefore, it is not a subclass of the ILR transformation and is adjacent to this approach.
 A similar method to CBEA would be phylofactor \cite{washburne2017}.
-However, instead of performing an optimization procedure to identify interesting balances, CBEA constructs balances apriori using pre-defined sets, and formulates the enrichment of a set as the scaled log-ratio between the center of the subcomposition represented by microbes within the set and the center of the subcomposition represented by remainder taxa.
+However, instead of performing an optimization procedure to identify interesting balances, CBEA constructs balances \emph{a priori} using pre-defined sets, and formulates the enrichment of a set as the scaled log-ratio between the center of the subcomposition represented by microbes within the set and the center of the subcomposition represented by remainder taxa.
 This formulation aligns with the $Q_1$ null hypothesis from the gene set testing literature \cite{tian2005}.
@@
 This assumption is also supported by preliminary simulation studies (detailed below).
-In panel A of Fig~\ref{fig:1}, we simulated microbiome taxonomic count data under the global null across different data features and compute raw CBEA scores and compute kurtosis and skewness.
-It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation.
+We simulated microbiome taxonomic count data under the global null across different data features and compute raw CBEA scores and compute kurtosis and skewness in Fig~\ref{fig:1}A.
+We found that the characteristics of the null change depending on sparsity and inter-taxa correlation.
 Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis).
@@
 To address this issue, we also evaluated a two-component normal mixture distribution.
-Panel B of Fig~\ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on CBEA scores in simulation scenarios under the global null.
+The goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on CBEA scores in simulation scenarios under the global null is shown in Fig~\ref{fig:1}B.
 We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.
@@
 When taxa within a set are highly correlated, the variance of the sample mean of taxon-wise statistics is inflated.
-Without loss of generalizability, for a set of taxa with taxa-specific statistics $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:
+Without loss of generalizability, for a set of taxa with taxon-specific statistics $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:
 \begin{equation} \label{vareq:1}
@@
 Due to CBEA's self-contained null hypothesis, we can perform inference at the sample level for the enrichment of a set.
-We evaluated this application by generating one random taxa of different sizes $S \in \{20, 50, 100, 150, 200\}$ across 500 iterations.
-Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample sets of randomly drawn taxa should not be significantly enriched compared to the remainder background taxa.
+We evaluated this application by generating one random taxon-set of different sizes $S \in \{20, 50, 100, 150, 200\}$ across 500 iterations.
+Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample, sets of randomly drawn taxa should not be significantly enriched compared to the remainder background taxa.
 For this evaluation, we estimated type I error as the fraction of samples where our random set is detected as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.
@@
 
-\item 16S rRNA gene sequencing of stool samples from IBD patients in the pediatric RISK cohort \cite{gevers2014}.
+\item 16S rRNA gene sequencing of stool samples from IBD patients in the RISK cohort \cite{gevers2014}.
 This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada.
@@
 This aspect is captured through computing AUROC values comparing computed enrichment scores and true labels.
-Consistent with prior approaches, adjusting for correlation did not improve performance, where the AUROC values are equivalent to using the Wilcoxon Rank sum statistic at around 0.5.
+Consistent with the previous type I error evaluation, adjusting for correlation did not improve performance, where obtained AUROC were around 0.5 and at the same level as the benchmark Wilcoxon rank sum statistic.
 Unadjusted methods were much better at ranking true enriched sets, however the mean AUROC values are lower than alternate single sample enrichment methods (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}) even though this difference is not significant due to overlapping confidence intervals.
@@
 \subsection*{Disease Prediction}   
-Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:4}), we want to assess whether they can also act as useful inputs to predictive models.
+Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:4}), we wanted to assess whether they can also act as useful inputs to predictive models.
 In this section we assessed the predictive performance of a standard baseline random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (CBEA, ssGSEA, and GSVA).
@@
 
-We fit our model to two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
+We fit our model to two data sets with a similar disease classification task of discriminating patients who were diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
 The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al.
@@
 \subsection*{Inference with CBEA} 
-CBEA is a microbiome-specific approach to generate sample specific enrichment scores for taxonomic sets defined apriori.
+CBEA is a microbiome-specific approach to generate sample specific enrichment scores for taxonomic sets defined \emph{a priori}.
 The formulation of CBEA as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}.
@@
 
-For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}) while also demonstrating respectable power (Fig~\ref{fig:4}).
+For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}) while also demonstrating adequate power (Fig~\ref{fig:4}).
 This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), where the mixture normal displayed superior fit to the null distribution.
@@
 
-We hypothesized that the discrepancy between simulation and real data evaluations can can arise due to differences in our assumptions regarding the data generating process that informed our simulation schema.
-For the non-zero component of each taxa, we sampled from the same negative binomial distribution where designated enriched taxa were generated with inflated means (but the same dispersion).
+We hypothesized that the discrepancy between simulation and real data evaluations could be due to differences in our assumptions regarding the data generating process that informed our simulation schema.
+For the non-zero component of each taxon, we sampled from the same negative binomial distribution where designated enriched taxa were generated with inflated means (but the same dispersion).
 These marginals were simulated to account for block exchangable correlation within the enriched set only.
-This might affect our results in two ways.
-First, our simulation scenario ensures that all designated non-enriched taxa is identical to each other.
+This might have affected our results in two ways.
+First, our simulation scenario ensures that all designated non-enriched taxa are identical to each other.
 This is not the case for real data, where our null scenario involves randomly sampled sets that might by chance all have taxa with inflated means compared to remainder taxa.
-This is represented in \nameref{S7_Fig}, where the distribution of type I error across 500 replications are right skewed for underperforming CBEA variants, indicating that these approaches are much more sensitive compared to the Wilcoxon rank sum test or unadjusted CBEA with mixture normal distribution.
+This is represented in \nameref{S7_Fig}, where the distribution of type I error across 500 replications is right skewed for underperforming CBEA variants, indicating that these approaches are much more sensitive compared to the Wilcoxon rank sum test or unadjusted CBEA with mixture normal distribution.
 Second, as described in the \nameref{introduction} section, we did not consider taxon-specific biases that distort the observed relative abundance of taxa compared to true values \cite{mclaren2019}.
-In the context of sum-based aggregations, the resulting bias of the aggregated taxon are dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}).
-Conceptually, this means that measurement error for a taxon aggregate is different across samples as relative abundance of contributing changes, leading to issues when attempting to perform inference.
+In the context of sum-based aggregations, the resulting bias of the aggregated taxon-set is dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}).
+Conceptually, this means that measurement error for a taxon-set is different across samples as relative abundance of contributing taxa changes, leading to issues when attempting to perform inference.
 As such, we expect methods like corncob or DESeq2 when performed on such sum aggregates in the presence of taxon-specific biases to have inflated type I error compared to our multiplication based approach.
@@
 This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}.
-Results showed that CBEA generate scores perform well across both real data settings and simulation scenarios.
+Results showed that CBEA generate scores perform well across both real data and simulation scenarios.
 Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that CBEA scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
@@
 These above results demonstrate the applicability of CBEA under different data analysis scenarios.
-If researchers are interested in perform inference, they can decide between an unsupervised sample level approach (i.e.
+If researchers are interested in performing inference, they can decide between an unsupervised sample level approach (i.e.
 screen samples for enrichment of certain characteristics) or a supervised population level approach (i.e.
@@
 Researchers might find CBEA useful under situations of high sparsity and inter-taxa correlation, or if the property of a singular covariance matrix (a byproduct of the CLR transformation \cite{gloor2017}) is undesired.
-Even though we only evaluated prediction models in this manuscript, researchers can benchmark their own usage of CBEA for other downstream tasks such as sample ordination.
+Even though we only evaluated prediction models, researchers can benchmark their own usage of CBEA for other downstream tasks such as sample ordination.
 
 However, there are various limitations to our evaluation of CBEA.
-First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome taxonomic data.
+First, our simulation analysis may not capture the appropriate data-generating distributions underlying microbiome taxonomic data.
 There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016a} have been used in the evaluation of prior studies.
@@
 Incorporation of taxa-specific weights (similar to PhILR \cite{silverman2017}) could reduce the influence of outliers, such as rare or highly invariant taxa.
-Finally, even though for a given set of apriori annotations CBEA can generate useful summary scores, such values are limited in their utility if the annotations themselves are not meaningful.
+Finally, even though for a given set of \emph{a priori} annotations CBEA can generate useful summary scores, such values are limited in their utility if the annotations themselves are not meaningful.
 As such, curating and validating sets (similar to MSigDB \cite{subramanian2005}) based on physiological or genomic characteristics of microbes \cite{weissman2021} or their association with human disease (in beta BugSigDB \url{https://bugsigdb.org/Main_Page}) can allow for incorporating functional insights into microbiome-outcome analyses.
@@
 \section*{Conclusion}
-Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomics data sets.
-However, limited work has been done developing set based methods specifically for microbiome relative abundance data.
+Gene set testing, or pathway analysis, is an important tool in the analysis of high-dimensional genomics data sets; however, limited work has been done developing set based methods specifically for microbiome relative abundance data.
 We introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level.
@@
 
-\bibliography{references}{}
+%\bibliography{references}{}
+\begin{thebibliography}{10}
+
+\bibitem{proctor2019}
+Proctor LM, Creasy HH, Fettweis JM, {Lloyd-Price} J, Mahurkar A, Zhou W, et~al.
+\newblock The {{Integrative Human Microbiome Project}}.
+\newblock Nature. 2019;569(7758):641--648.
+\newblock doi:{10.1038/s41586-019-1238-8}.
+
+\bibitem{sharma2019a}
+Sharma S, Tripathi P.
+\newblock Gut Microbiome and Type 2 Diabetes: Where We Are and Where to Go?
+\newblock The Journal of Nutritional Biochemistry. 2019;63:101--108.
+\newblock doi:{10.1016/j.jnutbio.2018.10.003}.
+
+\bibitem{aoun2020}
+Aoun A, Darwish F, Hamod N.
+\newblock The {{Influence}} of the {{Gut Microbiome}} on {{Obesity}} in
+  {{Adults}} and the {{Role}} of {{Probiotics}}, {{Prebiotics}}, and
+  {{Synbiotics}} for {{Weight Loss}}.
+\newblock Preventive Nutrition and Food Science. 2020;25(2):113--123.
+\newblock doi:{10.3746/pnf.2020.25.2.113}.
+
+\bibitem{cho2012}
+Cho I, Blaser MJ.
+\newblock The Human Microbiome: At the Interface of Health and Disease.
+\newblock Nature Reviews Genetics. 2012;13(4):260--270.
+\newblock doi:{10.1038/nrg3182}.
+
+\bibitem{callahan2016}
+Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
+\newblock {{DADA2}}: {{High-resolution}} Sample Inference from {{Illumina}}
+  Amplicon Data.
+\newblock Nature Methods. 2016;13(7):581--583.
+\newblock doi:{10.1038/nmeth.3869}.
+
+\bibitem{truong2015}
+Truong DT, Franzosa EA, Tickle TL, Scholz M, Weingart G, Pasolli E, et~al.
+\newblock {{MetaPhlAn2}} for Enhanced Metagenomic Taxonomic Profiling.
+\newblock Nature Methods. 2015;12(10):902--903.
+\newblock doi:{10.1038/nmeth.3589}.
+
+\bibitem{li2019a}
+Li H.
+\newblock Statistical and {{Computational Methods}} in {{Microbiome}} and
+  {{Metagenomics}}.
+\newblock In: Handbook of {{Statistical Genomics}}. {John Wiley \& Sons, Ltd};
+  2019. p. 977--550.
+
+\bibitem{li2015}
+Li H.
+\newblock Microbiome, {{Metagenomics}}, and {{High-Dimensional Compositional
+  Data Analysis}}.
+\newblock Annual Review of Statistics and Its Application. 2015;2(1):73--94.
+\newblock doi:{10.1146/annurev-statistics-010814-020351}.
+
+\bibitem{gloor2017}
+Gloor GB, Macklaim JM, {Pawlowsky-Glahn} V, Egozcue JJ.
+\newblock Microbiome {{Datasets Are Compositional}}: {{And This Is Not
+  Optional}}.
+\newblock Frontiers in Microbiology. 2017;8.
+\newblock doi:{10.3389/fmicb.2017.02224}.
+
+\bibitem{khatri2012}
+Khatri P, Sirota M, Butte AJ.
+\newblock Ten {{Years}} of {{Pathway Analysis}}: {{Current Approaches}} and
+  {{Outstanding Challenges}}.
+\newblock PLOS Computational Biology. 2012;8(2):e1002375.
+\newblock doi:{10.1371/journal.pcbi.1002375}.
+
+\bibitem{goeman2007}
+Goeman JJ, B{\"u}hlmann P.
+\newblock Analyzing Gene Expression Data in Terms of Gene Sets: Methodological
+  Issues.
+\newblock Bioinformatics. 2007;23(8):980--987.
+\newblock doi:{10.1093/bioinformatics/btm051}.
+
+\bibitem{subramanian2005}
+Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
+\newblock Gene Set Enrichment Analysis: {{A}} Knowledge-Based Approach for
+  Interpreting Genome-Wide Expression Profiles.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(43):15545--15550.
+\newblock doi:{10.1073/pnas.0506580102}.
+
+\bibitem{ashburner2000}
+Ashburner M, Ball CA, Blake JA, Botstein D, Butler H, Cherry JM, et~al.
+\newblock Gene {{Ontology}}: Tool for the Unification of Biology.
+\newblock Nature genetics. 2000;25(1):25--29.
+\newblock doi:{10.1038/75556}.
+
+\bibitem{irizarry2009}
+Irizarry RA, Wang C, Zhou Y, Speed TP.
+\newblock Gene {{Set Enrichment Analysis Made Simple}}.
+\newblock Statistical methods in medical research. 2009;18(6):565--575.
+\newblock doi:{10.1177/0962280209351908}.
+
+\bibitem{hanzelmann2013}
+H{\"a}nzelmann S, Castelo R, Guinney J.
+\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and {{RNA-Seq}}
+  Data.
+\newblock BMC Bioinformatics. 2013;14(1):7.
+\newblock doi:{10.1186/1471-2105-14-7}.
+
+\bibitem{frost2020}
+Frost HR.
+\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
+  Method for Cell-Specific Gene Set Scoring.
+\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
+\newblock doi:{10.1093/nar/gkaa582}.
+
+\bibitem{chong2020}
+Chong J, Liu P, Zhou G, Xia J.
+\newblock Using {{MicrobiomeAnalyst}} for Comprehensive Statistical,
+  Functional, and Meta-Analysis of Microbiome Data.
+\newblock Nature Protocols. 2020;15(3):799--821.
+\newblock doi:{10.1038/s41596-019-0264-1}.
+
+\bibitem{quinn2019}
+Quinn TP, Erb I, Gloor G, Notredame C, Richardson MF, Crowley TM.
+\newblock A Field Guide for the Compositional Analysis of Any-Omics Data.
+\newblock GigaScience. 2019;8(giz107).
+\newblock doi:{10.1093/gigascience/giz107}.
+
+\bibitem{quinn2018}
+Quinn TP, Erb I, Richardson MF, Crowley TM.
+\newblock Understanding Sequencing Data as Compositions: An Outlook and Review.
+\newblock Bioinformatics. 2018;34(16):2870--2878.
+\newblock doi:{10.1093/bioinformatics/bty175}.
+
+\bibitem{morton2019}
+Morton JT, Marotz C, Washburne A, Silverman J, Zaramela LS, Edlund A, et~al.
+\newblock Establishing Microbial Composition Measurement Standards with
+  Reference Frames.
+\newblock Nature Communications. 2019;10(1).
+\newblock doi:{10.1038/s41467-019-10656-5}.
+
+\bibitem{love2014}
+Love MI, Huber W, Anders S.
+\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA-seq}}
+  Data with {{DESeq2}}.
+\newblock Genome Biology. 2014;15(12):550.
+\newblock doi:{10.1186/s13059-014-0550-8}.
+
+\bibitem{weiss2017}
+Weiss S, Xu ZZ, Peddada S, Amir A, Bittinger K, Gonzalez A, et~al.
+\newblock Normalization and Microbial Differential Abundance Strategies Depend
+  upon Data Characteristics.
+\newblock Microbiome. 2017;5(1).
+\newblock doi:{10.1186/s40168-017-0237-y}.
+
+\bibitem{mcknight2019}
+McKnight DT, Huerlimann R, Bower DS, Schwarzkopf L, Alford RA, Zenger KR.
+\newblock Methods for Normalizing Microbiome Data: {{An}} Ecological
+  Perspective.
+\newblock Methods in Ecology and Evolution. 2019;10(3):389--400.
+\newblock doi:{10.1111/2041-210X.13115}.
+
+\bibitem{aitchison1994}
+Aitchison J.
+\newblock Principles of Compositional Data Analysis.
+\newblock Lecture Notes-Monograph Series. 1994; p. 73--81.
+
+\bibitem{mclaren2019}
+McLaren MR, Willis AD, Callahan BJ.
+\newblock Consistent and Correctable Bias in Metagenomic Sequencing
+  Experiments.
+\newblock eLife. 2019;8:e46923.
+\newblock doi:{10.7554/eLife.46923}.
+
+\bibitem{egozcue2005}
+Egozcue JJ, {Pawlowsky-Glahn} V.
+\newblock Groups of {{Parts}} and {{Their Balances}} in {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2005;37(7):795--828.
+\newblock doi:{10.1007/s11004-005-7381-9}.
+
+\bibitem{tian2005}
+Tian L, Greenberg SA, Kong SW, Altschuler J, Kohane IS, Park PJ.
+\newblock Discovering Statistically Significant Pathways in Expression
+  Profiling Studies.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(38):13544--13549.
+\newblock doi:{10.1073/pnas.0506577102}.
+
+\bibitem{rivera-pinto2018}
+{Rivera-Pinto} J, Egozcue JJ, {Pawlowsky-Glahn} V, Paredes R, {Noguera-Julian}
+  M, Calle ML.
+\newblock Balances: A {{New Perspective}} for {{Microbiome Analysis}}.
+\newblock mSystems. 2018;3(4):e00053--18.
+\newblock doi:{10.1128/mSystems.00053-18}.
+
+\bibitem{wu2012}
+Wu D, Smyth GK.
+\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
+  Correlation.
+\newblock Nucleic Acids Research. 2012;40(17):e133.
+\newblock doi:{10.1093/nar/gks461}.
+
+\bibitem{egozcue2003}
+Egozcue JJ, {Pawlowsky-Glahn} V, {Mateu-Figueras} G, {Barcel{\'o}-Vidal} C.
+\newblock Isometric {{Logratio Transformations}} for {{Compositional Data
+  Analysis}}.
+\newblock Mathematical Geology. 2003;35(3):279--300.
+\newblock doi:{10.1023/A:1023818214614}.
+
+\bibitem{quast2013}
+Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, et~al.
+\newblock The {{SILVA}} Ribosomal {{RNA}} Gene Database Project: Improved Data
+  Processing and Web-Based Tools.
+\newblock Nucleic Acids Research. 2013;41(D1):D590--D596.
+\newblock doi:{10.1093/nar/gks1219}.
+
+\bibitem{delignette-muller2015}
+{Delignette-Muller} ML, Dutang C.
+\newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
+\newblock Journal of Statistical Software. 2015;64(4):1--34.
+
+\bibitem{benaglia2009}
+Benaglia T, Chauveau D, Hunter DR, Young D.
+\newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
+\newblock Journal of Statistical Software. 2009;32(6):1--29.
+
+\bibitem{washburne2017}
+Washburne AD, Silverman JD, Leff JW, Bennett DJ, Darcy JL.
+\newblock Phylogenetic Factorization of Compositional Data Yields Lineage-Level
+  Associations in Microbiome Datasets.
+\newblock PeerJ. 2017; p.~26.
+
+\bibitem{silverman2017}
+Silverman JD, Washburne AD, Mukherjee S, David LA.
+\newblock A Phylogenetic Transform Enhances Analysis of Compositional
+  Microbiota Data.
+\newblock eLife. 2017;6:e21887.
+\newblock doi:{10.7554/eLife.21887}.
+
+\bibitem{morton2017}
+Morton JT, Sanders J, Quinn RA, McDonald D, Gonzalez A, {V{\'a}zquez-Baeza} Y,
+  et~al.
+\newblock Balance {{Trees Reveal Microbial Niche Differentiation}}.
+\newblock mSystems. 2017;2(1):e00162--16.
+\newblock doi:{10.1128/mSystems.00162-16}.
+
+\bibitem{aitchison1980}
+Aitchison J, Shen SM.
+\newblock Logistic-Normal Distributions:{{Some}} Properties and Uses.
+\newblock Biometrika. 1980;67(2):261--272.
+\newblock doi:{10.1093/biomet/67.2.261}.
+
+\bibitem{efron2004}
+Efron B.
+\newblock Large-{{Scale Simultaneous Hypothesis Testing}}.
+\newblock Journal of the American Statistical Association.
+  2004;99(465):96--104.
+\newblock doi:{10.1198/016214504000000089}.
+
+\bibitem{kurtz2015a}
+Kurtz ZD, M{\"u}ller CL, Miraldi ER, Littman DR, Blaser MJ, Bonneau RA.
+\newblock Sparse and {{Compositionally Robust Inference}} of {{Microbial
+  Ecological Networks}}.
+\newblock PLOS Computational Biology. 2015;11(5):e1004226.
+\newblock doi:{10.1371/journal.pcbi.1004226}.
+
+\bibitem{ackerman2009}
+Ackermann M, Strimmer K.
+\newblock A General Modular Framework for Gene Set Enrichment Analysis.
+\newblock BMC bioinformatics. 2009;10(1):1--20.
+
+\bibitem{geistlinger2021}
+Geistlinger L, Csaba G, Santarelli M, Ramos M, Schiffer L, Turaga N, et~al.
+\newblock Toward a Gold Standard for Benchmarking Gene Set Enrichment Analysis.
+\newblock Briefings in bioinformatics. 2021;22(1):545--556.
+
+\bibitem{pasolli2017}
+Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong DT, et~al.
+\newblock Accessible, Curated Metagenomic Data through {{ExperimentHub}}.
+\newblock Nature Methods. 2017;14(11):1023--1024.
+\newblock doi:{10.1038/nmeth.4468}.
+
+\bibitem{schiffer2019}
+Schiffer L, Azhar R, Shepherd L, Ramos M, Geistlinger L, Huttenhower C, et~al.
+\newblock {{HMP16SData}}: {{Efficient}} Access to the Human Microbiome Project
+  through Bioconductor.
+\newblock American Journal of Epidemiology. 2019;doi:{10.1093/aje/kwz006}.
+
+\bibitem{gonzalez2018}
+Gonzalez A, {Navas-Molina} JA, Kosciolek T, McDonald D, {V{\'a}zquez-Baeza} Y,
+  Ackermann G, et~al.
+\newblock Qiita: Rapid, Web-Enabled Microbiome Meta-Analysis.
+\newblock Nature Methods. 2018;15(10):796--798.
+\newblock doi:{10.1038/s41592-018-0141-9}.
+
+\bibitem{wickham2019}
+Wickham H, Averick M, Bryan J, Chang W, McGowan LD, Fran{\c c}ois R, et~al.
+\newblock Welcome to the Tidyverse.
+\newblock Journal of Open Source Software. 2019;4(43):1686.
+\newblock doi:{10.21105/joss.01686}.
+
+\bibitem{robin2011}
+Robin X, Turck N, Hainard A, Tiberti N, Lisacek F, Sanchez JC, et~al.
+\newblock {{pROC}}: An Open-Source Package for {{R}} and {{S}}+ to Analyze and
+  Compare {{ROC}} Curves.
+\newblock BMC Bioinformatics. 2011;12:77.
+
+\bibitem{mcmurdie2014}
+McMurdie PJ, Holmes S.
+\newblock Waste {{Not}}, {{Want Not}}: {{Why Rarefying Microbiome Data Is
+  Inadmissible}}.
+\newblock PLOS Computational Biology. 2014;10(4):e1003531.
+\newblock doi:{10.1371/journal.pcbi.1003531}.
+
+\bibitem{ernst2021}
+Ernst FGM, Shetty SA, Borman T, Lahti L. Mia: {{Microbiome}} Analysis; 2021.
+
+\bibitem{landau2021}
+Landau WM.
+\newblock The Targets {{R}} Package: A Dynamic {{Make-like}} Function-Oriented
+  Pipeline Toolkit for Reproducibility and High-Performance Computing.
+\newblock Journal of Open Source Software. 2021;6(57):2959.
+
+\bibitem{consortium2012}
+Consortium THMP, Huttenhower C, Gevers D, Knight R, Abubucker S, Badger JH,
+  et~al.
+\newblock Structure, Function and Diversity of the Healthy Human Microbiome.
+\newblock Nature. 2012;486(7402):207--214.
+\newblock doi:{10.1038/nature11234}.
+
+\bibitem{agresti1998}
+Agresti A, Coull BA.
+\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
+  Estimation}} of {{Binomial Proportions}}.
+\newblock The American Statistician. 1998;52(2):119--126.
+\newblock doi:{10.2307/2685469}.
+
+\bibitem{thurnheer2016}
+Thurnheer T, Bostanci N, Belibasakis GN.
+\newblock Microbial Dynamics during Conversion from Supragingival to
+  Subgingival Biofilms in an in Vitro Model.
+\newblock Molecular Oral Microbiology. 2016;31(2):125--135.
+\newblock doi:{10.1111/omi.12108}.
+
+\bibitem{beghini2019}
+Beghini F, Renson A, Zolnik CP, Geistlinger L, Usyk M, Moody TU, et~al.
+\newblock Tobacco Exposure Associated with Oral Microbiota Oxygen Utilization
+  in the {{New York City Health}} and {{Nutrition Examination Study}}.
+\newblock Annals of Epidemiology. 2019;34:18--25.e3.
+\newblock doi:{10.1016/j.annepidem.2019.03.005}.
+
+\bibitem{matteocalgaro2020}
+Calgaro M. Mcalgaro93/Sc2meta: {{Paper Release}}; 2020.
+\newblock Zenodo.
+
+\bibitem{delong1988}
+DeLong ER, DeLong DM, {Clarke-Pearson} DL.
+\newblock Comparing the Areas under Two or More Correlated Receiver Operating
+  Characteristic Curves: A Nonparametric Approach.
+\newblock Biometrics. 1988;44(3):837--845.
+
+\bibitem{nielsen2014}
+Nielsen HB, Almeida M, Juncker AS, Rasmussen S, Li J, Sunagawa S, et~al.
+\newblock Identification and Assembly of Genomes and Genetic Elements in
+  Complex Metagenomic Samples without Using Reference Genomes.
+\newblock Nature Biotechnology. 2014;32(8):822--828.
+\newblock doi:{10.1038/nbt.2939}.
+
+\bibitem{gevers2014}
+Gevers D, Kugathasan S, Denson LA, {V{\'a}zquez-Baeza} Y, Van~Treuren W, Ren B,
+  et~al.
+\newblock The {{Treatment-Naive Microbiome}} in {{New-Onset Crohn}}'s
+  {{Disease}}.
+\newblock Cell Host \& Microbe. 2014;15(3):382--392.
+\newblock doi:{10.1016/j.chom.2014.02.005}.
+
+\bibitem{breiman2001}
+Breiman L.
+\newblock Random {{Forests}}.
+\newblock Machine Learning. 2001;45(1):5--32.
+\newblock doi:{10.1023/A:1010933404324}.
+
+\bibitem{chawla2002}
+Chawla NV, Bowyer KW, Hall LO, Kegelmeyer WP.
+\newblock {{SMOTE}}: {{Synthetic Minority Over-sampling Technique}}.
+\newblock Journal of Artificial Intelligence Research. 2002;16:321--357.
+\newblock doi:{10.1613/jair.953}.
+
+\bibitem{kuhn2020}
+Kuhn M, Wickham H. Tidymodels: A Collection of Packages for Modeling and
+  Machine Learning Using Tidyverse Principles.; 2020.
+
+\bibitem{martin2020}
+Martin BD, Witten D, Willis AD.
+\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
+  Regression.
+\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
+\newblock doi:{10.1214/19-AOAS1283}.
+
+\bibitem{barbie2009}
+Barbie DA, Tamayo P, Boehm JS, Kim SY, Moody SE, Dunn IF, et~al.
+\newblock Systematic {{RNA}} Interference Reveals That Oncogenic
+  {{{\emph{KRAS}}}}-Driven Cancers Require {{TBK1}}.
+\newblock Nature. 2009;462(7269):108--112.
+\newblock doi:{10.1038/nature08460}.
+
+\bibitem{calgaro2020}
+Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
+\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA-seq}},
+  and Metagenomics Applied to Microbiome Data.
+\newblock Genome Biology. 2020;21(1):191.
+\newblock doi:{10.1186/s13059-020-02104-1}.
+
+\bibitem{wu2016a}
+Wu C, Chen J, Kim J, Pan W.
+\newblock An Adaptive Association Test for Microbiome Data.
+\newblock Genome Medicine. 2016;8(1):56.
+\newblock doi:{10.1186/s13073-016-0302-3}.
+
+\bibitem{morton2021}
+Morton JT, Silverman J, Tikhonov G, L{\"a}hdesm{\"a}ki H, Bonneau R.
+\newblock Scalable Estimation of Microbial Co-Occurrence Networks with
+  {{Variational Autoencoders}}.
+\newblock {Bioinformatics}; 2021.
+
+\bibitem{ma2021a}
+Ma S, Ren B, Mallick H, Moon YS, Schwager E, Maharjan S, et~al.
+\newblock A Statistical Model for Describing and Simulating Microbial Community
+  Profiles.
+\newblock PLOS Computational Biology. 2021;17(9):e1008913.
+\newblock doi:{10.1371/journal.pcbi.1008913}.
+
+\bibitem{naim2012}
+Naim I, Gildea D.
+\newblock Convergence of the {{EM Algorithm}} for {{Gaussian Mixtures}} with
+  {{Unbalanced Mixing Coefficients}}.
+\newblock Proceedings of the 29th International Coference on International
+  Conference on Machine Learning. 2012; p.~8.
+\newblock doi:{10.5555/3042573.3042756}.
+
+\bibitem{martin-fernandez2012}
+{Mart{\'i}n-Fern{\'a}ndez} JA, Hron K, Templ M, Filzmoser P,
+  {Palarea-Albaladejo} J.
+\newblock Model-Based Replacement of Rounded Zeros in Compositional Data:
+  {{Classical}} and Robust Approaches.
+\newblock Computational Statistics \& Data Analysis. 2012;56(9):2688--2704.
+\newblock doi:{10.1016/j.csda.2012.02.012}.
+
+\bibitem{kaul2017a}
+Kaul A, Davidov O, Peddada SD.
+\newblock Structural Zeros in High-Dimensional Data with Applications to
+  Microbiome Studies.
+\newblock Biostatistics. 2017;18(3):422--433.
+\newblock doi:{10.1093/biostatistics/kxw053}.
+
+\bibitem{weissman2021}
+Weissman JL, Dogra S, Javadi K, Bolten S, Flint R, Davati C, et~al.
+\newblock Exploring the Functional Composition of the Human Microbiome Using a
+  Hand-Curated Microbial Trait Database.
+\newblock BMC Bioinformatics. 2021;22(1):306.
+\newblock doi:{10.1186/s12859-021-04216-2}.
+
+\end{thebibliography}
 
@@
 \label{S7_Fig}
-{\bf Distribution of type I error values across all replications in real data random set evaluations for CBEA inference at the sample-level.} Placeholder text.
+{\bf Distribution of type I error values across all replications in real data random set evaluations for CBEA inference at the sample-level.} Density (\emph{y}-axis) for type I error values (\emph{x}-axis) of each evaluated approach for sample-level inference using real data across 500 replications.
+Here, type I error was estimated as the proportion of samples where a randomly sampled set of different sizes where identified to be statistically significant at $p$-value threshold of 0.05.
 

### 2022-02-15 supplemental.tex
Final manuscript

@@
 
-Fig~\ref{fig:s5} shows results for simulation studies as detailed in the \nameref{methods} section.
+Fig~\ref{fig:s5} shows results for simulation studies as detailed in the Methods section.
 Panel A presents results for a regression learning task with a continuous outcome.
@@
 
+\section{Single sample type I error diagnostic plot}
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.9\textwidth]{figures/data_ss_diag_ridges.png}
+    \caption{\textbf{Distribution of type I error values across app relications in real data random set evaluations for CBEA inference at the sample level}.
+Density (\emph{y}-axis) for type I error values (\emph{x}-axis) of each evaluated approach for sample-level inference using real data across 500 replications.
+Here, type I error was estimated as the proportion of samples where a randomly sampled set of different sizes where identified to be statistically significant at $p$-value threshold of 0.05}
+    \label{fig:s7}
+\end{figure}
+
 \clearpage
-\bibliography{references}{}
+%\bibliography{references}{}
+\begin{thebibliography}{10}
+
+\bibitem{delignette-muller2015}
+{Delignette-Muller} ML, Dutang C.
+\newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
+\newblock Journal of Statistical Software. 2015;64(4):1--34.
+
+\bibitem{benaglia2009}
+Benaglia T, Chauveau D, Hunter DR, Young D.
+\newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
+\newblock Journal of Statistical Software. 2009;32(6):1--29.
+
+\bibitem{wu2012}
+Wu D, Smyth GK.
+\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
+  Correlation.
+\newblock Nucleic Acids Research. 2012;40(17):e133.
+\newblock doi:{10.1093/nar/gks461}.
+
+\bibitem{calgaro2020}
+Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
+\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA-seq}},
+  and Metagenomics Applied to Microbiome Data.
+\newblock Genome Biology. 2020;21(1):191.
+\newblock doi:{10.1186/s13059-020-02104-1}.
+
+\bibitem{cario1997}
+Cario MC.
+\newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
+  Distributions}} and {{Correlation Matrix}}; 1997.
+
+\bibitem{agresti1998}
+Agresti A, Coull BA.
+\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
+  Estimation}} of {{Binomial Proportions}}.
+\newblock The American Statistician. 1998;52(2):119--126.
+\newblock doi:{10.2307/2685469}.
+
+\bibitem{frost2020}
+Frost HR.
+\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
+  Method for Cell-Specific Gene Set Scoring.
+\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
+\newblock doi:{10.1093/nar/gkaa582}.
+
+\bibitem{delong1988}
+DeLong ER, DeLong DM, {Clarke-Pearson} DL.
+\newblock Comparing the Areas under Two or More Correlated Receiver Operating
+  Characteristic Curves: A Nonparametric Approach.
+\newblock Biometrics. 1988;44(3):837--845.
+
+\bibitem{hawinkel2019}
+Hawinkel S, Mattiello F, Bijnens L, Thas O.
+\newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
+  Control the False Discovery Rate.
+\newblock Briefings in Bioinformatics. 2019;20(1):210--221.
+\newblock doi:{10.1093/bib/bbx104}.
+
+\bibitem{breiman2001}
+Breiman L.
+\newblock Random {{Forests}}.
+\newblock Machine Learning. 2001;45(1):5--32.
+\newblock doi:{10.1023/A:1010933404324}.
+
+\bibitem{kuhn2020}
+Kuhn M, Wickham H. Tidymodels: A Collection of Packages for Modeling and
+  Machine Learning Using Tidyverse Principles.; 2020.
+
+\bibitem{xiao2018}
+Xiao J, Chen L, Yu Y, Zhang X, Chen J.
+\newblock A {{Phylogeny-Regularized Sparse Regression Model}} for {{Predictive
+  Modeling}} of {{Microbial Community Data}}.
+\newblock Frontiers in Microbiology. 2018;9.
+\newblock doi:{10.3389/fmicb.2018.03112}.
+
+\bibitem{dong2020}
+Dong M, Li L, Chen M, Kusalik A, Xu W.
+\newblock Predictive Analysis Methods for Human Microbiome Data with
+  Application to {{Parkinson}}'s Disease.
+\newblock PLOS ONE. 2020;15(8):e0237779.
+\newblock doi:{10.1371/journal.pone.0237779}.
+
+\bibitem{hanzelmann2013}
+H{\"a}nzelmann S, Castelo R, Guinney J.
+\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and {{RNA-Seq}}
+  Data.
+\newblock BMC Bioinformatics. 2013;14(1):7.
+\newblock doi:{10.1186/1471-2105-14-7}.
+
+\bibitem{love2014}
+Love MI, Huber W, Anders S.
+\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA-seq}}
+  Data with {{DESeq2}}.
+\newblock Genome Biology. 2014;15(12):550.
+\newblock doi:{10.1186/s13059-014-0550-8}.
+
+\bibitem{martin2020}
+Martin BD, Witten D, Willis AD.
+\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
+  Regression.
+\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
+\newblock doi:{10.1214/19-AOAS1283}.
+
+\end{thebibliography}
+
 \bibliographystyle{plos2015}

### 2022-02-16 cbea_revised.tex
Updated funding

@@
 
-%\begin{figure}[!ht]
-%    \centering
-%    \includegraphics[width=\textwidth]{figures/sim_pred_rsq.png}
-%    \caption{Predictive R-squared of a random forest model for a continuous outcome trained on CBEA,     %ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data %across sparsity levels, correlation, and signal-to-noise ratio.
-Panel \textbf{(A)} and \textbf{(B)} %represent results across different levels of model saturation (proportion of sets associated with the %outcome).
-CBEA approaches outperformed GSVA and ssGSEA but not standard CLR.}
-%    \label{fig:sim_pred_rsq}
-%\end{figure}
-
 We fit our model to two data sets with a similar disease classification task of discriminating patients who were diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
@@
 The authors thank Becky Lebeaux, Modupe Coker, Erika Dade, Jie Zhou, and Weston Viles for insightful comments and suggestions that greatly improved the paper.
-This work is supported by the National Institutes of Health (grants NCI R21CA253408, NIGMS P20GM130454, NCI P30CA023108, and NLM R01LM012723).
+Funding for HRF provided by National Institutes of Health grants R21CA253408, P20GM130454 and P30CA023108.
+Funding for AGH provided by National Institutes of Health grant R01LM012723.
+Funding for QPN provide by National Institutes of Health grants K01LM012426,
+R21CA253408, P20GM130454, and R01LM012723. The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
 

### 2022-02-17 cbea_revised.tex
Final final

@@
 The authors thank Becky Lebeaux, Modupe Coker, Erika Dade, Jie Zhou, and Weston Viles for insightful comments and suggestions that greatly improved the paper.
-Funding for HRF provided by National Institutes of Health grants R21CA253408, P20GM130454 and P30CA023108.
-Funding for AGH provided by National Institutes of Health grant R01LM012723.
-Funding for QPN provide by National Institutes of Health grants K01LM012426,
-R21CA253408, P20GM130454, and R01LM012723. The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
 

### 2022-04-14 cbea_revised_v3.tex
Final version

--- 2022-02-17 cbea_revised.tex
@@
 \bigskip
-\textbf{1} Department of Biomedical Data Science, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA
+\textbf{1} Department of Biomedical Data Science, Geisel School of Medicine at Dartmouth College, Hanover, New Hampshire, United States
 \\
-\textbf{2} Department of Epidemiology, Geisel School of Medicine at Dartmouth College, Hanover, NH, USA
+\textbf{2} Department of Epidemiology, Geisel School of Medicine at Dartmouth College, Hanover, New Hampshire, United States
 \\
@@
 Finally, we discuss our results and the limitations of our method.
-An R package implementation of CBEA can be found on GitHub (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/CBEA}).
+An R package implementation of CBEA can be installed via Bioconductor.
+The development version can be found on GitHub (www.github.com/qpmnguyen/CBEA).
 
@@
 The CBEA method generates sample-specific enrichment scores for microbial sets using products of proportions \cite{egozcue2003}.
-Details on the computational implementation of CBEA can be found in the Supplementary Materials.
+Details on the computational implementation of CBEA can be found in \nameref{S1_File}.
 The CBEA method takes two inputs:
@@
     \centering
-    \includegraphics[width=\linewidth]{figures/kurtosis_skewness_gof.png}
+    %\includegraphics[width=\linewidth]{figures/kurtosis_skewness_gof.png}
     \caption{{\bf Properties of the null distribution of CBEA under the global null simulations}.
@@
 As such, this procedure assumes that the variance of the test statistic under the alternate hypothesis is the same as that of the null.
-Details of the computational implementation to this estimation process can be found in the Supplementary Materials.
+Details of the computational implementation to this estimation process can be found in \nameref{S1_File}.
 
@@
 In this section, we present results for evaluating statistical significance, phenotype relevance, and predictive performance.
-In addition to real data, we also evaluated models based on parametric simulations, where results can be found in the Supplemental Materials.
+In addition to real data, we also evaluated models based on parametric simulations, where results can be found in the Supplemental Materials (\nameref{S1_Fig}, \nameref{S2_Fig}, \nameref{S3_Fig}, \nameref{S4_Fig}, and \nameref{S5_Fig}).
 
@@
     \centering
-    \includegraphics[width = \textwidth]{figures/data_ss_fdr_new.png}
+    %\includegraphics[width = \textwidth]{figures/data_ss_fdr_new.png}
     \caption{Random taxa set analyses for inference at the sample level of CBEA under different parametric assumptions compared against a Wilcoxon rank-sum test.
@@
     \centering
-    \includegraphics[width = \textwidth]{figures/data_diffab_fdr.eps}
+    %\includegraphics[width = \textwidth]{figures/data_diffab_fdr.eps}
     \caption{Random sample label (\textbf{A}) and random set (\textbf{B}) analyses for population level inference.
@@
     \centering
-    \includegraphics[width = \textwidth]{figures/data_ss_pwr_new.png}
+    %\includegraphics[width = \textwidth]{figures/data_ss_pwr_new.png}
     \caption{Statistical power (\textbf{A}) and score rankings (\textbf{B}) to assess phenotype relevance.
@@
     \centering
-    \includegraphics[width = \textwidth]{figures/data_diffab_pwr.png}
+    %\includegraphics[width = \textwidth]{figures/data_diffab_pwr.png}
     \caption{Statistical power to assess phenotype relevance of inference tasks at the population level.
@@
     \centering
-    \includegraphics[width = \linewidth]{figures/data_pred.png}
+    %\includegraphics[width = \linewidth]{figures/data_pred.png}
     \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.

### 2022-04-15 cbea_revised_v3.tex
Fixing typos and grammar and some minor phrasing

@@
 \section*{Abstract}
-Research in human associated microbiomes often involves the analysis of taxonomic count tables generated via high-throughput sequencing.
+Research in human-associated microbiomes often involves the analysis of taxonomic count tables generated via high-throughput sequencing.
 It is difficult to apply statistical tools as the data is high-dimensional, sparse, and compositional.
 An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets.
-Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis.
+Set-based analysis is ubiquitous in the genomics literature and has demonstrable impact on improving interpretability and power of downstream analysis.
 Unfortunately, there is a lack of sophisticated set-based analysis methods specific to microbiome taxonomic data, where current practice often employs abundance summation as a technique for aggregation.
 This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias.
-Here, we attempt to fill this gap with a new single sample taxon enrichment method that uses a novel log-ratio formulation based on the competitive null hypothesis commonly used in the enrichment analysis literature.
-Our approach, titled competitive balances for taxonomic enrichment analysis (CBEA), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
+Here, we attempt to fill this gap with a new single-sample taxon enrichment method that uses a novel log-ratio formulation based on the competitive null hypothesis commonly used in the enrichment analysis literature.
+Our approach, titled competitive balances for taxonomic enrichment analysis (CBEA), generates sample-specific enrichment scores as the scaled log-ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement.
 We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values.
-Herein we demonstrate using both real data applications and simulations that CBEA controls for type I error even under high sparsity and high inter-taxa correlation scenarios.
-Additionally, it provides informative scores that can be inputs to downstream analyses such as prediction tasks.
+Herein, we demonstrate, using both real data applications and simulations, that CBEA controls for type I error, even under high sparsity and high inter-taxa correlation scenarios.
+Additionally, CBEA provides informative scores that can be inputs to downstream analyses such as prediction tasks.
 
@@
 \section*{Author summary}
-The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing.
-However, microbiome taxonomic data is sparse and high dimensional which prevents the application of standard statistical techniques.
+The study of human-associated microbiomes relies on genomic surveys via high-throughput sequencing.
+However, microbiome taxonomic data is sparse and high-dimensional which prevents the application of standard statistical techniques.
 One approach to address this problem is to perform analyses at the level of taxon sets.
-Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability.
+Set-based analysis has a long history in the genomics literature, with demonstrable impact on improving both power and interpretability.
 Unfortunately, there is limited interest in developing new set-based tools tailored for microbiome taxonomic data given its unique features compared to other 'omics data types.
 We developed a new tool to generate taxon set enrichment scores at the sample level through a novel log-ratio formulation based on the competitive null hypothesis.
-Our scores can be used for statistical inference at both the sample and population levels, as well as inputs to other downstream analyses such as prediction models.
+Our scores can be used for statistical inference at both the sample and population levels, and as inputs to other downstream analyses such as prediction models.
 We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.
@@
 \section*{Introduction} \label{introduction}
-The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host.
+The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exists with its host.
 Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019a}, and obesity \cite{aoun2020}.
@@
 whole-genome shotgun sequencing) \cite{cho2012}.
-Raw sequencing data is then processed through a variety of bioinformatic pipelines \cite{callahan2016, truong2015}, yielding various data products, one of which are taxonomic tables which can be used to study associations between members of the microbiome and an exposure or outcome of interest.
+Raw sequencing data is then processed through a variety of bioinformatic pipelines \cite{callahan2016, truong2015}, yielding various data products, including taxonomic tables which can be used to study associations between members of the microbiome and an exposure or outcome of interest.
 
@@
 As such, gene set testing methods have been shown to increase power and reproducibility of genomic analyses.
-Furthermore, through the usage of functionally informative sets defined \emph{a priori} based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analysis also allows for more biologically informative interpretations.
-
-A diverse set of methods have already been developed in this field.
+Furthermore, through the usage of functionally informative sets defined \emph{a priori} based on historical experiments (for example, MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analysis also allows for more biologically informative interpretations.
+
+A diverse set of methods has already been developed in this field.
 Traditional methods utilize the hypergeometric distribution to test for the overrepresentation of a gene set using a candidate list of genes screened based on a marginal model \cite{goeman2007}.
@@
 
-%First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present.
-For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016a}, allowing for simultaneous model fitting and variable selection.
-For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018c} to address hypothesis testing burden.
-
-%Second, the total number of reads varies across samples due to limitations of the sequencing instrument \cite{quinn2018}.
-
-%Third, the data are highly sparse, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment).
-Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015a}, or applying zero-inflated models \cite{chen2018c, kaul2017}.
-Newer methods developed recently have focused on understanding the different types of zeros in the data, providing more sophisticated heuristics around when pseudocounts can be utilized \cite{kaul2017a}.
-
 In microbiome research, even when no explicit enrichment analysis is performed, researchers often aggregate taxa to higher Linnean classification levels such as genus, family, or phylum.
@@
 Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}.
-However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used and p-value threshold.
+However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used and the p-value threshold.
 One of the primary challenges for adapting gene set analysis to the microbiome context is the compositional nature of the data.
 Sequencing technologies constrain the total number of reads, and samples are expected to have the same number of reads instead of DNA content \cite{quinn2019, quinn2018}.
-However, different samples still yield arbitrarily different total read counts \cite{morton2019, gloor2017}, suggesting the use of normalization methods to allow for proper comparison of feature abundances across samples.
+However, different samples still yield arbitrarily different total read counts \cite{morton2019, gloor2017}, which suggests the usage of normalization methods to allow for proper comparison of feature abundances across samples.
 However, microbiome data sets do not follow certain assumptions that enable the cross-application of methods from similar fields (such as RNA-seq) \cite{quinn2019, quinn2018}.
-For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes with constant expression levels across samples.
+For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes act as housekeeping genes with constant expression levels across samples.
 As such, practitioners often rely on total sum normalization to transform count data into relative proportions that sum to one \cite{weiss2017}.
-Some studies have provided emprical performance vealuations supporting this normalization schema \cite{mcknight2019}.
+Some studies have provided emprical performance evaluations supporting this normalization schema \cite{mcknight2019}.
 Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxon can only be interpreted relative to another.
@@
 
-Unfortunately, the standard practice for aggregating variables using element-wise summations (referred to as amalgamations in the CoDA literature), does not adequately address the compositional issue \cite{mclaren2019}.
+Unfortunately, the standard practice for aggregating variables using element-wise summations (referred to as ``amalgamations" in the CoDA literature), does not adequately address the compositional issue \cite{mclaren2019}.
 First, inter-sample Aitchison distances computed on original parts are not preserved after amalgamation \cite{egozcue2005}.
-This means that cluster analyses might show different results depending on the level of amalgamation and differs from the those computed from original variables.
+This means that cluster analyses might show different results depending on the level of amalgamation and differ from the those computed from original variables.
 Second, amalgamations do not allow for comparison between sets of different sizes within the same experimental condition since larger sets will have larger means and variances.
-Third, considering that each taxa has specific measurement biases \cite{mclaren2019}, an amalgamation based approach would make the bias of the amalgamated variable dependent on the relative abundance of the its constituents.
-In other words, if taxon $1$ has abundance $A_1$ and bias $B_1$, while taxon $2$ has abundance $A_2$ and bias $B_2$, then the bias of the aggregate variable (for example, a genera) is $(A_1B_1 + A_2B_2)/(A_1 + A_2)$ (see Appendix 1.
+Third, considering that each taxa has specific measurement biases \cite{mclaren2019}, an amalgamation-based approach would make the bias of the amalgamated variable dependent on the relative abundance of its constituents.
+In other words, if taxon $1$ has abundance $A_1$ and bias $B_1$, while taxon $2$ has abundance $A_2$ and bias $B_2$, then the bias of the aggregate variable (for example, a genus) is $(A_1B_1 + A_2B_2)/(A_1 + A_2)$ (see Appendix 1.
 from McLaren et al.
 \cite{mclaren2019}).
-This means that bias invariant approaches (such as analyses of ratios) would no longer be invariant when applied to amalgamated variables as bias now varies across samples.
+This means that bias invariant approaches (such as analyses of ratios) would no longer be invariant when applied to amalgamated variables, as bias now varies across samples.
 The alternative would be to multiply the proportions rather than to sum them \cite{egozcue2005}.
@@
 
-In the following sections, we present our approach titled competitive balances for taxonomic erichment analysis (CBEA).
+In the following sections, we present our approach, titled competitive balances for taxonomic erichment analysis (CBEA).
 First, we present the step-by-step formulation of CBEA and discuss its statistical properties.
-Second, we detail our evaluation strategy using both real data and parametric simulations, and the methods we are comparing against.
+Second, we detail our evaluation strategy using both real data and parametric simulations and the methods we are comparing.
 Third, we present results on enrichment testing using CBEA for single samples as well as at the population level.
@@
     \item $\mathbf{X}$: $n$ by $p$ matrix of positive proportions for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing.
-Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing.
-CBEA does not accept $X$ matrices with zeroes since it invalidates the log-ratio transformation.
-Users can generate a dense matrix $X$ using a method of choice, however by default mode CBEA will add a pseudocount of $10^{-5}$ if zeroes are detected in the matrix.
+Usually, $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing.
+CBEA does not accept $X$ matrices with zeroes since they invalidate the log-ratio transformation.
+Users can generate a dense matrix $X$ using their method of choice, however, by default CBEA will add a pseudocount of $10^{-5}$ if zeroes are detected in the matrix.
     \item $\mathbf{A}$: $p$ by $m$ indicator matrix annotating the membership of each taxon $p$ to $m$ sets of interest.
@@
 \begin{enumerate}
-    \item \textbf{Compute the CBEA statistic}: Let $\mathbf{M}$ be a $n$ by $m$ matrix of CBEA scores. Let $\mathbf{M}_{i,k}$ be CBEA score for set $k$ and sample $i$:   
+    \item \textbf{Compute the CBEA statistic}: Let $\mathbf{M}$ be a $n$ by $m$ matrix of CBEA scores. Let $\mathbf{M}_{i,k}$ be the CBEA score for set $k$ and sample $i$:   
     \begin{equation}\label{main_eq}
@@
     \end{equation}
-    where $g(.)$ is the geometric mean. This represents the ratio of the geometric mean of the relative abundance of taxa assigned to set $k$ and remainder taxa. 
-    \item \textbf{Estimate the empirical null distribution} Enrichment scores represent the test statistic for the $Q_1$ null hypothesis $H_o$ that relative abundances in $\mathbf{X}$ of members of set $k$ are not enriched compared to those not in set $k$.
-Since the distribution of CBEA under the null vary depending on data characteristics (Fig~\ref{fig:1}), an empirical null distribution will be estimated from data.
+    where $g(.)$ is the geometric mean. This represents the ratio of the geometric mean of the relative abundance of taxa assigned to set $k$ and the remainder taxa. 
+    \item \textbf{Estimate the empirical null distribution}: Enrichment scores represent the test statistic for the $Q_1$ null hypothesis $H_o$ that relative abundances in $\mathbf{X}$ of members of set $k$ are not enriched compared to those not in set $k$.
+Since the distribution of CBEA under the null varies depending on data characteristics (Fig~\ref{fig:1}), an empirical null distribution will be estimated from data.
     \begin{itemize}
@@
 Let $\mathbf{X}_{perm}$ be the column permuted relative abundance matrix, and $\mathbf{M}_{perm}$ be the corresponding CBEA scores generated from $\mathbf{X}_{perm}$.
-Similarly, we have $\mathbf{M}_{unperm}$ be CBEA scores generated from $\mathbf{X}$.
-        \item \textbf{Estimate correlation-adjusted empirical distribution for each set}.
+Similarly, we let $\mathbf{M}_{unperm}$ be CBEA scores generated from $\mathbf{X}$.
+        \item \textbf{Estimate the correlation-adjusted empirical distribution for each set}.
 For each set, a fit a parametric distribution to both $\mathbf{M}_{perm}$ and $\mathbf{M}_{unperm}$.
@@
     \end{itemize}
-    \item \textbf{Calculate finalized CBEA scores with respect to the empirical null}.
-Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution.
+    \item \textbf{Calculate finalized CBEA scores with respect to the empirical null}: Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution.
 Raw p-values can be calculated by subtracting $\mathbf{E}$ from 1.
@@
 The CBEA statistic is based on the multiplication-based aggregation approach used to calculate balances between groups of parts \cite{egozcue2005}.
-These balances are computed using the isometric log ratio (ILR) transformation \cite{egozcue2003} formula.
+These balances are computed using the isometric log-ratio (ILR) transformation \cite{egozcue2003} formula.
 For a given balance $i$ splitting variables across sets $R$ and $S$, we have the balance coordinate $x^{*}_{i}$ as:
@@
 
-CBEA belongs to a set of methods that seek to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto2018, silverman2017, morton2017}.
+CBEA belongs to a set of methods that seeks to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto2018, silverman2017, morton2017}.
 Unlike methods such as PhILR \cite{silverman2017}, CBEA does not present an orthonormal basis for the complete ILR transformation (such as a a sequential binary partition) \cite{egozcue2003}.
@@
 We can assume that the CBEA statistic, similar to other log-ratio based transforms, follows a normal distribution \cite{egozcue2003, aitchison1980}.
-However, when applying CBEA for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses.
-Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic directly (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies.
+However, when applying CBEA for hypothesis testing at the sample level, it is expected that the researcher will be testing a large number of hypotheses.
+Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic directly (termed the ``empirical null distribution") is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies.
 As such, to perform significance testing using CBEA, we also estimated the null distribution from observed raw CBEA variables.
@@
 This assumption is also supported by preliminary simulation studies (detailed below).
-We simulated microbiome taxonomic count data under the global null across different data features and compute raw CBEA scores and compute kurtosis and skewness in Fig~\ref{fig:1}A.
-We found that the characteristics of the null change depending on sparsity and inter-taxa correlation.
-Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis).
+We simulated microbiome taxonomic count data under the global null across different data features and compute raw CBEA scores as well as kurtosis and skewness of the distribution under the null in Fig~\ref{fig:1}A.
+We found that the characteristics of the distribution change depending on sparsity and inter-taxa correlation.
+Sparsity seems to drive it to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis) behaviour.
 The effect is most dramatic under both high inter-taxa correlation and sparsity.
@@
 To address this issue, we also evaluated a two-component normal mixture distribution.
-The goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on CBEA scores in simulation scenarios under the global null is shown in Fig~\ref{fig:1}B.
-We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.
+Fig~\ref{fig:1}B showed the goodness of fit each distribution form using the Kolmogorov-Smirnov (KS) test statistic comparing each fitted distribution to CBEA scores generated under global null simulation scenarios.
+We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution (across both sparsity and correlation settings).
 
@@
 When taxa within a set are highly correlated, the variance of the sample mean of taxon-wise statistics is inflated.
-Without loss of generalizability, for a set of taxa with taxon-specific statistics $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:
+Without loss of generalizability, for a set of taxa with taxon-specific statistics $x_1, ..., x_p$, we have the variance of the mean $\bar{x}$ to be:
 \begin{equation} \label{vareq:1}
@@
 \end{equation}
-where $p$ is the overall number of taxa, $j$ is the index of a taxa and $K$ is the set of indices of taxa in set $k$.
+where $p$ is the overall number of taxa, $j$ is the index of a taxa, and $K$ is the set of indices of taxa in set $k$.
 The CBEA statistic then looks similar to a t-statistic for difference in means of log-transformed proportions.
 As such, the pooled variance of CBEA is dependent on the variance inflation of both mean components $\overbar{\log{X_{i,j|j \in K}}}$ and $\overbar{\log{X_{i,j|j \notin K}}}$.
-The result of this variance inflation is inflated type I error since highly correlated sets are also detected as significantly enriched.
+The result of this variance inflation is an inflated type I error since highly correlated sets are also detected as significantly enriched.
 
 However, as Wu et al.
-\cite{wu2012} showed, performing column permutation to estimate the null distribution of a competitive test statistic doesn't allow for adequate capture of this variance inflation factor since the permutation procedure disrupts the natural correlation structure of the original variables.
+\cite{wu2012} showed, performing column permutations to estimate the null distribution of a competitive test statistic does not allow for adequate capture of this variance inflation factor since the permutation procedure disrupts the natural correlation structure of the original variables.
 It is important to address this problem since there is strong inter-taxa correlation within the microbiome \cite{kurtz2015a}.
-Our strategy for addressing this issue is to use the location (or mean) estimate from the column permuted raw score matrix with the spread (or variance) estimate taken from the original un-permuted scores.
+Our strategy for addressing this issue is to use the location (or mean) estimate from the column permuted raw score matrix, with the spread (or variance) estimate taken from the original un-permuted scores.
 This still allows us to leverage the null distribution generated via column permutation while using the proper variance estimate taken from scores where the correlation structure has not been disrupted.
@@
 However, set-based analysis is an exploratory approach that can help generate functionally informative hypotheses, and as such users might not want strict type I error control in favor of higher power.
-This is especially true for competitive hypotheses, where its stricter formulation compared to the self-contained approach implies that the test naturally has lower power \cite{goeman2007, ackerman2009}.
+This is especially true for competitive hypotheses where its stricter formulation compared to the self-contained approach implies that the test naturally has lower power \cite{goeman2007, ackerman2009}.
 Furthermore, sets that are highly correlated compared to background can be biologically relevant.
@@
 All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.
-All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).
+All code and data sets used for evaluation of this method are publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).
 Additional packages used to support this analysis includes: \emph{tidyverse} \cite{wickham2019}, \emph{pROC} \cite{robin2011}, \emph{phyloseq} \cite{mcmurdie2014}, \emph{mia} \cite{ernst2021}, \emph{targets} \cite{landau2021}.
@@
 \subsection*{Statistical significance}
-We evaluate the inference procedure of CBEA compared to alternate methods using two approaches: randomly sampled taxa sets and sample label permutation.
+We evaluated the inference procedure of CBEA compared to alternate methods using two approaches: randomly sampled taxa sets and sample label permutation.
 These analyses were performed on the 16S rRNA gene sequencing of the oral microbiome from the Human Microbiome Project \cite{consortium2012, proctor2019}.
@@
 We evaluated this application by generating one random taxon-set of different sizes $S \in \{20, 50, 100, 150, 200\}$ across 500 iterations.
-Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample, sets of randomly drawn taxa should not be significantly enriched compared to the remainder background taxa.
-For this evaluation, we estimated type I error as the fraction of samples where our random set is detected as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.
+Random sets can act as our estimate for type I error since they match the CBEA null hypothesis stated in \nameref{methods}, where we expect that within each sample sets of randomly drawn taxa should not be significantly enriched compared to the remainder background taxa.
+For this evaluation, we estimated type I error as the fraction of samples in which our random set is detected as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.
 Additionally, this analysis also tests whether CBEA is sensitive to different set sizes.
@@
 We can perform enrichment testing at the population level by generating corresponding sample level CBEA scores and performing a two-sample test such as Welch's t-test.
-In order to evaluate CBEA under this context, we generated CBEA scores of sets representing genus-level annotation in above gingival data set \cite{consortium2012, proctor2019} and applied a t-test to test for enrichment (similar to GSVA \cite{hanzelmann2013}) across a randomly generated variable indicating case/control status (repeated 500 times).
+In order to evaluate CBEA under this context, we generated CBEA scores of sets representing genus-level annotation in the above gingival data set \cite{consortium2012, proctor2019} and applied a t-test to test for enrichment (similar to GSVA \cite{hanzelmann2013}) across a randomly generated variable indicating case/control status (repeated 500 times).
 Type I error is estimated as the fraction of sets per iteration found to be significantly enriched with confidence bands computed from the standard error across all iterations.
-In addition, we also performed a random set analysis assessment, where we generated 100 sets of different set sizes $S \in \{20, 50, 100, 150, 200\}$ and evaluated the fraction of genera that were found to be differentially abundant across the original labels (supragingival versus subgingival subsite).
+In addition, we performed a random set analysis assessment, where we generated 100 sets of different set sizes $S \in \{20, 50, 100, 150, 200\}$ and evaluated the fraction of genera that were found to be differentially abundant across the original labels (supragingival versus subgingival subsite).
 95\% confidence intervals were computed using the Agresti-Couli approach \cite{agresti1998}.
@@
 This data set was chosen because its clear biological interpretation can serve as the ground truth.
-Specifically, we expect aerobic microbes to be enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}.
+Specifically, we expect aerobic microbes to be enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes should thrive in the subgingival site \cite{thurnheer2016}.
 Genus-level annotations for microbial metabolism from Beghini et al.
@@
 \cite{matteocalgaro2020}.
-For sample-level inference, we assessed power as the fraction of supragingival samples where aerobic microbes are significantly enriched.
-For population-level inference, power is the fraction of sets representing genus level taxonomic assignments that were significant across subsite labels.
-
-In addition to statistical power, we also assessed phenotype relevance through evaluating whether highly ranked sets based on CBEA scores are more likely to be enriched according to the ground truth.
+For sample-level inference, we assessed power as the fraction of supragingival samples whereby aerobic microbes are significantly enriched.
+For population-level inference, power is the fraction of sets representing genus-level taxonomic assignments that were significant across subsite labels.
+
+In addition to statistical power, we also assessed phenotype relevance through evaluating whether highly ranked sets based on CBEA scores were more likely to be enriched according to the ground truth.
 This is represented by the area under the receiving operator curve (AUROC/AUC) scores computed on CBEA scores against true labels (similar approach was used to evaluate VAM \cite{frost2020}).
@@
 \item Whole genome sequencing of stool samples from inflammatory bowel disease (IBD) patients in the MetaHIT consortium \cite{nielsen2014}.
-This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease).
-We processed this data by removing all samples with less than 1,000 total read counts as well as any OTU that was present (with non-zero proportions) in 10\% of the samples or less.
-Prior to model fitting, we back-transformed relative abundances into count data (to align the format with our 16S rRNA gene sequencing data set) using the provided total number of reads aligned to MetaPhlan marker genes (per sample).
+This data set contains 396 samples from a cohort of European adults, of whom 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease).
+We processed this data by removing all samples with less than 1,000 total read counts as well as any OTU that was present (with non-zero proportions) in 10\% of the samples or fewer.
+Prior to model fitting, we back-transformed relative abundances into count data (to align the format with our 16S rRNA gene sequencing data set), using the provided total number of reads aligned to MetaPhlan marker genes (per sample).
 
@@
 Of the 671 samples obtained, 500 samples belong to patients with IBD.
-We processed this data set by removing all samples with less than 1,000 total read counts as well as any OTU that was present (at least 1 count) in 10\% of the samples or less.
+We processed this data set by removing all samples with less than 1,000 total read counts as well as any OTU that was present (at least 1 count) in 10\% of the samples or fewer.
 \end{enumerate}
 
-We evaluate disease prediction performance by fitting a random forest model \cite{breiman2001} using as inputs CBEA scores to classify samples of patients with IBD and healthy controls.
+We evaluate disease prediction performance by fitting a random forest model \cite{breiman2001} using CBEA scores as inputs to classify samples of patients with IBD or as healthy controls.
 Random forest was chosen as a baseline learner due to its flexibility as an out-of-the-box model that is easy to fit.
-In this instance we evaluated predictive performance of a default random forest model (without hyperparameter tuning) AUROC after 10-fold cross validation.
+In this instance, we evaluated predictive performance of a default random forest model (without hyperparameter tuning) AUROC after 10-fold cross validation.
 Additionally, we utilized SMOTE to correct for class imbalances \cite{chawla2002}.
@@
 We benchmarked the statistical properties of CBEA against existing baseline approaches.
-For sample-level inference analyses, utilized the Wilcoxon rank-sum test, which non-parametrically tests the difference in mean counts between taxa from a pre-defined set and its remainder similar to CBEA.
-For assessments at the population level, we compared CBEA against performing a standard test for differential abundance with set-level features generated via element-wise summations instead.
-We chose DESeq2 \cite{love2014} and corncob \cite{martin2020} because they represent both methods extrapolated from RNA-seq \cite{mcmurdie2014} and those developed specifically for microbiome data.
-
-Since disease prediction models and rankings-based phenotype relevance analyses seek to evaluate the informativeness of CBEA scores instead of relying on computing p-values, we compared performance against other single sample based approaches from the gene set testing literature, specifically ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}.
-Additionally, for evaluating prediction, we also compared performance against a standard analysis plan where inputs are count-aggregated sets with the centered log-ratio (CLR) transformation.
+For sample-level inference analyses, we utilized the Wilcoxon rank-sum test, which non-parametrically tests the difference in mean counts between taxa from a pre-defined set and its remainder similar to CBEA.
+For assessments at the population level, we compared CBEA against the performance of a standard test for differential abundance with set-level features generated via element-wise summations instead.
+We chose DESeq2 \cite{love2014} and corncob \cite{martin2020} because they represent methods extrapolated from RNA-seq \cite{mcmurdie2014} and those developed specifically for microbiome data.
+
+Since disease prediction models and rankings-based phenotype relevance analyses seek to evaluate the informativeness of CBEA scores instead of relying on computing p-values, we compared performance against other single-sample-based approaches from the gene set testing literature, specifically ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}.
+Additionally, for evaluating predictions, we also compared performance against a standard analysis plan where inputs are count-aggregated sets with the centered log-ratio (CLR) transformation.
 
@@
 In this section, we present results for evaluating statistical significance, phenotype relevance, and predictive performance.
-In addition to real data, we also evaluated models based on parametric simulations, where results can be found in the Supplemental Materials (\nameref{S1_Fig}, \nameref{S2_Fig}, \nameref{S3_Fig}, \nameref{S4_Fig}, and \nameref{S5_Fig}).
+In addition to real data, we also evaluated models based on parametric simulations.
+Results can be found in the Supplemental Materials (\nameref{S1_Fig}, \nameref{S2_Fig}, \nameref{S3_Fig}, \nameref{S4_Fig}, and \nameref{S5_Fig}).
 
@@
 CBEA provides significance testing at the sample level through a self-contained competitive null hypothesis.
-Generating random sets approximate the global null setting where within each sample, sets generated by randomly sampling taxa should not be significantly more enriched than remainder taxa.
+Generating random sets approximates the global null setting where within each sample, sets generated by randomly sampling taxa should not be significantly more enriched than remainder taxa.
 
@@
     \caption{Random taxa set analyses for inference at the sample level of CBEA under different parametric assumptions compared against a Wilcoxon rank-sum test.
-Type I error (\emph{y}-axis) was evaluated by generating random sets of different sizes (\emph{x}-axis) (500 replications per size) and computing the fraction of samples where the set was found to be significantly enriched at $\alpha = 0.05$.
+Type I error (\emph{y}-axis) was evaluated by generating random sets of different sizes (\emph{x}-axis) (500 replications per size) and computing the fraction of samples in which the set was found to be significantly enriched at $\alpha = 0.05$.
 Error bars represent the mean type I error $\pm$ sample standard error computed across 500 replications of the experiment.
@@
 The Wilcoxon rank sum test and unadjusted CBEA under mixture normal assumption demonstrated good type I error control at the appropriate $\alpha$ level.
-This fits with our expectations since the mixture normal distribution has a much better fit than the normal distribution especially at the tails of the empirical distribution (Fig~\ref{fig:1}).
+This fits with our expectations since the mixture normal distribution, has a much better fit than the normal distribution especially at the tails of the empirical distribution (Fig~\ref{fig:1}).
 However, other variants of CBEA demonstrated inflated type I error, especially correlation adjusted variants compared to their unadjusted counter parts.
@@
 
-Similar to other single sample approaches to gene set testing such as GSVA \cite{hanzelmann2013}, we can perform inference at the population level by utilizing a two-sample difference in means test.
-Here, we evaluate using CBEA scores generated under different settings with Welch's t-test in a supervised manner to assess whether a set is enriched across case/control status.
+Similar to other single-sample approaches to gene set testing such as GSVA \cite{hanzelmann2013}, we can perform inference at the population level by utilizing a two-sample difference in means test.
+Here, we evaluated using CBEA scores generated under different settings with Welch's t-test in a supervised manner to assess whether a set is enriched across case/control status.
 
@@
 For sample label permutation (\textbf{A}), all CBEA approaches were able to control for type I error but not for corncob and DESeq2.
-For random set analyses (\textbf{B}), all approaches demonstrate similar rate of accepting significant sets and were invariant to overall set size.}
+For random set analyses (\textbf{B}), all approaches demonstrated similar rates of accepting significant sets and were invariant to overall set size.}
     \label{fig:3}
@@
 Fig~\ref{fig:3} shows results for this scenario using both random sample label and random set evaluations.
-The random sample label approach (Fig~\ref{fig:3}A) provides a controlled setting where we can estimate type I error rate controlled at $\alpha = 0.05$.
+The random sample label approach (Fig~\ref{fig:3}A) provided a controlled setting where we can estimate type I error rate controlled at $\alpha = 0.05$.
 Across all replications, CBEA methods were able to control for type I error at the nominal threshold of 0.05, with CBEA raw scores being the most performant.
 Neither output types, correlation adjustment, nor distributional assumption improved performance values.
-Surprisingly, DESeq2 and corncob both exhibit significantly inflated type I error.
-
-We also assessed the impact of set-size on the inference procedure by testing for enrichment using the original sample labels but with randomly sampled sets of different sizes (Fig~\ref{fig:3}B).
-Overall we observed very similar values across CBEA as well as corncob and DESeq2, suggesting that no individual method is systematically identifying too many significant sets.
+Surprisingly, DESeq2 and corncob both exhibited significantly inflated type I error.
+
+We also assessed the impact of set-size on the inference procedure by testing for enrichment using the original sample labels, but with randomly sampled sets of different sizes (Fig~\ref{fig:3}B).
+Overall, we observed very similar values across CBEA as well as corncob and DESeq2, suggesting that no individual method was systematically identifying too many significant sets.
 Additionally, similar to analogous analyses at the sample level, no approach was significantly sensitive to changes in set sizes.
@@
 
-In Fig~\ref{fig:4}, we evaluate whether sets found to be significant by CBEA are relevant to the phenotype of interest.
-We leveraged the gingival data set as stated in \nameref{evaluation} section where we know beforehand that aerobic microbes are more likely to be enriched in supragingival subsite samples and vice versa.
-
-We estimated statistical power using this data set as the fraction of supragingival samples where the set representing aerobic microbes were significantly enriched.
-We observed that adjusted CBEA approaches demonstrate much lower power compared to the Wilcoxon rank-sum test and unadjusted variants.
-This is surprising given the fact that in statistical significance analyses, the adjusted CBEA approach provides inflated type I error, especially if the normal distribution assumption was chosen, which indicates a mismatch in estimating the null distribution since a high type I error did not result in increased power.
+In Fig~\ref{fig:4}, we evaluated whether sets found to be significant by CBEA are relevant to the phenotype of interest.
+We leveraged the gingival data set as stated in \nameref{evaluation} section knowing beforehand that aerobic microbes were more likely to be enriched in supragingival subsite samples and vice versa.
+
+We estimated statistical power using this data set as the fraction of supragingival samples where the set representing aerobic microbes was significantly enriched.
+We observed that adjusted CBEA approaches demonstrated much lower power compared to the Wilcoxon rank-sum test and unadjusted variants.
+This is surprising given the fact that in statistical significance analyses, the adjusted CBEA approach provides inflated type I error, especially if the normal distribution assumption was chosen.
+This indicated a mismatch in estimating the null distribution since a high type I error did not result in increased power.
 
 We also evaluated phenotype relevance by assessing whether enriched sets according to ground truth are preferentially ranked higher using assigned continuous scores (instead of performing a hypothesis test).
-This aspect is captured through computing AUROC values comparing computed enrichment scores and true labels.
-Consistent with the previous type I error evaluation, adjusting for correlation did not improve performance, where obtained AUROC were around 0.5 and at the same level as the benchmark Wilcoxon rank sum statistic.
-Unadjusted methods were much better at ranking true enriched sets, however the mean AUROC values are lower than alternate single sample enrichment methods (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}) even though this difference is not significant due to overlapping confidence intervals.
-
-The above results were replicated in simulation studies where we observed that adjusted approaches were very conservative and demonstrated significantly lower power (\nameref{S3_Fig}) with increasing correlation even at the highest evaluated effect sizes.
-When assessing score rankings, the performance of CBEA was closer to ssGSEA and GSVA compared to real data evaluations, however all single sample approaches were much better than using the W statistic from the Wilcoxon Rank Sum test.
+This aspect was captured through computing AUROC values comparing computed enrichment scores and true labels.
+Consistent with the previous type I error evaluation, adjusting for correlation did not improve performance, whereas obtained AUROC were around 0.5 and at the same level as the benchmark Wilcoxon rank sum statistic.
+Unadjusted methods were much better at ranking true enriched sets, however the mean AUROC values were lower than alternate single-sample enrichment methods (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}) even though this difference is not significant due to overlapping confidence intervals.
+
+The above results were replicated in simulation studies where we observed that adjusted approaches were very conservative and demonstrated significantly lower power (\nameref{S3_Fig}), with increasing correlation even at the highest evaluated effect sizes.
+When assessing score rankings, the performance of CBEA was closer to ssGSEA and GSVA compared to real data evaluations, however all single-sample approaches were much better than using the W statistic from the Wilcoxon Rank Sum test.
 
@@
 Some CBEA variants, such as CDF outputs for the mixture normal distributional assumption, did not correctly detect as many significant sets as DESeq2 or corncob despite very close performance values.
-Using raw CBEA scores was best approach, however it did not exceed values obtained from DESeq2 and corncob.
+Using raw CBEA scores was best approach, however, it did not exceed values obtained from DESeq2 and corncob.
 
 \subsection*{Disease Prediction}   
-Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:4}), we wanted to assess whether they can also act as useful inputs to predictive models.
-In this section we assessed the predictive performance of a standard baseline random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (CBEA, ssGSEA, and GSVA).
-Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.
+Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:4}), we wanted to assess whether these scores can also act as useful inputs to predictive models.
+In this section we assessed the predictive performance of a standard baseline random forest model \cite{breiman2001} with different single-sample enrichment scoring methods as inputs (CBEA, ssGSEA, and GSVA).
+Additionally, we also compared the predictive performance of using these scores against a standard approach of using the centered log-ratio transformation (CLR) on taxon sets aggregated via abundance summations.
 
 We fit our model to two data sets with a similar disease classification task of discriminating patients who were diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition.
-The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al.
+The two data sets represent different microbiome sequencing approaches: the Gevers et al.
 \cite{gevers2014} data set uses 16S rRNA gene sequencing, while the Nielsen et al \cite{nielsen2014} data set uses whole genome shotgun sequencing.
@@
     %\includegraphics[width = \linewidth]{figures/data_pred.png}
-    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
-Data sets used span both 16S rRNA gene sequencing (Gevers et al.
+    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, and GSVA generated scores, as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles.
+The data sets used span both 16S rRNA gene sequencing (Gevers et al.
 \cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al.
@@
 In the 16S rRNA data set, the best performing CBEA variant (CDF values computed from an unadjusted mixture normal distribution) outperforms both GSVA and ssGSEA but not the standard CLR approach.
-Interestingly, in the whole genome sequencing data set, CBEA outperforms CLR, but was similar in performance to GSVA.
+Interestingly, in the whole genome sequencing data set, CBEA outperformed CLR, but was similar in performance to GSVA.
 However, due to large confidence intervals, no method significantly out-performed other evaluated approaches.
-As such, these results indicate that, for a given pre-determined collection of sets, CBEA generated scores are can be informative and provide competitive performance when acting as inputs to disease predictive models.
+As such, these results indicate that, for a given pre-determined collection of sets, CBEA generated scores can be informative and provide competitive performance when acting as inputs to disease predictive models.
 Simulation studies (\nameref{S5_Fig}) showed similar results, however CBEA more consistently underperformed compared to CLR across all scenarios.
-Interestingly, the performance gap decreases with increasing sparsity levels and correlation.
+Interestingly, the performance gap decreased with increasing sparsity levels and correlation.
 
@@
 \subsection*{Inference with CBEA} 
-CBEA is a microbiome-specific approach to generate sample specific enrichment scores for taxonomic sets defined \emph{a priori}.
+CBEA is a microbiome-specific approach to generating sample-specific enrichment scores for taxonomic sets defined \emph{a priori}.
 The formulation of CBEA as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}.
-Since this null hypothesis is self-contained per sample, this allows users perform enrichment testing at the sample level.
+Since this null hypothesis is self-contained per sample, it allows users perform enrichment testing at the sample level.
 Additionally, in combination with a difference in means test, CBEA can also test for enrichment at the population level across case/control status similar to GSVA \cite{hanzelmann2013}.
 
-For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}) while also demonstrating adequate power (Fig~\ref{fig:4}).
-This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), where the mixture normal displayed superior fit to the null distribution.
+For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}), while also demonstrating adequate power (Fig~\ref{fig:4}).
+This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), in which the mixture normal displayed superior fit to the null distribution.
 Unfortunately, other variants of CBEA demonstrated neither good type I error control nor power.
@@
 We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}.
-Alternately, corncob is a newer method developed specifically for microbiome taxonomic data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation.
+Alternately, corncob is a newer method developed specifically for microbiome taxonomic data sets, which models taxonomic counts directly using a beta-binomial distribution instead of relying on normalization via size factor estimation.
 We observed that using this approach resulted in an inflated type I error compared to all variants of CBEA (Fig~\ref{fig:2}), yet did not improve power (Fig~\ref{fig:4}).
-Results for CBEA approaches were replicated in simulation analyses, however for corncob and DESeq2 we observed an opposite effect: in simulation experiments, both methods show good type I error control but low power (\nameref{S2_Fig}, \nameref{S4_Fig}).
+Results for CBEA approaches were replicated in simulation analyses, however, for corncob and DESeq2 we observed an opposite effect: in simulation experiments, both methods show good type I error control but low power (\nameref{S2_Fig}, \nameref{S4_Fig}).
 
@@
 First, our simulation scenario ensures that all designated non-enriched taxa are identical to each other.
-This is not the case for real data, where our null scenario involves randomly sampled sets that might by chance all have taxa with inflated means compared to remainder taxa.
+This is not the case for real data, because our null scenario involves randomly sampled sets that might by chance all have taxa with inflated means compared to remainder taxa.
 This is represented in \nameref{S7_Fig}, where the distribution of type I error across 500 replications is right skewed for underperforming CBEA variants, indicating that these approaches are much more sensitive compared to the Wilcoxon rank sum test or unadjusted CBEA with mixture normal distribution.
@@
 In the context of sum-based aggregations, the resulting bias of the aggregated taxon-set is dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}).
-Conceptually, this means that measurement error for a taxon-set is different across samples as relative abundance of contributing taxa changes, leading to issues when attempting to perform inference.
-As such, we expect methods like corncob or DESeq2 when performed on such sum aggregates in the presence of taxon-specific biases to have inflated type I error compared to our multiplication based approach.
+Conceptually, this means that measurement error for a taxon-set is different across samples as the relative abundances of contributing taxa change, leading to issues when attempting to perform inference.
+As such, we expect methods like corncob or DESeq2, when performed on such sum aggregates in the presence of taxon-specific biases, to have inflated type I error compared to our multiplication based approach.
 This also explains why conversely in simulation studies, where taxon-specific biases are absent, corncob and DESeq2 performed better.
@@
 Similar to our inference analysis, we compared CBEA against both ssGSEA and GSVA.
-Additionally, we also evaluated CBEA with the approach where counts of a set were aggregated using sums and applied the centered log-ratio transformation (CLR).
+Additionally, we evaluated CBEA against a standard approach where counts of a set were aggregated using sums and applied the centered log-ratio transformation (CLR).
 This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}.
-Results showed that CBEA generate scores perform well across both real data and simulation scenarios.
+Results showed that CBEA generates scores perform well across both real data and simulation scenarios.
 Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that CBEA scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses.
-Comparatively, CBEA generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome taxonomic data sets.
-This is consistent with our sample ranking analysis (Fig~\ref{fig:4}), where CBEA scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
+Comparatively, CBEA generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that CBEA is more tailored for microbiome taxonomic data sets.
+This is consistent with our sample ranking analysis (Fig~\ref{fig:4}), showing that CBEA scores are on average more informative when used to rank samples based on their propensity to have inflated counts.
 However, CBEA did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data.
@@
 In the case where researchers only want to screen samples with mean-inflated taxon sets (instead of additionally detecting taxon sets with increased correlation), they can apply the adjusted approach, which can be effective at conserving type I error even for high correlation scenarios.
-However, the trade off for this adjustment is power, which decreases with increasing correlation.
+However, the trade-off for this adjustment is power, which decreases with increasing correlation.
 For the supervised analysis, all CBEA variants control for type I error and provide adequate statistical power.
-However, using raw CBEA scores with difference-in-means test such as Welch's t-test is preferable since is the least computationally expensive (no estimation process) while still outperforming the use of a sum-based approach with a standard differential abundance test.
+However, using raw CBEA scores with a difference-in-means test such as Welch's t-test is preferable since is the least computationally expensive (no estimation process) while still outperforming the use of a sum-based approach with a standard differential abundance test.
 
@@
 We attempted to perform this evaluation by leveraging the gingival data set similar to \cite{calgaro2020}.
-However, we acknowledge that this is not a perfect solution, since the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut.
-As such, results from power analyses using this data set is only relative between the comparison methods and cannot be treated as absolute measures of power or phenotype relevance.
-Third, fitting the mixture normal distribution to raw CBEA scores using the expectation-maximization algorithm is difficult, as the convergence rate is slow when there is high overlap between the mixtures, resulting in small mixing coefficient for one of the components and increased runtime (\nameref{S6_Fig}) \cite{naim2012}.
-In our implementation, we attempted to account for this by increasing the maximum number of iterations and relaxed the tolerance threshold.
+However, we acknowledge that this is not a perfect solution, since the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear-cut.
+As such, results from power analyses using this data set are only relative between the comparison methods and cannot be treated as absolute measures of power or phenotype relevance.
+Third, fitting the mixture normal distribution to raw CBEA scores using the expectation-maximization algorithm is difficult, as the convergence rate is slow when there is high overlap between the mixtures, resulting in a small mixing coefficient for one of the components and increased runtime (\nameref{S6_Fig}) \cite{naim2012}.
+In our implementation, we attempted to account for this by increasing the maximum number of iterations and relaxing the tolerance threshold.
 Finally, we assumed that taxa within a set are all equally associated with the outcome.
-This limits our ability to evaluate the performance of CBEA when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.
+This limits our ability to evaluate the performance of CBEA when only a small number of taxa within the set are associated with the outcome, or if there is variability in effect sizes or association direction of taxa within a set.
 
 Our evaluation also showed various drawbacks of the CBEA method itself.
-First, inference with CBEA at the sample level is limited, and can be affected by inter-taxa correlation if users wish to only detect mean-inflated sets.
+First, inference with CBEA at the sample level is limited, and can be affected by inter-taxa correlation if users wish to detect mean-inflated sets only.
 Second, for downstream analyses, CBEA might not always perform better than competing methods, especially when being used to generate inputs to predictive models.
-We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated the estimation procedure associated with adjusting the mixture normal distribution.
+We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated with the estimation procedure while adjusting the mixture normal distribution.
 As such, we hope to refine the null distribution estimating procedure by either choosing a better distributional form, or to further constrain the optimization procedure of the mixture normal distribution by fixing the third and fourth moments.
@@
 Performance was evaluated across different sparsity (\emph{x}-axis) and inter-taxa correlation levels.
-For CBEA methods, enrichment analysis was performed using a Welch's t-test across case/control status with single sample scores representing set-based features generated by CBEA (across different output types and distributional assumptions).
+For CBEA methods, enrichment analysis was performed using a Welch's t-test across case/control status with single-sample scores representing set-based features generated by CBEA (across different output types and distributional assumptions).
 For corncob and DESeq2, set-based features were constructed using element-wise summations.
@@
 {\bf Simulation results for phenotype relevance evaluation for CBEA sample-level inference.} \textbf{(A)} demonstrate statistical power (\emph{y}-axis) across different data sparsity levels (\emph{x}-axis) and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios.
-For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions).
+For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single-sample scores generated by CBEA (across different output types and distributional assumptions).
 CBEA associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.

### 2022-04-30 cbea_revised_v4.tex
V4

--- 2022-04-15 cbea_revised_v3.tex
@@
 Quang P. Nguyen\textsuperscript{1,2},
-Anne G. Hoen\textsuperscript{1,2\ddag},
-H. Robert Frost\textsuperscript{1*\ddag\dag},
+Anne G. Hoen\textsuperscript{1,2\dag},
+H. Robert Frost\textsuperscript{1*\dag},
 \\
@@
 % Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-\ddag These authors jointly supervised this work.\\
-\dag Corresponding author
+\dag These authors jointly supervised this work.\\
 % Current address notes
@@
 In addition to real data, we also evaluated models based on parametric simulations.
-Results can be found in the Supplemental Materials (\nameref{S1_Fig}, \nameref{S2_Fig}, \nameref{S3_Fig}, \nameref{S4_Fig}, and \nameref{S5_Fig}).
+Results can be found in the Supplemental Materials (S1-S5 Figs).
 
@@
 Unfortunately, other variants of CBEA demonstrated neither good type I error control nor power.
-Interestingly, while the adjusted methods showed poor performance in real data evaluations (Fig~\ref{fig:2}), in simulation studies (\nameref{S1_Fig}, \nameref{S3_Fig}) these approaches were able to control for type I error well with the trade-off of much lower power.
+Interestingly, while the adjusted methods showed poor performance in real data evaluations (Fig~\ref{fig:2}), in simulation studies (S1 and S3 Figs) these approaches were able to control for type I error well with the trade-off of much lower power.
 For the population-level inference task, CBEA also performed very well.
@@
 We observed that using this approach resulted in an inflated type I error compared to all variants of CBEA (Fig~\ref{fig:2}), yet did not improve power (Fig~\ref{fig:4}).
-Results for CBEA approaches were replicated in simulation analyses, however, for corncob and DESeq2 we observed an opposite effect: in simulation experiments, both methods show good type I error control but low power (\nameref{S2_Fig}, \nameref{S4_Fig}).
+Results for CBEA approaches were replicated in simulation analyses, however, for corncob and DESeq2 we observed an opposite effect: in simulation experiments, both methods show good type I error control but low power (S2 and S4 Figs).
 
